### 2020-12-09 paper.tex
lights v1

\documentclass[11pt]{article}

\usepackage[colorlinks=true,
            linkcolor=red,
            urlcolor=blue,
            citecolor=blue,
            breaklinks=true]{hyperref}
\usepackage{breakurl} 
\usepackage[a4paper, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\RequirePackage{amsmath,amsfonts, amssymb,amsthm}

\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}

\hypersetup{pdfauthor = {Simon Bussy}}

\usepackage{authblk}
\title{\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}}
\author[1]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
\author[2]{Antoine Barbieri}
\author[1]{Sarah Zohar}
\author[1,3]{Anne-Sophie Jannot}
\affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
\affil[2]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
\affil[3]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

\DeclareMathOperator{\argmin}{argmin}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

\date{}

\begin{document}

\maketitle

\vspace{-.5cm}

\begin{abstract}

This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available. Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a clinical perspective. Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.
\\

\noindent
\emph{Keywords.} High-dimensional estimation; Joint modeling; Multivariate longitudinal data; Survival analysis
\end{abstract}

\section{Introduction}

In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g. biomarkers) until the occurrence of an event of interest for a subject. The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{wulfsohn1997joint}. Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{tsiatis2004joint,yu2004joint}. They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features. In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories.

\paragraph{The high-dimensional longitudinal data context.}
With the development of electronic health records, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical. Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}. 
Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- computational power, limits in numerical estimation, or sample size for instance -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review). For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.

\paragraph{General framework.}

The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model. Let us consider the usual survival analysis framework.
Following~\citet{AndBorGilKei-93}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively. The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
\begin{equation*}
T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{}({\{T^\star \leq C\}})
\end{equation*}
respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{}(\cdot)$ the indicator function taking the value $1$ if the condition in $(\cdot)$ is satisfied and $0$ otherwise.

Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g. patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L > 0$.

\paragraph{Heterogeneity of the population.}

An assumption of heterogeneity within the patient population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}. To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk. Let us denote
\[ \pi_{\xi_k}(x) = \P[G=k|X=x] \]
the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
\[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group.  From now on, all computations are done conditionally on features $x$.

Then, the idea is to allow flexibility in modeling the dependency between the longitudinal features and the event time.
Joint models have the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features, thus offering a powerful tool for clinical decision making in patient monitoring. But those questions are never considered in a high-dimensional framework~\citep{proust2009development,rizopoulos2011dynamic}.

When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival}, or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power [REF!]. However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.

Once the model is train, posterior risk classification can be made through posterior estimates of the latent class membership probabilities, and latent class model may be particularly suited for prediction problems. We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters. Indeed, it makes sense from a practical point of view, since one want to obtain prognostic predictions as well as interpretations of the high-dimensional longitudinal features.

\paragraph{Main contribution.}

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) that \texttt{blabla... ajout description courte des caractéristiques du modèle une fois les maths terminées}.

\paragraph{Organization of the paper.}

A precise description of the model is given in Section~\ref{sec:Method}.
Section~\ref{sec:inference} focuses on a regularized version of the
model to exploit dimension reduction and prevent overfitting. Inference is
presented under this framework, as well as the convergence properties of the developed algorithm.
Section~\ref{sec:Performance evaluation} presents the simulation procedure used to evaluate the performance of our method and compares it with state-of-the-art ones. In Section~\ref{sec:application}, we apply our method to high-dimensional publicly available datasets. Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.

\paragraph{Notations.}

Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely
 $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$. We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
 For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element.
Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).

\section{Method}
\label{sec:Method}

In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.

\paragraph{Group-specific marker trajectories.}

We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
\begin{equation}
	\label{eq:link-function}
 	h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
 \end{equation}
where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
where $u^l(t) \in \R^{p_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq p_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$.
A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
\[ b^l \sim \cN(0, D_{ll}) \]
with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix that does not depend on the group membership. 
To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
\[ D = 
\begin{bmatrix}
  D_{11} & \cdots & D_{1L}\\
  \vdots &  \ddots & \vdots \\
  D_{1L}^\top & \cdots & D_{LL}
\end{bmatrix}
\]
the global variance-covariance matrix.
In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.

\paragraph{Group-specific risk of event.}

To quantify the effect of the longitudinal outcomes on the risk for an event, we use a relative risk model of the form
\begin{equation}
	\label{eq:intensity-model}
	\lambda(t|\cM_k(t), G = k, w) = \lambda_0(t) \exp \Big\{ w^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA \gamma_{k,a}^l \varphi_a(t, \beta_k^l, b^l) \Big\},
\end{equation}
where $\lambda_0(\cdot)$ denotes an unspecified baseline hazard function and given that $G = k$, $\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}$ denotes the history of the true unobserved longitudinal process up to time $t$, $w$ is a $q$-vector of time-independant features with corresponding fixed effects $\gamma_k \in \R^q$, and $\gamma_{k,a}^l \in \R$ is the joint model association parameter corresponding to a known functional $\varphi_a$ defining a shared association. Details on the choice of $(\varphi_a)_{a \in \cA}$ are given in Section~\ref{sec:numerical details}. From now on, all computations are done conditionally on features $w$.

\paragraph{Likelihood.}

Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
\[ \cD = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
for all $l=1, \ldots, L$. 
For the $i$-th subject, let us denote 
\[
\left\{
    \begin{array}{ll}
        y_i &= ({y_i^1}^\top, \ldots, {y_i^L}^\top)^\top \in \R^{n_i},\\
        b_i &= ({b_i^1}^\top, \ldots, {b_i^L}^\top)^\top \in \R^r,
    \end{array}
\right.
\]
with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
\[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times p}
\qquad \text{and} \qquad
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= [u_i^l(t_{i1}^l) \cdots u_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times p_l},\\
        V_{il} &= [v_i^l(t_{i1}^l) \cdots v_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
Note that from now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
Suppose that conditional on the random effects, each $Y_i^l$ is independent and the censoring times $T_i$ are independent of $b_i$; both are standard modelling assumptions.
For all $k = 0, \ldots, K-1$, we denote 
\[ \beta_k = ({\beta_k^1}^\top, \ldots, {\beta_k^L}^\top) \in \R^p \] 
with $p = \sum_{l=1}^L p_l$ and 
\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]

Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g. in the Gaussian case) or the counting measure (e.g. in the logistic cases).
The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
\begin{equation}
	\label{eq:exp-family-density}
	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
\end{equation}
with 
\[ \Phi_i = [\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top]^\top \in \R^{n_i} \] 
and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}. The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated. Note that $d_\phi(\cdot)$ is related to the normalizing constant.
\begin{table}[htb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\toprule
 Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_{\phi_l}(\cdot)$ \\
 \midrule
 Normal & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto \phi_l^{-1}\frac{z^2}{2}$ \\ 
 Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \phi_l^{-1} \frac{e^z}{1+e^z}$ \\
 Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto e^z$ & $z \mapsto \phi_l^{-1} e^z$ \\
 \bottomrule
\end{tabular}}
\caption[table]{\small Examples of standard distributions that fit in the considered setting.}
\label{table:glm}
\end{table}

Let us denote 
\[ \theta = (\xi_0, \ldots, \xi_{K-1}, \beta_0, \ldots, \beta_{K-1}, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0, \ldots, \gamma_{K-1})^\top \]
the collection of unknown parameters that we want to estimate, with $\gamma_k=$ \texttt{to be defined properly}.
To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us assume that the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
\begin{equation}
	\label{eq:ind-hyp}
	f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
\end{equation}
for all $i=1, \ldots, n$. Hypothesis~\eqref{eq:ind-hyp} is \texttt{faire le lien avec les hyps classiques de la litterature} [REFs].
Then, one has
\begin{align}
	\label{eq:log-lik}
	\ell_n(\theta) &= \ell_n(\theta ; \cD) \nonumber \\
	&= n^{-1} \sum_{i=1}^n \log \int_{-\infty}^{+\infty} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
\end{align}
where 
\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k, w_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(u|\cM_k(u), G_i = k, w_i) \dd u \Big\} \]
and
\[f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.\]

\section{Inference}
\label{sec:inference}

In order to avoid overfitting and to improve the prediction power of the lights model, we use Elastic-Net regularization~\citep{zou2005regularization} by minimizing the penalized objective
\begin{equation}
	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{E-N}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{E-N}, \eta} + \zeta_{3,k} \norm{\gamma_k}_{\text{E-N}, \eta}
\end{equation}
where for all $k=0, \ldots, K-1$, we add a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties of the vectors $\xi_k$, $\beta_k$ and $\gamma_k$ for tuning hyperparameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$, with a fixed $\eta \in [0, 1]$ and 
\[ \norm{v}_{\text{E-N}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \] 
for any vector $v$. \texttt{expliquer interpretation de $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})$}

One advantage of this regularization method is its ability to perform model selection (the lasso part) and pinpoint the most important
features relatively to the prediction objective. On the other hand, the ridge part allows to handle potential correlation between time-independent features or between longitudinal outcomes. Note that in practice, the intercept is not regularized.

\subsection{A Fast Stochastic Approximation of a Quasi-Newton EM}
\label{sec:FSA-QNEM}

\texttt{dérouler l'algo. idée: QNEM + \citet{karimi2020f}}

\subsection{Convergence to a stationary point}
\label{sec:convergence}

\texttt{théorème avec étude de convergence}

\subsection{Cross-validation procedure}
\label{sec:CV-procedure}

\paragraph{Posterior group membership probabilities.}

Les us denote
\begin{equation}
	\hat \pi_{ik}^{Y,T} = \P_{\hat \theta}[G_i = k | y_i, t_i, \delta_i]
\end{equation}

\texttt{utiliser ce marker pour cross-val les $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})$ avec du C-index + use randomized grid search~\citep{bergstra2012random}}

\section{Performance evaluation}
\label{sec:Performance evaluation}

In this section, we first give numerical details and briefly introduce the models we consider for performance comparisons. Then, we provide details regarding the simulation study and data generation. The chosen metrics for evaluating performances are then presented, followed by the results.

\subsection{Numerical details}
\label{sec:numerical details}

\paragraph{Specification of the design matrices.}

In many practical applications, subjects show highly nonlinear longitudinal trajectories. In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function. Hence, in order to produce a good estimate of $\cM(t)$, we consider a flexible representations for $u^l(t)$ and $v^l(t)$ using a high-dimensional vector of functions of time $t$ expressed in terms of splines.

\paragraph{Specification of the functionals $(\varphi_a)_{a\in \cA}$.}

\subsection{Competing models}
\label{sec:competing models}

State-of-the-art implementations in \texttt{R}: \\
\\
- recent package in multivariate settings: \texttt{joineRML}
\citep{hickey2018joinerml} \\
- reference one: \texttt{JM}
\citep{rizopoulos2010jm} \\
- MCMC methods: \texttt{JMbayes} 
\citep{rizopoulos2014r} (voir slide 132 \url{http://www.drizopoulos.com/courses/Int/JMwithR_CEN-ISBS_2017.pdf})\\ 
- \texttt{joineR}
\citep{philipson2018package} \\
- classical time-dependent Cox model: \texttt{survival}, explanations on how to do this in \citet{zhang2018time} \\
\\
> still nothing in \texttt{Python} ! I'll propose an implementation of the model in \texttt{Python} (today paramount in machine learning, it will be an additional asset of the paper).

\subsection{Simulation design}
\label{simulation design}

In order to assess the proposed method, we perform an extensive Monte Carlo simulation study. \\
\\
simulation of time-varying features in a Cox model: see \citet{therneau2017using}

\subsection{Metric}
\label{sec:Metrics}

We detail in this section the metric considered to evaluate risk prediction performances. Let us denote by $M$ the marker under study. We denote by $h$ the probability density function of marker $M$, and assume that the marker is measured once at $t = 0$.

A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
\begin{equation*}
  \cC =\P[M_i > M_j | T^\star_i < T^\star_j],
\end{equation*}
with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the $i.i.d.$ sample hypothesis). 
In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
\begin{equation*}
  \cC_\tau =\P[M_i > M_j | T_i < T_j , T_i < \tau],
\end{equation*}
with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}. A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{R} package \texttt{survival}.
Hence in the following, we consider the C-index metric to assess performances.

\subsection{Results of simulation}
\label{Results}
Let us present now the simulation results.

\section{Applications}
\label{sec:application}

\texttt{PBC dataset: open source, multidim mais dimension $\sim$15 (voir \url{https://rdrr.io/cran/JM/man/pbc.html}). \\ 
essayer de trouver 1 dataset en + grande dimension?}

\

\texttt{ other idea ++ use MIMIC II or III dataset ! \\
\url{https://www.nature.com/articles/sdata201635/}\\
\url{https://ieeexplore.ieee.org/abstract/document/1166854}\\
\url{https://www.sciencedirect.com/science/article/abs/pii/S2213260014702395}\\
\url{https://www.sciencedirect.com/science/article/pii/S2352556818302169}\\
}

\section{Conclusion}
\label{sec:conclusion}

\section*{Software}
All the methodology discussed in this paper is implemented in \texttt{Python} and \texttt{R}. The code is available from \href{https://github.com/SimonBussy/?}%
{https://github.com/SimonBussy/?} in the form of annotated programs, together with a notebook tutorial.

\section*{Acknowledgements}
\textit{Conflict of Interest}: None declared.

\appendix

\begin{center}
\LARGE \textbf{Appendices}
\end{center}

\section{Proofs}
\label{sec:proofs}

\bibliography{biblio}
\bibliographystyle{plainnat}{}
\end{document}

### 2020-12-09 paper.tex
lights v2

@@
 \RequirePackage{amsmath,amsfonts, amssymb,amsthm}
-
-\usepackage{mathtools}
-\usepackage{xcolor}
-\usepackage{mathrsfs} 
-\usepackage{natbib}
-\usepackage{setspace}
-\usepackage{dsfont}
-\usepackage{caption}
-\usepackage{algorithm}
-\usepackage{algorithmic}
-\usepackage{booktabs}
 
@@
 
+\usepackage{mathtools}
+\usepackage{mathrsfs}
+\usepackage{xcolor}
+\usepackage{mathrsfs} 
+\usepackage{natbib}
+\usepackage{setspace}
+\usepackage{dsfont}
+\usepackage{caption}
+\usepackage{booktabs}
+
+\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
+\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{orange}{#1}}
+\SetCommentSty{mycommfont}
+\SetKwInput{KwInput}{Input} 
+\SetKwInput{KwOutput}{Output}
+
+\newtheorem{assumption}{Assumption}{\bf}{\rm}
+
 \DeclareMathOperator{\argmin}{argmin}
@@
 \newcommand{\cA}{\mathcal A}
+\newcommand{\cQ}{\mathcal Q}
 \newcommand{\R}{\mathds R}
@@
 Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
-The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
+Features are extracted from the longitudinal processes and included as potential risk factor in a group-specific Cox model with high-dimensional shared associations.
+Appropriate penalties are then used during inference to allow flexibility in modeling the dependency between the longitudinal features and the event time.
+The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on publicly available datasets.
 On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
@@
 biomarkers) until the occurrence of an event of interest for a subject.
-The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{wulfsohn1997joint}.
-Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{tsiatis2004joint,yu2004joint}.
+The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}.
+Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{yu2004joint}.
 They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features.
 In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories.
+Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent subgroups sharing the same marker trajectories and the same prognostic.
 
@@
 Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}.
-Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- computational power, limits in numerical estimation, or sample size for instance -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
+Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g.
+computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
 For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.
@@
 \begin{equation*}
-T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{}({\{T^\star \leq C\}})
+T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
 \end{equation*}
-respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{}(\cdot)$ the indicator function taking the value $1$ if the condition in $(\cdot)$ is satisfied and $0$ otherwise.
+respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.
 
 Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g.
-patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L > 0$.
+patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
 
@@
 An assumption of heterogeneity within the patient population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}.
-To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk.
+To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}.
 Let us denote
-\[ \pi_{\xi_k}(x) = \P[G=k|X=x] \]
+\begin{equation}
+  \label{eq:pi}
+  \pi_{\xi_k}(x) = \P[G=k|X=x]
+\end{equation}
 the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
 \[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
-where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group.
+where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero.
+The intercept term is here omitted without loss of generality.
 From now on, all computations are done conditionally on features $x$.
-
-Then, the idea is to allow flexibility in modeling the dependency between the longitudinal features and the event time.
-Joint models have the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features, thus offering a powerful tool for clinical decision making in patient monitoring.
-But those questions are never considered in a high-dimensional framework~\citep{proust2009development,rizopoulos2011dynamic}.
-
-When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival}, or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
-Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power [REF!].
-However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.
-
-Once the model is train, posterior risk classification can be made through posterior estimates of the latent class membership probabilities, and latent class model may be particularly suited for prediction problems.
-We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters.
-Indeed, it makes sense from a practical point of view, since one want to obtain prognostic predictions as well as interpretations of the high-dimensional longitudinal features.
 
@@
 
-In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) that \texttt{blabla...
-ajout description courte des caractéristiques du modèle une fois les maths terminées}.
+In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
+To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
+for feature selection in the latent class membership as well as within the survival model, and sparse group lasso~\citep{simon2013sparse} for the fixed effect of each marker trajectories, allowing flexible representations of time.
+Inference is achieved using an efficient Quasi-Newton Expectation Maximization algorithm.
+The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
+Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.
 
@@
 model to exploit dimension reduction and prevent overfitting. Inference is
-presented under this framework, as well as the convergence properties of the developed algorithm.
-Section~\ref{sec:Performance evaluation} presents the simulation procedure used to evaluate the performance of our method and compares it with state-of-the-art ones.
+presented under this framework, as well as the developed algorithm.
+Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the competing methods considered.
+Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones.
 In Section~\ref{sec:application}, we apply our method to high-dimensional publicly available datasets.
@@
 For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element.
+We write $I_m$ for the identity matrix of $\R^{m \times m}$.
 Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).
@@
 
-\paragraph{Group-specific marker trajectories.}
+\subsection{Group-specific marker trajectories}
 
@@
 \[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
-where $u^l(t) \in \R^{p_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq p_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$.
+where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$ that does not depend on the group membership, which is not a strong modeling assumption.
+\begin{assumption}
+\label{assumption1}
+We suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
+\end{assumption}
 A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
 \[ b^l \sim \cN(0, D_{ll}) \]
-with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix that does not depend on the group membership. 
+with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
 To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
@@
 
-\paragraph{Group-specific risk of event.}
+\subsection{Group-specific risk of event}
 
-To quantify the effect of the longitudinal outcomes on the risk for an event, we use a relative risk model of the form
+To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
 \begin{equation}
 	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k, w) = \lambda_0(t) \exp \Big\{ w^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA \gamma_{k,a}^l \varphi_a(t, \beta_k^l, b^l) \Big\},
+	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
 \end{equation}
-where $\lambda_0(\cdot)$ denotes an unspecified baseline hazard function and given that $G = k$, $\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}$ denotes the history of the true unobserved longitudinal process up to time $t$, $w$ is a $q$-vector of time-independant features with corresponding fixed effects $\gamma_k \in \R^q$, and $\gamma_{k,a}^l \in \R$ is the joint model association parameter corresponding to a known functional $\varphi_a$ defining a shared association.
-Details on the choice of $(\varphi_a)_{a \in \cA}$ are given in Section~\ref{sec:numerical details}.
-From now on, all computations are done conditionally on features $w$.
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
+\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
+the history of the true unobserved longitudinal process up to time $t$, $x$ is the $p$-vector of time-independant features (also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during the inference, see~\eqref{eq:pen-log-lik}) with corresponding fixed effects $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\rho_a}$ the corresponding joint association parameters, and $\rho_a \in \N_+$ the dimension of the corresponding $\varphi_a$.
+This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
+\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
 
-\paragraph{Likelihood.}
+\paragraph{Specification of the functionals $(\varphi_a)_{a\in \cA}$.}
+
+The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework.
+In spite of that, rationale for selecting shared associations has received little attention.
+We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
+\begin{table}[htb]
+\centering
+\begin{tabular}{cccc}
+\toprule
+Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\rho_a$ & Reference \\
+\midrule
+Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
+Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\ 
+Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
+Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
+ \bottomrule
+\end{tabular}
+\caption[]{Description of the shared associations we include in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
+\label{table:shared_associations}
+\end{table}
+
+\subsection{Likelihood}
 
 Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
-\[ \cD = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
+\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
 where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
@@
     \begin{array}{ll}
-        y_i &= ({y_i^1}^\top, \ldots, {y_i^L}^\top)^\top \in \R^{n_i},\\
-        b_i &= ({b_i^1}^\top, \ldots, {b_i^L}^\top)^\top \in \R^r,
+        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
+        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
     \end{array}
@@
 \end{bmatrix} 
-\in \R^{n_i \times p}
+\in \R^{n_i \times q}
 \qquad \text{and} \qquad
@@
 \]
-where for all $l=1, \ldots, L$, one writes
+with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
 \[
@@
     \begin{array}{ll}
-        U_{il} &= [u_i^l(t_{i1}^l) \cdots u_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times p_l},\\
-        V_{il} &= [v_i^l(t_{i1}^l) \cdots v_i^l(t_{in_i^l}^l)]^\top \in \R^{n_i^l \times r_l}.
+        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
+        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
     \end{array}
@@
 \]
-Note that from now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
-Suppose that conditional on the random effects, each $Y_i^l$ is independent and the censoring times $T_i$ are independent of $b_i$; both are standard modelling assumptions.
+From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
+\begin{assumption}
+Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
+\end{assumption}
+This is a standard modeling assumption~\citep{tsiatis2004joint}.
 For all $k = 0, \ldots, K-1$, we denote 
-\[ \beta_k = ({\beta_k^1}^\top, \ldots, {\beta_k^L}^\top) \in \R^p \] 
-with $p = \sum_{l=1}^L p_l$ and 
+\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
+and 
 \[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]
@@
 	\label{eq:exp-family-density}
-	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
+	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \big\},
 \end{equation}
 with 
-\[ \Phi_i = [\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top]^\top \in \R^{n_i} \] 
+\[ \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i} \] 
 and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
@@
  \midrule
- Normal & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto \phi_l^{-1}\frac{z^2}{2}$ \\ 
- Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \phi_l^{-1} \frac{e^z}{1+e^z}$ \\
- Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto e^z$ & $z \mapsto \phi_l^{-1} e^z$ \\
+ Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2\sigma_l^2$ \\ 
+ Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
+ Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
  \bottomrule
 \end{tabular}}
-\caption[table]{\small Examples of standard distributions that fit in the considered setting.}
+\caption[]{Examples of standard distributions that fit in the considered setting.}
 \label{table:glm}
@@
 Let us denote 
-\[ \theta = (\xi_0, \ldots, \xi_{K-1}, \beta_0, \ldots, \beta_{K-1}, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0, \ldots, \gamma_{K-1})^\top \]
-the collection of unknown parameters that we want to estimate, with $\gamma_k=$ \texttt{to be defined properly}.
-To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us assume that the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
+\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
+the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
+To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
 \begin{equation}
@@
 \end{equation}
-for all $i=1, \ldots, n$. Hypothesis~\eqref{eq:ind-hyp} is \texttt{faire le lien avec les hyps classiques de la litterature} [REFs].
+for all $i=1, \ldots, n$. Hypothesis~\eqref{eq:ind-hyp} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
 Then, one has
@@
 	\label{eq:log-lik}
-	\ell_n(\theta) &= \ell_n(\theta ; \cD) \nonumber \\
-	&= n^{-1} \sum_{i=1}^n \log \int_{-\infty}^{+\infty} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
+	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
+	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
 \end{align}
 where 
-\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k, w_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(u|\cM_k(u), G_i = k, w_i) \dd u \Big\} \]
+\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
 and
-\[f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.\]
-
-\section{Inference}
-\label{sec:inference}
-
-In order to avoid overfitting and to improve the prediction power of the lights model, we use Elastic-Net regularization~\citep{zou2005regularization} by minimizing the penalized objective
 \begin{equation}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{E-N}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{E-N}, \eta} + \zeta_{3,k} \norm{\gamma_k}_{\text{E-N}, \eta}
+  \label{eq:b_density}
+  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
 \end{equation}
-where for all $k=0, \ldots, K-1$, we add a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties of the vectors $\xi_k$, $\beta_k$ and $\gamma_k$ for tuning hyperparameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$, with a fixed $\eta \in [0, 1]$ and
-\[ \norm{v}_{\text{E-N}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \] 
-for any vector $v$. \texttt{expliquer interpretation de $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})$}
-
-One advantage of this regularization method is its ability to perform model selection (the lasso part) and pinpoint the most important
-features relatively to the prediction objective.
-On the other hand, the ridge part allows to handle potential correlation between time-independent features or between longitudinal outcomes.
-Note that in practice, the intercept is not regularized.
-
-\subsection{A Fast Stochastic Approximation of a Quasi-Newton EM}
-\label{sec:FSA-QNEM}
-
-\texttt{dérouler l'algo. idée: QNEM + \citet{karimi2020f}}
-
-\subsection{Convergence to a stationary point}
-\label{sec:convergence}
-
-\texttt{théorème avec étude de convergence}
-
-\subsection{Cross-validation procedure}
-\label{sec:CV-procedure}
-
-\paragraph{Posterior group membership probabilities.}
-
-Les us denote
-\begin{equation}
-	\hat \pi_{ik}^{Y,T} = \P_{\hat \theta}[G_i = k | y_i, t_i, \delta_i]
-\end{equation}
-
-\texttt{utiliser ce marker pour cross-val les $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})$ avec du C-index + use randomized grid search~\citep{bergstra2012random}}
-
-\section{Performance evaluation}
-\label{sec:Performance evaluation}
-
-In this section, we first give numerical details and briefly introduce the models we consider for performance comparisons.
-Then, we provide details regarding the simulation study and data generation.
-The chosen metrics for evaluating performances are then presented, followed by the results.
-
-\subsection{Numerical details}
-\label{sec:numerical details}
 
@@
 In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function.
-Hence, in order to produce a good estimate of $\cM(t)$, we consider a flexible representations for $u^l(t)$ and $v^l(t)$ using a high-dimensional vector of functions of time $t$ expressed in terms of splines.
+Hence, in order to produce a good estimate of $\cM(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of polynomial functions of time $t$, namely
+\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top.\]
+% TODO: en pratique essayer d'ajouter X dans u^l(t) pr voir si ça améliore les perf ou non. Si oui l'ajouter au modèle.
+The idea here is to allow a wide range of orders for the polynomial so that a suitable one can be automatically chosen for each trajectory -- depending on its inherent complexity -- thanks to the regularization (see~\eqref{eq:pen-log-lik}).
+We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect.
+Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for for all $l=1, \ldots, L$.
 
-\paragraph{Specification of the functionals $(\varphi_a)_{a\in \cA}$.}
+\section{Inference}
+\label{sec:inference}
+
+In this section, we describe the procedure for estimating the parameters of the lights model.
+We begin by presenting the considered penalized objective, and then focus on the algorithm we use for inference.
+
+\subsection{Penalized objective}
+\label{sec:penalized-obj}
+
+In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
+\begin{equation}
+  \label{eq:pen-log-lik}
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
+\end{equation}
+where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vectors $\xi_k$ and $\gamma_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$.
+Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
+\[ \norm{v}_{\text{en}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \]
+for any vector $v$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
+\[ \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{\beta_k^l}_2 + \tilde{\eta} \norm{\beta_k}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+Hence, the resulting optimization problem is written
+\begin{equation}
+  \label{eq:optim-pb}
+   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
+ \end{equation}
+
+One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ (respectively $\gamma_k$) thus informs on the features involved in the $k$-th group membership (in the $k$-th group risk of event respectively).
+The ridge part allows to handle potential correlation between time-independent features or between longitudinal outcomes.
+Note that in practice, the intercept is not regularized.
+On the other hand, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory with the right complexity.
+
+\subsection{A Quasi-Newton Monte Carlo EM}
+\label{sec:QNMCEM}
+
+In order to derive an algorithm for this objective, we introduce a so-called QNMCEM algorithm, being a combination
+between an EM algorithm~\citep{dempster1977maximum} with Monte Carlo approximations~\citep{levine2001implementations}, and multiple L-BFGS-B algorithms~\citep{zhu1997algorithm}.
+EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
+
+We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
+It can be written
+\begin{align*}
+  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
+  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
+  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
+  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \Big].
+\end{align*}
+Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.
+
+\paragraph*{Monte Carlo E-step.}
+
+We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
+\begin{equation*}
+  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
+\end{equation*}
+Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
+\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
+for different functions $g$, where we denote 
+\begin{equation}
+  \label{eq:pi_ik-def}
+  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
+\end{equation}
+the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
+Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(Gi)$ with $g_k : G_i \mapsto \ind{\{G_i=k\}}$.
+
+In the case $g(b_i, G_i) = \tilde g(bi)$, one has
+\begin{equation*}
+  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1}  \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \Lambda_{ik,1}^{\theta^{(w)}}}
+\end{equation*}
+with
+\begin{equation}
+  \label{eq:Lambda_ik-def}
+  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
+\end{equation}
+and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
+The integral in~\eqref{eq:Lambda_ik-def} is intractable analytically, and some form of approximation must be used in practice.
+Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context.
+We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result.
+Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.
+
+\vspace{.5cm}
+
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\caption{Construction of the samples in $S^{(w)}$}
+\label{alg:construct-set-S}
+\KwInput{$S^{(w)} = \varnothing$}
+\KwOutput{$S^{(w)}$ filled with $2N$ samples}
+Compute $C^{(w)} \in \R^{r \times r}$ such that $C^{(w)}{C^{(w)}}^\top = D^{(w)}$ \tcp{Cholesky decomposition}
+\For{$\breve{n} = 1,\ldots,N$}    
+{ 
+  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
+  Compute $\breve{b}_{\breve{n}} = C^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
+  Update $S^{(w)} \leftarrow S^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
+}
+\textbf{Return:} {$S^{(w)}$}
+\end{algorithm}
+
+\vspace{.5cm}
+
+\noindent The approximation of~\eqref{eq:Lambda_ik-def} is finally obtained by
+\begin{equation}
+  \label{eq:hat_Lambda_ik}
+  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+\end{equation}
+
+It is common in Monte Carlo EM to increase $N$ with the iterations $w$~\citep{wei1990monte} since it is computationally inefficient to use a large $N$ in the early steps of the algorithm, when $\theta^{(w)}$ is far from $\hat \theta$.
+Multiple techniques have been proposed (see for instance~\citet{law2002joint} where a subjectively rule is followed, or~\citet{booth1999maximizing} that uses a rule based on confidence intervals for $\theta^{(w)}$ which requires additional variance estimation).
+We opt for a simple automated approach using relative differences of the objective function defined in~\eqref{eq:pen-log-lik} (see Algorithm~\ref{alg:QNMCEM}).
+
+Now in the case $g = g_k$, one has $\E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}}$ and we compute its approximation given
+\begin{equation}
+\label{eq:hat_pi_ik-def}
+  \hat \pi_{ik}^{\theta^{(w)}} = \dfrac{\hat \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \hat \Lambda_{ik,1}^{\theta^{(w)}}}.
+\end{equation}
+
+\paragraph*{Quasi-Newton M-step.}
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) .\]
+First, for the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
+\begin{equation}
+  \label{eq:update_lambda_0}
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} \ind{\{t_i \geq t\}} } ,
+\end{equation}
+with $\tilde g: b_i \mapsto \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big\}$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
+Then, the update of $D^{(w)}$ is naturally given in closed-form by
+\begin{equation}
+  \label{eq:update_D}
+  D^{(w+1)}= n^{-1} \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \breve{b} \breve{b}^\top f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+\end{equation}
+Concerning the update of $\phi^{(w)}$ now, and regarding Table~\ref{table:glm}, let us focus on the Gaussian case being particularly useful in practice.
+
+\paragraph*{Initialization.}
+Because our algorithm gives a local minimum, it is clever to choose an initial value $\theta^{(0)}$ relatively close to the final solution $\hat \theta$.
+
+\paragraph*{The QNMCEM algorithm.}
+Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimisation problem~\eqref{eq:optim-pb}.
+
+\vspace{.5cm}
+
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\caption{QNMCEM algorithm for inference of the lights model}
+\label{alg:QNMCEM}
+\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
+\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
+\tcc{Initialization}
+Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ following ?? \\
+\For{$w = 1,\ldots,$ until convergence}    
+{ 
+  \tcc{Monte Carlo E-step}
+  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik}\\
+  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
+  \If{Condition}
+  {
+      $N \leftarrow N + $
+  }
+  \tcc{Quasi-Newton M-step}
+  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
+  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
+
+}
+\textbf{Return:} {$\hat \theta$}
+\end{algorithm}
+
+\subsection{Practical details}
+\label{sec:practical details}
+
+Let us give some numerical details about the choices made for using the lights model in practice.
+
+\paragraph{Cross-validation procedure.}
+
+When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
+Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power in a prognostic prediction perspective~\citep{hickey2016joint}.
+However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.
+
+The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
+Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our QNMCEM algorithm introduced in Section~\ref{sec:QNMCEM}), posterior risk classification can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters.
+Indeed, it makes sense from a practical point of view, since one wants to obtain prognostic predictions as well as interpretation of the high-dimensional longitudinal features.
+
+\paragraph{The case $K=2$.}
+
+In many practical applications -- including those addressed in Section~\ref{sec:application} -- we are interested in identifying one subgroup of the population with a high risk of adverse event compared to the others.
+Then, in the following, we consider $G \in \{ 0, 1\}$ where $G=1$ means high-risk of early adverse event and $G=0$ means low risk.
+To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a patient belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
+In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\xi_k$ and $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$ and $\zeta_3 = \zeta_{3,0} = \zeta_{3,1}$.
+
+The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using marker $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) and the C-index score defined in Section~\ref{sec:Metrics}.
+Random search are indeed more efficient in terms of computing times than classical grid search for hyper-parameter optimization and finds better models by effectively searching a larger configuration space, which is appropriate to our high-dimensional problem.
+% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D
+
+\section{Performance evaluation}
+\label{sec:Performance evaluation}
+
+In this section, we first introduce the C-index metric used to assess performances and briefly introduce the models we consider for performance comparisons.
+Then, we provide details regarding the simulation study and data generation.
+The chosen metrics for evaluating performances are then presented, followed by the results.
+
+\subsection{The C-index metric}
+\label{sec:Metrics}
+
+We detail in this section the metric considered to evaluate risk prediction performances.
+Let us denote by $M$ the marker under study (e.g.
+the posterior group membership probabilities $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) for the lights model), and assume that it is measured once at $t = 0$.
+A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
+\begin{equation*}
+  \cC =\P[M_i > M_j | T^\star_i < T^\star_j],
+\end{equation*}
+with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
+In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
+\begin{equation*}
+  \cC_\tau =\P[M_i > M_j | T_i < T_j , T_i < \tau],
+\end{equation*}
+with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
+A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
+Hence in the following, we consider the C-index metric to assess performances.
+
+\subsection{Evaluation strategy}
+\label{sec:evaluation strategy}
 
@@
 
-State-of-the-art implementations in \texttt{R}: \\
-\\
-- recent package in multivariate settings: \texttt{joineRML}
-\citep{hickey2018joinerml} \\
-- reference one: \texttt{JM}
-\citep{rizopoulos2010jm} \\
-- MCMC methods: \texttt{JMbayes} 
-\citep{rizopoulos2014r} (voir slide 132 \url{http://www.drizopoulos.com/courses/Int/JMwithR_CEN-ISBS_2017.pdf})\\ 
-- \texttt{joineR}
-\citep{philipson2018package} \\
-- classical time-dependent Cox model: \texttt{survival}, explanations on how to do this in \citet{zhang2018time} \\
-\\
-> still nothing in \texttt{Python} ! I'll propose an implementation of the model in \texttt{Python} (today paramount in machine learning, it will be an additional asset of the paper).
+In this section, we briefly introduce the models we consider for performance comparisons in the simulation study as well as in the applications on real datasets in Section~\ref{sec:application}.
+
+\paragraph*{Penalized Cox model with time-independant features.}
+
+The first model we consider as a baseline is the well known Cox PH model with time-independant features.
+In this model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
+\begin{equation*}
+  \ell_n^{\text{cox}}(\beta) = n^{-1} \sum_{i=1}^n \delta_i \big( x_i^\top \beta - \log \sum_{i' : t_{i'} \geq t_i} \text{exp}(x_{i'}^\top \beta) \big).
+\end{equation*}
+We use respectively the \texttt{R} packages \texttt{survival} and \texttt{glmnet} \citep{simon2011regularization} for the partial log-likelihood and the minimization of the following quantity
+\begin{equation*}
+  - \ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big),
+\end{equation*}
+where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$.
+Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.
+We also choose to include basic time-independant features extracted from longitudinal processes, namely the average value for each time-dependant feature.
+
+\paragraph*{The time-dependent Cox model with \texttt{survival} package.}
+
+A classical extension of the Cox model supposes that features depend on time~\citep{sueyoshi1992semiparametric} (and then loosing the risk proportionality property), that is with our notations
+\[\lambda_i(t) = \lambda_0(t) \exp\big(y_i(t)^\top \beta \big). \]
+Using the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, blabla description...
+
+\paragraph*{Multivariate joint modeling with \texttt{joineRML} package.}
+In this \texttt{R} package~\citep{hickey2018joinerml}, blabla description...
+
+\paragraph*{The \texttt{JM} package.}
+In this \texttt{R} package~\citep{rizopoulos2010jm}, blabla description...
+
+\paragraph*{The \texttt{JMbayes} package.}
+In this \texttt{R} package~\citep{rizopoulos2014r}, blabla description...
+
+% TODO: voir slide 132 \url{http://www.drizopoulos.com/courses/Int/JMwithR_CEN-ISBS_2017.pdf})
+
+\paragraph*{The \texttt{joineR} package.}
+In this \texttt{R} package~\citep{philipson2018package}, blabla description...
+
+\section{High-dimensional simulation study}
+\label{sec:simulation study}
 
@@
 
-\subsection{Metric}
-\label{sec:Metrics}
-
-We detail in this section the metric considered to evaluate risk prediction performances.
-Let us denote by $M$ the marker under study.
-We denote by $h$ the probability density function of marker $M$, and assume that the marker is measured once at $t = 0$.
-
-A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
-\begin{equation*}
-  \cC =\P[M_i > M_j | T^\star_i < T^\star_j],
-\end{equation*}
-with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the $i.i.d.$ sample hypothesis). 
-In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
-\begin{equation*}
-  \cC_\tau =\P[M_i > M_j | T_i < T_j , T_i < \tau],
-\end{equation*}
-with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
-A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{R} package \texttt{survival}.
-Hence in the following, we consider the C-index metric to assess performances.
-
 \subsection{Results of simulation}
-\label{Results}
+\label{sec:simulation results}
 Let us present now the simulation results.
@@
 
-\texttt{PBC dataset: open source, multidim mais dimension $\sim$15 (voir \url{https://rdrr.io/cran/JM/man/pbc.html}). \\ 
-essayer de trouver 1 dataset en + grande dimension?}
+In this section, we apply our lights method on two publicly available datasets and compare its performance with state-of-the-art methods.
 
-\
+\subsection{The PBCseq dataset}
 
-\texttt{ other idea ++ use MIMIC II or III dataset ! \\
-\url{https://www.nature.com/articles/sdata201635/}\\
-\url{https://ieeexplore.ieee.org/abstract/document/1166854}\\
-\url{https://www.sciencedirect.com/science/article/abs/pii/S2213260014702395}\\
-\url{https://www.sciencedirect.com/science/article/pii/S2352556818302169}\\
-}
+This dataset is a follow-up to the original dataset~\citep{fleming2011counting, murtaugh1994primary}from the Mayo Clinic trial in primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984.
+A total of 424 PBC patients, referred to Mayo Clinic during that ten-year interval, met eligibility criteria for the randomized placebo controlled trial of the drug D-penicillamine.
+The first 312 cases in the data set participated in the randomized trial and contain largely complete data.
+The additional 112 cases did not participate in the clinical trial, but consented to have basic measurements recorded and to be followed for survival.
+Six of those cases were lost to follow-up shortly after diagnosis, so the data here are on an additional 106 cases as well as the 312 randomized participants.
+The dataset contains only baseline measurements of the laboratory paramters and contains multiple laboratory results, but only on the first 312 patients.
+
+\subsection{The MIMIC III dataset}
+
+The MIMIC III (Medical Information Mart for Intensive Care III) database is a large, freely-available hospital dataset containing de-identified data from over 40,000 patients.
+This data comes from patients who were admitted in critical care units to Beth Israel Deaconess Medical Center in Boston, Massachusetts from 2001 to 2012~\citep{johnson2016mimic}.
+The dataset was populated with data that had been acquired during routine hospital care, so there was no associated burden on caregivers and no interference with their workflow.
+
+% \url{https://www.nature.com/articles/sdata201635/}\\
+% \url{https://ieeexplore.ieee.org/abstract/document/1166854}\\
+% \url{https://www.sciencedirect.com/science/article/abs/pii/S2213260014702395}\\
+% \url{https://www.sciencedirect.com/science/article/pii/S2352556818302169}\\
+% }
 
@@
 
+In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
+algorithm (QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform covariate
+selection and to prevent overfitting.
+
 \section*{Software}
-All the methodology discussed in this paper is implemented in \texttt{Python} and \texttt{R}. The code is available from \href{https://github.com/SimonBussy/?}%
-{https://github.com/SimonBussy/?} in the form of annotated programs, together with a notebook tutorial.
+All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \href{https://github.com/SimonBussy/lights}%
+{https://github.com/SimonBussy/lights} in the form of annotated programs, together with a notebook tutorial.
 
@@
 
-\appendix
+% \appendix
 
-\begin{center}
-\LARGE \textbf{Appendices}
-\end{center}
+% \begin{center}
+% \LARGE \textbf{Appendices}
+% \end{center}
 
-\section{Proofs}
-\label{sec:proofs}
+% \section{Proofs}
+% \label{sec:proofs}
 

### 2020-12-09 paper.tex
lights v3

@@
 \DeclareMathOperator{\argmin}{argmin}
+\DeclareMathOperator{\Tr}{Tr}
 
@@
 Let us consider the usual survival analysis framework.
-Following~\citet{AndBorGilKei-93}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
+Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
 The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
@@
 To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
-for feature selection in the latent class membership as well as within the survival model, and sparse group lasso~\citep{simon2013sparse} for the fixed effect of each marker trajectories, allowing flexible representations of time.
-Inference is achieved using an efficient Quasi-Newton Expectation Maximization algorithm.
-The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
+for feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model, as well as for the fixed effect (allowing flexible representations of time).
+Indeed, this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model on one side.
+On the other hand, it yields sparsity for a given trajectory, namely feature selection.
+Inference is achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm.
+% NOOOOON
+% The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
 Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.
@@
 \begin{assumption}
+\label{indep-hyp-2}
 Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
 \end{assumption}
-This is a standard modeling assumption~\citep{tsiatis2004joint}.
+Assumption~\ref{indep-hyp-2} is a standard modeling assumption, see for instance~\citet{tsiatis2004joint}.
 For all $k = 0, \ldots, K-1$, we denote 
@@
 	\label{eq:exp-family-density}
-	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \big\},
+	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
 \end{equation}
@@
 \toprule
- Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_{\phi_l}(\cdot)$ \\
+ Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
  \midrule
- Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2\sigma_l^2$ \\ 
+ Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
  Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
@@
 \end{tabular}}
-\caption[]{Examples of standard distributions that fit in the considered setting.}
+\caption[]{Examples of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
 \label{table:glm}
@@
 the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
-To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
+To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
+\begin{assumption}
+\label{indep-hyp-3}
+Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
 \begin{equation}
@@
 \end{equation}
-for all $i=1, \ldots, n$. Hypothesis~\eqref{eq:ind-hyp} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
+for all $i=1, \ldots, n$.
+\end{assumption}
+Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
 Then, one has
@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
 \end{equation}
-where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vectors $\xi_k$ and $\gamma_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$.
+where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$ and $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$.
 Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
@@
 for any vector $v$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{\beta_k^l}_2 + \tilde{\eta} \norm{\beta_k}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+\[ \norm{v}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{v^l}_2 + \tilde{\eta} \norm{v}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
 Hence, the resulting optimization problem is written
@@
 
-One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ (respectively $\gamma_k$) thus informs on the features involved in the $k$-th group membership (in the $k$-th group risk of event respectively).
-The ridge part allows to handle potential correlation between time-independent features or between longitudinal outcomes.
+One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independant features involved in the $k$-th group membership.
+The ridge part allows to handle potential correlation between features.
+On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
+Finally, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory $l$ with the right complexity.
 Note that in practice, the intercept is not regularized.
-On the other hand, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory with the right complexity.
 
@@
   & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
-  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - \sum_{j=1}^{n_i} c_\phi(M_{ik, j}) + d_\phi(y_i) \Big].
+  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
 \end{align*}
@@
 the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(Gi)$ with $g_k : G_i \mapsto \ind{\{G_i=k\}}$.
+Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
 
-In the case $g(b_i, G_i) = \tilde g(bi)$, one has
-\begin{equation*}
-  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1}  \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \Lambda_{ik,1}^{\theta^{(w)}}}
-\end{equation*}
+In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
+\begin{equation}
+  \label{eq:def-E-g-tilde}
+  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
+\end{equation}
 with
@@
   \label{eq:Lambda_ik-def}
-  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
+  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
 \end{equation}
 and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
-The integral in~\eqref{eq:Lambda_ik-def} is intractable analytically, and some form of approximation must be used in practice.
+The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice.
 Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context.
@@
 
-\noindent The approximation of~\eqref{eq:Lambda_ik-def} is finally obtained by
+\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
+\begin{equation}
+  \label{eq:def-hat-E-g-tilde}
+  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
+\end{equation}
+with
 \begin{equation}
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \pi_{\xi_k^{(w)}}(x_i) \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}
@@
 \label{eq:hat_pi_ik-def}
-  \hat \pi_{ik}^{\theta^{(w)}} = \dfrac{\hat \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \hat \Lambda_{ik,1}^{\theta^{(w)}}}.
+  \hat \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}.
 \end{equation}
@@
 \paragraph*{Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) .\]
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
 First, for the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} \ind{\{t_i \geq t\}} } ,
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
 \end{equation}
-with $\tilde g: b_i \mapsto \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big\}$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
+which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model, and where 
+\begin{equation}
+  \label{eq:g-tilde-1-2}
+  \left\{
+    \begin{array}{ll}
+        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\},\\
+        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
+    \end{array}
+  \right.
+\end{equation}
 Then, the update of $D^{(w)}$ is naturally given in closed-form by
@@
   \label{eq:update_D}
-  D^{(w+1)}= n^{-1} \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \breve{b} \breve{b}^\top f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
 \end{equation}
-Concerning the update of $\phi^{(w)}$ now, and regarding Table~\ref{table:glm}, let us focus on the Gaussian case being particularly useful in practice.
+Concerning the update of $\phi^{(w)}$ now, and regarding Table~\ref{table:glm}, let us focus on the Gaussian case being particularly useful in practice. It is also obtained in closed-form by
+\begin{align}
+  \label{eq:update_phi}
+  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
+  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
+  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
+\end{align}
+Let us focus now on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. By denoting
+\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\begin{equation}
+  \label{eq:minimization-xi_k}
+  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
+\end{equation}
+It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}).
+We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
+of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
+In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
+Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
+\begin{equation}
+  \label{eq:sub-problem-xi_k}
+  \begin{split}
+    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
+    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
+  \end{split} 
+\end{equation}
+where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
+\begin{equation}
+  \label{eq:grad-xi_k}
+  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
+\end{equation}
+Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
+We then decrease the L-BFGS-B tolerance (which we denote $tol$) with every increase of $N$ (see Algorithm~\ref{alg:QNMCEM}).
+
+Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
+\begin{align*}
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
+\end{align*}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{s, \gamma_k, \beta_k}^1$ and $\tilde g_{s, \gamma_k, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N$, since $\lambda_0^{(w+1)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
+The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
+\begin{equation}
+  \label{eq:minimization-beta_k}
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{p+L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+\end{equation}
+We then rewrite problem~\eqref{eq:minimization-beta_k} as the following differentiable problem with box constraints
+\begin{equation}
+  \label{eq:sub-problem-beta_k}
+  \begin{split}
+    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 + \tilde \eta \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) \big) \\
+    \text{subject to}& \quad \quad \gamma_{k,j}^+ \geq 0 \text{ and } \gamma_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p+L\mathcal{A}.
+  \end{split} 
+\end{equation}
+% compute gradient
+
+Finally, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
+\begin{align*}
+  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
+  & + (y_i \odot \Phi_i^{(w+1)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
+  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w+1)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
+\end{align*}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, and solve The update for $\beta_k^{(w)}$ therefore requires to solve the following differentiable problem with box constraints
+\begin{equation}
+  \label{eq:sub-problem-beta_k}
+  \begin{split}
+    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
+    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
+  \end{split} 
+\end{equation}
+% calculer gradient
+
+% dire qqch in practice sur la dépendance de lambda, gamma et beta dans l'update
 
@@
 Because our algorithm gives a local minimum, it is clever to choose an initial value $\theta^{(0)}$ relatively close to the final solution $\hat \theta$.
+Then, let us give some details about the starting point of Algorithm~\ref{alg:QNMCEM}.
+For all $k = 0, \ldots, K-1$, we first use $\xi_k^{(0)}$ as the zero vector.
 
@@
   \tcc{Monte Carlo E-step}
-  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik}\\
+  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
   Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
@@
   {
-      $N \leftarrow N + $
+      $N \leftarrow N + $\\
+      $tol \leftarrow tol - $
   }
@@
   Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
-
+  Compute $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} with the L-BFGS-B algorithm\\
+  % \phi^{(w+1)} apres beta^{(w+1)} !!!!
+  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \quad \tcp{Gaussian case}
 }
@@
 
-Let us give some numerical details about the choices made for using the lights model in practice.
+Let us give some details about the choices made for using the lights model in practice.
 
@@
 
-% \appendix
+\appendix
 
-% \begin{center}
-% \LARGE \textbf{Appendices}
-% \end{center}
+\begin{center}
+\LARGE \textbf{Appendices}
+\end{center}
 
-% \section{Proofs}
-% \label{sec:proofs}
+\section{Numerical details}
+\label{sec:numerical-details}
+
+Concerning the 10-fold cross-validation procedure for tuning $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$, we then use a randomized search with the C-index metric (see Section~\ref{sec:practical details}).
+Let us precise that we choose $\zeta_{1,k}$ as the largest value such that error is within one standard error of the minimum, and that a randomized search is made during the cross-validation on an interval $[\zeta_{1,k}^{\text{max}} \times 10^{-4}, \zeta_{1,k}^{\text{max}} ]$, with $\zeta_{1,k}^{\text{max}}$ the interval upper bound computed hereafter.
+
+Considering the convex minimization problem~\eqref{eq:minimization-xi_k} at a given step $w$,
+and regarding the randomly chosen candidate values for $\zeta_{1,k}$, we consider $\zeta_{1,k}^1 \leq \zeta_{1,k}^2 \leq \dots \leq \zeta_{1,k}^{\text{max}}$.
+At $\zeta_{1,k}^{\text{max}}$, all coefficients $\hat \xi_{k, j}$ for all $j = 1, \dots, p$ are exactly zero.
+The KKT conditions \citep{boyd2004convex} claim that
+\begin{center}
+  $\left\{
+      \begin{aligned}
+        &\frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_j} =  \zeta_{1,k} \big( (\eta-1)\ \text{sgn}(\hat \xi_{k,j}) - \eta \hat \xi_{k,j} \big)  \quad \ \ \ \forall j \in \hat \cA_k \\
+        &\left| \frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_j} \right| < \zeta_{1,k} (1 - \eta) \quad \quad \quad \quad \quad \quad \quad \ \forall j \notin \hat \cA_k
+      \end{aligned}
+    \right.$,
+\end{center}
+where $\hat \cA_k = \big\{j = 1, \dots, p : \hat \xi_{k,j} \neq 0 \big\}$ is the active set of the $\hat \xi_k $ estimator, and for all $x \in~\R\setminus~\{0\},\ \text{sgn}(x) = \mathds{1}_{\{x>0\}} - \mathds{1}_{\{x<0\}} $.
+Then, using~\eqref{eq:grad-xi_k}, one obtains
+\begin{equation*}
+\hat \xi_{k, j} = 0 \ \Rightarrow \ \left| n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_{ij}  \right| < \zeta_{1,k} (1 - \eta)
+\end{equation*}
+for all $j = 1, \dots, p$.
+Hence, we choose the following upper bound for the randomly chosen candidate interval during the cross-validation procedure
+\begin{equation*}
+\zeta_{1,k}^{\text{max}} = \frac{1}{n(1-\eta)} \underset{j = 1, \dots, p}{\text{max}} \sum_{i=1}^n |x_{ij}|.
+\end{equation*}
 

### 2020-12-09 paper.tex
lights v4

@@
 
-In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g.
-biomarkers) until the occurrence of an event of interest for a subject.
+In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g., biomarkers) until the occurrence of an event of interest for a subject.
 The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}.
@@
 Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}.
-Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g.
-computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
+Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
 For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.
@@
 
-Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g.
-patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
+Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g., patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
 
@@
 presented under this framework, as well as the developed algorithm.
-Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the competing methods considered.
+Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the considered competing methods.
 Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones.
@@
 \[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
-the history of the true unobserved longitudinal process up to time $t$, $x$ is the $p$-vector of time-independant features (also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during the inference, see~\eqref{eq:pen-log-lik}) with corresponding fixed effects $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\rho_a}$ the corresponding joint association parameters, and $\rho_a \in \N_+$ the dimension of the corresponding $\varphi_a$.
+the history of the true unobserved longitudinal process up to time $t$.
+We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independant features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
+Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\rho_a}$ the corresponding joint association parameters, and $\rho_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
@@
 \centering
-\begin{tabular}{cccc}
+\begin{tabular}{ccccc}
 \toprule
-Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\rho_a$ & Reference \\
+Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$ & $\rho_a$ & Reference \\
 \midrule
-Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
-Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\ 
-Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
-Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
+Linear predictor & $m_k^l(t)$ & $u^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
+Random effects & $b^l$ & $\mathbf{0}_{q_l}$ & $r_l$ & \citet{hatfield2011joint} \\ 
+Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & $\dfrac{\dd}{\dd t} u^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
+Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & $\int_0^t u^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
  \bottomrule
 \end{tabular}
-\caption[]{Description of the shared associations we include in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
+\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.
+The gradient with respect to $\beta_k^l$ is also given, which will be useful for inference in Section~\ref{sec:QNMCEM}.}
 \label{table:shared_associations}
@@
 
-Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g.
-in the Gaussian case) or the counting measure (e.g.
-in the logistic cases).
+Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
 The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
@@
 In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function.
-Hence, in order to produce a good estimate of $\cM(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of polynomial functions of time $t$, namely
-\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top.\]
+Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely
+\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top\]
+with $\alpha \in \N_+$.
 % TODO: en pratique essayer d'ajouter X dans u^l(t) pr voir si ça améliore les perf ou non. Si oui l'ajouter au modèle.
-The idea here is to allow a wide range of orders for the polynomial so that a suitable one can be automatically chosen for each trajectory -- depending on its inherent complexity -- thanks to the regularization (see~\eqref{eq:pen-log-lik}).
+The idea here is to allow a wide range of polynomial orders for the representation so that a suitable one can be automatically chosen for each trajectory -- depending on its inherent complexity -- thanks to the regularization strategy proposed in~\eqref{eq:pen-log-lik}.
 We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect.
-Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for for all $l=1, \ldots, L$.
+Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.
 
@@
 In this section, we describe the procedure for estimating the parameters of the lights model.
-We begin by presenting the considered penalized objective, and then focus on the algorithm we use for inference.
+Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
 
@@
 Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
-\[ \norm{v}_{\text{en}, \eta} = (1-\eta)\norm{v}_1 + \dfrac\eta2 \norm{v}_2^2 \]
-for any vector $v$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{v}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{v^l}_2 + \tilde{\eta} \norm{v}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
+for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
+\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{z^l}_2 + \tilde{\eta} \norm{z}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
 Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
@@
 the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g(b_i, G_i) = \tilde g(bi)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
+Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
 
@@
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-First, for the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
-\begin{equation}
-  \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
-\end{equation}
-which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model, and where 
+Let us introduce the following useful function
 \begin{equation}
@@
     \begin{array}{ll}
-        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\},\\
-        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
+        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}\\
+        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l)
     \end{array}
@@
 \end{equation}
-Then, the update of $D^{(w)}$ is naturally given in closed-form by
+and present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
+First, the update of $D^{(w)}$ is naturally given in closed-form by
 \begin{equation}
@@
 \end{equation}
-Concerning the update of $\phi^{(w)}$ now, and regarding Table~\ref{table:glm}, let us focus on the Gaussian case being particularly useful in practice. It is also obtained in closed-form by
-\begin{align}
-  \label{eq:update_phi}
-  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
-  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
-  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
-\end{align}
-Let us focus now on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. By denoting
+Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
 \[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
@@
 
+Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
+\begin{align}
+  \label{eq:R-def}
+  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\ \nonumber
+  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\ \nonumber
+  & + (y_i \odot \Phi_i^{(w+1)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\ \nonumber
+  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w+1)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
+\end{align}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+Since $R^{(w)}_{n,k}$ depends on parameters estimates at iteration $w+1$ that have not been computed yet ($\gamma_k^{(w+1)}$, $\lambda_0^{(w+1)}$ and $\phi^{(w+1)}$), we denote $R^{(w^-)}_{n,k}$ the function obtained from~\eqref{eq:R-def} by replacing $(w+1)$ with $(w)$, that is using the last iteration estimates.
+Hence, we update $\beta_k^{(w)}$ by solving the following problem
+\begin{equation}
+  \label{eq:sub-problem-beta_k}
+  \begin{split}
+    \text{minimize}& \quad \quad R_{n, k}^{(w^-)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
+    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
+  \end{split} 
+\end{equation}
+The value of the gradient for the Gaussian case is here given by
+\begin{align*}
+  \label{eq:grad-beta_k}
+  \frac{\partial R_{n,k}^{(w^-)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\
+  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\
+  &+ U_{il}^\top \phi_l^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{align*}
+for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
+
+For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
+\begin{equation}
+  \label{eq:update_lambda_0}
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
+\end{equation}
+which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
+
 Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
@@
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{s, \gamma_k, \beta_k}^1$ and $\tilde g_{s, \gamma_k, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N$, since $\lambda_0^{(w+1)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w+1)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
 The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
@@
 \begin{equation}
-  \label{eq:sub-problem-beta_k}
+  \label{eq:sub-problem-gamma_k}
   \begin{split}
@@
 \end{equation}
-% compute gradient
+One can also compute the gradient since one has
+\begin{equation*}
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
+\end{equation*}
+and
+\begin{align*}
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\
+   &- \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+\end{align*}
+for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.
 
-Finally, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
-\begin{align*}
-  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
-  & + (y_i \odot \Phi_i^{(w+1)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
-  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w+1)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
-\end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, and solve The update for $\beta_k^{(w)}$ therefore requires to solve the following differentiable problem with box constraints
-\begin{equation}
-  \label{eq:sub-problem-beta_k}
-  \begin{split}
-    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
-    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
-  \end{split} 
-\end{equation}
-% calculer gradient
-
-% dire qqch in practice sur la dépendance de lambda, gamma et beta dans l'update
+Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
+\begin{align}
+  \label{eq:update_phi}
+  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
+  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
+  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
+\end{align}
 
@@
 \paragraph*{The QNMCEM algorithm.}
-Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimisation problem~\eqref{eq:optim-pb}.
+Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
 
@@
 \DontPrintSemicolon
-\caption{QNMCEM algorithm for inference of the lights model}
+\caption{QNMCEM algorithm for the lights model inference}
 \label{alg:QNMCEM}
@@
 \tcc{Initialization}
-Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ following ?? \\
+Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
 \For{$w = 1,\ldots,$ until convergence}    
@@
   \tcc{Quasi-Newton M-step}
+  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
+  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
+  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
   Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
-  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
-  Compute $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} with the L-BFGS-B algorithm\\
-  % \phi^{(w+1)} apres beta^{(w+1)} !!!!
-  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \quad \tcp{Gaussian case}
+  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
+  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
 }
@@
 We detail in this section the metric considered to evaluate risk prediction performances.
-Let us denote by $M$ the marker under study (e.g.
-the posterior group membership probabilities $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) for the lights model), and assume that it is measured once at $t = 0$.
+Let us denote by $M$ the marker under study (e.g., the posterior group membership probabilities $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) for the lights model), and assume that it is measured once at $t = 0$.
 A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by

### 2020-12-09 paper.tex
lights v5

@@
 \newcommand{\cQ}{\mathcal Q}
+\newcommand{\cR}{\mathcal R}
 \newcommand{\R}{\mathds R}
@@
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-Let us introduce the following useful function
+Let us introduce the following useful functions
 \begin{equation}
@@
     \begin{array}{ll}
-        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}\\
-        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l)
+        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
+        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
     \end{array}
@@
 \end{equation}
-and present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
+We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
 First, the update of $D^{(w)}$ is naturally given in closed-form by
@@
 Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-We then decrease the L-BFGS-B tolerance (which we denote $tol$) with every increase of $N$ (see Algorithm~\ref{alg:QNMCEM}).
+We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.
 
 Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
-\begin{align}
-  \label{eq:R-def}
-  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\ \nonumber
-  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\ \nonumber
-  & + (y_i \odot \Phi_i^{(w+1)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\ \nonumber
-  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w+1)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
-\end{align}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w+1)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
-Since $R^{(w)}_{n,k}$ depends on parameters estimates at iteration $w+1$ that have not been computed yet ($\gamma_k^{(w+1)}$, $\lambda_0^{(w+1)}$ and $\phi^{(w+1)}$), we denote $R^{(w^-)}_{n,k}$ the function obtained from~\eqref{eq:R-def} by replacing $(w+1)$ with $(w)$, that is using the last iteration estimates.
+\begin{align*}
+  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
+  & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
+  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
+\end{align*}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
 Hence, we update $\beta_k^{(w)}$ by solving the following problem
@@
   \begin{split}
-    \text{minimize}& \quad \quad R_{n, k}^{(w^-)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
+    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
     \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
@@
 The value of the gradient for the Gaussian case is here given by
-\begin{align*}
+\begin{align}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w^-)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\
-  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\
-  &+ U_{il}^\top \phi_l^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{align*}
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\ \nonumber
+  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\ \nonumber
+  &+ U_{il}^\top {\phi_l^{(w)}}^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{align}
 for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
-
-For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
-\begin{equation}
-  \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
-\end{equation}
-which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
 
@@
   Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
 \end{align*}
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w+1)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
 The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
@@
 One can also compute the gradient since one has
-\begin{equation*}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
-\end{equation*}
+\begin{equation}
+  \label{eq:grad-Q-1}
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
+\end{equation}
 and
-\begin{align*}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\
-   &- \sum_{j=1}^J \lambda_0^{(w+1)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
-\end{align*}
+\begin{align}
+  \label{eq:grad-Q-2}
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+\end{align}
 for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.
+
+For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
+\begin{equation}
+  \label{eq:update_lambda_0}
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
+\end{equation}
+which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
 
@@
 
-\begin{algorithm}[H]
+\begin{algorithm}[]
 \DontPrintSemicolon
@@
 
-\subsection{Practical details}
-\label{sec:practical details}
-
-Let us give some details about the choices made for using the lights model in practice.
-
-\paragraph{Cross-validation procedure.}
-
-When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
-Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness of fit nor generalisation power in a prognostic prediction perspective~\citep{hickey2016joint}.
-However, practitioners will naturally require predictive prognostic tools to evaluate model fits and compare models.
-
-The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
-Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our QNMCEM algorithm introduced in Section~\ref{sec:QNMCEM}), posterior risk classification can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
-We propose to use this probabilities as discriminative marker rule for risk prediction and use it in the cross-validation procedure for selecting the best regularization hyper-parameters.
-Indeed, it makes sense from a practical point of view, since one wants to obtain prognostic predictions as well as interpretation of the high-dimensional longitudinal features.
-
-\paragraph{The case $K=2$.}
+\subsection{The case $K=2$}
+\label{sec:K=2}
 
@@
 
-The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using marker $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) and the C-index score defined in Section~\ref{sec:Metrics}.
-Random search are indeed more efficient in terms of computing times than classical grid search for hyper-parameter optimization and finds better models by effectively searching a larger configuration space, which is appropriate to our high-dimensional problem.
+\paragraph{Cross-validation procedure.}
+
+The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
+Random search is indeed more efficient in terms of computing times than classical grid search for hyper-parameters optimization and finds better models by effectively searching in a larger configuration space, which is appropriate to our high-dimensional problem.
+Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:numerical-details}, while the search intervals for $\zeta_2$ and $\zeta_3$ are determined empirically.
 % TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D
@@
 
-In this section, we first introduce the C-index metric used to assess performances and briefly introduce the models we consider for performance comparisons.
-Then, we provide details regarding the simulation study and data generation.
-The chosen metrics for evaluating performances are then presented, followed by the results.
+In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances.
+We then introduce the appropriate metric used to this end: the C-index ; and finally, the considered models for performance comparisons.
+
+\subsection{Evaluation strategy: the real-time prediction paradigm}
+\label{sec:evaluation strategy}
+
+When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
+Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}.
+However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power.
+So we choose to put ourselves in this paradigm.
+
+\paragraph{Predictive marker.}
+
+The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
+The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
+Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our QNMCEM algorithm introduced in Section~\ref{sec:QNMCEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
+
+But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction.
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- we define
+\begin{equation}
+  \label{eq:marker-def}
+  \cR_i = \sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \exp \big\{ \tilde g_{t^{max}_i, \hat \gamma_k, \hat \beta_k}^1(b_i) \big\}
+\end{equation}
+as predictive marker rule of the lights model for subject $i$.
 
@@
 Hence in the following, we consider the C-index metric to assess performances.
-
-\subsection{Evaluation strategy}
-\label{sec:evaluation strategy}
 
@@
 
-\paragraph*{The time-dependent Cox model with \texttt{survival} package.}
+\paragraph*{The time-dependent Cox model.}
 
@@
 
-\paragraph*{Multivariate joint modeling with \texttt{joineRML} package.}
-In this \texttt{R} package~\citep{hickey2018joinerml}, blabla description...
+\paragraph*{Multivariate joint modeling.}
 
-\paragraph*{The \texttt{JM} package.}
-In this \texttt{R} package~\citep{rizopoulos2010jm}, blabla description...
+\texttt{joineRML} package : in this \texttt{R} package~\citep{hickey2018joinerml}, blabla description...
 
-\paragraph*{The \texttt{JMbayes} package.}
-In this \texttt{R} package~\citep{rizopoulos2014r}, blabla description...
+The \texttt{JM} package : in this \texttt{R} package~\citep{rizopoulos2010jm}, blabla description...
+
+The \texttt{JMbayes} package : n this \texttt{R} package~\citep{rizopoulos2014r}, blabla description...
 
@@
 
-\paragraph*{The \texttt{joineR} package.}
-In this \texttt{R} package~\citep{philipson2018package}, blabla description...
+The \texttt{joineR} package : in this \texttt{R} package~\citep{philipson2018package}, blabla description...
 
@@
 \label{sec:simulation study}
+
+In this section, we provide details regarding data generation, followed by the results of the extensive Monte Carlo simulation study to examine our method and compare it with state-of-the-art.
 
@@
 
-Concerning the 10-fold cross-validation procedure for tuning $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$, we then use a randomized search with the C-index metric (see Section~\ref{sec:practical details}).
-Let us precise that we choose $\zeta_{1,k}$ as the largest value such that error is within one standard error of the minimum, and that a randomized search is made during the cross-validation on an interval $[\zeta_{1,k}^{\text{max}} \times 10^{-4}, \zeta_{1,k}^{\text{max}} ]$, with $\zeta_{1,k}^{\text{max}}$ the interval upper bound computed hereafter.
+Concerning the cross-validation procedure for tuning 
+\[(\zeta_{1,0}, \zeta_{2,0}, \zeta_{3,0}, \ldots, \zeta_{1,K-1}, \zeta_{2,K-1}, \zeta_{3,K-1})^\top \in \R_+^{3K}, \]
+we use a randomized search with the C-index metric (see Section~\ref{sec:evaluation strategy}) where for all $k=0, \ldots, K-1$ and $j \in \{1, 2, 3\}$, each $\zeta_{j,k}$ finds its candidates in the interval
+\[[\zeta_{j,k}^{\text{max}} \times 10^{-4}, \zeta_{j,k}^{\text{max}} ] \subset \R, \]
+with $\zeta_{j,k}^{\text{max}}$ the interval upper bound computed hereafter.
 
 Considering the convex minimization problem~\eqref{eq:minimization-xi_k} at a given step $w$,
-and regarding the randomly chosen candidate values for $\zeta_{1,k}$, we consider $\zeta_{1,k}^1 \leq \zeta_{1,k}^2 \leq \dots \leq \zeta_{1,k}^{\text{max}}$.
-At $\zeta_{1,k}^{\text{max}}$, all coefficients $\hat \xi_{k, j}$ for all $j = 1, \dots, p$ are exactly zero.
+let us denote $\zeta_{1,k}^1 \leq \zeta_{1,k}^2 \leq \dots \leq \zeta_{1,k}^{\text{max}}$ the randomly chosen candidate values for $\zeta_{1,k}$, such that at $\zeta_{1,k}^{\text{max}}$, all coefficients $\hat \xi_{k, j}$ for all $j = 1, \dots, p$ are exactly zero.
 The KKT conditions \citep{boyd2004convex} claim that
@@
       \begin{aligned}
-        &\frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_j} =  \zeta_{1,k} \big( (\eta-1)\ \text{sgn}(\hat \xi_{k,j}) - \eta \hat \xi_{k,j} \big)  \quad \ \ \ \forall j \in \hat \cA_k \\
-        &\left| \frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_j} \right| < \zeta_{1,k} (1 - \eta) \quad \quad \quad \quad \quad \quad \quad \ \forall j \notin \hat \cA_k
+        &\frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_{k,j}} =  \zeta_{1,k} \big( (\eta-1)\ \text{sgn}(\hat \xi_{k,j}) - \eta \hat \xi_{k,j} \big)  \quad \ \ \ \forall j \in \hat \cA_k \\
+        &\left| \frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_{k,j}} \right| < \zeta_{1,k} (1 - \eta) \quad \quad \quad \quad \quad \quad \quad \ \forall j \notin \hat \cA_k
       \end{aligned}
@@
 \end{equation*}
+Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ or $\zeta_{3,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
+We then choose those interval upper bounds empirically.
 

### 2020-12-09 paper.tex
lights v6

@@
 \title{\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}}
-\author[1]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
-\author[2]{Antoine Barbieri}
+\author[1,2]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
+\author[3]{Antoine Barbieri}
 \author[1]{Sarah Zohar}
-\author[1,3]{Anne-Sophie Jannot}
+\author[1,4]{Anne-Sophie Jannot}
 \affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
-\affil[2]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
-\affil[3]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
+\affil[2]{Califrais' Machine Learning Lab, Paris, France}
+\affil[3]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
+\affil[4]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
 \setcounter{Maxaffil}{0}
@@
 \usepackage{booktabs}
+\usepackage{subfigure}
 
@@
 \newcommand{\cC}{\mathcal C}
+\newcommand{\cG}{\mathcal G}
 \newcommand{\cN}{\mathcal N}
@@
 \newcommand{\cQ}{\mathcal Q}
+\newcommand{\cU}{\mathcal U}
 \newcommand{\cR}{\mathcal R}
+\newcommand{\cH}{\mathcal H}
+\newcommand{\cB}{\mathcal B}
+\newcommand{\cS}{\mathcal S}
 \newcommand{\R}{\mathds R}
@@
 On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
-In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a clinical perspective.
-Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.
-\\
+In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a practical perspective.
+Thus, we propose a powerfull tool with the ability of automatically determining significant prognostic longitudinal features, which is of increasing importance in many areas: for instance personalized medicine, or churn prediction in a customer profile and activity monitoring setting, to name but a few.\\
 
@@
 
-In many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g., biomarkers) until the occurrence of an event of interest for a subject.
+With the increasing expectations to know their customers from account opening throughout the duration of the business relationship, web companies have the luxury of building elaborate systems to help them keep everything on track.
+The amount of recorded data per client is often tremendous and growing through time.
+There is no tool today to take into account simultaneously a huge number of longitudinal signals in a high-dimensional context to perform real-time churn (or satisfaction) risk prediction.
+Similarly, in many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g., biomarkers) until the occurrence of an event of interest for a subject.
 The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}.
@@
 \paragraph{The high-dimensional longitudinal data context.}
-With the development of electronic health records, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
+With the exploding number of daily internet users, or with the development of electronic health records in a medical context, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
 Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical.
@@
 Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
-The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression.
+The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression in a medical context; or the time when a client stops using a company's product or service in a churn prediction setting.
 We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
@@
 
-Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g., patients characteristics, therapeutic strategy, or omics features recorded at the begining of a study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
+Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g., patients characteristics, therapeutic strategy, or omics features recorded at the begining of a medical study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
 
@@
 
-An assumption of heterogeneity within the patient population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}.
+An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}.
 To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}.
@@
 Inference is achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm.
-% NOOOOON
-% The posterior estimates of the latent class membership probabilities are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters to adapt to the data complexity.
-Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring.
+Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters (to perform feature selection and adapt to the longitudinal data complexity).
+Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring for instance, or else for real-time decision support in a customer's satisfaction monitoring context.
 
@@
 
-Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely
- $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$. We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
+Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$.
+We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
+$\lfloor a \rfloor$ denotes the largest integer less than or equal to $a \in \R$.
 For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
@@
 A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
-\[ b^l \sim \cN(0, D_{ll}) \]
+\begin{equation}
+   \label{eq:random-effect-law}
+   b^l \sim \cN(0, D_{ll})
+ \end{equation} 
 with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
@@
 We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independant features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
-Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\rho_a}$ the corresponding joint association parameters, and $\rho_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
+Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
@@
 \toprule
-Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$ & $\rho_a$ & Reference \\
+Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$ & $\imath_a$ & Reference \\
 \midrule
@@
 \label{indep-hyp-2}
-Suppose that conditional on the random effects, each $Y_i^l$ is independent, and that the censoring times $T_i$ are independent of $b_i$.
+Suppose that for a given subject $i$ and conditional on the random effects and the group membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
 \end{assumption}
-Assumption~\ref{indep-hyp-2} is a standard modeling assumption, see for instance~\citet{tsiatis2004joint}.
+Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
 For all $k = 0, \ldots, K-1$, we denote 
@@
 
-Given~\eqref{eq:link-function}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
+Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
 The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
@@
 for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\sum_{l=1}^L\norm{z^l}_2 + \tilde{\eta} \norm{z}_1 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
 Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
@@
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-Let us introduce the following useful functions
+Let us introduce the following useful notations
 \begin{equation}
@@
 \paragraph*{Initialization.}
-Because our algorithm gives a local minimum, it is clever to choose an initial value $\theta^{(0)}$ relatively close to the final solution $\hat \theta$.
-Then, let us give some details about the starting point of Algorithm~\ref{alg:QNMCEM}.
-For all $k = 0, \ldots, K-1$, we first use $\xi_k^{(0)}$ as the zero vector.
+Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced.
+We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}.
+For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
+Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using \texttt{R} package \texttt{survival}.
+Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm, being itself initialized with univariates fits using \texttt{R} package \texttt{nlme}.
 
@@
   Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
+  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
   Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
-  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
   Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
@@
 Then, in the following, we consider $G \in \{ 0, 1\}$ where $G=1$ means high-risk of early adverse event and $G=0$ means low risk.
-To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a patient belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
+To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
 In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\xi_k$ and $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$ and $\zeta_3 = \zeta_{3,0} = \zeta_{3,1}$.
@@
 Random search is indeed more efficient in terms of computing times than classical grid search for hyper-parameters optimization and finds better models by effectively searching in a larger configuration space, which is appropriate to our high-dimensional problem.
-Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:numerical-details}, while the search intervals for $\zeta_2$ and $\zeta_3$ are determined empirically.
+Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ and $\zeta_3$ are determined empirically.
 % TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D
@@
 
-But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction.
-Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- we define
+But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction.
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in subgroup $k$ would rather be
 \begin{equation}
   \label{eq:marker-def}
-  \cR_i = \sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \exp \big\{ \tilde g_{t^{max}_i, \hat \gamma_k, \hat \beta_k}^1(b_i) \big\}
+  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)},
 \end{equation}
-as predictive marker rule of the lights model for subject $i$.
+with 
+\[ \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
+where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
+Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
+\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
 
@@
 We detail in this section the metric considered to evaluate risk prediction performances.
-Let us denote by $M$ the marker under study (e.g., the posterior group membership probabilities $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}) for the lights model), and assume that it is measured once at $t = 0$.
+Let us denote by $M$ the marker under study (e.g., the marker~\eqref{eq:marker-def} for the lights model), and assume that it is measured once at $t = t^{max}$.
 A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
@@
 Hence in the following, we consider the C-index metric to assess performances.
+
+Note that other metrics have been proposed to compare individual risk predictions.
+In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$.
+But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~\citep{bussy2019comparison}), which is not the case of the C-index metric.
+Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.
 
@@
 
-In order to assess the proposed method, we perform an extensive Monte Carlo simulation study. \\
-\\
-simulation of time-varying features in a Cox model: see \citet{therneau2017using}
+In order to assess the proposed method, we run an extensive Monte Carlo simulation study in the case $K=2$ (see Section~\ref{sec:K=2}).
+We first generate $G_i \sim \cB\big(\pi_1\big)$, with $\pi_1 \in [0,1]$ a desired high-risk subjects proportion and where $\cB(\alpha)$ denotes the Bernoulli distribution with parameter $\alpha \in [0,1]$.
+We then choose a coefficient vector 
+\begin{equation}
+  \label{eq:xi-sparse}
+  \xi = (\underbrace{\varsigma_1,\ldots,\varsigma_1}_s,0,\ldots,0) \in \R^p,
+\end{equation} 
+with $\varsigma_1\in\R$ being the value of the active coefficients and $s\in \{1,\dots,p\}$ a sparsity parameter chosen such that a proportion $r_s$ of time-independent features are actually active, that is $s = \lfloor p \times r_s \rfloor$.
+For the generation of the time-independent features matrix, we take 
+\[ [x_{ij}] \in \R^{n \times p} \sim \cN \big(0, \bSigma_1(\rho_1)\big),\] 
+with $\bSigma_1(\rho_1)$ a $(p \times p)$ Toeplitz covariance matrix \citep{mukherjee1988some} with correlation $\rho_1 \in (0, 1)$. 
+We then add a $gap \in \R^+$ value for high-risk subjects and subtract it for low-risk subjects, only on active features, that is
+\begin{center}
+$x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{G_i=0\}}} gap\ \text{for } j = 1, \dots, s$.
+\end{center}
+Note that this is equivalent to generate the time-independent features according to a gaussian mixture.
+
+Now, concerning the simulation of the longitudinal data, the idea is to sample from multivariate normal distributions. 
+We use linear time-varying features such that
+\[ Y_i^l(t) = \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \big( (1, t)^\top \beta_k^l + (1, t)^\top b_i^l + \epsilon_i^l(t) \big) \]
+where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to
+\[ \beta_k^l \sim \cN\Big( 
+\mu_k, 
+\begin{bmatrix}
+  \rho_3 & 0\\
+  0 & \rho_3
+\end{bmatrix} 
+\Big) \]
+for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that $\beta_k^l$ coefficients are both positive for high-risk subjects ($G_i=1$), which favours increasing trajectories with positive values, and negative otherwise ($G_i=0$), which favours decreasing trajectories with negative values.
+This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.
+
+Most joint modeling simulation studies use fixed time points~\citep{hickey2018joinerml}, that is $n_i^l = n_i^{l'}$ and $t_{ij}^l = t_{ij}^{l'}$ for all $(l, l') \in \{1, \ldots, L\}^2$ and $j = 1, \ldots, n_i^l$ for a given subject $i$, but it does not reflect actual behavior in practice, since data measurements of the different longitudinal processes are barely done simultaneously.
+For this reason, the simulation of realistic longitudinal features is not trivial and we use Hawkes processes~\citep{hawkes1974cluster} with exponential kernels to generate measurement times, which is a family of counting process with an autoregressive intensity.
+Hawkes processes model self-exciting behavior, that is in our case, when the measurement of one longitudinal process makes future measurements of the process -- as well as those of others processes, in a multivariate way -- more likely to happen.
+Namely, for a subject $i$, times $\{t_{ij}^l\}_{j \geq 1}$ for processes $l=1, \ldots, L$ are simulated using a multivariate Hawkes process $N_{it} = [N_{it}^1 \cdots N_{it}^L]$ with $t \geq 0$ and $N_{it}^l = \sum_{j \geq 1} \ind{\{t_{ij}^l \leq t\}}$.
+The process $N_{it}$ is a multivariate counting process, whose components $N_{it}^l$ have intensities
+\[ \lambda_i^l(t) = \Upsilon_{l} + \sum_{l'=1}^L \sum_{j \geq 1} A_{ll'} \upsilon \exp\big(-\upsilon(t - t_{ij}^{l'}) \big) \]
+for $l=1, \ldots, L$.
+$\Upsilon_{l} \geq 0$ is called baseline intensity, and corresponds to the exogenous probability of having a measurement for process $l$.
+We sample $\Upsilon_{l} \sim \cU\big([0.1,0.5]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
+The matrix $A = [A_{ll'}]_{1 \leq l,l' \leq L}$ is the adjacency matrix such that $A_{ll'} \geq 0$ quantifies the impact of past measurement time of process $l'$ on the measurement time of process $l$, and $\upsilon \geq 0$ is a memory parameter.
+We sample $A_{ll'} \sim \cU\big([0.05,0.1]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
+Simulation of the Hawkes processes
+is achieved using the \texttt{tick} library~\citep{bacry2017tick}. An example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
+
+To generate survival times, we choose a risk model of the form
+\begin{align*}
+\lambda_i(t|G_i = k) = \lambda_0(t) \exp \Big\{ x_i^\top \xi + \sum_{l=1}^L &\gamma_{k,1}^l (\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t) \\
+&+ (\gamma_{k,2,1}^l b_{i,1}^l + \gamma_{k,2,2}^l b_{i,2}^l) + \gamma_{k,3}^l (\beta_{k,2}^l + b_{i,2}^l) \Big\},
+\end{align*}
+which corresponds to~\eqref{eq:intensity-model} taking $\gamma_{k,0} = \xi$ and the first three functionals described in Table~\ref{table:shared_associations} as shared associations. 
+Now, for the choice of the association parameters, we want to induce sparsity (similarly to~\eqref{eq:xi-sparse}), so we set 
+\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] and
+\begin{equation}
+  \label{eq:gamma-sparse}
+  \gamma_{k,a}^l = \varsigma_2 \big(\ind{\{l \leq (k + 1) \lfloor L r_s/K \rfloor \}} + \ind{\{l \in \cS_k \}} \big) \mathbf{1}_{\imath_a},
+\end{equation}
+so that there is a proportion $r_s$ of active longitudinal features among the $L$, and the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$.
+Note that one can write
+\[ \lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\},\]
+being a Cox model with a linear time-varying feature.
+We choose a Gompertz distribution~\citep{gompertz1825xxiv} for the baseline, that is
+\begin{equation}
+  \label{eq:baseline}
+  \lambda_0(t) = \kappa_1 \kappa_2 \exp(\kappa_2t)
+\end{equation}
+with $\kappa_1 > 0$ and $\kappa_2 \in \R$ the scale and shape parameters respectively, which is a common distribution choice in survival analysis~\citep{klein2005survival} with a rich history in describing mortality curves.
+One can now generate survival times explicitly as
+\begin{equation}
+  \label{eq:time-generation}
+  T_i^\star | G_i=k \sim \dfrac{1}{\iota_{i,k,2} + \kappa_2} \log \Big(1 - \dfrac{(\iota_{i,k,2} + \kappa_2) \log U_i}{\kappa_1 \kappa_2 \exp\iota_{i,k,1}} \Big)
+\end{equation}
+where $U_i \sim \cU\big([0,1]\big)$ (see Appendix~\ref{sec:event-time-simu} for the derivation of this expression).
+The distribution of the censoring variable $C_i$ is the geometric distribution $\cG(\alpha_c)$, where $\alpha_c \in (0, 1)$ is empirically tuned to maintain a desired censoring rate $r_c \in [0,1]$. 
+Finally, the prediction time for subject $i$ is draw according to 
+\begin{equation}
+  \label{eq:law-t_i^max}
+  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
+\end{equation}
+with $(\nu_1, \nu_2) \in \R_+^2$ two shape parameters. Figure~\ref{fig:timax-pdf} shows the corresponding probability density function for $t_i^{max}$ with some explanations.
+We then truncate all $Y_i$ measurements being after $t_i^{max}$, since it represents the ``present'' time, so the time up to which one has longitudinal data.
+The choice of all hyper-parameters is driven by the applications on real data presented in Section~\ref{sec:application}, and summarized in Table~\ref{table:parameters choice}.
+Figure~\ref{fig:example-data} gives an example of data generated according to the design we have just described.
+\begin{table}[htb]
+\caption{Hyper-parameter choices for simulation. Let us also precise that we take $n \in [200, 4000]$ and $(p, L) \in [3, 300]^2$.
+\label{table:parameters choice}}
+\centering
+\resizebox{\textwidth}{!}{
+\begin{tabular}{ccccccccccccc}
+\toprule
+$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\nu_1, \nu_2)$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
+\midrule
+0.4 & $(, 10^{-3},)$ & 
+$\begin{pmatrix}
+  1 \\ 0.3
+\end{pmatrix}$ & 
+$\begin{pmatrix}
+  1 \\ 0.5
+\end{pmatrix}$ 
+& 0.5 & 0.5 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
+\bottomrule
+\end{tabular}}
+\end{table}
+\begin{figure}[!htb]
+\centering
+\subfigure[Density of the law used to simulate $t_i^{max}$ according to~\eqref{eq:law-t_i^max} with $(\nu_1, \nu_2) = (2,5)$, to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.]{
+\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
+\label{fig:timax-pdf}}
+\hspace{.02\textwidth}
+\subfigure[exemple simu]{
+% \includegraphics[width=.38\textwidth]{./figures/time2.pdf}
+\label{fig:example-data}}
+\caption{Illustrations of simulated data.}
+\end{figure}
+
+Following the real-time prediction paradigm presented in Section~\ref{sec:evaluation strategy}, we compare the predictive performances of the lights model with the competing ones introduced in Section~\ref{sec:competing models} in terms of C-index (see Section~\ref{sec:Metrics}).
+We also want to assess the stability of the lights model in terms of feature selection, but we do not compare this aspect with state-of-the-art since other models are not designed to perform feature selection, and comparisons would be unfair.
+To this end, we follow the same simulation procedure explained in the previous lines and for each simulation case, we use the following approach to evaluate the variable selection power of the model.
+Denoting
+\[\tilde{\xi}_j = \dfrac{|\hat{\xi}_j|}{\max_{j=1,\ldots,p}{|\hat{\xi}_j|}} \qquad \text{and} \qquad \tilde{\gamma}_{k,a}^l = \dfrac{|\hat{\gamma}_{k,a}^l|}{\max_{a=1, \ldots, \cA}{|\hat{\gamma}_{k,a}^l|}}\]
+for $k=0,\ldots,K-1$ and $l=1, \ldots, L$, and considering that $\tilde{\xi}_j$ and $\tilde{\gamma}_{k,a}^l$ are the predicted probability that the true $\xi_j$ equals $\varsigma_1$ and the predicted probability that the true $\gamma_{k,a}^l$ equals $\varsigma_2$ respectively, then we are in a binary prediction setting and we use the two resulting AUC scores of this problem to evaluate the selection power of the time-independent and the time-dependent features respectively.
 
@@
 In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
-algorithm (QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform covariate
+algorithm (QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform feature
 selection and to prevent overfitting.
@@
 \section*{Software}
-All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \href{https://github.com/SimonBussy/lights}%
-{https://github.com/SimonBussy/lights} in the form of annotated programs, together with a notebook tutorial.
+All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \href{https://github.com/Califrais/lights}%
+{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.
 
@@
 
-\section{Numerical details}
-\label{sec:numerical-details}
+\section{Automatic cross-validation grid for $\xi$}
+\label{sec:auto-grid-cv}
 
@@
 
+\section{Event time simulation}
+\label{sec:event-time-simu}
+
+Let us derive here the expression~\eqref{eq:time-generation} used to generate event times.
+Indeed, we can not refer to~\citet{austin2012generating} since the intensity~\eqref{eq:baseline} of the baseline is not correct in this paper.
+Note that other errors had already been pointed in~\citep{austin2013correction}, for instance the fact that there is no closed-form expression with a Weibull baseline.
+Suppose that
+\[\lambda(t) = \lambda_0(t) \exp (a + b t)\]
+with $(a, b) \in \R^2$ and $\lambda_0$ defined in~\eqref{eq:baseline}, then the cumulative hazard function writes
+\[H(t) = \kappa_1 \kappa_2 e^a \int_0^t \exp\{(\kappa_2 + b)s \} \dd s = \dfrac{\kappa_1 \kappa_2 e^a}{\kappa_2 + b} \big( \exp\{(\kappa_2 + b) t\} - 1 \big).\]
+Then, on has
+\[t = \dfrac{1}{\kappa_2 + b} \log \Big( 1 + \dfrac{H(t)(\kappa_2 + b)}{\kappa_1 \kappa_2 e^a} \Big),\]
+and one can generate event times according to
+\[ T \sim \dfrac{1}{\kappa_2 + b} \log \Big( 1 - \dfrac{(\kappa_2 + b)\log U}{\kappa_1 \kappa_2 e^a} \Big) \]
+with $U \sim \cU\big([0,1]\big)$.
+$\hfill\square$
+
 \bibliography{biblio}

### 2020-12-09 paper.tex
lights v7

@@
   \label{eq:law-t_i^max}
-  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
+  1 - t_i^{max} \sim T_i \times \text{Beta}(\nu_1, \nu_2)
 \end{equation}
@@
 
+\section{Multivariate linear mixed model}
+We re-write the conditional distribution of $y_i|b_i; \theta $ is in form
+\begin{align*}
+	f(y_i|b_i, \theta) &= \exp \big\{(y_i \odot \Phi_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \big\} \\
+	&= \exp \big\{( \Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \big\}
+\end{align*}
+where $\Sigma_i$ is a diagonal matrix whose diagonal is $\Phi_i$. And if the distribution $b_i; \theta$ is Gaussian, then its density form is
+
+\begin{equation*}
+  \label{eq:b_density}
+  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
+\end{equation*}
+\\
+The complete log-likelihood
+\begin{align*}
+  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}) \\ 
+  &= \text{log} (f(y, b ; \theta) \\ 
+  &= \sum_{i=1}^n \text{log}(f(y_i|b_i, \theta)) + \text{log}(f(b_i ; \theta))\\
+  &= \sum_{i=1}^n (\Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \\
+  &  \quad -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i).
+\end{align*}
+
+\paragraph*{E-step.}
+
+We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
+\begin{equation*}
+  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
+\end{equation*}
+Here, computing $\E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$  reduces to computing $E_{\theta^{(w)}}[b_i | y_i, \theta]$ and $E_{\theta^{(w)}}[b_ib_i^T | y_i, \theta]$ for $i = 1,..,n$\\
+\\
+If the marginal distributions of $y_i$ and $b_i$ are both Gaussian, then conditional distribution of $b_i$ is also Gaussian with a mean and a variance that can be computed. Indeed, from Bayes Theorem,
+\begin{align*}
+    f(b_i|y_i; \theta) &= \frac{f(y_i|b_i; \theta)f(b_i; \theta)}{f(y_i; \theta} \\
+    &= C_1 \text{ x } \exp \big\{( S_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \big\} \text{ x } \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}\\
+    &= C_2  \text{ x } \exp \big\{-\dfrac12 {b_i}^\top ({V_i}^\top \Sigma_iV_i)b_i + {b_i}^\top {V_i}^\top S(y_i - U_i\beta) \big\} \text{ x } \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}\\
+    &= C_2  \text{ x } \exp \big\{-\dfrac12 {b_i}^\top ({V_i}^\top \Sigma_iV_i + D^{-1})b_i + {b_i}^\top {V_i}^\top \Sigma_i(y_i - U_i\beta) \big\}\\
+    &= C_2 \text{ x } \exp \big\{ -\dfrac12 (b_i - \mu_i)^\top \Omega_i^{-1} (b_i - \mu_i) \big\}
+\end{align*}
+where,
+\begin{align*}
+    \Omega_i = ({V_i}^\top \Sigma_iV_i + D^{-1})^{-1} \text{  ;   } \mu_i = \Omega_i{V_i}^\top \Sigma_i(y_i - U_i\beta)
+\end{align*}
+Then,
+\begin{align*}
+    \E_{\theta^{(w)}}[b_i | y_i, \theta] &= \mu_i \\
+    \E_{\theta^{(w)}}[b_ib_i^T | y_i, \theta] &= \text{Var}_{\theta^{(w)}}[b_i | y_i, \theta] + \E_{\theta^{(w)}}[b_i | y_i, \theta]\E_{\theta^{(w)}}[b_i | y_i, \theta]^\top\\
+                                        &= \Omega_i + \mu_i\mu_i^\top
+\end{align*}
+
+\paragraph*{M-step.}
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) .\]
+The value of the gradient for the Gaussian case is here given by
+\begin{align*}
+  \label{eq:grad-beta}
+  \frac{\partial \cQ_n(\theta, \theta^{(w)})(\beta)}{\partial \beta} &= \sum_{i=1}^n 
+  \Big[ {y_i}^\top \Sigma U_{i} - U_{i}^\top \Sigma V_{i}\hat \E_{\theta^{(w)}}[ b_i | y_i; \theta]) - U_{i}^\top \Sigma U_{i}\beta \Big] \\
+  \frac{\partial \cQ_n(\theta, \theta^{(w)})(\phi)}{\partial \phi^l} &= \sum_{i=1}^n 
+  \dfrac12 {\phi_l}^{-2}\Big[ (y_i^l - U_{il}\beta_l)^\top (y_i^l - U_{il}\beta_l - 2V_{il}\hat \E_{\theta^{(w)}}[ b_i^l | y_i^l; \theta])) \\
+  & \quad \quad +  \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | y_i^l; \theta]\big) \Big] - \frac{n_i^l}{2} {\phi_l}^{-1}\\
+  \frac{\partial \cQ_n(\theta, \theta^{(w)})(D)}{\partial D} &= \sum_{i=1}^n 
+   \dfrac12 \big( D - \E_{\theta^{(w)}}[ b_i {b_i}^\top | y_i; \theta]\big)
+\end{align*}
+\\
+Then, the updated parameter $\theta^{(w)}$ will be
+\begin{align*}
+  \beta^{(w+1)} &= \Big(\sum_{i=1}^n U_{i}^\top U_{i}\Big)^{-1}  \Big( \sum_{i=1}^n \big[ U_{i}^\top  y_i - U_{i} V_{i}\hat \E_{\theta^{(w)}}[ b_i | y_i; \theta]) \big] \Big)\\
+  \phi_l^{(w+1)} &= \big(\sum_{i=1}^n n_i^l \big)^{-1}  \sum_{i=1}^n \Big[ (y_i^l - U_{il}\beta_l^{(w+1)})^\top (y_i^l - U_{il}\beta_l^{(w+1)} - 2V_{il}\hat \E_{\theta^{(w)}}[ b_i^l | y_i^l; \theta])) \\
+  & \hspace{3cm}+ \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | y_i^l; \theta]\big) \Big] \\
+  D^{(w+1)} &= n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i {b_i}^\top | y_i; \theta]
+\end{align*}
+
 \bibliography{biblio}

### 2020-12-09 paper.tex
lights v8

--- 2020-12-09 paper.tex
@@
 \documentclass[11pt]{article}
+
+\usepackage{xcolor}
+\definecolor{darkpastelgreen}{rgb}{0.01, 0.75, 0.24}
 
@@
             linkcolor=red,
-            urlcolor=blue,
+            urlcolor=darkpastelgreen,
             citecolor=blue,
@@
 \author[1,2]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
+\author[2]{Van Tuan Nguyen}
 \author[3]{Antoine Barbieri}
@@
 \usepackage{mathrsfs}
-\usepackage{xcolor}
 \usepackage{mathrsfs} 
@@
 the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
-\[ \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}} \]
+\begin{equation}
+  \label{eq:pi-softmax} 
+  \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}
+\end{equation}
 where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero.
@@
 the history of the true unobserved longitudinal process up to time $t$.
-We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independant features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independant regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
+We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
 Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
+\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,L\cA}^\top)= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
 
@@
 Let us denote 
-\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
-the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
+\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
+the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
 To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_{k,0}}_{\text{en}, \eta}\big) + \zeta_{2,k} \norm{\gamma_{k,L\cA}}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
 \end{equation}
@@
 \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
-Note that in the $\gamma_k$ penalty, $l$ actually starts at 0 (and not 1), but for simplicity we do not change the notation. 
 Hence, the resulting optimization problem is written
@@
 
-One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independant features involved in the $k$-th group membership.
+One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independent features involved in the $k$-th group membership.
 The ridge part allows to handle potential correlation between features.
@@
 Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
-\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
-for different functions $g$, where we denote 
+\[ \E_{\theta^{(w)}}[ g_i(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
+for different functions $g_i$, where we denote 
 \begin{equation}
@@
 the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
-
-In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
+Then, either $g_i(b_i, G_i) = \tilde g_i(b_i)$ (for instance $\tilde g_i: b_i \mapsto b_i^\top D^{-1}b_i$), or $g_i(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
+
+In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
 \begin{equation}
   \label{eq:def-E-g-tilde}
-  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
+  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
 \end{equation}
@@
   \label{eq:Lambda_ik-def}
-  \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
-\end{equation}
-and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
+  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
+\end{equation}
+and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
 The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice.
@@
   \label{eq:def-hat-E-g-tilde}
-  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
+  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
 \end{equation}
@@
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}
@@
     \begin{array}{ll}
-        \tilde g_{s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
+        \tilde g_{i, s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
         \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
@@
   R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
   & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
@@
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
@@
   \begin{split}
-    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 + \tilde \eta \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) \big) \\
+    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) + \tilde \eta \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 \big) \\
     \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
@@
   \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\ \nonumber
-  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\ \nonumber
-  &+ U_{il}^\top {\phi_l^{(w)}}^{-2}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\ \nonumber
+  &+ U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
 \end{align}
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
 The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
@@
   \begin{split}
-    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 + \tilde \eta \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) \big) \\
+    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) + \tilde \eta \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 \big) \\
     \text{subject to}& \quad \quad \gamma_{k,j}^+ \geq 0 \text{ and } \gamma_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p+L\mathcal{A}.
@@
   \label{eq:grad-Q-1}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
 \end{equation}
@@
   \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \Lambda_{ik,\tilde g_{t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1}^{\theta^{(w)}} \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}
@@
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
-Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using \texttt{R} package \texttt{survival}.
-Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm, being itself initialized with univariates fits using \texttt{R} package \texttt{nlme}.
+Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
+Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
 
@@
   \tcc{Monte Carlo E-step}
-  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
+  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
   Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
@@
 Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
-\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
+\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i, x_i].\]
 
@@
 We detail in this section the metric considered to evaluate risk prediction performances.
-Let us denote by $M$ the marker under study (e.g., the marker~\eqref{eq:marker-def} for the lights model), and assume that it is measured once at $t = t^{max}$.
+Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}.
+Then, let us denote by $M$ the marker under study (e.g., the marker~\eqref{eq:marker-def} for the lights model), and assume that it is measured once at $t = t^{max}$.
 A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
@@
 
-\paragraph*{Penalized Cox model with time-independant features.}
-
-The first model we consider as a baseline is the well known Cox PH model with time-independant features.
+\paragraph*{Penalized Cox model with time-independent features.}
+
+The first model we consider as a baseline is the well known Cox PH model with time-independent features.
 In this model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
@@
 Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.
-We also choose to include basic time-independant features extracted from longitudinal processes, namely the average value for each time-dependant feature.
+We also choose to include basic time-independent features extracted from longitudinal processes, namely the average value for each time-dependant feature.
 
@@
 
-In order to assess the proposed method, we run an extensive Monte Carlo simulation study in the case $K=2$ (see Section~\ref{sec:K=2}).
-We first generate $G_i \sim \cB\big(\pi_1\big)$, with $\pi_1 \in [0,1]$ a desired high-risk subjects proportion and where $\cB(\alpha)$ denotes the Bernoulli distribution with parameter $\alpha \in [0,1]$.
-We then choose a coefficient vector 
+In order to assess the proposed method, we run an extensive Monte Carlo simulation study in the case $K=2$ (see Section~\ref{sec:K=2}). We first choose a coefficient vector 
 \begin{equation}
@@
 with $\varsigma_1\in\R$ being the value of the active coefficients and $s\in \{1,\dots,p\}$ a sparsity parameter chosen such that a proportion $r_s$ of time-independent features are actually active, that is $s = \lfloor p \times r_s \rfloor$.
+For a desired high-risk subjects proportion $\pi_1 \in [0,1]$, the high-risk subjects index set is given by
+\[\cH = \big\{\lfloor \pi_1 n \rfloor \text{ random samples without replacement} \big\} \subset \{1, \ldots,n\}.\]
 For the generation of the time-independent features matrix, we take 
 \[ [x_{ij}] \in \R^{n \times p} \sim \cN \big(0, \bSigma_1(\rho_1)\big),\] 
-with $\bSigma_1(\rho_1)$ a $(p \times p)$ Toeplitz covariance matrix \citep{mukherjee1988some} with correlation $\rho_1 \in (0, 1)$. 
-We then add a $gap \in \R^+$ value for high-risk subjects and subtract it for low-risk subjects, only on active features, that is
+with $\bSigma_1(\rho_1)$ a $(p \times p)$ Toeplitz covariance matrix \citep{mukherjee1988some} with correlation $\rho_1 \in (0, 1)$, that is $\bSigma_1(\rho_1)_{jj'} = \rho_1^{|j - j'|}$. 
+We then add a $gap \in \R^+$ value for subjects $i \in \cH$ and subtract it for subjects $i \notin \cH$, only on active features, that is
 \begin{center}
-$x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{G_i=0\}}} gap\ \text{for } j = 1, \dots, s$.
+$x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{i \notin \cH\}}} gap\ \text{for } j = 1, \dots, s$.
 \end{center}
-Note that this is equivalent to generate the time-independent features according to a gaussian mixture.
+Note that this is equivalent to generate the time-independent features according to a gaussian mixture. We finally normalize each time-independent features for practical interpretations.
+Then, we generate $G_i \sim \cB\big(\pi_\xi(x_i)\big)$, where $\pi_\xi(x_i)$ is computed given Equation~\eqref{eq:pi-softmax} and $\cB(\alpha)$ denotes the Bernoulli distribution with parameter $\alpha \in [0,1]$.
 
@@
 $\Upsilon_{l} \geq 0$ is called baseline intensity, and corresponds to the exogenous probability of having a measurement for process $l$.
-We sample $\Upsilon_{l} \sim \cU\big([0.1,0.5]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
+We sample $\Upsilon_{l} \sim \cU\big([0.1,1]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
 The matrix $A = [A_{ll'}]_{1 \leq l,l' \leq L}$ is the adjacency matrix such that $A_{ll'} \geq 0$ quantifies the impact of past measurement time of process $l'$ on the measurement time of process $l$, and $\upsilon \geq 0$ is a memory parameter.
-We sample $A_{ll'} \sim \cU\big([0.05,0.1]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
+We sample $A_{ll'} \sim \cU\big([0.1,0.2]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
 Simulation of the Hawkes processes
@@
 \end{pmatrix}$ 
-& 0.5 & 0.5 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
+& 0.5 & 4 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
 \bottomrule
@@
 \section*{Software}
-All the methodology discussed in this paper is implemented in \texttt{Python}. The code is available from \href{https://github.com/Califrais/lights}%
-{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.
+All the methodology discussed in this paper is implemented in \texttt{Python}.
+The code is available from \url{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.
 
@@
 
+\section{Multivariate linear mixed model}
+\label{sec:MLMM}
+
+Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the QNMCEM algorithm in Section~\ref{sec:QNMCEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
+For the sake of simplicity, let us denote here 
+\[\theta = (\beta^\top, \text{vech}(D), \phi^\top)^\top \in \R^\vartheta\]
+the parameter vector to infer.
+The conditional distribution of $y_i|b_i$ then writes
+\[f(y_i|b_i ; \theta) = \exp \big\{( \Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \big\},\]
+where $\Sigma_i$ is the diagonal matrix whose diagonal is $\Phi_i$. 
+The complete log-likelihood in this context writes
+\begin{align*}
+\ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}) \\
+&= \sum_{i=1}^n (\Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) -\dfrac12 \big(r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i\big).
+\end{align*}
+
+\paragraph*{E-step.}
+Supposing that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$, we need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters, which is given by
+\[\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].\]
+Here, computing this quantity reduces to computing $\E_{\theta^{(w)}}[b_i | y_i]$ and $\E_{\theta^{(w)}}[b_ib_i^T | y_i]$ for $i = 1,\ldots,n$.
+The marginal distributions of $y_i$ and $b_i$ being both Gaussian, one has from Bayes Theorem
+\[ f(b_i|y_i; \theta) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i)^\top \Omega_i^{-1} (b_i - \mu_i) \big\}\]
+where
+\[\Omega_i = ({V_i}^\top \Sigma_iV_i + D^{-1})^{-1} \quad \text{ and } \quad \mu_i = \Omega_i{V_i}^\top \Sigma_i(y_i - U_i\beta).\]
+Then, on has
+\begin{center}
+  $\left\{
+    \begin{aligned}
+      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i, \\
+      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i + \mu_i\mu_i^\top.
+    \end{aligned}
+    \right.$
+\end{center}
+
+\paragraph*{M-step.}
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) .\]
+The parameters updates are then naturally given in closed form by zeroing the gradient. One obtains
+\[\beta^{(w+1)} = \Big(\sum_{i=1}^n U_{i}^\top U_{i}\Big)^{-1}  \sum_{i=1}^n \big[ U_{i}^\top  y_i - U_{i} V_{i} \E_{\theta^{(w)}}[ b_i | y_i]) \big],\]
+\begin{align*}
+\phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1}  \sum_{i=1}^n \Big[(&y_i^l - U_{il}\beta_l^{(w+1)})^\top (y_i^l - U_{il}\beta_l^{(w+1)} - 2V_{il}\E_{\theta^{(w)}}[b_i^l | y_i^l])) \\
+&+\Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | y_i^l]\big) \Big]
+\end{align*}
+and \[D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i {b_i}^\top | y_i].\]
+
 \bibliography{biblio}

### 2021-01-15 paper.tex
update competing models

@@
 
-We detail in this section the metric considered to evaluate risk prediction performances.
-Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}.
-Then, let us denote by $M$ the marker under study (e.g., the marker~\eqref{eq:marker-def} for the lights model), and assume that it is measured once at $t = t^{max}$.
+We detail in this section the metric considered to evaluate risk prediction performances. Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}. 
+In this context, the predictive marker for the lights model is
+\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}.\]
+Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
 A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
 \begin{equation*}
-  \cC =\P[M_i > M_j | T^\star_i < T^\star_j],
+  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
 \end{equation*}
@@
 \begin{equation*}
-  \cC_\tau =\P[M_i > M_j | T_i < T_j , T_i < \tau],
+  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
 \end{equation*}
@@
 In this section, we briefly introduce the models we consider for performance comparisons in the simulation study as well as in the applications on real datasets in Section~\ref{sec:application}.
-
-\paragraph*{Penalized Cox model with time-independent features.}
-
-The first model we consider as a baseline is the well known Cox PH model with time-independent features.
-In this model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
+We also give the corresponding marker rule for each model, using our notations.
+
+\paragraph*{Landmark Cox model.}
+
+The first model we consider as a baseline is the well known Cox PH model with time-independent features, also known in the joint modeling context as the ``landmark'' model, in which we include basic time-independent features extracted from longitudinal processes and computed at the prediction time [REF].
+In the Cox PH model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
 \begin{equation*}
@@
 \end{equation*}
-We use respectively the \texttt{R} packages \texttt{survival} and \texttt{glmnet} \citep{simon2011regularization} for the partial log-likelihood and the minimization of the following quantity
-\begin{equation*}
-  - \ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big),
-\end{equation*}
+For each time-dependant feature, we include subject-specific random effects, values of longitudinal processes at baseline and at time $t^{max}$, slope of longitudinal processes at time $t^{max}$ and area under of longitudinal processes from baseline up to $t^{max}$.
+The predictive marker for a new subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with
+\[\hat\beta^{\text{cox}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta).\]
+We use the \texttt{Python} package \texttt{tick} for the estimation of the previous quantity.
+Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.
+
+\paragraph*{Penalized Cox model.}
+
+We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}.
+The predictive marker for the a subject $i$ is $\hat \cR_i^{\text{cox EN}} = \exp (x_i^\top\hat\beta^{\text{cox EN}})$, with 
+\[\hat\beta^{\text{cox EN}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big)\] 
+the estimate obtained using elastic net penalization (still with the \texttt{tick} package), and
 where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$.
-Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.
-We also choose to include basic time-independent features extracted from longitudinal processes, namely the average value for each time-dependant feature.
 
@@
 
-\paragraph*{Multivariate joint modeling.}
-
-\texttt{joineRML} package : in this \texttt{R} package~\citep{hickey2018joinerml}, blabla description...
-
-The \texttt{JM} package : in this \texttt{R} package~\citep{rizopoulos2010jm}, blabla description...
-
-The \texttt{JMbayes} package : n this \texttt{R} package~\citep{rizopoulos2014r}, blabla description...
-
-% TODO: voir slide 132 \url{http://www.drizopoulos.com/courses/Int/JMwithR_CEN-ISBS_2017.pdf})
-
-The \texttt{joineR} package : in this \texttt{R} package~\citep{philipson2018package}, blabla description...
+\paragraph*{Multivariate joint latent class model.}
+
+We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
+Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model.
+Given the group membership, each submodel is assumed to be independent.
+In this context, the predictive marker is
+\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) \hat f(t^{max}_i, y_i | b_i, G_i=0 ; \hat \theta)}.\]
+
+\paragraph*{Multivariate shared random effect model.}
+
+In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}.
+For a new subject $i$, the predictive marker associated with this model is naturally given by the corresponding risk score
+\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta_k^l, b_i^l) \Big\}.\]
+
+\texttt{@Antoine : est-ce qu'il peut prendre des time-indep feat ? et quelles sont les assos ?}
 

### 2021-01-15 paper.tex
update markers

@@
   \label{eq:marker-def}
-  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)},
+  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
 \end{equation}
 with 
-\[ \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
+\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
 where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
@@
 In this context, the predictive marker for the lights model is
-\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}.\]
+\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
 Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
@@
 For each time-dependant feature, we include subject-specific random effects, values of longitudinal processes at baseline and at time $t^{max}$, slope of longitudinal processes at time $t^{max}$ and area under of longitudinal processes from baseline up to $t^{max}$.
-The predictive marker for a new subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with
+The predictive marker for subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with
 \[\hat\beta^{\text{cox}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta).\]
@@
 
-We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}.
+We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}~\citep{christ2018time}.
 The predictive marker for the a subject $i$ is $\hat \cR_i^{\text{cox EN}} = \exp (x_i^\top\hat\beta^{\text{cox EN}})$, with 
@@
 A classical extension of the Cox model supposes that features depend on time~\citep{sueyoshi1992semiparametric} (and then loosing the risk proportionality property), that is with our notations
-\[\lambda_i(t) = \lambda_0(t) \exp\big(y_i(t)^\top \beta \big). \]
-Using the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, blabla description...
+\[\lambda_i(t) = \lambda_0(t) \exp\big(x_i^\top \gamma_0 + y_i(t_i^{max})^\top \beta \big). \]
+We use the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, and we consider the natural predictive marker for a subject $i$, that is the corresponding risk score given by
+\[\hat \cR_i^\text{cox t-d} = \exp \big(x_i^\top \hat\gamma_0 + y_i(t)^\top \hat\beta \big).\]
 
@@
 Given the group membership, each submodel is assumed to be independent.
-In this context, the predictive marker is
-\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) \hat f(t^{max}_i, y_i | b_i, G_i=0 ; \hat \theta)}.\]
+In this context, the predictive marker for subject $i$ is
+\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
 
@@
 
-In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}.
-For a new subject $i$, the predictive marker associated with this model is naturally given by the corresponding risk score
-\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta_k^l, b_i^l) \Big\}.\]
+In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
+\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta^l, b_i^l) \Big\}.\]
 

### 2021-01-18 paper.tex
add missing steps in MLMM
technical docu for the MLMM class

@@
 The code is available from \url{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.
+A technical documentation is also provided in Appendix~\ref{sec:package}.
 
@@
 The marginal distributions of $y_i$ and $b_i$ being both Gaussian, one has from Bayes Theorem
-\[ f(b_i|y_i; \theta) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i)^\top \Omega_i^{-1} (b_i - \mu_i) \big\}\]
+\[ f(b_i|y_i; \theta^{(w)}) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i^{(w)})^\top {\Omega_i^{(w)}}^{-1} (b_i - \mu_i^{(w)}) \big\}\]
 where
-\[\Omega_i = ({V_i}^\top \Sigma_iV_i + D^{-1})^{-1} \quad \text{ and } \quad \mu_i = \Omega_i{V_i}^\top \Sigma_i(y_i - U_i\beta).\]
+\[\Omega_i^{(w)} = ({V_i}^\top \Sigma_i^{(w)}V_i + {D^{(w)}}^{-1})^{-1} \quad \text{ and } \quad \mu_i^{(w)} = \Omega_i^{(w)}{V_i}^\top \Sigma_i^{(w)}(y_i - U_i\beta^{(w)}).\]
 Then, on has
@@
     \begin{aligned}
-      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i, \\
-      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i + \mu_i\mu_i^\top.
+      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i^{(w)}, \\
+      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i^{(w)} + \mu_i^{(w)}{\mu_i^{(w)}}^\top.
     \end{aligned}
@@
 
+\section{Technical documentation of the \texttt{lights} package}
+\label{sec:package}
+
+\subsection{\texttt{MLMM class}}
+
+This class implements an EM algorithm for fitting a multivariate linear mixed model used to initialize the MCQNEM algorithm.
+Let us introduce the list $\Omega^{(w)} = [\Omega_1^{(w)}, \ldots, \Omega_n^{(w)}]$, the matrices $\mu = [\mu_1, \ldots, \mu_n] \in \R^{r \times n}$, $U^l = [{U_{1l}}^\top \cdots {U_{nl}}^\top]^\top \in \R^{n_l \times q_l}$, $U = [U_1^\top \cdots U_n^\top]^\top \in \R^{\cN \times q}$,
+\[ V^l = \begin{bmatrix}
+    V_{1l} & \cdots & 0\\
+    \vdots &  \ddots & \vdots \\
+    0 & \cdots & V_{nl}
+\end{bmatrix},
+\quad
+V = \begin{bmatrix}
+  V_{1} & \cdots & 0\\
+  \vdots &  \ddots & \vdots \\
+  0 & \cdots & V_{n}
+\end{bmatrix}
+\quad \text{ and } \quad
+{\Omega^l}^{(w)} = \begin{bmatrix}
+    {\Omega_{1}^l}^{(w)} & \cdots & 0\\
+    \vdots &  \ddots & \vdots \\
+    0 & \cdots & {\Omega_{n}^l}^{(w)}
+\end{bmatrix}\]
+that belong respectively in $\R^{n_l \times nr_l}$, $\R^{\cN \times nr}$ and $\R^{nr_l \times nr_l}$, as well as the vectors $\tilde \mu^{(w)} = ({\mu_1^{(w)}}^\top \cdots {\mu_n^\top)^{(w)}}^\top \in \R^{nr}$, $(\tilde \mu^l)^{(w)} = \big({(\mu_1^l)^{(w)}}^\top \cdots {(\mu_n^l)^{(w)}}^\top \big)^\top \in \R^{nr_l}$, $y^l = ({y_1^l}^\top  \cdots {y_n^l}^\top)^\top \in \R^{n_l}$
+with $n_l = \sum_{i=1}^{n} n_i^l$ and $y = (y_1^\top \cdots y_n^\top)^\top \in \R^\cN$ with $\cN = \sum_{i=1}^{n} n_i$.
+The $\beta$ update then rewrites
+\[\beta^{(w+1)} = ({U}^\top U)^{-1}{U}^\top(y-V \tilde \mu^{(w)}).\]
+For the $D$ update, one has
+\[D^{(w+1)} = n^{-1} \big( \text{sum}(\Omega^{(w)}) + \mu^{(w)}{\mu^{(w)}}^\top\big).\]
+And finally for the $\phi$ update, one has
+\begin{align*}
+\phi_l^{(w+1)} = n_l^{-1} \big[&(y^l - U^l{\beta_l^{(w+1)})}^\top \big(y^l - U^l\beta_l^{(w+1)} - 2 V^l (\tilde\mu^l)^{(w)}\big) \\
+&+ \text{Tr}\big\{{V^l}^\top V^l \big({\Omega^l}^{(w)} + (\tilde\mu^l)^{(w)} {(\tilde\mu^l)^{(w)}}^\top\big)\big\} \big].
+\end{align*}
+
 \bibliography{biblio}

### 2021-03-04 paper.tex
fix error in lambda_0 update
Update the complete log likelihood formula
Update the expected Lambda and Monte-Carlo sample set

@@
 	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
-	&= n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i,
+	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
 \end{align}
@@
 and
-\begin{equation}
+\begin{align}
   \label{eq:b_density}
-  f(b_i ; \theta) = (2 \pi)^{-\dfrac r2} |D|^{-\dfrac12} \exp \Big\{ -\dfrac12 b_i^\top D^{-1}b_i \Big\}.
-\end{equation}
+  b_i | y_i, G_i=k; \theta &\sim \cN\Big((V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}V_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}\Big) \nonumber \\
+  y_i | G_i = k; \theta &\sim \cN\Big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\Big)
+\end{align}
 
@@
   \label{eq:Lambda_ik-def}
-  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i,
+  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i,
 \end{equation}
@@
 \DontPrintSemicolon
-\caption{Construction of the samples in $S^{(w)}$}
+\caption{Construction of the samples in $S_i^{(w)}$}
 \label{alg:construct-set-S}
-\KwInput{$S^{(w)} = \varnothing$}
-\KwOutput{$S^{(w)}$ filled with $2N$ samples}
-Compute $C^{(w)} \in \R^{r \times r}$ such that $C^{(w)}{C^{(w)}}^\top = D^{(w)}$ \tcp{Cholesky decomposition}
+\KwInput{$S_i^{(w)} = \varnothing$}
+\KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
+Compute $C_i^{(w)} \in \R^{r \times r}$ such that $C_i^{(w)}{C_i^{(w)}}^\top = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$ \tcp{Cholesky decomposition}
 \For{$\breve{n} = 1,\ldots,N$}    
@@
   Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
-  Compute $\breve{b}_{\breve{n}} = C^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
-  Update $S^{(w)} \leftarrow S^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
+  Compute $\breve{b}_{\breve{n}} = C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
+  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
 }
-\textbf{Return:} {$S^{(w)}$}
+\textbf{Return:} {$S_i^{(w)}$}
 \end{algorithm}
@@
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_i^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}

### 2021-03-08 paper.tex
Update the condition of increasing number of Monte Carlo samples

@@
 Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
+The $Condition$ in algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w)})$ is coefficient of variation at the $(w + 1)$-th iteration computed as
+\begin{equation*}
+  \label{eq:coefficient_variation}
+  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
+\end{equation*}
+where sd(.) and mean(.) are the sample standard deviation and mean functions respectively and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
+\begin{equation*}
+  \label{eq:relative_difference}
+  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}
+\end{equation*}
 

### 2021-03-15 paper.tex
update b_i | y_i and add ref
ISTA update pour gamma

@@
 \newtheorem{assumption}{Assumption}{\bf}{\rm} 
+\newtheorem{lemma}{Lemma}
 
@@
 	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
+	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
 \end{equation}
@@
 We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
-Corresponding fixed effects are denoted $\gamma_{k,0} \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
+Corresponding fixed effects are denoted $\gamma_k^0 \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[\gamma_k= (\gamma_{k,0}^\top, \gamma_{k,L\cA}^\top)= (\gamma_{k,0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
+\[({\gamma_k^0}^\top, \gamma_k^\top)^\top= ({\gamma_k^0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
 
@@
 where 
-\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\} \]
-and
-\begin{align}
-  \label{eq:b_density}
-  b_i | y_i, G_i=k; \theta &\sim \cN\Big((V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}V_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}\Big) \nonumber \\
-  y_i | G_i = k; \theta &\sim \cN\Big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\Big)
-\end{align}
+\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}. \]
 
@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_{k,0}}_{\text{en}, \eta}\big) + \zeta_{2,k} \norm{\gamma_{k,L\cA}}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta}\big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
 \end{equation}
@@
 for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \] for the sparse group lasso penalty, for which we do not have to account for group sizes since they all have the same one.
+\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty with $\norm{z}_{2,l}=\norm{z^l}_2$, for which we do not have to account for group sizes since they all have the same one.
 Hence, the resulting optimization problem is written
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
-  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
+  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
 and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
+
+in the Gaussian case used in practice, one can also sample from
+\begin{align*}
+  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
+  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
+\end{align*}
+with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$
+
 The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice.
@@
 \KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
-Compute $C_i^{(w)} \in \R^{r \times r}$ such that $C_i^{(w)}{C_i^{(w)}}^\top = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$ \tcp{Cholesky decomposition}
+Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
+$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i^{-1}$ \tcp{Cholesky decomposition} \nonumber
 \For{$\breve{n} = 1,\ldots,N$}    
@@
 
+$\text{M}_i= V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1}$ or $\text{M}_i= D^{(w)}$
+
+\texttt{@VT: weird that $D^{(w)}$ and ${D^{(w)}}^{-1}$ ?}
+
 \noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
@@
 \paragraph*{Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta} \big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
 Let us introduce the following useful notations
@@
     \begin{array}{ll}
-        \tilde g_{i, s, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_{k,0} + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
+        \tilde g_{i, s, \gamma_k^0, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
         \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
@@
 \end{equation}
+In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}.
 Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
@@
   R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
   & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
@@
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
-Hence, we update $\beta_k^{(w)}$ by solving the following problem
-\begin{equation}
-  \label{eq:sub-problem-beta_k}
-  \begin{split}
-    \text{minimize}& \quad \quad R_{n, k}^{(w)}(\beta_k^+ - \beta_k^-) + \zeta_{3,k} \big((1 - \tilde \eta) \sum_{j=1}^q (\beta_{k,j}^+ + \beta_{k,j}^-) + \tilde \eta \sum_{l=1}^L \norm{{\beta_k^l}^+ - {\beta_k^l}^-}_2 \big) \\
-    \text{subject to}& \quad \quad \beta_{k,j}^+ \geq 0 \text{ and } \beta_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, q.
-  \end{split} 
-\end{equation}
-The value of the gradient for the Gaussian case is here given by
+The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\begin{equation}
+  \label{eq:minimization-beta_k}
+  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
+\end{equation}
+The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
+Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}.
+
+\vspace{.5cm}
+
+\begin{algorithm}[H]
+\DontPrintSemicolon
+\caption{Proximal gradient descent ISTA$_{\text{sg} l_1}$($v$, $f$, $\tilde \eta$, $\zeta$)}
+\label{alg:ISTA}
+\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$ and $\zeta$}
+\KwOutput{$\flat^{\breve{w}^{max}}$}
+Initialize $\flat^{(0)}=v$ \\
+\For{$\breve{w} = 0,\ldots,\breve{w}^{max}$}    
+{ 
+  \vspace{-.8cm}
+  \begin{align*}
+  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
+  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
+  \end{align*}
+  \vspace{-.3cm}
+}
+\textbf{Return:} {$\flat^{\breve{w}^{max}}$}
+\end{algorithm}
+
+\vspace{.5cm}
+TODO w/ VT: note on $\breve{w}^{max}$. Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
+\begin{lemma}
+\label{lemma:prox-sgl1}
+The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
+\normalfont
+\begin{equation*}
+  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
+\end{equation*}
+\end{lemma}
+The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable.
+We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
+A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
+Hence, we update $\beta_k^{(w)}$ with the following step
+\begin{equation}
+  \label{eq:beta-update}
+  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
+\end{equation}
+Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
+The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
 \begin{align}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \\ \nonumber
-  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \\ \nonumber
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \nonumber \\
+  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \nonumber \\
   &+ U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
@@
 
-Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
+Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
+  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
-The update for $\gamma_k^{(w)}$ therefore requires to solve the following minimization problem
-\begin{equation}
-  \label{eq:minimization-beta_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{p+L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
-\end{equation}
-We then rewrite problem~\eqref{eq:minimization-beta_k} as the following differentiable problem with box constraints
-\begin{equation}
-  \label{eq:sub-problem-gamma_k}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
+\begin{equation}
+  \label{eq:minimization-gamma_k^0}
+  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
+\end{equation}
+We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
+\begin{equation}
+  \label{eq:sub-problem-gamma_k^0}
   \begin{split}
-    \text{minimize}& \quad \quad Q_{n, k}^{(w)}(\gamma_k^+ - \gamma_k^-) + \zeta_{2,k} \big((1 - \tilde \eta) \sum_{j=1}^{p+L\mathcal{A}} (\gamma_{k,j}^+ + \gamma_{k,j}^-) + \tilde \eta \sum_{l=1}^L \norm{{\gamma_k^l}^+ - {\gamma_k^l}^-}_2 \big) \\
-    \text{subject to}& \quad \quad \gamma_{k,j}^+ \geq 0 \text{ and } \gamma_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p+L\mathcal{A}.
+    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
+    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
   \end{split} 
 \end{equation}
-One can also compute the gradient since one has
-\begin{equation}
+The gradient is here given by
+\begin{align}
   \label{eq:grad-Q-1}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,0}} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i
-\end{equation}
-and
+  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
+  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
+\end{align}
+Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
+\begin{equation}
+  \label{eq:minimization-gamma_k}
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+\end{equation}
+We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
+\begin{equation}
+  \label{eq:gamma-update}
+  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
+\end{equation}
+with the gradient being known since
 \begin{align}
   \label{eq:grad-Q-2}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  \frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w)}, \gamma_k)}{\partial \gamma_{k,a}^l} &= -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t_i, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}
@@
 \end{align}
+Appendix~\ref{sec:?} gives some details about the way we implement the equations and their computation in practice.
 
@@
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
-Then, we initialize $\lambda_0^{(0)}$ and $\gamma_{k,0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
+Then, we initialize $\lambda_0^{(0)}$ and ${\gamma_k^0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
 Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
@@
   Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
-  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-beta_k} \\
-  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k} \\
+  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
+  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
+  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
   Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
@@
 \end{align*}
-which corresponds to~\eqref{eq:intensity-model} taking $\gamma_{k,0} = \xi$ and the first three functionals described in Table~\ref{table:shared_associations} as shared associations. 
+which corresponds to~\eqref{eq:intensity-model} taking $\gamma_k^0 = \xi$ and the first three functionals described in Table~\ref{table:shared_associations} as shared associations. 
 Now, for the choice of the association parameters, we want to induce sparsity (similarly to~\eqref{eq:xi-sparse}), so we set 
@@
 
+\section{Proof of Lemma~\ref{lemma:prox-sgl1}}
+\label{sec:proof-lemma-prox-sgl1}
+
 \section{Automatic cross-validation grid for $\xi$}
@@
 let us denote $\zeta_{1,k}^1 \leq \zeta_{1,k}^2 \leq \dots \leq \zeta_{1,k}^{\text{max}}$ the randomly chosen candidate values for $\zeta_{1,k}$, such that at $\zeta_{1,k}^{\text{max}}$, all coefficients $\hat \xi_{k, j}$ for all $j = 1, \dots, p$ are exactly zero.
-The KKT conditions \citep{boyd2004convex} claim that
+The KKT conditions~\citep{boyd2004convex} claim that
 \begin{center}

### 2021-03-16 paper.tex
2 versions of \Lambda_ik

@@
 with
-\begin{equation}
-  \label{eq:Lambda_ik-def}
-  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i,
-\end{equation}
+\begin{align}
+  \label{eq:Lambda_ik-b_i|y_i}
+  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i \\
+  \label{eq:Lambda_ik-b_i}
+  &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i
+\end{align}
 and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
@@
 Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
-$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i^{-1}$ \tcp{Cholesky decomposition} \nonumber
+$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i$ \tcp{Cholesky decomposition} \nonumber
 \For{$\breve{n} = 1,\ldots,N$}    
@@
 
-$\text{M}_i= V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1}$ or $\text{M}_i= D^{(w)}$
-
-\texttt{@VT: weird that $D^{(w)}$ and ${D^{(w)}}^{-1}$ ?}
+$\text{M}_i= W_i$ or $\text{M}_i= D^{(w)}$
 
@@
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
@@
 { 
-  \vspace{-.8cm}
-  \begin{align*}
+  $\begin{aligned}
   \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
   &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
-  \end{align*}
-  \vspace{-.3cm}
+  \end{aligned}$
 }
@@
 \vspace{.5cm}
-TODO w/ VT: note on $\breve{w}^{max}$. Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
+TODO w/ VT: note on $\breve{w}^{max}$ and $\varrho_{\breve{w}}$.
+Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
 \begin{lemma}

### 2021-03-18 paper.tex
{\gamma_k^0}^{(w+1)} for the \gamma_k update
Update the proximal gradient descent expression
Add back-tracking line search in ISTA algorithm

@@
 \usepackage{subfigure}
-
+\usepackage{algpseudocode}
 \usepackage[linesnumbered,ruled,vlined]{algorithm2e}
@@
 \KwOutput{$\flat^{\breve{w}^{max}}$}
-Initialize $\flat^{(0)}=v$ \\
+Define $p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho \dfrac{\partial f}{\partial u}(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}$\\
+Initialize $\flat^{(0)}=v$, $\varrho_0 > 0$, some $\epsilon > 1$ \\
 \For{$\breve{w} = 0,\ldots,\breve{w}^{max}$}    
-{ 
+{   Find the smallest nonnegative integer $i^{\breve{w}}$ such that with $\tilde{\varrho}_{\breve{w}} = \epsilon^{i^{\breve{w}}}\varrho_{\breve{w}}$
+	$f(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})})) \leq f(\flat^{(\breve{w})}) + \big<\big(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}\big), \dfrac{\partial f}{\partial u}(\flat^{(\breve{w})})\big> + \dfrac{1}{2\tilde{\varrho}_{\breve{w}}}\norm{p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}}_2^2$\\
+  Set $\varrho_{(\breve{w} + 1)} = \tilde{\epsilon}^{i^{\breve{w}}}  \varrho_{\breve{w}}$
   $\begin{aligned}
-  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
-  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
+  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{(\breve{w} + 1)}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{(\breve{w} + 1)} \dfrac{\partial f}{\partial u}(\flat^{(\breve{w})}) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
+  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{(\breve{w} + 1)} \dfrac{\partial f}{\partial u}(\flat^{(\breve{w})}) \Big)
   \end{aligned}$
@@
   \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
 \end{equation}
@@
   \label{eq:gamma-update}
-  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
+  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
 \end{equation}
@@
   \label{eq:grad-Q-2}
-  \frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w)}, \gamma_k)}{\partial \gamma_{k,a}^l} &= -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}

### 2021-03-21 paper.tex
Add stopping criterion in ISTA
algo 2 update ongoing

--- 2021-03-16 paper.tex
@@
 Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
-$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i$ \tcp{Cholesky decomposition} \nonumber
+$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i^1$ \tcp{Cholesky decomposition} \nonumber
 \For{$\breve{n} = 1,\ldots,N$}    
@@
   Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
-  Compute $\breve{b}_{\breve{n}} = C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
-  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
+  Compute $\breve{b}_{\breve{n}}^\pm = \text{M}_{ik}^2 \pm C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
+  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}^+, \breve{b}_{\breve{n}}^- \}$
 }
@@
 
-$\text{M}_i= W_i$ or $\text{M}_i= D^{(w)}$
+$\text{M}_i^1 = W_i$ $\text{M}_{ik}^2 = W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k)$ or $\text{M}_i^1 = D^{(w)}$ $\text{M}_{ik}^2 = 0$
 

### 2021-03-21 paper.tex

--- 2021-03-18 paper.tex
@@
 \label{alg:ISTA}
-\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$ and $\zeta$}
+\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$, $tol$ and $\zeta$}
 \KwOutput{$\flat^{\breve{w}^{max}}$}
@@
   \end{aligned}$
+	\If{$\dfrac{\norm{\flat^{(\breve{w}+1)} - \flat^{(\breve{w})}}_2}{\varrho_{(\breve{w} + 1)}} < tol$}
+	{
+		Convergence
+	}
 }

### 2021-03-21 paper.tex
update ISTA

--- 2021-03-21 paper.tex
@@
 \label{alg:ISTA}
-\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$ and $\zeta$}
+\KwInput{A vector $v$, a function $f$ (its gradient $\nabla f$), hyper-parameters $\tilde \eta$, $tol$ and $\zeta$}
 \KwOutput{$\flat^{\breve{w}^{max}}$}
-Initialize $\flat^{(0)}=v$ \\
-\For{$\breve{w} = 0,\ldots,\breve{w}^{max}$}    
+Define $p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}$\\
+Initialize $\flat^{(0)}=v$, $\varrho^0 > 0$, some $\epsilon > 1$ \\
+\For{$\breve{w} = 0,\ldots,$ until convergence}    
+{Backtracking line search : find the smallest nonnegative integer $i^{\breve{w}}$ such that
+$f\big(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})})\big) \leq f(\flat^{(\breve{w})}) + \big<\big(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}\big), \nabla f(\flat^{(\breve{w})})\big> + \dfrac{1}{2\tilde{\varrho}_{\breve{w}}}\norm{p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}}_2^2$
+with $\tilde{\varrho}_{\breve{w}} = \epsilon^{i^{\breve{w}}}\varrho^{\breve{w}}$\\
+Set $\varrho^{(\breve{w} + 1)} = \tilde{\varrho}_{\breve{w}}$ \\
+$\flat^{(\breve{w}+1)} = p_{\varrho^{(\breve{w} + 1)}}(\flat^{(\breve{w})}) \coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \big(\flat^{(\breve{w})} - \varrho^{(\breve{w} + 1)} \nabla f(\flat^{(\breve{w})}) \big)$ \\
+\If{$\dfrac{\norm{\flat^{(\breve{w}+1)} - \flat^{(\breve{w})}}_2}{\varrho^{(\breve{w} + 1)}} < tol$}
 	{
-  $\begin{aligned}
-  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
-  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
-  \end{aligned}$
+    \textbf{Return:} {$\flat^{\breve{w}}$}
 	}
-\textbf{Return:} {$\flat^{\breve{w}^{max}}$}
+}
 \end{algorithm}
@@
 \vspace{.5cm}
-TODO w/ VT: note on $\breve{w}^{max}$ and $\varrho_{\breve{w}}$.
 Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.

### 2021-03-21 paper.tex
define p_{\varrho}(y) outside the algo

@@
 The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
-Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}.
+Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
+\[p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}.\]
 
@@
 \KwOutput{$\flat^{\breve{w}^{max}}$}
-Define $p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}$\\
 Initialize $\flat^{(0)}=v$, $\varrho^0 > 0$, some $\epsilon > 1$ \\

### 2021-04-08 paper.tex
precisions on the 2 sampling options

@@
   \label{eq:Lambda_ik-b_i|y_i}
-  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i \\
+  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
   \label{eq:Lambda_ik-b_i}
-  &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i
+  &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
 \end{align}
 and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
-
-in the Gaussian case used in practice, one can also sample from
-\begin{align*}
-  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
-  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
-\end{align*}
-with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$
-
-The integral in~\eqref{eq:Lambda_ik-def} is not tractable analytically, and some form of approximation must be used in practice.
+The integral in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} is not tractable analytically, and some form of approximation must be used in practice.
 Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context.
 We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result.
-Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S^{(w)}$ used for the approximation.
+Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S_{ik}^{(w)}$ used for the approximation.
 
@@
 \DontPrintSemicolon
-\caption{Construction of the samples in $S_i^{(w)}$}
+\caption{Construction of the samples in $S_{ik}^{(w)}$}
 \label{alg:construct-set-S}
-\KwInput{$S_i^{(w)} = \varnothing$}
-\KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
+\KwInput{$S_{ik}^{(w)} = \varnothing$}
+\KwOutput{$S_{ik}^{(w)}$ filled with $2N$ samples}
 Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
@@
   Compute $\breve{b}_{\breve{n}}^\pm = \text{M}_{ik}^2 \pm C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
-  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}^+, \breve{b}_{\breve{n}}^- \}$
+  Update $S_{ik}^{(w)} \leftarrow S_{ik}^{(w)} \cup \{\breve{b}_{\breve{n}}^+, \breve{b}_{\breve{n}}^- \}$
 }
-\textbf{Return:} {$S_i^{(w)}$}
+\textbf{Return:} {$S_{ik}^{(w)}$}
 \end{algorithm}
@@
 
-$\text{M}_i^1 = W_i$ $\text{M}_{ik}^2 = W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k)$ or $\text{M}_i^1 = D^{(w)}$ $\text{M}_{ik}^2 = 0$
-
-\noindent The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
+\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i}, where in the Gaussian case used in practice, one can sample from
+\begin{align*}
+  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
+  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
+\end{align*}
+with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$.
+The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances, even though the variance is bigger than the first option (see Appendix~\ref{sec:?}).
+We then let the practitioner choose the option in our implementation, since the second one could be more appropriated for a cross-validation phase for instance.
+
+The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
 \begin{equation}
@@
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_i^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}

### 2021-04-29 paper.tex
Keep beta fixed in survival model and change the formula for updating beta

@@
 Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
-\begin{align*}
-  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
-  & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
-  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
-\end{align*}
+\begin{equation*}
+  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
+\end{equation*}
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
@@
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \nonumber \\
-  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \nonumber \\
-  &+ U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
 \end{align}

### 2021-05-10 paper.tex
add graphical model

@@
 \usepackage{subfigure}
+\usepackage{bm}
+\usepackage{tikz}
+\usetikzlibrary{fit,shapes,shadows,arrows,positioning,graphs}
 \usepackage{algpseudocode}
@@
 \end{table}
+
+\subsection{Graphical model for the lights}
+Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
+\begin{figure}[!htb]
+\centering
+\resizebox{.9\textwidth}{!}{
+\begin{tikzpicture}
+\hspace{3cm}
+\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
+\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
+\tikzstyle{connect}=[-latex, thick]
+\tikzstyle{box}=[rectangle, draw=black!100]
+  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
+  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
+  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
+  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
+  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
+  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
+  \node[main, fill = white!100] (b) [below=0.7cm of Y] {\Large$\bm{b_i^l}$};
+  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
+  \node[var, fill = white!100] (Sigma) [right=of b] {\Large$\bm{D}$};
+  \node[main, fill = white!100] (W) [below=of b] {\Large$\bm{\varphi_i(t)}$};
+  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
+  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
+  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};
+
+  \path [->,draw,ultra thick]
+        (beta) edge (mu)
+        (G) edge (mu)
+        (b) edge (mu)
+        (mu) edge (Y)
+        (Sigma) edge (b)
+        (phi) edge (Y)
+        (xi) edge (G)
+        (X) edge (G)
+        (gamma) edge (lambda)
+        (W) edge (lambda)
+        (G) edge (lambda)
+        (X) edge (lambda)
+        (lambda) edge (T)
+        ;
+
+  \node[rectangle, inner sep=8mm,draw=black!100, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.5cm, xshift=-10mm]{\Large$\bm{k = 1...K}$}}] at (0, -4.6) {};
+  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
+  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (12, 3.5) {\Large$\bm{l = 1...L}$};
+  \node[rectangle, dashed, inner sep=3mm, draw=red!100, fit= (b) (W),ultra thick] {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100, label=center:\rotatebox{90}{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,0.5) {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:\rotatebox{90}{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,-8.2) {};
+
+\end{tikzpicture}}
+\caption{Graphical representation of the lights model.
+Square boxes denote observed data while circles denote unobserved (including random) terms.
+The red-dashed box indicates that the association features and the random effects are correlated, which, in addition to the link with the latent group membership, gives rise to the joint model.}
+\label{fig:graphical-model}
+\end{figure}
 

### 2021-05-10 paper.tex
update convergence criterion

@@
 The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{align}
+\begin{equation}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{align}
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{equation}
 for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
@@
 Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
-The $Condition$ in algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w)})$ is coefficient of variation at the $(w + 1)$-th iteration computed as
-\begin{equation*}
-  \label{eq:coefficient_variation}
-  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
-\end{equation*}
-where sd(.) and mean(.) are the sample standard deviation and mean functions respectively and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
-\begin{equation*}
-  \label{eq:relative_difference}
-  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}
-\end{equation*}
-
-\vspace{.5cm}
-
-\begin{algorithm}[]
+
+% \vspace{.5cm}
+
+\begin{algorithm}[!htb]
 \DontPrintSemicolon
@@
 \end{algorithm}
+The convergence condition in Algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w+1)})$ represents the variation at the $(w + 1)$-th iteration computed as
+\begin{equation*}
+  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
+\end{equation*}
+where sd(.) and mean(.) are the empirical standard deviation and mean functions respectively, and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
+\begin{equation*}
+  \label{eq:relative_difference}
+  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}.
+\end{equation*}
 

### 2021-05-10 paper.tex
Remove dependent of beta in survival formulas

--- 2021-04-29 paper.tex
@@
 	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\},
+	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
 \end{equation}
@@
 \toprule
-Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$ & $\imath_a$ & Reference \\
+Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\imath_a$ & Reference \\
 \midrule
-Linear predictor & $m_k^l(t)$ & $u^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
-Random effects & $b^l$ & $\mathbf{0}_{q_l}$ & $r_l$ & \citet{hatfield2011joint} \\ 
-Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & $\dfrac{\dd}{\dd t} u^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
-Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & $\int_0^t u^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
+Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
+Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
+Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
+Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
  \bottomrule
 \end{tabular}
-\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.
-The gradient with respect to $\beta_k^l$ is also given, which will be useful for inference in Section~\ref{sec:QNMCEM}.}
+\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
 \label{table:shared_associations}
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, \beta_k^l, b_i^l) \Big) \\
-  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l) \Big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
+  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
     \begin{array}{ll}
-        \tilde g_{i, s, \gamma_k^0, \gamma_k, \beta_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k, \beta_k}^2(b_i) \big\}, \\
-        \tilde g_{s, \gamma_k, \beta_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, \beta_k^l, b_i^l).
+        \tilde g_{i, s, \gamma_k^0, \gamma_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
+        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
     \end{array}
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] 
+  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align*}
@@
   \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
-  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
+  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
 \end{align}
@@
   \label{eq:grad-Q-2}
-  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, {\beta_k^l}^{(w+1)}, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, {\beta_k^l}^{(w+1)}, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}, \beta_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}

### 2021-05-19 paper.tex
update stopping rule

@@
 \end{algorithm}
-The convergence condition in Algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w+1)})$ represents the variation at the $(w + 1)$-th iteration computed as
-\begin{equation*}
-  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
-\end{equation*}
-where sd(.) and mean(.) are the empirical standard deviation and mean functions respectively, and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
-\begin{equation*}
-  \label{eq:relative_difference}
-  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}.
-\end{equation*}
+Let us define the relative difference at step $(w+1)$ as
+\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
+Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
+\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2 \]
+is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
 

### 2021-05-19 paper.tex
update beta dependence

@@
 \toprule
-Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\imath_a$ & Reference \\
+Description & $\varphi_a(t, b^l)$ & $\imath_a$ & Reference \\
 \midrule
@@
 \end{equation}
-for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
+for all $l = 1, \ldots, L$.
 
@@
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}.
 The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
@@
 Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
-\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2 \]
+\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
 is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
@@
 In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
-\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, \hat \beta^l, b_i^l) \Big\}.\]
+\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, b_i^l) \Big\}.\]
 

### 2021-05-24 paper.tex
no more g_1 nor g_2 in R

@@
 \end{equation*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.

### 2021-07-28 paper.tex
Remove fixed-effect features out of time-to-event model

--- 2021-05-19 paper.tex
@@
 \usepackage{subfigure}
-\usepackage{bm}
-\usepackage{tikz}
-\usetikzlibrary{fit,shapes,shadows,arrows,positioning,graphs}
 \usepackage{algpseudocode}
@@
 	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
+	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
 \end{equation}
@@
 We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
-Corresponding fixed effects are denoted $\gamma_k^0 \in \R^p$, and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
+Corresponding each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[({\gamma_k^0}^\top, \gamma_k^\top)^\top= ({\gamma_k^0}^\top, \gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
 
@@
 \end{table}
-
-\subsection{Graphical model for the lights}
-Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
-\begin{figure}[!htb]
-\centering
-\resizebox{.9\textwidth}{!}{
-\begin{tikzpicture}
-\hspace{3cm}
-\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
-\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
-\tikzstyle{connect}=[-latex, thick]
-\tikzstyle{box}=[rectangle, draw=black!100]
-  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
-  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
-  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
-  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
-  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
-  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
-  \node[main, fill = white!100] (b) [below=0.7cm of Y] {\Large$\bm{b_i^l}$};
-  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
-  \node[var, fill = white!100] (Sigma) [right=of b] {\Large$\bm{D}$};
-  \node[main, fill = white!100] (W) [below=of b] {\Large$\bm{\varphi_i(t)}$};
-  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
-  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
-  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};
-
-  \path [->,draw,ultra thick]
-        (beta) edge (mu)
-        (G) edge (mu)
-        (b) edge (mu)
-        (mu) edge (Y)
-        (Sigma) edge (b)
-        (phi) edge (Y)
-        (xi) edge (G)
-        (X) edge (G)
-        (gamma) edge (lambda)
-        (W) edge (lambda)
-        (G) edge (lambda)
-        (X) edge (lambda)
-        (lambda) edge (T)
-        ;
-
-  \node[rectangle, inner sep=8mm,draw=black!100, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.5cm, xshift=-10mm]{\Large$\bm{k = 1...K}$}}] at (0, -4.6) {};
-  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (12, 3.5) {\Large$\bm{l = 1...L}$};
-  \node[rectangle, dashed, inner sep=3mm, draw=red!100, fit= (b) (W),ultra thick] {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100, label=center:\rotatebox{90}{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,0.5) {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:\rotatebox{90}{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,-8.2) {};
-
-\end{tikzpicture}}
-\caption{Graphical representation of the lights model.
-Square boxes denote observed data while circles denote unobserved (including random) terms.
-The red-dashed box indicates that the association features and the random effects are correlated, which, in addition to the link with the latent group membership, gives rise to the joint model.}
-\label{fig:graphical-model}
-\end{figure}
 
@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta}\big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
 \end{equation}
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
-  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ x_i^\top \gamma_k^0 + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
+  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
 \paragraph*{Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \big(\norm{\xi_k}_{\text{en}, \eta} + \norm{\gamma_k^0}_{\text{en}, \eta} \big) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
 Let us introduce the following useful notations
@@
     \begin{array}{ll}
-        \tilde g_{i, s, \gamma_k^0, \gamma_k}^1&: b_i \mapsto \exp \big\{ x_i^\top \gamma_k^0 + \tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
+        \tilde g_{i, s, \gamma_k}^1&: b_i \mapsto \exp \big\{\tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
         \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
@@
 \end{equation*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
@@
 The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{equation}
+\begin{align}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{equation}
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{align}
 for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
 
-Concerning the update of ${\gamma_k^0}^{(w)}$ and $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
+Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k^0, \gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k^0$ and $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k^0, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
-The update for ${\gamma_k^0}^{(w)}$ therefore requires to solve the following minimization problem
-\begin{equation}
-  \label{eq:minimization-gamma_k^0}
-  {\gamma_k^0}^{(w+1)} \in \argmin_{\gamma_k^0 \in \R^p} Q^{(w)}_{n,k}(\gamma_k^0, {\gamma_k}^{(w)}) + \zeta_{1,k} \norm{\gamma_k^0}_{\text{en}, \eta}.
-\end{equation}
-We then rewrite problem~\eqref{eq:minimization-gamma_k^0} as the following differentiable problem with box constraints
-\begin{equation}
-  \label{eq:sub-problem-gamma_k^0}
-  \begin{split}
-    \text{minimize}& \quad \quad Q_{n, k}^{(w)}({\gamma_k^0}^+ - {\gamma_k^0}^-, {\gamma_k}^{(w)}) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p ({\gamma_{k,j}^0}^+ + {\gamma_{k,j}^0}^-) + \dfrac \eta 2 \norm{{\gamma_k^0}^+ - {\gamma_k^0}^-}_2^2 \big) \\
-    \text{subject to}& \quad \quad {\gamma_{k,j}^0}^+ \geq 0 \text{ and } {\gamma_{k,j}^0}^- \geq 0 \text{ for } j = 1, \ldots, p.
-  \end{split} 
-\end{equation}
-The gradient is here given by
-\begin{align}
-  \label{eq:grad-Q-1}
-  \frac{\partial Q_{n,k}^{(w)}(\gamma_k^0, {\gamma_k}^{(w)})}{\partial \gamma_k^0} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \nonumber \\
-  &\times \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k^0, {\gamma_k}^{(w)}}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big] x_i.
-\end{align}
-Then, the update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
 \begin{equation}
   \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
 \end{equation}
@@
   \label{eq:gamma-update}
-  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}({\gamma_k^0}^{(w+1)}, \cdot), \tilde \eta, \zeta_{3,k}\big)
+  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{3,k}\big)
 \end{equation}
@@
   \label{eq:grad-Q-2}
-  &\frac{\partial Q_{n,k}^{(w)}({\gamma_k^0}^{(w+1)}, \gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, {\gamma_k^0}^{(w+1)}, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
-Then, we initialize $\lambda_0^{(0)}$ and ${\gamma_k^0}^{(0)}$ like if there is no subgroup ($\gamma_{0,0}^{(0)} = \cdots = \gamma_{K-1,0}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
+Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
 Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
@@
 Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
-
-% \vspace{.5cm}
-
-\begin{algorithm}[!htb]
+The $Condition$ in algorithm~\ref{alg:QNMCEM} is cv$(\Delta_{\text{rel}}^{(w+1)}) >$ cv$(\Delta_{\text{rel}}^{(w)})$, where cv$(\Delta_{\text{rel}}^{(w)})$ is coefficient of variation at the $(w + 1)$-th iteration computed as
+\begin{equation*}
+  \label{eq:coefficient_variation}
+  \text{cv}(\Delta_{\text{rel}}^{(w+1)})= \dfrac{\text{sd}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})}{\text{mean}(\Delta_{\text{rel}}^{(w-1)}, \Delta_{\text{rel}}^{(w)}, \Delta_{\text{rel}}^{(w+1)})},
+\end{equation*}
+where sd(.) and mean(.) are the sample standard deviation and mean functions respectively and relative difference $\Delta_{\text{rel}}^{(w+1)}$ is given by 
+\begin{equation*}
+  \label{eq:relative_difference}
+  \Delta_{\text{rel}}^{(w+1)} = \text{max} \Big\{\dfrac{|\theta^{(w+1)} - \theta^{(w)}|}{|\theta^{(w)}| + \epsilon } \Big\}
+\end{equation*}
+
+\vspace{.5cm}
+
+\begin{algorithm}[]
 \DontPrintSemicolon
@@
   Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
-  Update $({\gamma_k^0}^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-gamma_k^0} \\
   Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
@@
 \end{algorithm}
-Let us define the relative difference at step $(w+1)$ as
-\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
-Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
-\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2 \]
-is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
 
@@
 \begin{align*}
-\lambda_i(t|G_i = k) = \lambda_0(t) \exp \Big\{ x_i^\top \xi + \sum_{l=1}^L &\gamma_{k,1}^l (\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t) \\
+\lambda_i(t|G_i = k) = \lambda_0(t) \exp \Big\{ \sum_{l=1}^L &\gamma_{k,1}^l (\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t) \\
 &+ (\gamma_{k,2,1}^l b_{i,1}^l + \gamma_{k,2,2}^l b_{i,2}^l) + \gamma_{k,3}^l (\beta_{k,2}^l + b_{i,2}^l) \Big\},
 \end{align*}
-which corresponds to~\eqref{eq:intensity-model} taking $\gamma_k^0 = \xi$ and the first three functionals described in Table~\ref{table:shared_associations} as shared associations. 
+which corresponds to~\eqref{eq:intensity-model} taking the first three functionals described in Table~\ref{table:shared_associations} as shared associations.
 Now, for the choice of the association parameters, we want to induce sparsity (similarly to~\eqref{eq:xi-sparse}), so we set 

### 2021-07-28 paper.tex

@@
 \end{equation*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
@@
 The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{align}
+\begin{equation}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{align}
-for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{equation}
+for all $l = 1, \ldots, L$.
 
@@
 \end{algorithm}
+Let us define the relative difference at step $(w+1)$ as
+\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
+Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
+\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
+is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
 

### 2021-07-28 paper.tex
Remove penaty on beta

@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
-\end{equation}
-where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$ and $\beta_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})^\top \in \R_+^3$.
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
+\end{equation}
+where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{3,k})^\top \in \R_+^3$.
 Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
@@
 \paragraph*{Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
 Let us introduce the following useful notations
@@
 
-Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
+Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
 \begin{equation*}
@@
 \end{equation*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\begin{equation}
+  \label{eq:minimization-beta_k}
+  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
+\end{equation}
+
+The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
+\begin{align}
+  \label{eq:grad-beta_k}
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
+\end{align}
+for all $l = 1, \ldots, L$. The closed-form update is then given by
+
+\begin{equation}
+  \label{eq:update_beta}
+  {\beta_k^l}^{(w+1)}= \dfrac{\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
+\end{equation}
+
+Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
+\begin{align*}
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+\end{align*}
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
-The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
-\begin{equation}
-  \label{eq:minimization-beta_k}
-  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
-\end{equation}
+The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
+\begin{equation}
+  \label{eq:minimization-gamma_k}
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+\end{equation}
+
 The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
@@
 A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
-Hence, we update $\beta_k^{(w)}$ with the following step
-\begin{equation}
-  \label{eq:beta-update}
-  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
-\end{equation}
-Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
-The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{equation}
-  \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{equation}
-for all $l = 1, \ldots, L$.
-
-Concerning the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
-\begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
-\end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k, \beta_k^{(w+1)}}^1$ defined in~\eqref{eq:g-tilde-1-2}.
-The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
-\begin{equation}
-  \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
-\end{equation}
-We treat problem~\eqref{eq:minimization-gamma_k} in the same way as~\eqref{eq:minimization-beta_k}, that is
+Hence, we update $\gamma_k^{(w)}$ with the following step
 \begin{equation}
@@
 \end{equation}
-with the gradient being known since
+Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
+The gradient of $Q_{n,k}^{(w)}$ is here given by
 \begin{align}
@@
 \label{alg:QNMCEM}
-\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
+\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
 \KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
@@
 \end{equation*}
-Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ or $\zeta_{3,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
+Similar strategy to automatically compute $\zeta_{3,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
 We then choose those interval upper bounds empirically.

### 2021-07-28 paper.tex
Minor update on beta's formala

@@
   \label{eq:update_beta}
-  {\beta_k^l}^{(w+1)}= \dfrac{\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
+  {\beta_k^l}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
 \end{equation}
@@
   Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
-  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:beta-update} \\
+  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
   Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\

### 2021-08-17 paper.tex
Missing typo
review for removing time-independant features out of time-to-event model

@@
 the history of the true unobserved longitudinal process up to time $t$.
-We choose to incorporate $x$, also used in~\eqref{eq:pi}, with no \textit{a priori} on the choice of time-independent features involved in the definition of $\pi_{\xi_k}$ nor $\lambda$, since independent regularizations are used during inference, see~\eqref{eq:pen-log-lik}.
-Corresponding each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{p+L\cA}. \]
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}. \]
 
@@
 with $\alpha \in \N_+$.
-% TODO: en pratique essayer d'ajouter X dans u^l(t) pr voir si ça améliore les perf ou non. Si oui l'ajouter au modèle.
-The idea here is to allow a wide range of polynomial orders for the representation so that a suitable one can be automatically chosen for each trajectory -- depending on its inherent complexity -- thanks to the regularization strategy proposed in~\eqref{eq:pen-log-lik}.
+The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
 We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect.
@@
   \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
-\end{equation}
-where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{3,k})^\top \in \R_+^3$.
+	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
+\end{equation}
+where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$.
 Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
@@
 \paragraph*{Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
 Let us introduce the following useful notations
@@
 
-Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
+Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
 \begin{equation*}
@@
 \end{equation}
-
 The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
@@
 \end{equation}
-
 Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
@@
   \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{3,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
 \end{equation}
@@
   \label{eq:gamma-update}
-  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{3,k}\big)
+  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
 \end{equation}
@@
 \label{alg:QNMCEM}
-\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{3,k})_{k=0,\ldots,K-1}$}
+\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
 \KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
@@
 \end{equation*}
-Similar strategy to automatically compute $\zeta_{3,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
+Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
 We then choose those interval upper bounds empirically.

### 2021-08-17 paper.tex
remove tilde g from main paper

@@
 \newcommand{\cA}{\mathcal A}
+\newcommand{\sA}{\mathscr A}
 \newcommand{\cQ}{\mathcal Q}
@@
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}. \]
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\sA}, \] with $\sA = \sum_{a=1}^\cA \imath_a$ and so that
+\begin{equation}
+  \label{eq:def-asso-features}
+  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
+\end{equation}
+where $\Psi(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.
 
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t_i, b_i^l) \Big) \\
-  & \quad -\int_0^{t_i} \lambda_0(s) \exp \Big\{ \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l) \Big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
 Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
-\[ \E_{\theta^{(w)}}[ g_i(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
-for different functions $g_i$, where we denote 
+\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
+for different functions $g$, where we denote 
 \begin{equation}
@@
 the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g_i(b_i, G_i) = \tilde g_i(b_i)$ (for instance $\tilde g_i: b_i \mapsto b_i^\top D^{-1}b_i$), or $g_i(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
-
-In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
+Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
+
+In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
 \begin{equation}
   \label{eq:def-E-g-tilde}
-  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
+  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
 \end{equation}
@@
   \label{eq:Lambda_ik-b_i|y_i}
-  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
+  \Lambda_{ik,\tilde g}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
   \label{eq:Lambda_ik-b_i}
-  &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
+  &= \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
 \end{align}
-and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.
+and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
 The integral in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} is not tractable analytically, and some form of approximation must be used in practice.
@@
 
+\texttt{define $\Sigma_i$}
+
 \noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i}, where in the Gaussian case used in practice, one can sample from
@@
   \label{eq:def-hat-E-g-tilde}
-  \hat \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
+  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
 \end{equation}
@@
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g_i(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}
@@
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-Let us introduce the following useful notations
-\begin{equation}
-  \label{eq:g-tilde-1-2}
-  \left\{
-    \begin{array}{ll}
-        \tilde g_{i, s, \gamma_k}^1&: b_i \mapsto \exp \big\{\tilde g_{s, \gamma_k}^2(b_i) \big\}, \\
-        \tilde g_{s, \gamma_k}^2&: b_i \mapsto \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(s, b_i^l).
-    \end{array}
-  \right. 
-\end{equation}
 We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
@@
 The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{align}
+\begin{equation}
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
-\end{align}
-for all $l = 1, \ldots, L$. The closed-form update is then given by
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ U_{i}^\top \Sigma_i^{(w)} I_{n_{i}} (y_i - U_{i}\beta_k - V_{i}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big].
+\end{equation}
+The closed-form update is then given by
 
@@
   \label{eq:update_beta}
-  {\beta_k^l}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} (y_i^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[U_{il}^\top I_{n_{il}} U_{il} \Big]}.
+  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
 \end{equation}
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \log \circ \tilde g_{i, t_i, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$, with $\tilde g_{i, t_j, \gamma_k}^1$ defined in~\eqref{eq:g-tilde-1-2}.
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
@@
   \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\mathcal{A}}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
 \end{equation}
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g_{i, t_j, \gamma_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g^1_{t_j, \gamma_k}(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g_{i, t, \gamma_k^{(w+1)}}^1(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g^1_{t, \gamma_k^{(w+1)}}(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}
@@
   \tcc{Monte Carlo E-step}
-  Compute $\hat \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g_i$\\
+  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
   Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
@@
 
+\subsection{\texttt{QNMCEM class}}
+
+Let us introduce the following useful notations
+\begin{equation}
+  \label{eq:g-tilde}
+  \left\{
+    \begin{array}{ll}
+    \tilde g^1&: b_i \mapsto b_i,\\
+    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
+    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
+    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
+    \tilde g^5_{s, \gamma_k}&: b_i \mapsto b_i g^4_{s, \gamma_k}(b_i), \\
+    \tilde g^6_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
+    \end{array}
+  \right. 
+\end{equation}
+
 \bibliography{biblio}

### 2021-08-19 paper.tex
update tilde g notations
minor updates

@@
 \end{equation}
-where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vectors $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$.
+where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$.
 Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
 \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
-for any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty with $\norm{z}_{2,l}=\norm{z^l}_2$, for which we do not have to account for group sizes since they all have the same one.
+for the elastic net penalty of any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
+\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty of any vector $z$, with $\norm{z}_{2,l}=\norm{z^l}_2$, and for which we do not have to account for group sizes since they all have the same one.
 Hence, the resulting optimization problem is written
@@
 \end{equation}
-Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we denote
+Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
 \begin{align*}
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \tilde g^1_{t_j, \gamma_k}(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[ \tilde g^1_{t, \gamma_k^{(w+1)}}(b_i) | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}
@@
 \end{align}
-Appendix~\ref{sec:?} gives some details about the way we implement the equations and their computation in practice.
+Appendix~\ref{sec:QNMCEM class} gives some details about the way we implement the equations and their computation in practice.
 
@@
 Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
-Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (the details are given in Appendix~\ref{sec:MLMM}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
+Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
 
@@
 \subsection{\texttt{MLMM class}}
+\label{sec:MLMM class}
 
@@
 \subsection{\texttt{QNMCEM class}}
+\label{sec:QNMCEM class}
 
@@
     \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
-    \tilde g^5_{s, \gamma_k}&: b_i \mapsto b_i g^4_{s, \gamma_k}(b_i), \\
-    \tilde g^6_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
+    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
     \end{array}

### 2021-08-19 paper.tex
piesewise constant estimator for the baseline hazard

@@
   Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+  & - \sum_{j=1}^{J-1} (t_{j+1}-t_j) \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j < t_i\}} \Big]
 \end{align*}
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is piecewise-constant with jumps at observed event times.
 The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}-t_j) \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j < t_i\}} \Big]
 \end{align}
@@
 
-For the update of $\lambda_0^{(w)}$, we treat the jump size at the observed event times as parameters to be estimated~\citep{klein1992semiparametric}. The closed-form update is then given by
+Then we consider a piecewise-constant estimator $\lambda_0^{(w)}$ for the baseline hazard with jumps at each uncensored observed event times.
+Hence, we treat the jump size as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
 \begin{equation}
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
-\end{equation}
-which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
+  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
+\end{equation}
+for all $j=1, \ldots, n$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
 

### 2021-08-19 paper.tex
update graphical reprensentation

@@
         (G) edge (lambda)
-        (X) edge (lambda)
         (lambda) edge (T)
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
   & - \sum_{j=1}^{J-1} (t_{j+1}-t_j) \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j < t_i\}} \Big]
@@
   \label{eq:update_phi}
-  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
-  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) \nonumber \\
-  & + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
+  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
+  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
 \end{align}

### 2021-08-19 paper.tex
get version 1 of the lemma 1 proof, to be refactored

@@
 \newcommand{\cS}{\mathcal S}
+\newcommand{\cI}{\mathcal I}
 \newcommand{\R}{\mathds R}
@@
 
-\texttt{define $\Sigma_i$}
+\texttt{TODO Van Tuan : define $\Sigma_i$}
 
@@
 
+\texttt{TODO Simon : rewrite the proof properly}
+
+The proximal operator for any function $f : R^p \to R$ is defined by
+\begin{equation}
+    \label{eq:prox-def}
+    \text{prox}_{f}(u) = \argmin_{x \in \R^p} \{\frac{1}{2}\norm{x - u}_2^2 + f(x)\}.
+\end{equation}
+\begin{lemma} 
+The following holds
+\begin{equation*}
+    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L  \norm{\cdot}_{2,l}}
+    = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}
+\end{equation*}
+where $prox_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_2}$ is the proximal operator for the sparse group lasso regulariser, $prox_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_2}$ is the proximal operator for the group lasso regulariser and $prox_{\lambda_1 \norm{\cdot}_1}$ is the proximal operator for the lasso regulariser.
+$\lambda_1 > 0$, $\lambda_2 > 0$ are regularization parameters, $\norm{\cdot}_{2,l}$ denotes the Euclidean norm on features in l-th group.
+\end{lemma}
+
+\begin{proof}
+Denote 
+\begin{equation*}
+    k(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
+\end{equation*}
+then we have the proximal operator (see~\eqref{eq:prox-def}) for the group lasso regulariser with $u = \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)$ is given by
+\begin{equation*}
+    \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_x k(x).
+\end{equation*}
+The sub-differential \texttt{[REF]} of $k(\cdot)$ at $x$ is given by
+\begin{equation*}
+    \partial k(x) = x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} )
+\end{equation*}
+It is clear that $k(\cdot)$ is convex function. Denote the unique minimizer of $k(\cdot)$ is $x^\star$. The sufficient and necessary condition for the optimality of $x^\star$ is 
+\begin{equation}
+    \label{eq:opt_gl}
+    0 \in x^* - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l} ).
+\end{equation}
+\texttt{[REF]}.
+Denote $y = \text{prox}_{\lambda_1\norm{\cdot}_1}(\beta)$. From Lemma 1 in ~\citet{yuan2011efficient}, one has that $\text{SGN}(y) \subseteq \text{SGN}(x^*)$\\
+where \[\text{SGN}(x) = 
+\left\{
+  \begin{array}{ll}
+    \{1\} &\text{if } x > 0, \\ 
+    $[-1, 1]$ &\text{if } x = 0, \\ 
+    \{-1\} &\text{if } x < 0. 
+  \end{array}
+\right. \]
+which is the sub-differential of the l1 norm. \texttt{[REF]} \\
+\noindent Denote 
+\begin{equation*}
+    h(x) =  \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
+\end{equation*}
+then we have the  proximal  operator  for  the  sparse  group  lasso  regulariser is given by
+\begin{equation*}
+    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}}(\beta) = \argmin_{x} \{ h(x)\}.
+\end{equation*}
+Denote the sub-differential of $h(\cdot)$ at $x$ is given by
+\begin{align*}
+    \partial(h(x)) &= x - \beta + \partial(\lambda_1 \norm{x}_1) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l} ) \\
+    &= x - \beta + \lambda_1 \text{SGN}(x) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l} )
+\end{align*}
+
+\noindent We need to prove that $0 \in \partial\big(h(x^\star)\big)$. We can easily prove that 
+\begin{equation*}
+    y = \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) 
+    = \argmin_{u} \{\norm{u - \beta}_2^2 + \lambda_1 \norm{u}_1\} 
+    = \beta - \lambda_1 \text{SGN}(y) \in \beta - \lambda_1 \text{SGN}(x^*).
+\end{equation*}
+Hence
+\begin{equation}
+    \label{eq:opt_gl_in_sgl}
+    x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ) \in x^\star- \beta + \lambda_1 \text{SGN}(x^\star) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} )
+\end{equation}
+It follows from ~\eqref{eq:opt_gl}- ~\eqref{eq:opt_gl_in_sgl} that $0 \in \partial(h(x^\star))$. Hence
+\begin{equation*}
+    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}} = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}
+\end{equation*}
+\end{proof}
+
 \section{Automatic cross-validation grid for $\xi$}
@@
 \end{equation}
+We also denote 
+\[ \cI_1 = 
+\begin{bmatrix}
+  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
+  \vdots &  \ddots & \vdots \\
+  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
+\end{bmatrix}
+\in \R^{n \times J}
+\quad \text{and} \quad 
+\cI_2 = 
+\begin{bmatrix}
+  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
+  \vdots &  \ddots & \vdots \\
+  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
+\end{bmatrix}
+\in \R^{n \times J} \]
+some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
+\begin{align*}
+f(t_i, \delta_i| b_i, G_i = k ; \theta) &= \Big( \sum_{j=1}^J \lambda_0(t_j) \tilde g_{i, t_j, \gamma_k^, \beta_k}^1 \ind{\{t_j = t_i\}} \Big)^{\delta_i} \exp \Big( - \sum_{j=1}^J \lambda_0(t_j) \tilde g_{i, t_j, \gamma_k, \beta_k}^1 \ind{\{t_j \leq t_i\}} \Big) \\
+& = \Big( \text{sum}(\mathbf{\lambda_0} \tilde g_{\gamma_k, \beta_k}^1 \cI_1) \Big)^{\delta_i} \exp \Big( - \text{sum} (\mathbf{\lambda_0} \tilde g_{\gamma_k, \beta_k}^1\cI_2) \Big)
+\end{align*}
+
+\texttt{TODO Van Tuan : rewrite the two previous lines addind the $^{(w)}$ (else it's not true ! since it's only ok for the estimator) + the rigth $\tilde g$ + introduce undefined notations.}
 

### 2021-08-19 paper.tex
adjust distinct times notations

@@
   Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^{J-1} (t_{j+1}-t_j) \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j < t_i\}} \Big]
+  & - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
 \end{align*}
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is piecewise-constant with jumps at observed event times.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is piecewise-constant with jumps at observed event times.
 The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}-t_j) \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i] \ind{\{t_j < t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
 \end{align}

### 2021-08-20 paper.tex
Update the definition of Sigma_i
Update matrix formulation

@@
 
-\texttt{TODO Van Tuan : define $\Sigma_i$}
+\noindent where $\Sigma_i$ is a diagonal matrix whose diagonal is $\Phi_i$.
 
@@
 \begin{align*}
-f(t_i, \delta_i| b_i, G_i = k ; \theta) &= \Big( \sum_{j=1}^J \lambda_0(t_j) \tilde g_{i, t_j, \gamma_k^, \beta_k}^1 \ind{\{t_j = t_i\}} \Big)^{\delta_i} \exp \Big( - \sum_{j=1}^J \lambda_0(t_j) \tilde g_{i, t_j, \gamma_k, \beta_k}^1 \ind{\{t_j \leq t_i\}} \Big) \\
-& = \Big( \text{sum}(\mathbf{\lambda_0} \tilde g_{\gamma_k, \beta_k}^1 \cI_1) \Big)^{\delta_i} \exp \Big( - \text{sum} (\mathbf{\lambda_0} \tilde g_{\gamma_k, \beta_k}^1\cI_2) \Big)
+f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
+& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big)
 \end{align*}
 
-\texttt{TODO Van Tuan : rewrite the two previous lines addind the $^{(w)}$ (else it's not true ! since it's only ok for the estimator) + the rigth $\tilde g$ + introduce undefined notations.}
+\texttt{TODO Van Tuan : update matrix formulation.}
 

### 2021-08-20 paper.tex
Update the tex label of grad-Q

@@
 \begin{align}
-  \label{eq:grad-Q-2}
+  \label{eq:grad-Q}
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
@@
 \end{equation*}
-Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k},~\ref{eq:grad-Q-1} and~\ref{eq:grad-Q-2} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
+Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k} and~\ref{eq:grad-Q} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
 We then choose those interval upper bounds empirically.

### 2021-08-27 paper.tex
Update missing index
Update missing time interval in baseline hazard formula
Add one more tidle g function
Update matrix notation
remove MC_sep option idea in the paper for now

@@
 with 
-\[ \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i} \] 
+\begin{equation}
+  \label{eq:def-Phi_i} 
+  \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}
+\end{equation} 
 and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
@@
 
-\noindent where $\Sigma_i$ is a diagonal matrix whose diagonal is $\Phi_i$.
-
-\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i}, where in the Gaussian case used in practice, one can sample from
+\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i} and where $\Sigma_i$ is the diagonal matrix whose diagonal is $\Phi_i$ (defined in~\eqref{eq:def-Phi_i}), where in the Gaussian case used in practice, one can sample from
 \begin{align*}
@@
 with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$.
-The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances, even though the variance is bigger than the first option (see Appendix~\ref{sec:?}).
-We then let the practitioner choose the option in our implementation, since the second one could be more appropriated for a cross-validation phase for instance.
+The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances.
+We then implement this option in practice, being more appropriated in our high-dimensional context.
 
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
+  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} (t_{j+1}^u-t_j^u) \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
 \end{equation}
@@
 f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
-& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big)
+& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big) \\
+f(T, \Delta| S^{(w)}, G = k ; \theta^{(w)}) &= \big\{ \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \cI_1 \big] \big\}^{\Delta} \exp \Big( - \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \Delta T\cI_2 \big] \Big)
 \end{align*}
-
-\texttt{TODO Van Tuan : update matrix formulation.}
+where $\Delta T = (t^u_2 - t^u_1  \cdots t^u_J - t^u_{J - 1}) \in \R^{J - 1}$.
 

### 2021-09-01 paper.tex
update matrix formulation of survival density
go back to a dirac comb form for the baseline estimate

@@
 Let us denote 
-\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
-the collection of the $\vartheta \in \N_+$ unknown parameters to estimate. Note that for ease of notations, we include $\lambda_0$ in $\theta$ although it is a function.
+\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
+the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
+Note that we include $\lambda_0$ in $\theta$.
+Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time.
+So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
 To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
@@
   Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align*}
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is piecewise-constant with jumps at observed event times.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
 The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
@@
   &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u < t_i\}} \Big]
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align}
@@
 
-Then we consider a piecewise-constant estimator $\lambda_0^{(w)}$ for the baseline hazard with jumps at each uncensored observed event times.
-Hence, we treat the jump size as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
+For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
 \begin{equation}
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t_j)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t_j=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} (t_{j+1}^u-t_j^u) \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_j)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t_j\}} },
-\end{equation}
-for all $j=1, \ldots, n$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+\end{equation}
+for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
 
@@
 Let us introduce the following useful notations
-\begin{equation}
-  \label{eq:g-tilde}
+\begin{equation*}
   \left\{
@@
   \right. 
-\end{equation}
+\end{equation*}
 We also denote 
@@
 \begin{align*}
-f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) &= \big[\lambda_0^{(w)}(t_i) \exp \big\{{\gamma_k^{(w)}}^\top\Psi_i(t_i) \big\} \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) \exp \big\{{\gamma_k^{(w)}}^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u < t_i\}} \Big) \\
-& = \big[\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k}(b_i) \big]^{\delta_i} \exp \Big( - \sum_{j=1}^{J-1} (t_{j+1}^u-t_j^u) \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k}(b_i) \ind{\{t_j^u < t_i\}} \Big) \\
-f(T, \Delta| S^{(w)}, G = k ; \theta^{(w)}) &= \big\{ \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \cI_1 \big] \big\}^{\Delta} \exp \Big( - \underset{J}{\text{sum }} \big[g^4_{\gamma_k} \lambda_0^{(w)} \Delta T\cI_2 \big] \Big)
+f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(b_i) \big)^{\delta_i} \\ 
+&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(b_i) \ind{\{t_j^u \leq t_i\}} \Big\}
 \end{align*}
-where $\Delta T = (t^u_2 - t^u_1  \cdots t^u_J - t^u_{J - 1}) \in \R^{J - 1}$.
+so that one can compute the tensor
+\[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
+where 
+\begin{equation*}
+  \left\{
+    \begin{array}{ll}
+    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
+    \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
+    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
+    \end{array}
+  \right. 
+\end{equation*}
 

### 2021-09-01 paper.tex
Add the matrix form of the updated baseline hazard function

@@
 \[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
+or the updated baseline hazard function~\eqref{eq:update_lambda_0} can be rewrote as
+\begin{equation*}
+  \label{eq:update_lambda_0_matrix}
+  \lambda_0^{(w+1)}(t)= \dfrac{\underset{n}{\text{sum }} \boldsymbol{\delta} \cI_1 }{\underset{n}{\text{sum }} \underset{K = 2}{\text{sum }} \hat \E_{\theta^{(w)}}[\boldsymbol{g^4}] \cI_2 \boldsymbol{\hat \pi^{\theta^{(w)}}}  },
+\end{equation*}
 where 
@@
     \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
-    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
+    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n, \\
+    \boldsymbol{\hat \pi^{\theta^{(w)}}} \in \R^{2 \times n} \text{ with }  \boldsymbol{\hat \pi_1^{\theta^{(w)}}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n \text{ and } \boldsymbol{\hat\pi_0^{\theta^{(w)}}} = 1 - \boldsymbol{\hat\pi_1^{\theta^{(w)}}}.
     \end{array}

### 2021-09-02 paper.tex
update algo name

--- 2021-09-01 paper.tex
@@
 \affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
-\affil[2]{Califrais' Machine Learning Lab, Paris, France}
+\affil[2]{LOPF, Califrais' Machine Learning Lab, Paris, France}
 \affil[3]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
@@
 On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
-Finally, the sparse group lasso of $\beta_k$ allows to consider a flexible representation of time for the design matrices $(U_i)_{i=1, \dots, n}$ and lets the model automatically fit each trajectory $l$ with the right complexity.
 Note that in practice, the intercept is not regularized.
@@
 \subsection{A Quasi-Newton Monte Carlo EM}
-\label{sec:QNMCEM}
-
-In order to derive an algorithm for this objective, we introduce a so-called QNMCEM algorithm, being a combination
-between an EM algorithm~\citep{dempster1977maximum} with Monte Carlo approximations~\citep{levine2001implementations}, and multiple L-BFGS-B algorithms~\citep{zhu1997algorithm}.
+\label{sec:prox-QNMCEM}
+
+In order to derive an algorithm for this objective, we introduce a so-called prox-QNMCEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum} with Monte Carlo approximations~\citep{levine2001implementations}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}.
 EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
@@
 Multiple techniques have been proposed (see for instance~\citet{law2002joint} where a subjectively rule is followed, or~\citet{booth1999maximizing} that uses a rule based on confidence intervals for $\theta^{(w)}$ which requires additional variance estimation).
-We opt for a simple automated approach using relative differences of the objective function defined in~\eqref{eq:pen-log-lik} (see Algorithm~\ref{alg:QNMCEM}).
+We opt for a simple automated approach using relative differences of the objective function defined in~\eqref{eq:pen-log-lik} (see Algorithm~\ref{alg:prox-QNMCEM}).
 
@@
 
-\paragraph*{Quasi-Newton M-step.}
+\paragraph*{Proximal Quasi-Newton M-step.}
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-We then present the parameters updates in the order given in Algorithm~\ref{alg:QNMCEM}.
+We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNMCEM}.
 First, the update of $D^{(w)}$ is naturally given in closed-form by
@@
 Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:QNMCEM}) with every increase of $N$.
+We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNMCEM}) with every increase of $N$.
 
@@
 \end{align}
-Appendix~\ref{sec:QNMCEM class} gives some details about the way we implement the equations and their computation in practice.
+Appendix~\ref{sec:prox-QNMCEM class} gives some details about the way we implement the equations and their computation in practice.
 
@@
 Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced.
-We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:QNMCEM}.
+We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNMCEM}.
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
@@
 
-\paragraph*{The QNMCEM algorithm.}
-Algorithm~\ref{alg:QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
+\paragraph*{The prox-QNMCEM algorithm.}
+Algorithm~\ref{alg:prox-QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
 
@@
 \DontPrintSemicolon
-\caption{QNMCEM algorithm for the lights model inference}
-\label{alg:QNMCEM}
+\caption{prox-QNMCEM algorithm for the lights model inference}
+\label{alg:prox-QNMCEM}
 \KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
@@
 \[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
-Then in practice, the convergence for Algorithm~\ref{alg:QNMCEM} is declared when the condition
+Then in practice, the convergence for Algorithm~\ref{alg:prox-QNMCEM} is declared when the condition
 \[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
@@
 The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
-Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our QNMCEM algorithm introduced in Section~\ref{sec:QNMCEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNMCEM algorithm introduced in Section~\ref{sec:prox-QNMCEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
 Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
@@
 In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
-algorithm (QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform feature
+algorithm (prox-QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform feature
 selection and to prevent overfitting.
@@
 
-Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the QNMCEM algorithm in Section~\ref{sec:QNMCEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
+Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNMCEM algorithm in Section~\ref{sec:prox-QNMCEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
 For the sake of simplicity, let us denote here 
@@
 
-\subsection{\texttt{QNMCEM class}}
-\label{sec:QNMCEM class}
+\subsection{\texttt{prox-QNMCEM class}}
+\label{sec:prox-QNMCEM class}
 

### 2021-09-02 paper.tex
no more Hawkes in the lights paper

@@
 This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.
-
-Most joint modeling simulation studies use fixed time points~\citep{hickey2018joinerml}, that is $n_i^l = n_i^{l'}$ and $t_{ij}^l = t_{ij}^{l'}$ for all $(l, l') \in \{1, \ldots, L\}^2$ and $j = 1, \ldots, n_i^l$ for a given subject $i$, but it does not reflect actual behavior in practice, since data measurements of the different longitudinal processes are barely done simultaneously.
-For this reason, the simulation of realistic longitudinal features is not trivial and we use Hawkes processes~\citep{hawkes1974cluster} with exponential kernels to generate measurement times, which is a family of counting process with an autoregressive intensity.
-Hawkes processes model self-exciting behavior, that is in our case, when the measurement of one longitudinal process makes future measurements of the process -- as well as those of others processes, in a multivariate way -- more likely to happen.
-Namely, for a subject $i$, times $\{t_{ij}^l\}_{j \geq 1}$ for processes $l=1, \ldots, L$ are simulated using a multivariate Hawkes process $N_{it} = [N_{it}^1 \cdots N_{it}^L]$ with $t \geq 0$ and $N_{it}^l = \sum_{j \geq 1} \ind{\{t_{ij}^l \leq t\}}$.
-The process $N_{it}$ is a multivariate counting process, whose components $N_{it}^l$ have intensities
-\[ \lambda_i^l(t) = \Upsilon_{l} + \sum_{l'=1}^L \sum_{j \geq 1} A_{ll'} \upsilon \exp\big(-\upsilon(t - t_{ij}^{l'}) \big) \]
-for $l=1, \ldots, L$.
-$\Upsilon_{l} \geq 0$ is called baseline intensity, and corresponds to the exogenous probability of having a measurement for process $l$.
-We sample $\Upsilon_{l} \sim \cU\big([0.1,1]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
-The matrix $A = [A_{ll'}]_{1 \leq l,l' \leq L}$ is the adjacency matrix such that $A_{ll'} \geq 0$ quantifies the impact of past measurement time of process $l'$ on the measurement time of process $l$, and $\upsilon \geq 0$ is a memory parameter.
-We sample $A_{ll'} \sim \cU\big([0.1,0.2]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
-Simulation of the Hawkes processes
-is achieved using the \texttt{tick} library~\citep{bacry2017tick}. An example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
+We simulate measurement times using a regular grid, and an example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
 

### 2021-09-09 paper.tex
update baseline update matrix formulation
Update the P function and its gradient

--- 2021-09-02 paper.tex
@@
 Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
-\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\[ P^{(w)}_{n}(\xi_0, \ldots, \xi_K) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on all of $\xi_k$ for $k=0, \ldots, K-1$. Then we can also write
+\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \Big[\hat \pi_{ik}^{\theta^{(w)}} x_i^\top \xi_k - \log \sum_{k^{\prime}=0}^{K-1} e^{x_i^\top\xi_{k^{\prime}}}\Big] \]
+based on the quantities involved in $P^{(w)}_{n}(\xi_0, \ldots, \xi_K)$ that depend on only one $\xi_k$.
+The update for each $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
 \begin{equation}
@@
   \label{eq:grad-xi_k}
-  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_i.
+  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big( \hat \pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
 \end{equation}
@@
 \end{equation*}
+Another example is the update of the baseline hazard~\eqref{eq:update_lambda_0} as
+\[\boldsymbol{\lambda_0^{(w+1)}} = \underset{n}{\text{sum}}( \boldsymbol{\delta} \cI_1) / \underset{n}{\text{sum }} \underset{K}{\text{sum }} (\boldsymbol{g^4_{\gamma_k}} \cI_2 \boldsymbol{\hat \pi^{(w)}})\]
+with $\boldsymbol{\hat \pi^{(w)}} \in \R^{2 \times n} \text{ and }  \left\{
+    \begin{array}{ll}
+    \boldsymbol{\hat \pi_{1,\bullet}^{(w)}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n, \\
+    \boldsymbol{\hat\pi_{0,\bullet}^{(w)}} = 1 - \boldsymbol{\hat\pi_{1,\bullet}^{(w)}}.
+    \end{array}
+  \right.$
 

### 2021-09-09 paper.tex
updates
update P function definition

--- 2021-09-02 paper.tex
@@
   \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i | \breve{b}, G_i = k ; \theta^{(w)}).
+  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
 \end{equation}
@@
 Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
-\[ P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
+\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
@@
 \in \R^{n \times J} \]
-some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance
+some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance, the quantities
 \begin{align*}
-f(t_i, \delta_i| b_i, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(b_i) \big)^{\delta_i} \\ 
-&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(b_i) \ind{\{t_j^u \leq t_i\}} \Big\}
+f(t_i, \delta_i| \breve{b}, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(\breve{b}) \big)^{\delta_i} \\ 
+&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}) \ind{\{t_j^u \leq t_i\}} \Big\}
 \end{align*}
-so that one can compute the tensor
+for $i = 1, \ldots, n$ and $\breve{b} \in S^{(w)}$ can be represented in the following matrix 
 \[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
@@
     \begin{array}{ll}
-    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
+    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = \tilde g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
     \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
@@
 \end{equation*}
+Another example is the update of the baseline hazard~\eqref{eq:update_lambda_0} as
+\[\boldsymbol{\lambda_0^{(w+1)}} = \underset{n}{\text{sum}}( \boldsymbol{\delta} \cI_1) / \underset{n}{\text{sum }} \underset{k}{\text{sum }} (\boldsymbol{g^4_{\gamma_k}} \cI_2 \boldsymbol{\hat \pi^{(w)}})\]
+with $\boldsymbol{\hat \pi^{(w)}} \in \R^{2 \times n} \text{ and }  \left\{
+    \begin{array}{ll}
+    \boldsymbol{\hat \pi_{1,\bullet}^{(w)}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n, \\
+    \boldsymbol{\hat\pi_{0,\bullet}^{(w)}} = 1 - \boldsymbol{\hat\pi_{1,\bullet}^{(w)}}.
+    \end{array}
+  \right.$
 

### 2021-10-28 paper.tex
update beta simulation formula
rewrite lemma 1 proof

@@
 where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to
-\[ \beta_k^l \sim \cN\Big( 
+\[ \beta_k^l \sim \ind{\{l \in \cS_k \}} \cN\Big( 
 \mu_k, 
@@
   0 & \rho_3
+\end{bmatrix} 
+\Big)
++ \ind{\{l \notin \cS_k \}} \cN\Big( 
+\begin{pmatrix}
+  0 \\ 0
+\end{pmatrix}
+, 
+\begin{bmatrix}
+  \rho_4 & 0\\
+  0 & \rho_4
 \end{bmatrix} 
@@
 
-\texttt{TODO Simon : rewrite the proof properly}
-
-The proximal operator for any function $f : R^p \to R$ is defined by
+The proximal operator~\citep{moreau1962fonctions} for any function $f : \R^p \to \R$ is defined by
 \begin{equation}
     \label{eq:prox-def}
-    \text{prox}_{f}(u) = \argmin_{x \in \R^p} \{\frac{1}{2}\norm{x - u}_2^2 + f(x)\}.
-\end{equation}
-\begin{lemma} 
-The following holds
+    \text{prox}_{f}(u) = \argmin_{x \in \R^p} \big\{\frac{1}{2}\norm{x - u}_2^2 + f(x)\big\}.
+\end{equation}
+To prove Lemma~\ref{lemma:prox-sgl1}, we need to show that
+\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L  \norm{\cdot}_{2,l}} = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}\]
+where $\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}$ is the proximal operator for the sparse group lasso regularizer, $\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}$ is the proximal operator for the group lasso regularizer and $\text{prox}_{\lambda_1 \norm{\cdot}_1}$ is the proximal operator for the lasso regularizer.
+$\lambda_1 > 0$ and $\lambda_2 > 0$ are regularization parameters, and $\norm{\cdot}_{2,l}$ denotes the Euclidean norm on features in the $l$-th group.
+
+\begin{proof}
+Fix any $\beta \in \R^p$ and denote 
 \begin{equation*}
-    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L  \norm{\cdot}_{2,l}}
-    = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}
+    k_\beta(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
 \end{equation*}
-where $prox_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_2}$ is the proximal operator for the sparse group lasso regulariser, $prox_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_2}$ is the proximal operator for the group lasso regulariser and $prox_{\lambda_1 \norm{\cdot}_1}$ is the proximal operator for the lasso regulariser.
-$\lambda_1 > 0$, $\lambda_2 > 0$ are regularization parameters, $\norm{\cdot}_{2,l}$ denotes the Euclidean norm on features in l-th group.
-\end{lemma}
-
-\begin{proof}
-Denote 
-\begin{equation*}
-    k(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
-\end{equation*}
-then we have the proximal operator (see~\eqref{eq:prox-def}) for the group lasso regulariser with $u = \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)$ is given by
-\begin{equation*}
-    \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_x k(x).
-\end{equation*}
-The sub-differential \texttt{[REF]} of $k(\cdot)$ at $x$ is given by
-\begin{equation*}
-    \partial k(x) = x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} )
-\end{equation*}
-It is clear that $k(\cdot)$ is convex function. Denote the unique minimizer of $k(\cdot)$ is $x^\star$. The sufficient and necessary condition for the optimality of $x^\star$ is 
+for all $x \in \R^p$.
+Given that $\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big)_j=\max(0, 1-\lambda_1/|\beta_j|)\beta_j$~\citep{bach2011optimization}, $k_\beta$ is clearly strictly convex.
+Denoting $x^\star$ its unique minimizer,~\eqref{eq:prox-def} entails
+\[\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_{x \in \R^p} k_\beta(x) = \{x^\star\}.\]
+Then, it follows that
 \begin{equation}
     \label{eq:opt_gl}
-    0 \in x^* - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l} ).
-\end{equation}
-\texttt{[REF]}.
-Denote $y = \text{prox}_{\lambda_1\norm{\cdot}_1}(\beta)$. From Lemma 1 in ~\citet{yuan2011efficient}, one has that $\text{SGN}(y) \subseteq \text{SGN}(x^*)$\\
-where \[\text{SGN}(x) = 
+    0 \in \partial k(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
+\end{equation}
+where $\partial k$ denotes the subgradient~\citep{rockafellar1974} of $k$.
+Now denoting
+\[h_\beta(x) = \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} \] 
+for all $x \in \R^p$ such that
+\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}(\beta) = \argmin_{x\in \R^p} h_\beta(x),\]
+it remains to be proved that $0 \in \partial h_\beta(x^\star)$.
+
+Then, it turns out that 
+\[ \partial h_\beta(x) = x - \beta + \lambda_1 \text{SGN}(x) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l})\]
+for all $x \in \R^p$, where SGN($x$) is the subgradient of the $\ell_1$-norm at $x$~\citep{bach2011optimization} defined in a component wise fashion as
+\[\text{SGN}(x) = 
 \left\{
@@
 \right. \]
-which is the sub-differential of the l1 norm. \texttt{[REF]} \\
-\noindent Denote 
-\begin{equation*}
-    h(x) =  \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
-\end{equation*}
-then we have the  proximal  operator  for  the  sparse  group  lasso  regulariser is given by
-\begin{equation*}
-    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}}(\beta) = \argmin_{x} \{ h(x)\}.
-\end{equation*}
-Denote the sub-differential of $h(\cdot)$ at $x$ is given by
-\begin{align*}
-    \partial(h(x)) &= x - \beta + \partial(\lambda_1 \norm{x}_1) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l} ) \\
-    &= x - \beta + \lambda_1 \text{SGN}(x) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l} )
-\end{align*}
-
-\noindent We need to prove that $0 \in \partial\big(h(x^\star)\big)$. We can easily prove that 
-\begin{equation*}
-    y = \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) 
-    = \argmin_{u} \{\norm{u - \beta}_2^2 + \lambda_1 \norm{u}_1\} 
-    = \beta - \lambda_1 \text{SGN}(y) \in \beta - \lambda_1 \text{SGN}(x^*).
-\end{equation*}
-Hence
-\begin{equation}
-    \label{eq:opt_gl_in_sgl}
-    x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ) \in x^\star- \beta + \lambda_1 \text{SGN}(x^\star) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} )
-\end{equation}
-It follows from ~\eqref{eq:opt_gl}- ~\eqref{eq:opt_gl_in_sgl} that $0 \in \partial(h(x^\star))$. Hence
-\begin{equation*}
-    \text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}} = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}
-\end{equation*}
+Denoting $y = \text{prox}_{\lambda_1\norm{\cdot}_1}(\beta)$, Lemma 1.4) in~\citet{yuan2011efficient} entails $\text{SGN}(y) \subseteq \text{SGN}(x^*)$. Hence,
+\[ y = \argmin_{u \in \R^p} \big\{\frac{1}{2} \norm{u - \beta}_2^2 + \lambda_1 \norm{u}_1\big\} 
+    = \beta - \lambda_1 \text{SGN}(y) \subseteq \beta - \lambda_1 \text{SGN}(x^*), \]
+which gives
+\[0 \in x^\star - y + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ) \subseteq x^\star- \beta + \lambda_1 \text{SGN}(x^\star) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ).\]
 \end{proof}

### 2021-10-28 paper.tex
put back hawkes process for simulated times

@@
 This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.
-We simulate measurement times using a regular grid, and an example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
+
+Most joint modeling simulation studies use fixed time points~\citep{hickey2018joinerml}, that is $n_i^l = n_i^{l'}$ and $t_{ij}^l = t_{ij}^{l'}$ for all $(l, l') \in \{1, \ldots, L\}^2$ and $j = 1, \ldots, n_i^l$ for a given subject $i$, but it does not reflect actual behavior in practice, since data measurements of the different longitudinal processes are barely done simultaneously.
+For this reason, the simulation of realistic longitudinal features is not trivial and we use Hawkes processes~\citep{hawkes1974cluster} with exponential kernels to generate measurement times, which is a family of counting process with an autoregressive intensity.
+Hawkes processes model self-exciting behavior, that is in our case, when the measurement of one longitudinal process makes future measurements of the process -- as well as those of others processes, in a multivariate way -- more likely to happen.
+Namely, for a subject $i$, times $\{t_{ij}^l\}_{j \geq 1}$ for processes $l=1, \ldots, L$ are simulated using a multivariate Hawkes process $N_{it} = [N_{it}^1 \cdots N_{it}^L]$ with $t \geq 0$ and $N_{it}^l = \sum_{j \geq 1} \ind{\{t_{ij}^l \leq t\}}$.
+The process $N_{it}$ is a multivariate counting process, whose components $N_{it}^l$ have intensities
+\[ \lambda_i^l(t) = \Upsilon_{l} + \sum_{l'=1}^L \sum_{j \geq 1} A_{ll'} \upsilon \exp\big(-\upsilon(t - t_{ij}^{l'}) \big) \]
+for $l=1, \ldots, L$.
+$\Upsilon_{l} \geq 0$ is called baseline intensity, and corresponds to the exogenous probability of having a measurement for process $l$.
+We sample $\Upsilon_{l} \sim \cU\big([0.1,1]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
+The matrix $A = [A_{ll'}]_{1 \leq l,l' \leq L}$ is the adjacency matrix such that $A_{ll'} \geq 0$ quantifies the impact of past measurement time of process $l'$ on the measurement time of process $l$, and $\upsilon \geq 0$ is a memory parameter.
+We sample $A_{ll'} \sim \cU\big([0.1,0.2]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
+Simulation of the Hawkes processes
+is achieved using the \texttt{tick} library~\citep{bacry2017tick}. An example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
 

### 2021-11-01 paper.tex
add group dependance on association vector Psi

@@
 	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, b^l) \Big\},
+	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_{k,a}(t, b^l) \Big\},
 \end{equation}
@@
 the history of the true unobserved longitudinal process up to time $t$.
-For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_a)$.
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_{k,a}$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_{k,a})$.
 This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
@@
   \label{eq:def-asso-features}
-  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
-\end{equation}
-where $\Psi(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.
-
-\paragraph{Specification of the functionals $(\varphi_a)_{a\in \cA}$.}
+  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi_k(t) \big\},
+\end{equation}
+where $\Psi_k(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.
+
+\paragraph{Specification of the functionals $(\varphi_{k,a})_{a\in \cA}$.}
 
@@
 \toprule
-Description & $\varphi_a(t, b^l)$ & $\imath_a$ & Reference \\
+Description & $\varphi_{k,a}(t, b^l)$ & $\imath_a$ & Reference \\
 \midrule
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_{k,i}(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_{k,i}(s) \big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_i(t_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_{k,i}(t_i) | t_i, \delta_i, y_i] \\
+  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align*}
@@
   \label{eq:grad-Q}
-  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_a(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_a(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
+   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_{k,i}(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
 \end{equation}
@@
 In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
-\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_a(t_i^{max}, b_i^l) \Big\}.\]
+\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\}.\]
 
@@
     \label{eq:opt_gl}
-    0 \in \partial k(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
+    0 \in \partial k_\beta(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
 \end{equation}
@@
     \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
-    \tilde g^3_{s}&: b_i \mapsto \Psi_i(s), \\
-    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s}(b_i) \big\}, \\
-    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s}(b_i)^\top g^4_{s}(b_i).
+    \tilde g^3_{s, k}&: b_i \mapsto \Psi_{k,i}(s), \\
+    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s, k}(b_i) \big\}, \\
+    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s, k}(b_i)^\top g^4_{s, k}(b_i).
     \end{array}

### 2021-11-04 paper.tex
update simulated data generation process

@@
 Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely
-\[u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top\]
+\begin{equation}
+  \label{eq:alpha-def}
+  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
+\end{equation}
 with $\alpha \in \N_+$.
@@
 where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to
-\[ \beta_k^l \sim \ind{\{l \in \cS_k \}} \cN\Big( 
+\[ \beta_k^l \sim \cN\Big( 
 \mu_k, 
@@
 \end{bmatrix} 
-\Big)
-+ \ind{\{l \notin \cS_k \}} \cN\Big( 
-\begin{pmatrix}
-  0 \\ 0
-\end{pmatrix}
-, 
-\begin{bmatrix}
-  \rho_4 & 0\\
-  0 & \rho_4
-\end{bmatrix} 
 \Big) \]
-for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that $\beta_k^l$ coefficients are both positive for high-risk subjects ($G_i=1$), which favours increasing trajectories with positive values, and negative otherwise ($G_i=0$), which favours decreasing trajectories with negative values.
+for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that trajectories start on average higher for high-risk subjects ($G_i=1$), and with a higher slope than low-risk subjects ($G_i=0$).
 This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.
@@
 
-To generate survival times, we choose a risk model of the form
-\begin{align*}
-\lambda_i(t|G_i = k) = \lambda_0(t) \exp \Big\{ \sum_{l=1}^L &\gamma_{k,1}^l (\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t) \\
-&+ (\gamma_{k,2,1}^l b_{i,1}^l + \gamma_{k,2,2}^l b_{i,2}^l) + \gamma_{k,3}^l (\beta_{k,2}^l + b_{i,2}^l) \Big\},
-\end{align*}
-which corresponds to~\eqref{eq:intensity-model} taking the first three functionals described in Table~\ref{table:shared_associations} as shared associations.
-Now, for the choice of the association parameters, we want to induce sparsity (similarly to~\eqref{eq:xi-sparse}), so we set 
-\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] and
+Now to generate survival times, we want to induce sparsity in the longitudinal data effects, so we set 
+\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] 
+and we choose a risk model~\eqref{eq:intensity-model} with
+\begin{equation*}
+  \left\{
+    \begin{array}{ll}
+    \big(\varphi_{k,a}(t, b_i^l)\big)_{a \in \{1,2\}} = \big(\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t\ , {b_i^l}^\top \big)^\top \ind{\{l \in \cS_k\}} + \cN \big(0, \bSigma_3(\rho_3)\big) \ind{\{l \notin \cS_k\}}  \\
+    \big( \varphi_{k,a}(t, b_i^l)\big)_{a=3,\ldots,\cA} = \cN \big(0, \bSigma_4(\rho_3)\big),
+    \end{array}
+  \right.
+\end{equation*}
+and 
 \begin{equation}
   \label{eq:gamma-sparse}
-  \gamma_{k,a}^l = \varsigma_2 \big(\ind{\{l \leq (k + 1) \lfloor L r_s/K \rfloor \}} + \ind{\{l \in \cS_k \}} \big) \mathbf{1}_{\imath_a},
-\end{equation}
-so that there is a proportion $r_s$ of active longitudinal features among the $L$, and the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$.
-Note that one can write
+  \gamma_{k,a}^l = \varsigma_2 \mathbf{1}_{\imath_a} \ind{\big\{l \in \cS_k,\ a\in \{1, 2\} \big\}}
+\end{equation}
+for the choice of the association parameters, with $\cA = 3 + 3(1-r_s)/r_s$ so that $(i)$ there is a proportion $r_s$ of active longitudinal features among the $L$ (with the $\cS_k$), $(ii)$ we induce sparsity within each group adding a proportion of $(1-r_s)$ non-active association features ($a \geq 3$) in each group (similarly to~\eqref{eq:xi-sparse} for the time-independent features), and $(iii)$ the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$ (thanks to the $\cS_k$ definition).
+
+In terms of active association features, it corresponds to~\eqref{eq:intensity-model} taking the first two functionals described in Table~\ref{table:shared_associations} as shared associations with $\alpha = 1$ in~\eqref{eq:alpha-def} so that one can write
 \[ \lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\},\]
-being a Cox model with a linear time-varying feature.
+being a Cox model with a linear time-varying feature that allows the following explicit survival times generation process.
+Then, we do not consider the time-dependent slope as association feature since here with $\alpha = 1$, it would be collinear with the random effects.
 We choose a Gompertz distribution~\citep{gompertz1825xxiv} for the baseline, that is

### 2021-11-07 paper.tex
update cv

@@
 To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
-In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\xi_k$ and $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$ and $\zeta_3 = \zeta_{3,0} = \zeta_{3,1}$.
+In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\xi_k$ and $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$.
 
@@
 
-The hyper-parameters $(\zeta_1, \zeta_2, \zeta_3)^\top \in \R_+^3$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
+The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
 Random search is indeed more efficient in terms of computing times than classical grid search for hyper-parameters optimization and finds better models by effectively searching in a larger configuration space, which is appropriate to our high-dimensional problem.
-Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ and $\zeta_3$ are determined empirically.
-% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2, \zeta_3) en 3D
+Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
+% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D
 
@@
 Concerning the cross-validation procedure for tuning 
-\[(\zeta_{1,0}, \zeta_{2,0}, \zeta_{3,0}, \ldots, \zeta_{1,K-1}, \zeta_{2,K-1}, \zeta_{3,K-1})^\top \in \R_+^{3K}, \]
-we use a randomized search with the C-index metric (see Section~\ref{sec:evaluation strategy}) where for all $k=0, \ldots, K-1$ and $j \in \{1, 2, 3\}$, each $\zeta_{j,k}$ finds its candidates in the interval
+\[(\zeta_{1,0}, \zeta_{2,0}, \ldots, \zeta_{1,K-1}, \zeta_{2,K-1})^\top \in \R_+^{2K}, \]
+we use a randomized search with the C-index metric (see Section~\ref{sec:evaluation strategy}) where for all $k=0, \ldots, K-1$ and $j \in \{1, 2\}$, each $\zeta_{j,k}$ finds its candidates in the interval
 \[[\zeta_{j,k}^{\text{max}} \times 10^{-4}, \zeta_{j,k}^{\text{max}} ] \subset \R, \]

### 2022-01-03 paper.tex
minor updates after global review

@@
 \author[1,2]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
-\author[2]{Van Tuan Nguyen}
-\author[3]{Antoine Barbieri}
+\author[2,3]{Van Tuan Nguyen}
+\author[4]{Antoine Barbieri}
 \author[1]{Sarah Zohar}
-\author[1,4]{Anne-Sophie Jannot}
+\author[1,5]{Anne-Sophie Jannot}
 \affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
 \affil[2]{LOPF, Califrais' Machine Learning Lab, Paris, France}
-\affil[3]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
-\affil[4]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
+\affil[3]{LPSM, UMR 8001, CNRS, Sorbonne University, Paris, France}
+\affil[4]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
+\affil[5]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
 \setcounter{Maxaffil}{0}
@@
 
-This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available.
+This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of both longitudinal and time-independent features are available.
 Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
-Features are extracted from the longitudinal processes and included as potential risk factor in a group-specific Cox model with high-dimensional shared associations.
+Shared association features are learned from the longitudinal processes and included as potential risk factor in a group-specific Cox model.
 Appropriate penalties are then used during inference to allow flexibility in modeling the dependency between the longitudinal features and the event time.
-The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on publicly available datasets.
-On these high-dimensional datasets, our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
+The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
+Our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
 In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a practical perspective.
@@
 Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
-The event of interest could be for instance survival time, re-hospitalization, relapse or disease progression in a medical context; or the time when a client stops using a company's product or service in a churn prediction setting.
+The event of interest could be for instance the time when a client stops using a company's product or service in a churn prediction setting~ ; or survival time, re-hospitalization, relapse or disease progression in a medical context.
 We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
@@
 To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
-for feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model, as well as for the fixed effect (allowing flexible representations of time).
-Indeed, this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model on one side.
-On the other hand, it yields sparsity for a given trajectory, namely feature selection.
-Inference is achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm.
-Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters (to perform feature selection and adapt to the longitudinal data complexity).
-Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for clinical decision making in patient monitoring for instance, or else for real-time decision support in a customer's satisfaction monitoring context.
+for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
+Inference is achieved using an efficient and novel proximal Quasi-Newton Monte Carlo Expectation Maximization algorithm (prox-QNMCEM).
+Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
+Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.
 
@@
 Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones.
-In Section~\ref{sec:application}, we apply our method to high-dimensional publicly available datasets.
+In Section~\ref{sec:application}, we apply our method to a publicly available dataset.
 Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.
@@
 
-Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic cases).
+Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic case).
 The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
@@
 The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}.
-The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameters $\phi_l$, while parameters $\beta_k$ have to be estimated.
+The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameter $\phi_l$, while parameter $\beta_k$ has to be estimated.
 Note that $d_\phi(\cdot)$ is related to the normalizing constant.
@@
 \end{tabular}}
-\caption[]{Examples of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
+\caption[]{Example of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
 \label{table:glm}
@@
 One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independent features involved in the $k$-th group membership.
-The ridge part allows to handle potential correlation between features.
+The ridge part allows to handle potential correlation between time-independent features.
 On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
@@
 and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
-The integral in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} is not tractable analytically, and some form of approximation must be used in practice.
+The integrals in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} are not tractable analytically, and some form of approximation must be used in practice.
 Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context.
@@
 with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$.
-The advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances.
+The huge advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances.
 We then implement this option in practice, being more appropriated in our high-dimensional context.
@@
 
-The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
-Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
+The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-gamma_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
+Then, we choose to solve problem~\eqref{eq:minimization-gamma_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
 \[p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}.\]
@@
 We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNMCEM}.
-For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_{k,a}^l}^{(0)} = 0$ for all $l=1, \ldots, L$ and $a=1, \ldots, \cA$.
+For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\sA}$.
 Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
@@
 To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
-In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\xi_k$ and $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$.
+In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$.
 
@@
 Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
-\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i, x_i].\]
+\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
 

### 2022-01-03 slides/IBC2020/abstract/v1/abstract.tex
restructuration : include oral pres slides

\documentclass[11pt]{article}

\usepackage[colorlinks=true,
            linkcolor=red,
            urlcolor=blue,
            citecolor=blue,
            breaklinks=true]{hyperref}
\usepackage{breakurl} 
\usepackage[a4paper, left=3cm, right=3cm, top=3cm, bottom=3cm]{geometry}
\RequirePackage{amsmath,amsfonts, amssymb,amsthm}

\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{mathrsfs} 
\usepackage{natbib}
\usepackage{setspace}
\usepackage{dsfont}
\usepackage{caption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{booktabs}

\hypersetup{pdfauthor = {Simon Bussy}}

\usepackage{authblk}
\title{\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}}
\author[1]{Simon Bussy\thanks{Corresponding author: \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}}}
\author[2]{Antoine Barbieri}
\author[1]{Sarah Zohar}
\author[1,3]{Anne-Sophie Jannot}
\affil[1]{INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France}
\affil[2]{INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France}
\affil[3]{Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France}
\setcounter{Maxaffil}{0}
\renewcommand\Affilfont{\itshape\small}

\DeclareMathOperator{\argmin}{argmin}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\cM}{\mathcal M}
\newcommand{\cD}{\mathcal D}
\newcommand{\cC}{\mathcal C}
\newcommand{\cN}{\mathcal N}
\newcommand{\cA}{\mathcal A}
\newcommand{\R}{\mathds R}
\newcommand{\N}{\mathds N}
\newcommand{\E}{\mathds E}
\renewcommand{\P}{\mathds P}
\newcommand{\bSigma}{\textbf{$\Sigma$}}

\date{}

\begin{document}

\maketitle

\vspace{-.5cm}

\begin{abstract}

In many clinical studies, it has become increasingly common to record the values of longitudinal covariates until the occurrence of an event of interest. 
With the development of electronic health records and an increased focus on personalized medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes simultaneously is critical. Despite this, the state-of-the-art methods in this context have predominantly focused on univariate data, or on problems where the number of longitudinal outcomes are very low.
Our paper introduces a prognostic method called lights (generaLized joInt hiGH-dimensional longiTudinal Survival) to deal with the problem of joint modeling of longitudinal data and censored durations, in a high-dimensional context. 
The later introduces a latent variable modeling the heterogeneity within the patient population, with subgroups of different risk, and
supposes a group-specific marker trajectory with a generalized linear mixed model for each longitudinal marker given the subgroup, and a group-specific Cox risk of event with multiple shared association defined through a known functional family.
Inference is achieved using a novel fast stochastic approximation of a quasi-newton EM algorithm, by minimizing the negative log-likelihood penalized with elastic-net or group lasso regularization on the different parameter vectors of the model, depending on the desired interpretability power.
The estimated latent class membership posterior probabilities are  used as discriminative marker rule in the cross-validation procedure for selecting the best regularization hyper-parameters.
The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
On this high-dimensional dataset, our proposed method outperforms the state-of-the-art models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster. In addition, it provides powerful interpretability by automatically pinpointing significant covariates being relevant from a clinical perspective. Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.

\end{abstract}

\end{document}

### 2022-01-03 slides/IBC2020/abstract/v2/abstract.tex
restructuration : include oral pres slides

--- 2022-01-03 slides/IBC2020/abstract/v1/abstract.tex
@@
 
-In many clinical studies, it has become increasingly common to record the values of longitudinal covariates until the occurrence of an event of interest. 
+In many clinical studies, it has become increasingly common to record the values of longitudinal outcomes until the occurrence of an event of interest. 
 With the development of electronic health records and an increased focus on personalized medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes simultaneously is critical.
@@
 Our paper introduces a prognostic method called lights (generaLized joInt hiGH-dimensional longiTudinal Survival) to deal with the problem of joint modeling of longitudinal data and censored durations, in a high-dimensional context.
-The later introduces a latent variable modeling the heterogeneity within the patient population, with subgroups of different risk, and
-supposes a group-specific marker trajectory with a generalized linear mixed model for each longitudinal marker given the subgroup, and a group-specific Cox risk of event with multiple shared association defined through a known functional family.
+The later introduces a latent variable modeling the heterogeneity within the patient population, with subgroups at different risks, and supposes a group-specific marker trajectory with a generalized linear mixed model for each longitudinal marker given the subgroup.
+Furthermore, a group-specific Cox risk of event includes multiple shared associations defined through a known functional family.
 Inference is achieved using a novel fast stochastic approximation of a quasi-newton EM algorithm, by minimizing the negative log-likelihood penalized with elastic-net or group lasso regularization on the different parameter vectors of the model, depending on the desired interpretability power.
@@
 The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
-On this high-dimensional dataset, our proposed method outperforms the state-of-the-art models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
-In addition, it provides powerful interpretability by automatically pinpointing significant covariates being relevant from a clinical perspective.
+On this high-dimensional dataset, our proposed method is compared to the state-of-the-art models regarding risk prediction in terms of C-index, and regarding computing times.
+It provides powerful interpretability by automatically pinpointing significant covariates being relevant from a clinical perspective.
 Thus, we propose a powerfull tool for personalized medicine, with the ability of automatically determining significant prognostic longitudinal biomarkers, which is of increasing importance in many areas of medicine.

### 2022-01-03 slides/IBC2020/oral/v2/pres.tex
restructuration : include oral pres slides

\documentclass{beamer}

\mode<presentation>
{
  \usetheme{Goettingen}      
  \usecolortheme{default}
  \usefonttheme{default}
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{booktabs,multirow,array}
\usepackage{algorithm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xcolor}
\usepackage{moresize}
\usepackage{lmodern}
\usepackage{mathrsfs}

\usepackage[numbers]{natbib}
\renewcommand\bibsection{\section[]{\refname}}

\definecolor{blue_pres}{RGB}{51,50,178}
\definecolor{darkpastelgreen}{rgb}{0.01, 0.75, 0.24}

\hypersetup{
    colorlinks,
    citecolor=orange,
    linkcolor=blue_pres,
    backref=true,
    urlcolor=darkpastelgreen
}

\makeatletter
\let\@@magyar@captionfix\relax
\makeatother

\DeclareCaptionFormat{myformat}{#3}
\captionsetup[algorithm]{format=myformat}
\usepackage{algpseudocode}

%%%%%%%%%%%%%%%%%%%%
%%% New commands %%%
%%%%%%%%%%%%%%%%%%%%

\DeclareMathOperator{\TV}{TV}
\DeclareMathOperator{\bina}{bina}
\DeclareMathOperator{\argmin}{argmin}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\bC}{\mathbb C}
\newcommand{\bD}{\mathbb D}
\newcommand{\bP}{\mathbb P}
\newcommand{\bX}{\boldsymbol X}
\newcommand{\XB}{{\boldsymbol X}^B}
\newcommand{\E}{\mathbb E}
\newcommand{\N}{\mathbb N}
\newcommand{\bigO}{\mathcal{O}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\R}{\mathbb R}
\newcommand{\cA}{\mathcal A}
\newcommand{\cB}{\mathcal B}
\newcommand{\cH}{\mathcal H}
\newcommand{\cS}{\mathcal S}
\newcommand{\cM}{\mathcal M}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cU}{\mathcal U}
\renewcommand{\P}{\mathds{P}}
\newcommand{\Gb}{\bar G}
\newcommand{\Fb}{\bar F}
\newcommand{\Ub}{\bar U}
\newcommand{\by}{\boldsymbol y}
\newcommand{\bz}{\boldsymbol z}
\newcommand{\bdelta}{\boldsymbol \delta}
\newcommand{\bSigma}{\textbf{$\Sigma$}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\pluseq}{\mathrel{+}=}
\newcommand{\minuseq}{\mathrel{-}=}
\newcommand{\bcdot}{\raisebox{-0.25ex}{\scalebox{1.2}{$\cdot$}}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}} 
\newcommand{\cD}{\mathcal D}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cR}{\mathcal R}
\renewcommand{\P}{\mathds P}

\definecolor{blue_pres}{RGB}{51,50,178}

\newcommand\Algphase[1]{%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
\vspace*{-.4\baselineskip}\Statex\hspace*{\algorithmicindent}\textbf{#1}%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
}
\newcommand*\rot{\rotatebox{90}}
\newcommand*\OK{\ding{51}}

\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newtheorem{hyp}{Hypothesis}

\title[IBC 2020 \\ Lights \\ \insertframenumber/\inserttotalframenumber \vspace{-1cm}]{}

\institute[shortinst]{
  \vspace{-2.2cm} \\
  \includegraphics[width=2.5cm]{figures/SU} \hspace{1cm}
  \includegraphics[width=2.2cm]{figures/inserm} \hspace{1cm}
  \includegraphics[width=1.7cm]{figures/Califrais_logo} \\
  \vspace{.6cm}
  \textcolor{blue_pres}{\Large IBC 2020 Online Learning Series}\\
  \vspace{.5cm}
  \large
  \textbf{Simon Bussy}\\
  \vspace{.25cm}
  \scriptsize
  \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}\\
  \vspace{.4cm}
  \small
  \textbf{Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations} \\
  \vspace{.3cm}
  S. Bussy $^{(1,2)}$, A. Barbieri $^{(3)}$, S. Zohar $^{(1)}$, A.S. Jannot $^{(1,4)}$\\

  \vspace{.4cm}

  \tiny

  $^{(1)}$INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France. \\
  $^{(2)}$Califrais' Machine Learning Lab, Paris, France. \\
  $^{(3)}$INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France. \\
  $^{(4)}$Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France. 
}

\date{}

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\section{Introduction}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{I.} Introduction
\end{frame}

\subsection{Overview}

\begin{frame}{Overview}

\begin{itemize}
  \item<1-> Deal with the problem of joint modeling of longitudinal data and censored durations
  \item<2-> Large number of both longitudinal and time-independent features are available
  \item<3-> Flexibility in modeling the dependency between the longitudinal features and the event time with appropriate penalties
  \item<4-> Inference achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm
\end{itemize}

\end{frame}

\subsection{Use cases}

\begin{frame}{Use cases}

\small
\begin{itemize}
  \item<1-> Predict the risk for an event of interest to occur quickly, taking into account simultaneously a huge number of longitudinal signals in a high-dimensional context
  \item<2-> Provides powerful interpretability by automatically pinpointing significant time-dependent and time-independent features
\end{itemize}

\begin{block}<3->{Real-time decision support}
\begin{itemize}
  \item<3-> Medical context $\rightarrow$ event of interest: survival time, re-hospitalization, relapse or disease progression ; longitudinal data: biomarkers or vital parameters measurements
  \item<4-> Customer's satisfaction monitoring context $\rightarrow$ event of interest: time when a client churns ; longitudinal data: the client's activity recorded from account opening throughout the duration of the business relationship
\end{itemize}
\end{block}

\end{frame}

\subsection{Framework}

\begin{frame}{High-dimensional framework}

\small
\begin{itemize}
  \item<1-> Survival analysis
  \begin{center}
  $T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}$
  \end{center}
  \item<2-> Time-independent features $X \in \R^p$ with $p \gg n$
  \item<3-> $L$ longitudinal outcomes such that $L \gg n$ and \[Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L\]
  \item<4-> Heterogeneity of the population: latent subgroups \[G \in \{ 0, \ldots, K-1\}\]
  \item<5-> Softmax link function for the latent class membership probability given time-independent features
  \begin{center}
  $\pi_{\xi_k}(x) = \P[G=k|X=x] = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}$
  \end{center}
\end{itemize}

\end{frame}

\section{Model}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{II.} Model
\end{frame}

\subsection{Submodels}

\begin{frame}{Submodels}

\small
\begin{block}<1->{Group-specific marker trajectories}
\begin{itemize}
  \item<1-> \footnotesize $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
  with fixed effect parameters $\beta_k^l \in \R^{q_l}$ and subject-and-longitudinal outcome specific random effects $b^l \in \R^{r_l} \sim \cN(0, D_{ll})$
  \item<2-> $\text{Cov}[b^l,b^{l'}] = D_{ll'}$
  \only<2>{ and
  \[ D = 
  \begin{bmatrix}
    D_{11} & \cdots & D_{1L}\\
    \vdots &  \ddots & \vdots \\
    D_{1L}^\top & \cdots & D_{LL}
  \end{bmatrix}
  \]
  the global variance-covariance matrix}

\end{itemize}
\end{block}

\begin{block}<3->{Group-specific risk of event}
\begin{itemize}
  \item<3-> \footnotesize $\lambda(t|G = k) = \lambda_0(t) \exp \Big\{ x^\top \gamma_{k,0} + \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_a(t, \beta_k^l, b^l) \Big\}$
  \item<4-> Functionals $(\varphi_a)_{a\in \cA}$
\begin{table}[htb]
\centering
\resizebox{.9\textwidth}{!}{%
\begin{tabular}{cccc}
\toprule
Description & $\varphi_a(t, \beta_k^l, b^l)$ & $\frac{\partial \varphi_a(t, \beta_k^l, b^l)}{\partial \beta_k^l}$  & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & $u^l(t)$ & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $\mathbf{0}_{q_l}$ & \citet{hatfield2011joint} \\ 
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & $\dfrac{\dd}{\dd t} u^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & $\int_0^t u^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}}
\end{table}

\end{itemize}
\end{block}

\end{frame}

\subsection{Likelihood}

\begin{frame}{Likelihood}

\footnotesize
\begin{itemize}
  \item<1-> $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
  \item<2-> $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ with $n_i = \sum_{l=1}^L n_i^l$
  \item<3-> $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
  \item<4-> \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0(t), \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$ 
  \item<5-> Survival part: 
  \begin{align*}f(t_i, \delta_i| b_i, G_i = k ; \theta) = &\big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}
  \end{align*}
  \item<6-> Then, the likelihood writes
  \begin{align*}\ell_n(\theta) = n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} &\pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) \\ & \times f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i
  \end{align*}
\end{itemize}

\end{frame}

\section{Inference}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{III.} Inference
\end{frame}

\subsection{Penalization}

\begin{frame}{Penalization}

\small

\begin{itemize}
  \item<1-> Penalized objective
  \begin{equation*}
  \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}
  \end{equation*}
  with the elasticnet penalty \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \] and the sparse group lasso penalty \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \]
  \item<2-> Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
\end{itemize}

\end{frame}

\subsection{MCQNEM}

\begin{frame}{MCQNEM algorithm}

\small

\begin{itemize}
  \item<1-> $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
\end{itemize}

\begin{block}<2->{Monte Carlo E-step}
\begin{itemize}
  \item<2-> $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
\end{itemize}
\end{block}

\begin{block}<3->{Quasi-Newton M-step}
\begin{itemize}
  \item<3-> \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}} + \zeta_{3,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}$
\end{itemize}
\end{block}

\begin{itemize}
\item<4-> Predictive marker \[\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)},\] which is an estimate of \[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]\]
\end{itemize}

\end{frame}

\section{Conclusion}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{V.} Conclusion
\end{frame}

\begin{frame}{Conclusion}

\begin{itemize}
  \item<1-> Prognostic method called lights introduced to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available
  \item<2-> Penalization of the likelihood in order to perform feature selection and to prevent overfitting
  \item<3-> New efficient estimation algorithm (QNMCEM) has been derived
  \item<4-> Automatically determines significant prognostic longitudinal features
\end{itemize}

\begin{block}<4->{\texttt{Python} 3 package}
\begin{itemize}
  \item<4-> Available at \small{\url{https://github.com/Califrais/lights}}
  \item<5-> Applications of the model available soon on an \texttt{arXiv} paper.
\end{itemize}
\end{block}

\end{frame}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{} Thank you!
\end{frame}

\section{References}
\frame{
  \small
  \frametitle{References}
  \bibliographystyle{plainnat}
  \bibliography{refs}
}

\end{document}

### 2022-01-31 manuscript/paper.tex
Update the survival model with extracted association features from tsfresh

--- 2022-01-03 paper.tex
@@
 In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.
+
+\subsection{Group-specific risk of event}
+
+To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
+\begin{equation}
+  \label{eq:intensity-model}
+  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \Psi_a^l \Big\},
+\end{equation}
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function.
+
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known features $\Psi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\cA}$ the corresponding joint association parameters.
+These features are extracted from $Y$ by the \texttt{Python} library \texttt{tsfresh}.
+This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}.
+Let us finally denote
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
+\begin{equation}
+  \label{eq:def-asso-features}
+  \lambda(t|\Psi, G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi \big\}.
+\end{equation}
 

### 2022-01-31 manuscript/paper.tex
Update the assumption about dependency of latent variables

@@
 \label{assumption1}
-We suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
+We suppose that the random effects are conditional independent of the group membership on the association features, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
 \end{assumption}

### 2022-01-31 manuscript/paper.tex
Remove out-of-date survival model

@@
 In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.
-
-\subsection{Group-specific risk of event}
-
-To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
-\begin{equation}
-	\label{eq:intensity-model}
-	\lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_{k,a}(t, b^l) \Big\},
-\end{equation}
-where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
-\[\cM_k(t) = \{m_k^l(u), 0 \leq u < t\}\] 
-the history of the true unobserved longitudinal process up to time $t$.
-For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\varphi_{k,a}$ defining a shared association with $\gamma_{k,a}^l \in \R^{\imath_a}$ the corresponding joint association parameters, and $\imath_a \in \N_+$ the dimension of the corresponding $\text{Im}(\varphi_{k,a})$.
-This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}. Let us finally denote
-\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\sA}, \] with $\sA = \sum_{a=1}^\cA \imath_a$ and so that
-\begin{equation}
-  \label{eq:def-asso-features}
-  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi_k(t) \big\},
-\end{equation}
-where $\Psi_k(t) \in \R^{L\sA}$ denotes the association features vector at time $t$.
-
-\paragraph{Specification of the functionals $(\varphi_{k,a})_{a\in \cA}$.}
-
-The association structure between the longitudinal and the time-to-event submodels is key to the joint modeling framework.
-In spite of that, rationale for selecting shared associations has received little attention.
-We then propose to include some of the most common parameterization with no \textit{a priori}, set out in Table~\ref{table:shared_associations}, and let the model select the relevant ones through the regularization strategy described in Section~\ref{sec:penalized-obj}.
-\begin{table}[htb]
-\centering
-\begin{tabular}{ccccc}
-\toprule
-Description & $\varphi_{k,a}(t, b^l)$ & $\imath_a$ & Reference \\
-\midrule
-Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
-Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
-Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
-Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
- \bottomrule
-\end{tabular}
-\caption[]{Description of the shared associations included in the group-specific risk of event submodel~\eqref{eq:intensity-model}.}
-\label{table:shared_associations}
-\end{table}
 

### 2022-01-31 manuscript/paper.tex
Update the graphical representation

@@
   \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
-  \node[main, fill = white!100] (b) [below=0.7cm of Y] {\Large$\bm{b_i^l}$};
+  \node[main, fill = white!100] (b) [above=of mu] {\Large$\bm{b_i^l}$};
   \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
-  \node[var, fill = white!100] (Sigma) [right=of b] {\Large$\bm{D}$};
-  \node[main, fill = white!100] (W) [below=of b] {\Large$\bm{\varphi_i(t)}$};
+  \node[var, fill = white!100] (Sigma) [left=of b] {\Large$\bm{D}$};
+  \node[main, fill = black!30] (W) [below=3cm of Y] {\Large$\bm{\Psi_i}$};
   \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
@@
         (lambda) edge (T)
+        (Y) edge (W)
         ;
@@
   \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (12, 3.5) {\Large$\bm{l = 1...L}$};
-  \node[rectangle, dashed, inner sep=3mm, draw=red!100, fit= (b) (W),ultra thick] {};
+  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (8, 4.1) {\Large$\bm{l = 1...L}$};
 

### 2022-01-31 manuscript/paper.tex
Update the dataset

@@
 Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
-\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
+\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1, \Psi_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n, \Psi_n) \big\} \]
 where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that

### 2022-01-31 manuscript/paper.tex
Update the assumption 3

@@
 \label{indep-hyp-3}
-Assume that both the random effects vector $b_i$ and the group membership account for the association between the longitudinal and event outcomes, that is
+Assume that the random effects vector $b_i$, the group membership account for the association between the longitudinal and event outcomes and the association features $\Psi_i$, that is
 \begin{equation}
   \label{eq:ind-hyp}
-  f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta) = f(t_i, \delta_i| b_i, G_i = k ; \theta) f(y_i | b_i, G_i = k ; \theta)
+  f(t_i, \delta_i, y_i | b_i, G_i = k, \Psi_i ; \theta) = f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k ; \theta)
 \end{equation}

### 2022-01-31 manuscript/paper.tex
Update the likelihood

@@
 	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
-	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(y_i | G_i = k)\int_{\R^r} f(t_i, \delta_i| b_i, G_i = k ; \theta) f(b_i | y_i, G_i = k; \theta) \dd b_i,
+	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) \int_{\R^r} f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
+  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i,
 \end{align}
 where 
-\[f(t_i, \delta_i| b_i, G_i = k ; \theta) = \big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}. \]
+\[f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = \big[\lambda(t_i|G_i = k, \Psi_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|G_i = k, \Psi_i) \dd s \Big\}. \]
 
@@
 
-In order to derive an algorithm for this objective, we introduce a so-called prox-QNMCEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum} with Monte Carlo approximations~\citep{levine2001implementations}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}.
+In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}.
 EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_{k,i}(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_{k,i}(s) \big\} \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i \Big) - \exp \big\{ \gamma_k^\top \Psi_i \big\} \int_0^{t_i} \lambda_0(s) \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].

### 2022-01-31 manuscript/paper.tex
Update the Lambda_g

@@
 \begin{align}
-  \label{eq:Lambda_ik-b_i|y_i}
-  \Lambda_{ik,\tilde g}^{\theta^{(w)}} &= f(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i \\
+  \Lambda_{ik,\tilde g}^{\theta^{(w)}} &= \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i | b_i, G_i = k, \Psi_i ; \theta^{(w)}) f(y_i | b_i, G_i = k, \Psi_i ; \theta^{(w)}) f(b_i | G_i = k, \Psi_i; \theta^{(w)}) \dd b_i \\
   \label{eq:Lambda_ik-b_i}
-  &= \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i
+  &=  f(t_i, \delta_i | G_i = k, \Psi_i ; \theta^{(w)}) f(y_i | G_i = k ; \theta^{(w)}) \int_{\R^r} \tilde g(b_i) f(b_i | y_i, G_i = k ; \theta^{(w)}) \dd b_i,
 \end{align}
-and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$.
-The integrals in~\eqref{eq:Lambda_ik-b_i|y_i} or~\eqref{eq:Lambda_ik-b_i} are not tractable analytically, and some form of approximation must be used in practice.
-Standard numerical integration techniques (such as Gaussian quadrature) are not well suited in our high-dimensional context.
-We then propose to use Monte Carlo approximation, which has already been applied for generalized linear mixed models~\citep{booth1999maximizing}, with antithetic normal variates method~\citep{hammersley1956new} to reduce variance of the simulation result.
-Algorithm~\ref{alg:construct-set-S} describes how to construct the set $S_{ik}^{(w)}$ used for the approximation.
-
-\vspace{.5cm}
-
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\caption{Construction of the samples in $S_{ik}^{(w)}$}
-\label{alg:construct-set-S}
-\KwInput{$S_{ik}^{(w)} = \varnothing$}
-\KwOutput{$S_{ik}^{(w)}$ filled with $2N$ samples}
-Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
-$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i^1$ \tcp{Cholesky decomposition} \nonumber
-\For{$\breve{n} = 1,\ldots,N$}    
-{ 
-  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
-  Compute $\breve{b}_{\breve{n}}^\pm = \text{M}_{ik}^2 \pm C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
-  Update $S_{ik}^{(w)} \leftarrow S_{ik}^{(w)} \cup \{\breve{b}_{\breve{n}}^+, \breve{b}_{\breve{n}}^- \}$
-}
-\textbf{Return:} {$S_{ik}^{(w)}$}
-\end{algorithm}
-
-\vspace{.5cm}
-
-\noindent One can either choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= W_i \\ \text{M}_{ik}^2 &= W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k) \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i|y_i} and where $\Sigma_i$ is the diagonal matrix whose diagonal is $\Phi_i$ (defined in~\eqref{eq:def-Phi_i}), where in the Gaussian case used in practice, one can sample from
+where
 \begin{align*}
@@
 \end{align*}
-with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$, or one can choose $\left\{ \begin{array}{ll} \text{M}_i^1 &= D^{(w)} \\ \text{M}_{ik}^2 &= 0 \end{array} \right.$ based on~\eqref{eq:Lambda_ik-b_i} using a sampling $b_i \sim \cN(0, D^{(w)})$.
-The huge advantage of this second option is that $S_{ik}^{(w)}$ does not depend on $i$ nor $k$, which significantly improves computational performances.
-We then implement this option in practice, being more appropriated in our high-dimensional context.
-
-The approximation of~\eqref{eq:def-E-g-tilde} is finally obtained by
-\begin{equation}
-  \label{eq:def-hat-E-g-tilde}
-  \hat \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}
-\end{equation}
-with
-\begin{equation}
-  \label{eq:hat_Lambda_ik}
-  \hat \Lambda_{ik,\tilde g}^{\theta^{(w)}} = \dfrac{1}{2N} \sum_{\breve{b} \in S_{ik}^{(w)}} \tilde g(\breve{b}) f(t_i, \delta_i, y_i | \breve{b}, G_i = k ; \theta^{(w)}).
-\end{equation}
-
-It is common in Monte Carlo EM to increase $N$ with the iterations $w$~\citep{wei1990monte} since it is computationally inefficient to use a large $N$ in the early steps of the algorithm, when $\theta^{(w)}$ is far from $\hat \theta$.
-Multiple techniques have been proposed (see for instance~\citet{law2002joint} where a subjectively rule is followed, or~\citet{booth1999maximizing} that uses a rule based on confidence intervals for $\theta^{(w)}$ which requires additional variance estimation).
-We opt for a simple automated approach using relative differences of the objective function defined in~\eqref{eq:pen-log-lik} (see Algorithm~\ref{alg:prox-QNMCEM}).
+with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$.
 

### 2022-01-31 manuscript/paper.tex
Update Monte-Carlo stuff

@@
 for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
-Inference is achieved using an efficient and novel proximal Quasi-Newton Monte Carlo Expectation Maximization algorithm (prox-QNMCEM).
+Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
 Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
@@
 
-\subsection{A Quasi-Newton Monte Carlo EM}
-\label{sec:prox-QNMCEM}
+\subsection{A Quasi-Newton EM}
+\label{sec:prox-QNEM}
 
@@
 
-\paragraph*{Monte Carlo E-step.}
+\paragraph*{E-step.}
 
@@
 Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNMCEM}.
+We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
 First, the update of $D^{(w)}$ is naturally given in closed-form by
@@
 Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNMCEM}) with every increase of $N$.
+We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNEM}) with every increase of $N$.
 
@@
 \begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_{k,i}(t_i) | t_i, \delta_i, y_i] \\
-  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
+  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \gamma_k^\top \Psi_i -  \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align*}
@@
   \label{eq:grad-Q}
-  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
-   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Psi_{i, a}^l  \Big[ \delta_i - \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_{k,i}(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i\big\} \ind{\{t_i \geq t\}} },
 \end{equation}
@@
 \end{align}
-Appendix~\ref{sec:prox-QNMCEM class} gives some details about the way we implement the equations and their computation in practice.
+Appendix~\ref{sec:prox-QNEM class} gives some details about the way we implement the equations and their computation in practice.
 
@@
 Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced.
-We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNMCEM}.
+We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNEM}.
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\sA}$.
@@
 
-\paragraph*{The prox-QNMCEM algorithm.}
-Algorithm~\ref{alg:prox-QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
+\paragraph*{The prox-QNEM algorithm.}
+Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
 
@@
 \DontPrintSemicolon
-\caption{prox-QNMCEM algorithm for the lights model inference}
-\label{alg:prox-QNMCEM}
+\caption{prox-QNEM algorithm for the lights model inference}
+\label{alg:prox-QNEM}
 \KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
@@
 { 
-  \tcc{Monte Carlo E-step}
-  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
+  \tcc{E-step}
+  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for the required functions $\tilde g$\\
   Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
@@
 \[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
-Then in practice, the convergence for Algorithm~\ref{alg:prox-QNMCEM} is declared when the condition
+Then in practice, the convergence for Algorithm~\ref{alg:prox-QNEM} is declared when the condition
 \[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
-is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.
+is satisfied for 3 consecutive iterations.
 
@@
 The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
-Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNMCEM algorithm introduced in Section~\ref{sec:prox-QNMCEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
 Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
@@
 In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
-algorithm (prox-QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform feature
+algorithm (prox-QNEM) has been derived, that considers a penalization of the likelihood in order to perform feature
 selection and to prevent overfitting.
@@
 
-Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNMCEM algorithm in Section~\ref{sec:prox-QNMCEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
+Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNEM algorithm in Section~\ref{sec:prox-QNEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
 For the sake of simplicity, let us denote here 
@@
 
-This class implements an EM algorithm for fitting a multivariate linear mixed model used to initialize the MCQNEM algorithm.
+This class implements an EM algorithm for fitting a multivariate linear mixed model used to initialize the QNEM algorithm.
 Let us introduce the list $\Omega^{(w)} = [\Omega_1^{(w)}, \ldots, \Omega_n^{(w)}]$, the matrices $\mu = [\mu_1, \ldots, \mu_n] \in \R^{r \times n}$, $U^l = [{U_{1l}}^\top \cdots {U_{nl}}^\top]^\top \in \R^{n_l \times q_l}$, $U = [U_1^\top \cdots U_n^\top]^\top \in \R^{\cN \times q}$,

### 2022-01-31 manuscript/paper.tex
Update g functions

@@
 
-\subsection{\texttt{prox-QNMCEM class}}
-\label{sec:prox-QNMCEM class}
+\subsection{\texttt{prox-QNEM class}}
+\label{sec:prox-QNEM class}
 
@@
     \tilde g^1&: b_i \mapsto b_i,\\
-    \tilde g^2&: b_i \mapsto b_ib_i^\top, \\
-    \tilde g^3_{s, k}&: b_i \mapsto \Psi_{k,i}(s), \\
-    \tilde g^4_{s, \gamma_k}&: b_i \mapsto \exp \big\{\gamma_k^\top g^3_{s, k}(b_i) \big\}, \\
-    \tilde g^5_{s, \gamma_k}&: b_i \mapsto g^3_{s, k}(b_i)^\top g^4_{s, k}(b_i).
+    \tilde g^2&: b_i \mapsto b_ib_i^\top.
     \end{array}
@@
 \end{equation*}
-We also denote 
-\[ \cI_1 = 
-\begin{bmatrix}
-  \ind{\{t_{1} = t_1^u\}}  & \cdots & \ind{\{t_{1} = t_J^u\}}\\
-  \vdots &  \ddots & \vdots \\
-  \ind{\{t_{n} = t_1^u\}} & \cdots & \ind{\{t_{n} = t_J^u\}}
-\end{bmatrix}
-\in \R^{n \times J}
-\quad \text{and} \quad 
-\cI_2 = 
-\begin{bmatrix}
-  \ind{\{t_{1} \leq t_1^u\}} & \cdots & \ind{\{t_{1} \leq t_J^u\}}\\
-  \vdots &  \ddots & \vdots \\
-  \ind{\{t_{n} \leq t_1^u\}} & \cdots & \ind{\{t_{n} \leq t_J^u\}}
-\end{bmatrix}
-\in \R^{n \times J} \]
-some matrices to compare each event times with the vector of distinct event times. Those matrices are widely used in practice to replace indicators. For instance, the quantities
-\begin{align*}
-f(t_i, \delta_i| \breve{b}, G_i = k ; \theta^{(w)}) = &\big(\lambda_0^{(w)}(t_i) g^4_{t_i, \gamma_k^{(w)}}(\breve{b}) \big)^{\delta_i} \\ 
-&\times \exp \Big\{ - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}) \ind{\{t_j^u \leq t_i\}} \Big\}
-\end{align*}
-for $i = 1, \ldots, n$ and $\breve{b} \in S^{(w)}$ can be represented in the following matrix 
-\[ \underset{J}{\text{sum}} \big( \boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_1 \big)^{\boldsymbol{\delta}} \exp \big\{- \underset{J}{\text{sum}} \big(\boldsymbol{g^4_{\gamma_k}} \boldsymbol{\lambda_0^{(w)}} \cI_2 \big) \big\} \in \R^{n \times 2N} \]
-where 
-\begin{equation*}
-  \left\{
-    \begin{array}{ll}
-    \boldsymbol{g^4_{\gamma_k}} \in \R^{2N \times J} \text{ with } \big[\boldsymbol{g^4_{\gamma_k}}\big]_{i, j} = \tilde g^4_{t_j^u, \gamma_k^{(w)}}(\breve{b}_i) , \\
-    \boldsymbol{\lambda_0^{(w)}} = \big(\lambda_0^{(w)}(t_1^u), \ldots, \lambda_0^{(w)}(t_J^u) \big)^\top \in \R^J, \\
-    \boldsymbol{\delta} = (\delta_1, \ldots, \delta_n)^\top \in \R^n.
-    \end{array}
-  \right.
-\end{equation*}
-Another example is the update of the baseline hazard~\eqref{eq:update_lambda_0} as
-\[\boldsymbol{\lambda_0^{(w+1)}} = \underset{n}{\text{sum}}( \boldsymbol{\delta} \cI_1) / \underset{n}{\text{sum }} \underset{k}{\text{sum }} (\boldsymbol{g^4_{\gamma_k}} \cI_2 \boldsymbol{\hat \pi^{(w)}})\]
-with $\boldsymbol{\hat \pi^{(w)}} \in \R^{2 \times n} \text{ and }  \left\{
-    \begin{array}{ll}
-    \boldsymbol{\hat \pi_{1,\bullet}^{(w)}}= (\hat \pi_{11}^{\theta^{(w)}}, \ldots, \hat \pi_{n1}^{\theta^{(w)}})^\top \in \R^n, \\
-    \boldsymbol{\hat\pi_{0,\bullet}^{(w)}} = 1 - \boldsymbol{\hat\pi_{1,\bullet}^{(w)}}.
-    \end{array}
-  \right.$
 

### 2022-02-02 slides/IASD/pres.tex
Initial commit for IASD

\documentclass{beamer}

\mode<presentation>
{
  \usetheme{Goettingen}      
  \usecolortheme{default}
  \usefonttheme{default}
  \setbeamertemplate{navigation symbols}{}
  \setbeamertemplate{caption}[numbered]
}

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{booktabs,multirow,array}
\usepackage{algorithm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{url}
\usepackage{xcolor}
\usepackage{moresize}
\usepackage{lmodern}
\usepackage{mathrsfs}

\usepackage[numbers]{natbib}
\renewcommand\bibsection{\section[]{\refname}}

\definecolor{blue_pres}{RGB}{51,50,178}
\definecolor{darkpastelgreen}{rgb}{0.01, 0.75, 0.24}

\hypersetup{
    colorlinks,
    citecolor=orange,
    linkcolor=blue_pres,
    backref=true,
    urlcolor=darkpastelgreen
}

\makeatletter
\let\@@magyar@captionfix\relax
\makeatother

\DeclareCaptionFormat{myformat}{#3}
\captionsetup[algorithm]{format=myformat}
\usepackage{algpseudocode}

%%%%%%%%%%%%%%%%%%%%
%%% New commands %%%
%%%%%%%%%%%%%%%%%%%%

\DeclareMathOperator{\TV}{TV}
\DeclareMathOperator{\bina}{bina}
\DeclareMathOperator{\argmin}{argmin}

\newcommand{\dd}{\mathrm{d}}
\newcommand{\bC}{\mathbb C}
\newcommand{\bD}{\mathbb D}
\newcommand{\bP}{\mathbb P}
\newcommand{\bX}{\boldsymbol X}
\newcommand{\XB}{{\boldsymbol X}^B}
\newcommand{\E}{\mathbb E}
\newcommand{\N}{\mathbb N}
\newcommand{\bigO}{\mathcal{O}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\R}{\mathbb R}
\newcommand{\cA}{\mathcal A}
\newcommand{\cB}{\mathcal B}
\newcommand{\cH}{\mathcal H}
\newcommand{\cS}{\mathcal S}
\newcommand{\cM}{\mathcal M}
\newcommand{\cC}{\mathcal C}
\newcommand{\cG}{\mathcal G}
\newcommand{\cN}{\mathcal N}
\newcommand{\cU}{\mathcal U}
\renewcommand{\P}{\mathds{P}}
\newcommand{\Gb}{\bar G}
\newcommand{\Fb}{\bar F}
\newcommand{\Ub}{\bar U}
\newcommand{\by}{\boldsymbol y}
\newcommand{\bz}{\boldsymbol z}
\newcommand{\bdelta}{\boldsymbol \delta}
\newcommand{\bSigma}{\textbf{$\Sigma$}}
\newcommand{\norm}[1]{\|#1\|}
\newcommand{\pluseq}{\mathrel{+}=}
\newcommand{\minuseq}{\mathrel{-}=}
\newcommand{\bcdot}{\raisebox{-0.25ex}{\scalebox{1.2}{$\cdot$}}}
\newcommand{\ind}[1]{\mathds{1}_{#1}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}} 
\newcommand{\cD}{\mathcal D}
\newcommand{\cQ}{\mathcal Q}
\newcommand{\cR}{\mathcal R}
\renewcommand{\P}{\mathds P}

\definecolor{blue_pres}{RGB}{51,50,178}

\newcommand\Algphase[1]{%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
\vspace*{-.4\baselineskip}\Statex\hspace*{\algorithmicindent}\textbf{#1}%
\vspace*{-.7\baselineskip}\Statex\hspace*{\dimexpr-\algorithmicindent-2pt\relax}\rule{\textwidth}{0.4pt}%
}
\newcommand*\rot{\rotatebox{90}}
\newcommand*\OK{\ding{51}}

\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newtheorem{hyp}{Hypothesis}

\begin{document}

\section{Introduction}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{I.} Introduction
\end{frame}

\subsection{Overview}

\begin{frame}{Overview}

\begin{itemize}
  \item Deal with the problem of joint modeling of longitudinal data and censored durations
  \item Large number of both longitudinal and time-independent features are available
  \item Flexibility in modeling the dependency between the longitudinal features and the event time with appropriate penalties
  \item Inference achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm
\end{itemize}

\end{frame}

\subsection{Use cases}

\begin{frame}{Use cases}

\small
\begin{itemize}
  \item Predict the risk for an event of interest to occur quickly, taking into account simultaneously a huge number of longitudinal signals in a high-dimensional context
  \item Provides powerful interpretability by automatically pinpointing significant time-dependent and time-independent features
\end{itemize}

\begin{block}{Real-time decision support}
\begin{itemize}
  \item Medical context $\rightarrow$ event of interest: survival time, re-hospitalization, relapse or disease progression ; longitudinal data: biomarkers or vital parameters measurements
  \item Customer's satisfaction monitoring context $\rightarrow$ event of interest: time when a client churns ; longitudinal data: the client's activity recorded from account opening throughout the duration of the business relationship
\end{itemize}
\end{block}

\end{frame}

\subsection{Framework}

\begin{frame}{High-dimensional framework}

\small
\begin{itemize}
  \item Survival analysis
  \begin{center}
  $T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}$
  \end{center}
  \item Time-independent features $X \in \R^p$ with $p \gg n$
  \item $L$ longitudinal outcomes such that $L \gg n$ and \[Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L\]
  \item Heterogeneity of the population: latent subgroups \[G \in \{ 0, \ldots, K-1\}\]
  \item Softmax link function for the latent class membership probability given time-independent features
  \begin{center}
  $\pi_{\xi_k}(x) = \P[G=k|X=x] = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}$
  \end{center}
\end{itemize}

\end{frame}

\section{Model}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{II.} Model
\end{frame}

\subsection{Submodels}

\begin{frame}{Submodels}

\small
\begin{block}{Group-specific marker trajectories}
\begin{itemize}
  \item \footnotesize $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
  with fixed effect parameters $\beta_k^l \in \R^{q_l}$ and subject-and-longitudinal outcome specific random effects $b^l \in \R^{r_l} \sim \cN(0, D_{ll})$
  \item $\text{Cov}[b^l,b^{l'}] = D_{ll'}$
  \only{ and
  \[ D = 
  \begin{bmatrix}
    D_{11} & \cdots & D_{1L}\\
    \vdots &  \ddots & \vdots \\
    D_{1L}^\top & \cdots & D_{LL}
  \end{bmatrix}
  \]
  the global variance-covariance matrix}

\end{itemize}
\end{block}

\end{frame}

\begin{frame}{Submodels}

\small
\begin{block}{Group-specific risk of event}
\begin{itemize}
  \item \footnotesize $\lambda(t|G = k) = \lambda_0(t) \exp \Big\{ \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_{k, a}(t, b^l) \Big\}$
  \item Functionals $(\varphi_a)_{a\in \cA}$
\begin{table}[htb]
\centering
\resizebox{.9\textwidth}{!}{%
\begin{tabular}{ccccc}
\toprule
Description & $\varphi_{k,a}(t, b^l)$ & $\imath_a$ & Reference \\
\midrule
Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
 \bottomrule
\end{tabular}}
\end{table}

\end{itemize}
\end{block}

\end{frame}

\begin{frame}{Submodels}
\begin{itemize}
\item Graphical representation of a joint model of a time-to-event submodel and K-multivariate longitudinal outcomes submodel
\includegraphics[scale=0.42]{figures/lights_graphical_representation.png}
\end{itemize}

\end{frame}
\subsection{Notations}

\begin{frame}{Notations}

\footnotesize
\begin{itemize}
  \item $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
  \item $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ with $n_i = \sum_{l=1}^L n_i^l$
  \item $b_i = ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r$ with $r = \sum_{l=1}^L r_l$
  \item Design matrices
  \[ U_i = 
\begin{bmatrix}
  U_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & U_{iL}
\end{bmatrix} 
\in \R^{n_i \times q}
\text{ and }
V_i = 
\begin{bmatrix}
  V_{i1} & \cdots & 0\\
  \vdots &  \ddots & \vdots \\
  0 & \cdots & V_{iL}
\end{bmatrix}
\in \R^{n_i \times r}
\]
with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
\[
\left\{
    \begin{array}{ll}
        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
    \end{array}
\right.
\]
  \item $\beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q$
  \item $M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}$
\end{itemize}

\end{frame}

\subsection{Likelihood}

\begin{frame}{Likelihood}

\footnotesize
\begin{itemize}
  \item \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$
  \item $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
  \item Survival part:
  \begin{align*}f(t_i, \delta_i| b_i, G_i = k ; \theta) = &\big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}
  \end{align*}
  \item Then, the likelihood writes
  \begin{align*}\ell_n(\theta) = n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} &\pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) \\ & \times f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i
  \end{align*}
\end{itemize}

\end{frame}

\section{Inference}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{III.} Inference
\end{frame}

\subsection{Penalization}

\begin{frame}{Penalization}

\small

\begin{itemize}
  \item Penalized objective
  \begin{equation*}
  \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
  \end{equation*}
  with the elasticnet penalty \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \] and the sparse group lasso penalty \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \]
  \item Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
\end{itemize}

\end{frame}

\subsection{QNMCEM}

\begin{frame}{QNMCEM algorithm (1/2)}

\small

\begin{itemize}
  \item $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
\end{itemize}

\begin{block}{Monte Carlo E-step}
\begin{itemize}
  \item $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
  \item Requires to compute expectations of the form
  \footnotesize
\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \] 
\small for different functions $g$, where we denote 
\[\pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] \]
  \item Monte Carlo approximations used for untractable integrals
\end{itemize}
\end{block}

\end{frame}

\begin{frame}{QNMCEM algorithm (2/2)}

\scriptsize

\begin{block}<1->{Quasi-Newton M-step}
\begin{itemize}
  \item \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
  \item $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
  \item \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
  \item $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
  \item $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
  \item $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
  \item L-BFGS-B to solve the problem
  \item Proximal gradient method to estimate $\gamma_k^{(w+1)}$
  \item Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
\end{itemize}
\end{block}

\end{frame}

\section{Evaluation}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{V.} Evaluation
\end{frame}

\begin{frame}{Experiments}
  \centering
  \includegraphics[height= 3cm, width=5cm]{figures/likelihood.png} \hspace{1cm} \\
  \vspace{0.5cm}
  \includegraphics[height= 2cm, width=4cm]{figures/beta_learning.png} \hspace{0.7cm}
  \includegraphics[height= 2cm, width=4cm]{figures/beta_estimation.png} \\
  \vspace{0.5cm}
  \includegraphics[height= 2cm, width=4cm]{figures/xi_learning.png} \hspace{0.7cm}
  \includegraphics[height= 2cm, width=4cm]{figures/xi_estimation.png} \\
\end{frame}

\section{Conclusion}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{VI.} Conclusion
\end{frame}

\begin{frame}{Conclusion}

\begin{itemize}
  \item Prognostic method called lights introduced to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available
  \item Penalization of the likelihood in order to perform feature selection and to prevent overfitting
  \item New efficient estimation algorithm (QNMCEM) has been derived
  \item Automatically determines significant prognostic longitudinal features
\end{itemize}

\begin{block}{\texttt{Python} 3 package}
\begin{itemize}
  \item Available at \small{\url{https://github.com/Califrais/lights}}
  \item Applications of the model available soon on an \texttt{arXiv} paper.
\end{itemize}
\end{block}

\end{frame}

\begin{frame}[noframenumbering]
\Large \centering
\textcolor{blue_pres}{} Thank you!
\end{frame}

\section{References}
\frame{
  \small
  \frametitle{References}
  \bibliographystyle{plainnat}
  \bibliography{refs}
}

\end{document}

### 2022-02-02 slides/IASD/pres.tex
Refactor the structure of itemize

@@
 \begin{itemize}
-  \item Deal with the problem of joint modeling of longitudinal data and censored durations
-  \item Large number of both longitudinal and time-independent features are available
-  \item Flexibility in modeling the dependency between the longitudinal features and the event time with appropriate penalties
-  \item Inference achieved using an efficient and novel Quasi-Newton Monte Carlo Expectation Maximization algorithm
+  \item<1-> Deal with the problem of joint modeling of longitudinal data and censored durations
+  \item<2-> Large number of both longitudinal and time-independent features are available
+  \item<3-> Flexibility in modeling the dependency between the longitudinal features and the event time with appropriate penalties
+  \item<4-> Inference achieved using an efficient and novel Quasi-Newton Expectation Maximization algorithm
 \end{itemize}
@@
 \begin{itemize}
-  \item Predict the risk for an event of interest to occur quickly, taking into account simultaneously a huge number of longitudinal signals in a high-dimensional context
-  \item Provides powerful interpretability by automatically pinpointing significant time-dependent and time-independent features
+  \item<1-> Predict the risk for an event of interest to occur quickly, taking into account simultaneously a huge number of longitudinal signals in a high-dimensional context
+  \item<2-> Provides powerful interpretability by automatically pinpointing significant time-dependent and time-independent features
 \end{itemize}
 
-\begin{block}{Real-time decision support}
+\begin{block}<3->{Real-time decision support}
 \begin{itemize}
-  \item Medical context $\rightarrow$ event of interest: survival time, re-hospitalization, relapse or disease progression ; longitudinal data: biomarkers or vital parameters measurements
-  \item Customer's satisfaction monitoring context $\rightarrow$ event of interest: time when a client churns ; longitudinal data: the client's activity recorded from account opening throughout the duration of the business relationship
+  \item<3-> Medical context $\rightarrow$ event of interest: survival time, re-hospitalization, relapse or disease progression ; longitudinal data: biomarkers or vital parameters measurements
+  \item<4-> Customer's satisfaction monitoring context $\rightarrow$ event of interest: time when a client churns ; longitudinal data: the client's activity recorded from account opening throughout the duration of the business relationship
 \end{itemize}
@@
 \begin{itemize}
-  \item Survival analysis
+  \item<1-> Survival analysis
   \begin{center}
@@
   \end{center}
-  \item Time-independent features $X \in \R^p$ with $p \gg n$
-  \item $L$ longitudinal outcomes such that $L \gg n$ and \[Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L\]
-  \item Heterogeneity of the population: latent subgroups \[G \in \{ 0, \ldots, K-1\}\]
-  \item Softmax link function for the latent class membership probability given time-independent features
+  \item<2-> Time-independent features $X \in \R^p$ with $p \gg n$
+  \item<3-> $L$ longitudinal outcomes such that $L \gg n$ and \[Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L\]
+  \item<4-> Heterogeneity of the population: latent subgroups \[G \in \{ 0, \ldots, K-1\}\]
+  \item<5-> Softmax link function for the latent class membership probability given time-independent features
   \begin{center}
@@
 \begin{itemize}
-  \item \footnotesize $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
+  \item<1-> \footnotesize $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
   with fixed effect parameters $\beta_k^l \in \R^{q_l}$ and subject-and-longitudinal outcome specific random effects $b^l \in \R^{r_l} \sim \cN(0, D_{ll})$
-  \item $\text{Cov}[b^l,b^{l'}] = D_{ll'}$
+  \item<2-> $\text{Cov}[b^l,b^{l'}] = D_{ll'}$
   \only{ and
@@
 \begin{itemize}
-  \item $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
-  \item $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ with $n_i = \sum_{l=1}^L n_i^l$
-  \item $b_i = ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r$ with $r = \sum_{l=1}^L r_l$
-  \item Design matrices
+  \item<1-> $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1, \Psi_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n, \Psi_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
+  \item<2-> $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ with $n_i = \sum_{l=1}^L n_i^l$
+  \item<3-> $b_i = ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r$ with $r = \sum_{l=1}^L r_l$
+  \item<4-> Design matrices
   \[ U_i = 
@@
 \]
-  \item $\beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q$
-  \item $M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}$
+  \item<5-> $\beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q$
+  \item<6-> $M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}$
 \end{itemize}
@@
 \begin{itemize}
-  \item Prognostic method called lights introduced to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available
-  \item Penalization of the likelihood in order to perform feature selection and to prevent overfitting
-  \item New efficient estimation algorithm (QNMCEM) has been derived
-  \item Automatically determines significant prognostic longitudinal features
+  \item<1-> Prognostic method called lights introduced to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of longitudinal features are available
+  \item<2-> Penalization of the likelihood in order to perform feature selection and to prevent overfitting
+  \item<3-> New efficient estimation algorithm (QNEM) has been derived
+  \item<4-> Automatically determines significant prognostic longitudinal features
 \end{itemize}
 
-\begin{block}{\texttt{Python} 3 package}
+\begin{block}<5->{\texttt{Python} 3 package}
 \begin{itemize}
-  \item Available at \small{\url{https://github.com/Califrais/lights}}
-  \item Applications of the model available soon on an \texttt{arXiv} paper.
+  \item<5-> Available at \small{\url{https://github.com/Califrais/lights}}
+  \item<6-> Applications of the model available soon on an \texttt{arXiv} paper.
 \end{itemize}

### 2022-02-02 slides/IASD/pres.tex
Update the survival model

@@
 \begin{itemize}
-  \item \footnotesize $\lambda(t|G = k) = \lambda_0(t) \exp \Big\{ \sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \varphi_{k, a}(t, b^l) \Big\}$
-  \item Functionals $(\varphi_a)_{a\in \cA}$
-\begin{table}[htb]
-\centering
-\resizebox{.9\textwidth}{!}{%
-\begin{tabular}{ccccc}
-\toprule
-Description & $\varphi_{k,a}(t, b^l)$ & $\imath_a$ & Reference \\
-\midrule
-Linear predictor & $m_k^l(t)$ & 1 & \citet{chi2006joint} \\ [.15cm]
-Random effects & $b^l$ & $r_l$ & \citet{hatfield2011joint} \\
-Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & 1 & \citet{rizopoulos2011bayesian} \\ [.3cm]
-Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & 1 & \citet{andrinopoulou2017combined} \\ [.1cm]
- \bottomrule
-\end{tabular}}
-\end{table}
+  \item<1-> \footnotesize $\lambda(t|\Psi, G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi \big\}$
+  \item<2-> Fetures $\Psi$ are extracted from $Y$ by the \texttt{Python} library \texttt{tsfresh}
+  \item<2-> \color{red}{TODO: Add descriptions}
 

### 2022-02-02 slides/IASD/pres.tex
Update the graphical representation
Update the likelihood
Update the predictive marker and minor refactore the slide

@@
 \item Graphical representation of a joint model of a time-to-event submodel and K-multivariate longitudinal outcomes submodel
-\includegraphics[scale=0.42]{figures/lights_graphical_representation.png}
+\includegraphics[scale=0.29]{figures/lights_graphical_representation.png}
 \end{itemize}
-
 \end{frame}
@@
 \begin{itemize}
-  \item \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$
-  \item $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
-  \item Survival part:
-  \begin{align*}f(t_i, \delta_i| b_i, G_i = k ; \theta) = &\big[\lambda(t_i|\cM_k(t_i), G_i = k)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s|\cM_k(s), G_i = k) \dd s \Big\}
+  \item<1-> \scriptsize $\theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big) \in \R^\vartheta$
+  \item<2-> $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$ with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$
+  \item<3-> Survival part:
+  \begin{align*}f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = &\big[\lambda(t_i| G_i = k, \Psi_i)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s| G_i = k, \Psi_i) \dd s \Big\}
   \end{align*}
-  \item Then, the likelihood writes
-  \begin{align*}\ell_n(\theta) = n^{-1} \sum_{i=1}^n \log \int_{\R^r} \sum_{k=0}^{K-1} &\pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k ; \theta) \\ & \times f(y_i | b_i, G_i = k ; \theta) f(b_i ; \theta) \dd b_i
+  \item<4-> Then, the likelihood writes
+  \begin{align*}
+  \ell_n(\theta) &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \int_{\R^r} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) \\ & \qquad\qquad\qquad\qquad \times f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
+  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i
   \end{align*}
@@
 \begin{itemize}
-  \item Penalized objective
+  \item<1-> Penalized objective
   \begin{equation*}
@@
   with the elasticnet penalty \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \] and the sparse group lasso penalty \[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z^l}_2 \]
-  \item Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
+  \item<2-> Resulting optimization problem \[\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta)\]
 \end{itemize}
@@
 
-\subsection{QNMCEM}
+\subsection{QNEM}
 
-\begin{frame}{QNMCEM algorithm (1/2)}
+\begin{frame}{QNEM algorithm (1/2)}
 
@@
 \begin{itemize}
-  \item $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
+  \item<1-> $\ell_n^\text{comp}(\theta) = \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}})$
 \end{itemize}
 
-\begin{block}{Monte Carlo E-step}
+\begin{block}{E-step}
 \begin{itemize}
-  \item $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
-  \item Requires to compute expectations of the form
+  \item<2-> $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$
+  \item<3-> Requires to compute expectations of the form
   \footnotesize
@@
 \[\pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] \]
-  \item Monte Carlo approximations used for untractable integrals
 \end{itemize}
@@
 
-\begin{frame}{QNMCEM algorithm (2/2)}
+\begin{frame}{QNEM algorithm (2/2)}
 
@@
 \begin{itemize}
-  \item \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
-  \item $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
-  \item \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
-  \item $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
-  \item $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
-  \item $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
-  \item L-BFGS-B to solve the problem
-  \item Proximal gradient method to estimate $\gamma_k^{(w+1)}$
-  \item Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | b_i, G_i=k ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
+  \item<1-> \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
+  \item<2-> $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
+  \item<3-> \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
+  \item<4-> $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
+  \item<5-> $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
+  \item<6-> $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
+  \item<7-> L-BFGS-B to solve the problem
+  \item<8-> Proximal gradient method to estimate $\gamma_k^{(w+1)}$
+  \item<9-> Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
 \end{itemize}

### 2022-02-02 slides/IASD/pres.tex
Update the figures of estimated params

@@
 \begin{frame}{Experiments}
-  \centering
-  \includegraphics[height= 3cm, width=5cm]{figures/likelihood.png} \hspace{1cm} \\
+  \includegraphics[height= 1.5cm, width=4cm]{figures/beta_estimation.png} \hspace{0.1cm}
+  \includegraphics[height= 1.5cm, width=2.2cm]{figures/xi_estimation.png} \hspace{0.1cm}
+  \includegraphics[height= 1.4cm, width=2.2cm]{figures/baseline_estimation.png} \hspace{0.1cm} \\
   \vspace{0.5cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/beta_learning.png} \hspace{0.7cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/beta_estimation.png} \\
+  \includegraphics[height= 2cm, width=4cm]{figures/beta_mlmm_learning.png} \hspace{0.7cm}
+  \includegraphics[height= 1.5cm, width=4cm]{figures/gamma_estimation.png} \\
   \vspace{0.5cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/xi_learning.png} \hspace{0.7cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/xi_estimation.png} \\
+  \hspace{0.2cm}
+  \includegraphics[height= 2cm, width=3cm]{figures/beta_learning.png}
+  \\
 \end{frame}

### 2022-02-28 slides/Staff Heka/pres.tex
staff heka slides

--- 2022-02-02 slides/IASD/pres.tex
@@
 \newcommand{\cS}{\mathcal S}
-\newcommand{\cM}{\mathcal M}
+\newcommand{\cY}{\mathcal Y}
 \newcommand{\cC}{\mathcal C}
@@
 
+\title[Lights \\ \insertframenumber/\inserttotalframenumber \vspace{-1cm}]{}
+
+\institute[shortinst]{
+  \vspace{-2.2cm} \\
+  \includegraphics[width=2cm]{figures/SU} \hspace{.2cm}
+  \includegraphics[width=1cm]{figures/cnrs} \hspace{.1cm}
+  \includegraphics[width=2.3cm]{figures/UniversiteParis.png}
+  \includegraphics[width=1.7cm]{figures/lpsm_logo.jpeg}
+  \includegraphics[width=1.4cm]{figures/Califrais_logo} \\
+  \vspace{.6cm}
+  \textcolor{blue_pres}{\Large HeKA Staff Meeting}\\
+  \vspace{.5cm}
+  \large
+  \textbf{Simon Bussy}\\
+  \vspace{.25cm}
+  \scriptsize
+  \href{mailto:simon.bussy@gmail.com}{\texttt{simon.bussy@gmail.com}}\\
+  \vspace{.4cm}
+  \small
+  \textbf{Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations} \\
+  \vspace{.3cm}
+  \tiny
+  S. Bussy $^{(1,2)}$, V.T. Nguyen$^{(1,2)}$, A. Barbieri $^{(3)}$, S. Zohar $^{(4)}$, A.S. Jannot $^{(4,5)}$\\
+
+  \vspace{.4cm}
+
+  \tiny
+
+  $^{(1)}$LOPF, Califrais’ Machine Learning Lab, Paris, France \\
+  $^{(2)}$LPSM, UMR 8001, CNRS, Sorbonne University, Paris, France \\
+  $^{(3)}$INSERM, UMR 1219, Bordeaux Population Health Research Center, Univ. Bordeaux, France \\
+  $^{(4)}$INSERM, UMRS 1138, Centre de Recherche des Cordeliers, Paris, France \\
+  $^{(5)}$Biomedical Informatics and Public Health Department, EGPH, APHP, Paris, France
+}
+
+\date{}
+
 \begin{document}
+
+\begin{frame}
+  \titlepage
+\end{frame}
 
@@
   \item<3-> Flexibility in modeling the dependency between the longitudinal features and the event time with appropriate penalties
-  \item<4-> Inference achieved using an efficient and novel Quasi-Newton Expectation Maximization algorithm
+  \item<4-> Inference achieved using an efficient and novel Proximal Quasi-Newton Expectation Maximization algorithm
 \end{itemize}
@@
 \small
-\begin{block}{Group-specific marker trajectories}
+\begin{block}<1->{Group-specific marker trajectories}
 \begin{itemize}
-  \item<1-> \footnotesize $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
+  \footnotesize
+  \item<1-> $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$\\
   with fixed effect parameters $\beta_k^l \in \R^{q_l}$ and subject-and-longitudinal outcome specific random effects $b^l \in \R^{r_l} \sim \cN(0, D_{ll})$
   \item<2-> $\text{Cov}[b^l,b^{l'}] = D_{ll'}$
-  \only{ and
+  \only<2>{ and
   \[ D = 
@@
 
-\end{frame}
-
-\begin{frame}{Submodels}
-
-\small
-\begin{block}{Group-specific risk of event}
+\begin{block}<3->{Group-specific risk of event}
 \begin{itemize}
-  \item<1-> \footnotesize $\lambda(t|\Psi, G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi \big\}$
-  \item<2-> Fetures $\Psi$ are extracted from $Y$ by the \texttt{Python} library \texttt{tsfresh}
-  \item<2-> \color{red}{TODO: Add descriptions}
+  \item<3-> $\lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\underbrace{\sum_{l=1}^L \sum_{a=1}^\cA \gamma_{k,a}^l \Psi_a^l(t)}_{\gamma_k^\top \Psi(t)} \Big\}$ \\
+    \vspace{.1cm}
+    with $\cY(t) = \{Y(u), 0 \leq u < t\}$ and for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping
+  \item<4-> $\Psi(t) \in \R^{L\cA}$ : high-dim representation features vector, highly flexible and independent of any modeling assumption !
 
@@
 
-\begin{frame}{Submodels}
+\begin{frame}{Generalization of SREMs and JLCMs}
+
 \begin{itemize}
-\item Graphical representation of a joint model of a time-to-event submodel and K-multivariate longitudinal outcomes submodel
-\includegraphics[scale=0.29]{figures/lights_graphical_representation.png}
+  \item<1-> JLCMs~\citep{vermunt2003latent} : homogeneous latent subgroups sharing the same marker trajectories and the same prognostic
+  \item<2-> SREMs~\citep{wulfsohn1997joint} : characteristics of the longitudinal processes (e.g. functions of the random effects) are included as features in the survival model 
+  \item<3-> Usual functionals would be
+    \begin{table}[htb]
+    \centering
+    \resizebox{.9\textwidth}{!}{
+    \begin{tabular}{ccc}
+    \toprule
+    Description & $\tilde \Psi_a^l(t, \beta_k^l, b^l)$ & Reference \\
+    \midrule
+    Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
+    Random effects & $b^l$ & \citet{hatfield2011joint} \\ 
+    Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
+    Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
+     \bottomrule
+    \end{tabular}}
+    \end{table}
+  \item<4-> Depend on $\beta_k$ (then on $k$) which leads to complicated updates, and on $b$ which leads to untracktable integrals that requires approximation methods : computationally intensive and not scalable
+
 \end{itemize}
+
 \end{frame}
+
+\begin{frame}{Graphical model}
+
+\centering
+\includegraphics[width=7cm]{figures/graphical_model}
+
+\end{frame}
+
 \subsection{Notations}
@@
 \begin{itemize}
-  \item<1-> $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1, \Psi_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n, \Psi_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
+  \item<1-> $\cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}$ with $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \text{ and } y_{ij}^l=Y_i^l(t_{ij}^l)$
   \item<2-> $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ with $n_i = \sum_{l=1}^L n_i^l$
@@
   \item<3-> Survival part: 
-  \begin{align*}f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = &\big[\lambda(t_i| G_i = k, \Psi_i)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s| G_i = k, \Psi_i) \dd s \Big\}
+  \begin{align*}f(t_i, \delta_i| G_i = k ; \theta) = &\big[\lambda(t_i|\cY(t_i), G_i = k)\big]^{\delta_i} \\ & \times \exp \Big\{-\int_0^{t_i} \lambda(s|\cY(s), G_i = k) \dd s \Big\}
   \end{align*}
-  \item<4-> Then, the likelihood writes
-  \begin{align*}
-  \ell_n(\theta) &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \int_{\R^r} \pi_{\xi_k}(x_i) f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) \\ & \qquad\qquad\qquad\qquad \times f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
-  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i
+  \item<4-> Then, the log-likelihood (JLCMs type) writes
+  \begin{align*}\ell_n(\theta) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k) \dd b_i
   \end{align*}
@@
 
-\subsection{QNEM}
+\subsection{prox-QNEM}
 
-\begin{frame}{QNEM algorithm (1/2)}
+\begin{frame}{prox-QNEM algorithm (1/2)}
 
@@
 
-\begin{block}{E-step}
+\begin{block}<2->{Monte Carlo E-step}
 \begin{itemize}
@@
   \footnotesize
-\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \] 
+\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i \] 
 \small for different functions $g$, where we denote 
@@
 
-\begin{frame}{QNEM algorithm (2/2)}
+\begin{frame}{prox-QNEM algorithm (2/2)}
 
@@
 
-\begin{block}<1->{Quasi-Newton M-step}
+\begin{block}<1->{Proximal Quasi-Newton M-step}
 \begin{itemize}
-  \item<1-> \tiny $\theta^{(w+1)} \in \argmin_{\theta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
+  \item<1-> \tiny $\theta^{(w+1)} \in \argmin_{\theta\in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}$
   \item<2-> $D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$
-  \item<3-> \tiny $R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]$
-  \item<4-> $\beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k)$
-  \item<5-> $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
-  \item<6-> $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$
-  \item<7-> L-BFGS-B to solve the problem
-  \item<8-> Proximal gradient method to estimate $\gamma_k^{(w+1)}$
-  \item<9-> Predictive marker \scriptsize $\hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k, \Psi_i ; \hat \theta)}$, which is an estimate of $\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i]$
+  \item<3-> $P^{(w)}_{n,k}(\xi_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$
+  \item<4-> $\xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}$ 
+  \item<5-> L-BFGS-B to solve the problem
+  \tiny
+  \begin{equation*}
+  \begin{split}
+    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
+    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
+  \end{split} 
+  \end{equation*}
+  \item<6-> Closed-form update for $\beta_k^{(w+1)}$
+  \item<7-> Proximal gradient (ISTA) for the $\gamma_k^{(w+1)}$ update, based on Lemma 1 that states $\text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}$
+  \item<8-> $\lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t_i)\big\} \ind{\{t_i \geq t\}}}$
 \end{itemize}
@@
 
-\section{Evaluation}
+\subsection{Evaluation}
 
-\begin{frame}[noframenumbering]
-\Large \centering
-\textcolor{blue_pres}{V.} Evaluation
+\begin{frame}{Performance evaluation}
+
+\scriptsize
+\begin{block}<1->{Real-time prediction paradigm}
+
+\begin{itemize}
+ \item<1-> Practitioners need predictive prognostic tools to be used in real-time
+ \item<2-> Obtain $\hat\theta$ on training data
+ \item<3-> On test data : sample $t^{max}_i \sim T_i \big(1 - \text{Beta}(2, 5)\big)$ the time for subject $i$ when one wants to perform the risk prediction -- the ``present'' time
+ \only<3>{
+    \centering
+    \\
+    \vspace{.2cm}
+    \includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
+  }
+  \item<4-> Truncate $y_i$ up to $t^{max}_i$ to get $\tilde y_i$
+\end{itemize}
+
+\end{block}
+
+\begin{block}<5->{Predictive marker}
+
+\begin{itemize}
+ \item<5-> $\P_{\hat \theta}[G_i=k | T_i > t^{max}_i, \tilde y_i] = \dfrac{\pi_{\hat \xi_k}(x_i) f_{\hat \theta}(t^{max}_i, \delta_i=0, \tilde y_i | G_i=k)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) f_{\hat \theta}(t^{max}_i, \delta_i=0, \tilde y_i | G_i=k)}$
+ \item<6-> In practice $K=2$
+ \item<7-> C-index metric $\cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau]$
+\end{itemize}
+
+\end{block}
+
 \end{frame}
 
-\begin{frame}{Experiments}
+\section{Application}
+
+\subsection{Competitors}
+
+\begin{frame}{Competing models}
+
+\begin{itemize}
+ \item<1-> Landmark Cox model : \texttt{survival} \texttt{R} package~\citep{zhang2018time}
+ \item<2-> The time-dependent Cox model~\citep{sueyoshi1992semiparametric} : \texttt{survival} \texttt{R} package
+ \item<3-> Multivariate joint latent class model : \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}
+ \item<4-> Multivariate shared random effect model : \texttt{JMbayes} package~\citep{2017_JMbayes}
+\end{itemize}
+
+\end{frame}
+
+\subsection{Simulations}
+
+\begin{frame}{High-dimensional simulation study (1/2)}
+
+\begin{itemize}
+ \small
+ \item<1-> High-risk subjects proportion $\pi_1 \in [0,1]$
+ \item<2-> $\cH = \big\{\lfloor \pi_1 n \rfloor \text{ random samples without replacement} \big\}$
+ \item<3-> $[x_{ij}] \in \R^{n \times p} \sim \cN \big(0, \bSigma_1(\rho_1)\big)$
+ \item<4-> $x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{i \notin \cH\}}} gap\ \text{for } j = 1, \dots, s$
+ \item<5-> $\xi = (\underbrace{\varsigma_1,\ldots,\varsigma_1}_s,0,\ldots,0) \in \R^p$
+ \item<6-> $G_i \sim \cB\big(\pi_\xi(x_i)\big)$
+ \item<7-> $\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}$
+ \item<8-> $\beta_k^l \sim \ind{\{l \in \cS_k \}} \cN\Big( 
+\mu_k, 
+\begin{bmatrix}
+  \rho_3 & 0\\
+  0 & \rho_3
+\end{bmatrix} 
+\Big)$
+ \item<9-> $Y_i^l(t) = \epsilon_i^l(t) + \ind{\{l \in \cS_k \}}\sum_{k=0}^{K-1} \ind{\{G_i=k\}} \big( (1, t)^\top \beta_k^l + (1, t)^\top b_i^l \big)$
+ \item<10-> $\big(\tilde \Psi_{k,a}(t, \beta_k, b_i^l)\big)_{a \in \{1,2\}} = \big(\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t\ , {b_i^l}^\top \big)$
+ \item<11-> $\tilde \gamma_{k,a}^l = \varsigma_2 \ind{\{l \in \cS_k,\ a\in \{1, 2\} \}}$
+ \item<12-> $\lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\}$
+ \item<13-> Gompertz baseline $\lambda_0(t) = \kappa_1 \kappa_2 \exp(\kappa_2t)$
+ \item<14-> $T_i^\star | G_i=k \sim \dfrac{1}{\iota_{i,k,2} + \kappa_2} \log \Big(1 - \dfrac{(\iota_{i,k,2} + \kappa_2) \log U_i}{\kappa_1 \kappa_2 \exp\iota_{i,k,1}} \Big)$
+\end{itemize}
+
+\end{frame}
+
+\begin{frame}{High-dimensional simulation study (2/2)}
+
 \centering
-  \includegraphics[height= 3cm, width=5cm]{figures/likelihood.png} \hspace{1cm} \\
-  \vspace{0.5cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/beta_learning.png} \hspace{0.7cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/beta_estimation.png} \\
-  \vspace{0.5cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/xi_learning.png} \hspace{0.7cm}
-  \includegraphics[height= 2cm, width=4cm]{figures/xi_estimation.png} \\
+\only<1>{\includegraphics[width=1\textwidth]{./figures/simu_param}}
+\only<2>{
+  \includegraphics[width=.9\textwidth]{./figures/simu_time}
+  \includegraphics[width=.7\textwidth]{./figures/simu_surv}
+  }
+\only<3>{\includegraphics[width=1\textwidth]{./figures/simu_long}}
+
+\end{frame}
+
+\subsection{Results}
+
+\begin{frame}{First results (1/2)}
+
+\centering
+\includegraphics[width=1\textwidth]{./figures/obj_vcg}
+\uncover<2->{\includegraphics[width=.5\textwidth]{./figures/xi_estimation}}
+\uncover<3->{\includegraphics[width=.45\textwidth]{./figures/baseline_estimation}}
+\uncover<4->{\includegraphics[width=.8\textwidth]{./figures/beta_estimation}}
+
+\end{frame}
+
+\begin{frame}{First results (2/2)}
+
+\centering
+\includegraphics[width=1\textwidth]{./figures/gamma_estimation}
+\uncover<2->{\includegraphics[width=1\textwidth]{./figures/predictions}}
+
 \end{frame}
@@
 \Large \centering
-\textcolor{blue_pres}{VI.} Conclusion
+\textcolor{blue_pres}{V.} Conclusion
 \end{frame}
@@
   \item<2-> Penalization of the likelihood in order to perform feature selection and to prevent overfitting
-  \item<3-> New efficient estimation algorithm (QNEM) has been derived
+  \item<3-> New efficient estimation algorithm (prox-QNEM) has been derived
   \item<4-> Automatically determines significant prognostic longitudinal features
@@
   \item<5-> Available at \small{\url{https://github.com/Califrais/lights}}
-  \item<6-> Applications of the model available soon on an \texttt{arXiv} paper.
+  \item<6-> Applications of the model available soon
 \end{itemize}
@@
 \frame{
-  \small
+  \tiny
   \frametitle{References}

### 2022-02-28 manuscript/paper.tex
update 1

@@
 \newcommand{\norm}[1]{\|#1\|}
-\newcommand{\cM}{\mathcal M}
+\newcommand{\cY}{\mathcal Y}
 \newcommand{\cD}{\mathcal D}
@@
 \newcommand{\cA}{\mathcal A}
-\newcommand{\sA}{\mathscr A}
 \newcommand{\cQ}{\mathcal Q}
@@
 This paper introduces a prognostic method called \textit{lights} to deal with the problem of joint modeling of longitudinal data and censored durations, where a large number of both longitudinal and time-independent features are available.
-Yet there is no standard model so far to learn from such high-dimensional multivariate longitudinal data in a survival analysis setting.
-Shared association features are learned from the longitudinal processes and included as potential risk factor in a group-specific Cox model.
-Appropriate penalties are then used during inference to allow flexibility in modeling the dependency between the longitudinal features and the event time.
+In the literature, standard joint models are either of type shared random-effect or joint latent class ones ; where the association structure between the longitudinal and the time-to-event submodels takes respectively the form of either shared association features learned from the longitudinal processes and included as potential risk factor in the survival model, or latent classes modeling population heterogeneity.
+We pick modeling ideas from both worlds and use appropriate penalties during inference for being able to learn from a high-dimensional context.
 The statistical performance of the method is examined on an extensive Monte Carlo simulation study, and finally illustrated on a publicly available dataset.
-Our proposed method significantly outperforms the state-of-the-art survival models regarding risk prediction in terms of C-index, with a computing time orders of magnitude faster.
+Our proposed method significantly outperforms the state-of-the-art joint models regarding risk prediction in terms of C-index in a so-called real-time prediction paradigm, with a computing time orders of magnitude faster.
 In addition, it provides powerful interpretability by automatically pinpointing significant features being relevant from a practical perspective.
@@
 
-In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
+In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
 To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
@@
 
-\subsection{Group-specific risk of event}
-
-To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
-\begin{equation}
-  \label{eq:intensity-model}
-  \lambda(t|\cM_k(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l}^\top \Psi_a^l \Big\},
-\end{equation}
-where $\lambda_0(\cdot)$ is an unspecified baseline hazard function.
-
-For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known features $\Psi_a$ defining a shared association with $\gamma_{k,a}^l \in \R^{\cA}$ the corresponding joint association parameters.
-These features are extracted from $Y$ by the \texttt{Python} library \texttt{tsfresh}.
-This can be viewed as a generalization of SREMs~\citep{rizopoulos2010jm}.
-Let us finally denote
-\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
-\begin{equation}
-  \label{eq:def-asso-features}
-  \lambda(t|\Psi, G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi \big\}.
-\end{equation}
-
 \subsection{Group-specific marker trajectories}
@@
 \[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
-where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$ and where the corresponding subject-and-longitudinal outcome specific random effects $b^l$ that does not depend on the group membership, which is not a strong modeling assumption.
-\begin{assumption}
-\label{assumption1}
-We suppose that the random effects are conditional independent of the group membership on the association features, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
-\end{assumption}
+where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
 A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
@@
 
-\subsection{Graphical model for the lights}
+\subsection{Group-specific risk of event}
+
+To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
+\begin{equation}
+  \label{eq:intensity-model}
+  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
+\end{equation}
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
+\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
+the history of the true longitudinal process up to time $t$.
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.
+
+Let us finally denote
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
+\begin{equation}
+  \label{eq:def-asso-features}
+  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
+\end{equation}
+where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.
+
+\subsection{Generalization of SREMs and JLCMs}
+
+Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way.
+Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
+\begin{table}[htb]
+\centering
+\begin{tabular}{cccc}
+\toprule
+Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
+\midrule
+Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
+Random effects & $b^l$  & \citet{hatfield2011joint} \\
+Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
+Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
+ \bottomrule
+\end{tabular}
+\caption[]{Examples of classical shared associations in SREMs.}
+\label{table:shared_associations}
+\end{table}
+
+Our proposition differs from standard SREMs in two respects: $(i)$ representation vector $\Psi(t)$ does not depend on the modeling assumptions in the longitudinal submodel and $(ii)$ it allows to choose multiple high-dimensional representation mappings that characterize time series and concatenate them, since we perform features selection through the regularization strategy described in Section~\ref{sec:penalized-obj}.
+Point $(i)$ is key since for instance the $\beta_k^l$ dependence in $\Psi_a^l$ would lead to complicated updates, while the $b^l$ dependence would lead to untracktable integrals that are common in SREMs and require approximation methods such as Monte Carlo~\citep{hickey2018joinerml}, being computationally intensive and not scalable in a high-dimensional context.
+For point $(ii)$ in practice, we use the \texttt{Python} library \texttt{tsfresh}~\citep{christ2018time} and include many extracted features, such as absolute energy of the time series, statistics on autocorrelation, or Fourier and wavelet basis projections, to name but a few.
+
+\subsection{Graphical model}
 Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
@@
   \node[var, fill = white!100] (Sigma) [left=of b] {\Large$\bm{D}$};
-  \node[main, fill = black!30] (W) [below=3cm of Y] {\Large$\bm{\Psi_i}$};
+  \node[main, fill = white!100] (W) [below=3cm of Y] {\Large$\bm{\Psi_i(t)}$};
   \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
@@
 Square boxes denote observed data while circles denote unobserved (including random) terms.
-The red-dashed box indicates that the association features and the random effects are correlated, which, in addition to the link with the latent group membership, gives rise to the joint model.}
+One can then observe the two links in blue between the submodels that gives rise to the joint model.}
 \label{fig:graphical-model}
@@
 Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
-\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1, \Psi_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n, \Psi_n) \big\} \]
+\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
 where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
@@
 From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
+Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
+Now, let us make two first hypothesis.
+\begin{assumption}
+\label{assumption1}
+Suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
+\end{assumption}
+Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the group membership, which is not a strong modeling assumption.
 \begin{assumption}
@@
 Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time.
-So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to the number of unique failure times (denoted $J$ thereafter).
+So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
 To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
@@
 \label{indep-hyp-3}
-Assume that the random effects vector $b_i$, the group membership account for the association between the longitudinal and event outcomes and the association features $\Psi_i$, that is
+Assume that
 \begin{equation}
   \label{eq:ind-hyp}
-  f(t_i, \delta_i, y_i | b_i, G_i = k, \Psi_i ; \theta) = f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k ; \theta)
+  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
 \end{equation}
@@
 Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
+Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$.
+Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the group membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
 Then, one has
-\begin{align}
+\begin{equation}
 	\label{eq:log-lik}
-	\ell_n(\theta) &= \ell_n(\theta ; \cD_n) \nonumber \\
-	&= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) \int_{\R^r} f(t_i, \delta_i| b_i, G_i = k, \Psi_i ; \theta) f(y_i | b_i, G_i = k, \Psi_i; \theta) f(b_i ; \theta) \dd b_i \\
-  &= n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) f(y_i | G_i = k; \theta) \dd b_i,
-\end{align}
+	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
+\end{equation}
 where 
-\[f(t_i, \delta_i| G_i = k, \Psi_i ; \theta) = \big[\lambda(t_i|G_i = k, \Psi_i)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|G_i = k, \Psi_i) \dd s \Big\}. \]
+\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
+As expected, one has a JLCMs like log-likelihood.
 
@@
 
-In many practical applications, subjects show highly nonlinear longitudinal trajectories.
-In~\eqref{eq:log-lik}, the complete longitudinal history is required for the computation of the survival function.
-Hence, in order to produce a good estimate of $\cM_k(t)$, we consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely
+In many practical applications, subjects show highly nonlinear longitudinal trajectories. We consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
 \begin{equation}
@@
 
-\subsection{A Quasi-Newton EM}
+\subsection{A Proximal Quasi-Newton EM}
 \label{sec:prox-QNEM}
@@
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
+The integration over the survival process has been replaced with a finite sum over the process evaluated at the $J$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
 The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
@@
   \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\cA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
 \end{equation}
@@
 We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNEM}.
-For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\sA}$.
+For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\cA}$.
 Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
@@
 \[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
+
+dire ici qu'on tronque $y$ à $t^{max}$ et le noter $|t^{max}$
+
 where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
@@
 We then add a $gap \in \R^+$ value for subjects $i \in \cH$ and subtract it for subjects $i \notin \cH$, only on active features, that is
-\begin{center}
-$x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{i \notin \cH\}}} gap\ \text{for } j = 1, \dots, s$.
-\end{center}
+\[x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{i \notin \cH\}}} gap\ \text{for } j = 1, \dots, s.\]
 Note that this is equivalent to generate the time-independent features according to a gaussian mixture. We finally normalize each time-independent features for practical interpretations.
@@
 \end{equation*}
+
+rmq : plus besoin du sk dans phi vu qu'on l'ajoute directement ds le Y non ?
+
 and 

### 2022-03-02 manuscript/paper.tex
update 2

@@
 In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories.
-Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent subgroups sharing the same marker trajectories and the same prognostic.
+Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent classes sharing the same marker trajectories and the same prognostic.
 
@@
 An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}.
-To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ subgroups of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}.
+To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ classes of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}.
 Let us denote
@@
 \end{equation}
-where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th group, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero.
+where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th latent class, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero.
 The intercept term is here omitted without loss of generality.
@@
 
-In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
+In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a class-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
 To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
@@
 
-\subsection{Group-specific marker trajectories}
-
-We suppose a group-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given subgroup $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
+\subsection{Class-specific marker trajectories}
+
+We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
 \begin{equation}
@@
 
-\subsection{Group-specific risk of event}
+\subsection{Class-specific risk of event}
 
@@
 \label{assumption1}
-Suppose that the random effects are independent of the group membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
+Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
 \end{assumption}
-Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the group membership, which is not a strong modeling assumption.
+Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
 \begin{assumption}
 \label{indep-hyp-2}
-Suppose that for a given subject $i$ and conditional on the random effects and the group membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
+Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
 \end{assumption}
@@
 Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$.
-Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the group membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
+Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
 Then, one has
@@
 where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$.
-Here, $(\eta, \tilde{\eta}) \in [0, 1]^2$ are fixed and we denote
+Here, for fixed $(\eta, \tilde{\eta}) \in [0, 1]^2$, one has
 \[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
@@
 
-One advantage of the considered regularization method is its ability to perform feature selection (the lasso part) and pinpoint the most important features relatively to the prediction objective : the support of $\xi_k$ thus informs on the time-independent features involved in the $k$-th group membership.
+One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important (time-dependent and time-independent) features relatively to the prediction objective.
+The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership.
 The ridge part allows to handle potential correlation between time-independent features.
-On the other hand, the sparse group lasso allows to perform feature selection for each trajectory : the support of $\gamma_k^l$ informs on the features involved in the $k$-th group risk of event for the $l$-th longitudinal outcome.
-Note that in practice, the intercept is not regularized.
+On the other hand, for the sparse group lasso penalty, a group $l$ corresponds to a trajectory, that is one longitudinal outcome.
+So if $\hat \gamma_k^l$ is entirely zeroed (thanks to the group lasso part), it means that the $l$-th longitudinal process is discarded by the model in terms of risk effect for the $k$-th latent class.
+Then, the sparse part of the penalty allows to perform representation features selection for each trajectory : for $\hat \gamma_k^l$ that are not entirely zeroed, its support informs on the representation features involved in the $k$-th latent class risk of event for the $l$-th longitudinal outcome.
+Note that in practice, the intercept terms are not regularized.
 
@@
 In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}.
-EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
+EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different and original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
 
@@
   &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i \Big) - \exp \big\{ \gamma_k^\top \Psi_i \big\} \int_0^{t_i} \lambda_0(s) \dd s \\
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
   & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
@@
 the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g(b_i, G_i) = \tilde g(b_i)$ (for instance $\tilde g: b_i \mapsto b_i^\top D^{-1}b_i$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
+Then, either $g(b_i, G_i) = \tilde g(b_i)$ (either $\tilde g_1: b_i \mapsto b_i$ or $\tilde g_2: b_i \mapsto b_i b_i^\top$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
 
 In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
-\begin{equation}
-  \label{eq:def-E-g-tilde}
-  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
-\end{equation}
-with
-\begin{align}
-  \Lambda_{ik,\tilde g}^{\theta^{(w)}} &= \int_{\R^r} \tilde g(b_i) f(t_i, \delta_i | b_i, G_i = k, \Psi_i ; \theta^{(w)}) f(y_i | b_i, G_i = k, \Psi_i ; \theta^{(w)}) f(b_i | G_i = k, \Psi_i; \theta^{(w)}) \dd b_i \\
+\begin{equation*}
+  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
+\end{equation*}
+where
+\begin{equation}
   \label{eq:Lambda_ik-b_i}
-  &=  f(t_i, \delta_i | G_i = k, \Psi_i ; \theta^{(w)}) f(y_i | G_i = k ; \theta^{(w)}) \int_{\R^r} \tilde g(b_i) f(b_i | y_i, G_i = k ; \theta^{(w)}) \dd b_i,
-\end{align}
-where
+  \Lambda_{ik,\tilde g}^{\theta^{(w)}} =  f_{\theta^{(w)}}(t_i, \delta_i | G_i = k) f_{\theta^{(w)}}(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | y_i, G_i = k) \dd b_i,
+\end{equation}
+and in the Gaussian case used in practice one has
 \begin{align*}
-  y_i | G_i = k &\sim \cN\big(U_i\beta_k, V_iDV_i^\top + \Sigma_i\big) \\
-  b_i | y_i, G_i=k &\sim \cN \big(W_iV_i^T\Sigma_i^{-1}(y_i - U_i\beta_k), W_i\big)  
+  y_i | G_i = k ; \theta^{(w)} &\sim \cN\big(U_i\beta_k^{(w)}, V_iD^{(w)}V_i^\top + \Sigma_i^{(w)}\big), \\
+  b_i | y_i, G_i=k ; \theta^{(w)} &\sim \cN \big(O_{i,k}^{(w)}, W_i^{(w)}\big),  
 \end{align*}
-with $W_i = (V_i^\top\Sigma_i^{-1}V_i + D^{-1})^{-1}$.
-
-Now in the case $g = g_k$, one has $\E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}}$ and we compute its approximation given
+with $O_{i,k}^{(w)}=W_i^{(w)}V_i^T{\Sigma_i^{(w)}}^{-1}(y_i - U_i\beta_k^{(w)})$ and $W_i^{(w)} = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$.
+Hence, what is interesting here is that~\eqref{eq:Lambda_ik-b_i} is actually closed-formed since
+\[ \E_{\theta^{(w)}}[ \tilde g_1(b_i) | y_i, G_i = k] = O_{i,k}^{(w)} \quad \text{and} \quad \E_{\theta^{(w)}}[ \tilde g_2(b_i) | y_i, G_i = k] = W_i^{(w)} + O_{i,k}^{(w)}{O_{i,k}^{(w)}}^\top\]
+and then tractable analytically.
+Whereas in classical SREMs, some form of approximation must be used in practice to compute this quantity, like numerical integration techniques (for instance Gaussian quadrature) or Monte Carlo approximation.
+
+Now in the case $g = g_k$, one has
 \begin{equation}
 \label{eq:hat_pi_ik-def}
-  \hat \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \hat \Lambda_{ik,1}^{\theta^{(w)}}}.
+  \E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}.
 \end{equation}
@@
   \label{eq:update_D}
-  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
+  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
 \end{equation}
@@
 For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\cA}$.
-Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
-Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
+Then, we initialize $\lambda_0^{(0)}$ like if there is no latent classes ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
+Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no latent classes ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
 
@@
   \tcc{E-step}
+  A RE-ECRIRE
   Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for the required functions $\tilde g$\\
@@
 
-In many practical applications -- including those addressed in Section~\ref{sec:application} -- we are interested in identifying one subgroup of the population with a high risk of adverse event compared to the others.
+In many practical applications -- including those addressed in Section~\ref{sec:application} -- we are interested in identifying one latent class of the population with a high risk of adverse event compared to the others.
 Then, in the following, we consider $G \in \{ 0, 1\}$ where $G=1$ means high-risk of early adverse event and $G=0$ means low risk.
-To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the group with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
+To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the latent class with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
 In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$.
@@
 The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
-Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), posterior risk classification for subject $i$ and latent class $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
 Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
@@
 But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction.
-Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in subgroup $k$ would rather be
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ would rather be
 \begin{equation}
@@
 Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model.
-Given the group membership, each submodel is assumed to be independent.
+Given the latent class membership, each submodel is assumed to be independent.
 In this context, the predictive marker for subject $i$ is
@@
 \end{equation}
-for the choice of the association parameters, with $\cA = 3 + 3(1-r_s)/r_s$ so that $(i)$ there is a proportion $r_s$ of active longitudinal features among the $L$ (with the $\cS_k$), $(ii)$ we induce sparsity within each group adding a proportion of $(1-r_s)$ non-active association features ($a \geq 3$) in each group (similarly to~\eqref{eq:xi-sparse} for the time-independent features), and $(iii)$ the longitudinal effects are distinct for each subgroup $k=0, \ldots, K-1$ (thanks to the $\cS_k$ definition).
+for the choice of the association parameters, with $\cA = 3 + 3(1-r_s)/r_s$ so that $(i)$ there is a proportion $r_s$ of active longitudinal features among the $L$ (with the $\cS_k$), $(ii)$ we induce sparsity within each latent class adding a proportion of $(1-r_s)$ non-active association features ($a \geq 3$) in each latent class (similarly to~\eqref{eq:xi-sparse} for the time-independent features), and $(iii)$ the longitudinal effects are distinct for each latent class $k=0, \ldots, K-1$ (thanks to the $\cS_k$ definition).
 
@@
 
-Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNEM algorithm in Section~\ref{sec:prox-QNEM}, acting as if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
+Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNEM algorithm in Section~\ref{sec:prox-QNEM}, acting as if there is no latent classes ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
 For the sake of simplicity, let us denote here 

### 2022-03-02 manuscript/paper.tex
Update the graph representation of joint model

@@
 \begin{tikzpicture}
-\hspace{3cm}
 \tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
@@
   \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
-  \node[var, fill = white!100] (xi) [below=4cm of beta] {\Large$\bm{\xi_k}$};
-  \node[var, fill = white!100] (gamma) [below=3cm of xi] {\Large$\bm{\gamma_k}$};
-  \node[main, fill = white!100] (G) [right=4cm of xi] {\Large$\bm{G_i}$};
-  \node[main, fill = white!100] (mu) [right=4cm of beta] {\Large$\bm{M_{ik}^l(t)}$};
-  \node[main, fill = black!30] (Y) [right=of mu] {\Large$\bm{Y_i^l(t)}$};
-  \node[main, fill = white!100] (b) [above=of mu] {\Large$\bm{b_i^l}$};
-  \node[var, fill = white!100] (phi) [above=of Y] {\Large$\bm{\phi_l}$};
-  \node[var, fill = white!100] (Sigma) [left=of b] {\Large$\bm{D}$};
-  \node[main, fill = white!100] (W) [below=3cm of Y] {\Large$\bm{\Psi_i(t)}$};
-  \node[main, fill = white!100] (lambda) [right=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
-  \node[main, fill = black!30] (T) [right=of lambda] {\Large$\bm{T_i}$};
-  \node[main, fill = black!30] (X) at (3, -8.3) {\Large$\bm{X_i}$};
-
+  \node[main, fill = white!100] (mu) [below=4cm of beta] {$\bm{M_{ik}^l(t)}$};
+  \node[main, fill = black!30] (Y) [below=of mu] {\Large$\bm{Y_i^l(t)}$};
+  \node[main, fill = white!100] (b) [left=of mu] {\Large$\bm{b_i^l}$};
+  \node[var, fill = white!100] (phi) [left=of Y] {\Large$\bm{\phi_l}$};
+  \node[var, fill = white!100] (Sigma) [above=of b] {\Large$\bm{D}$};
+  \node[var, fill = white!100] (xi) [right=4.5cm of beta] {\Large$\bm{\xi_k}$};
+  \node[var, fill = white!100] (gamma) [right=5cm of xi] {\Large$\bm{\gamma_k}$};
+  \node[main, fill = white!100] (G) [below=4cm of xi] {\Large$\bm{G_i}$};
+  \node[main, fill = white!100] (W) [right=8.5cm of Y] {\Large$\bm{\Psi_i(t)}$};
+  \node[main, fill = white!100] (lambda) [below=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
+  \node[main, fill = black!30] (T) [below=1.3cm of lambda] {\Large$\bm{T_i}$};
+  \node[main, fill = black!30] (X) at (4, -3.3) {\Large$\bm{X_i}$};
+
+  \path [->,draw,ultra thick, blue]
+        (G) edge (mu)
+        (Y) edge (W);
   \path [->,draw,ultra thick]
         (beta) edge (mu)
-        (G) edge (mu)
         (b) edge (mu)
@@
         (lambda) edge (T)
-        (Y) edge (W)
         ;
 
-  \node[rectangle, inner sep=8mm,draw=black!100, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.5cm, xshift=-10mm]{\Large$\bm{k = 1...K}$}}] at (0, -4.6) {};
+  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1...K}$}}] at (6.5,0.4) {};
   \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick] at (8, 4.1) {\Large$\bm{l = 1...L}$};
-
-  \node[rectangle,inner sep=6mm, draw=black!100, label=center:\rotatebox{90}{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,0.5) {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:\rotatebox{90}{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=1cm, minimum height = 6cm, fill = black!50] at (-2.7,-8.2) {};
+  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.7cm]{\Large$\bm{l = 1...L}$}}] {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100, label=center:{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (-1.5,2.8) {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5,2.8) {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (11.5,2.8) {};
 

### 2022-03-02 manuscript/paper.tex
update 3

@@
 
-  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1...K}$}}] at (6.5,0.4) {};
+  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (6.5,0.4) {};
   \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.7cm]{\Large$\bm{l = 1...L}$}}] {};
+  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.7cm]{\Large$\bm{l = 1, \ldots,L}$}}] {};
 

### 2022-03-02 manuscript/paper.tex
Minor update

@@
 
-  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (6.5,0.4) {};
+  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (7,0.4) {};
   \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
@@
 
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5,2.8) {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (11.5,2.8) {};
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5.5,2.8) {};
+
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (12,2.8) {};
 

### 2022-03-09 manuscript/paper.tex
Minor update the graph

@@
 
-  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (7,0.4) {};
-  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi)] {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (b) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.7cm]{\Large$\bm{l = 1, \ldots,L}$}}] {};
+  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (6.8,0.4) {};
+  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (Sigma) (phi)] at (-1.6,-4.5) {};
+  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.8cm]{\Large$\bm{l = 1, \ldots,L}$}}] {};
 
@@
 
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5.5,2.8) {};
+  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5.2,2.8) {};
 

### 2022-03-15 manuscript/paper.tex
review #3

--- 2022-03-02 manuscript/paper.tex
@@
 Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
-\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
+\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
@@
 \end{equation}
-It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}).
+It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}).
 We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
@@
   \label{eq:grad-xi_k}
-  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big( \hat \pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
+  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
 \end{equation}
 In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}.
-Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNEM}) with every increase of $N$.
+Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
+To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.
 
@@
 \begin{equation*}
-  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
+  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
 \end{equation*}
@@
   \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ U_{i}^\top \Sigma_i^{(w)} I_{n_{i}} (y_i - U_{i}\beta_k - V_{i}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big].
+  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k} = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[ U_{i}^\top \Sigma_i^{(w)} I_{n_{i}} \big(y_i - U_{i}\beta_k - V_{i} \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i]\big)\Big].
 \end{equation}
@@
   \label{eq:update_beta}
-  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
+  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
 \end{equation}
 Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
-\begin{align*}
-  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \gamma_k^\top \Psi_i -  \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
-\end{align*}
+\[Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[\delta_i \gamma_k^\top \Psi_i(t_i) - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u \leq t_i\}} \Big]\]
 based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
 The integration over the survival process has been replaced with a finite sum over the process evaluated at the $J$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
-The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
+The update for $\gamma_k^{(w)}$ requires to solve the following convex minimization problem
 \begin{equation}
@@
 The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable.
-We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
+We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our \texttt{lights} package a first \texttt{Python} implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
 A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
@@
   \label{eq:grad-Q}
-  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Psi_{i, a}^l  \Big[ \delta_i - \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
+  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Psi_{i, a}^l(t_i) \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \exp \big\{\gamma_k^\top \Psi_i(t_j^u) \big\} \ind{\{t_j^u \leq t_i\}} \Big]
 \end{align}
@@
   \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i\big\} \ind{\{t_i \geq t\}} },
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} \ind{\{t_i \geq t\}} },
 \end{equation}
@@
   \label{eq:update_phi}
-  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
-  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
+  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
+  & - 2V_{il} \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
 \end{align}
@@
 \paragraph*{The prox-QNEM algorithm.}
-Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
+Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting prox-QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
 
@@
 \label{alg:prox-QNEM}
-\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
+\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
 \KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
@@
   \tcc{E-step}
-  A RE-ECRIRE
-  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for the required functions $\tilde g$\\
-  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
-  \If{Condition}
-  {
-      $N \leftarrow N + $\\
-      $tol \leftarrow tol - $
-  }
-  \tcc{Quasi-Newton M-step}
+  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
+  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
+  \tcc{Proximal Quasi-Newton M-step}
   Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
@@
 \end{algorithm}
-Let us define the relative difference at step $(w+1)$ as
-\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
-Then in practice, the convergence for Algorithm~\ref{alg:prox-QNEM} is declared when the condition
-\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
-is satisfied for 3 consecutive iterations.
+
+Then in practice, the convergence for Algorithm~\ref{alg:prox-QNEM} is declared when the relative objective $\big(\ell_n^\text{pen}(\theta^{(w+1)}) - \ell_n^\text{pen}(\hat\theta^{(w)}) \big) / \ell_n^\text{pen}(\hat\theta^{(w)})$ goes below a chosen threshold.
 
@@
 
-The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned during a 10-fold randomized search cross-validation procedure~\citep{bergstra2012random} using predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
-Random search is indeed more efficient in terms of computing times than classical grid search for hyper-parameters optimization and finds better models by effectively searching in a larger configuration space, which is appropriate to our high-dimensional problem.
+The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
+In practice, we use the snoek2012practical\texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
+SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
 Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
@@
 However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power.
-So we choose to put ourselves in this paradigm.
+So we choose to put ourselves in what we call a ``real-time prediction paradigm''.
+
+\paragraph{Real-time prediction paradigm.}
+
+Once the learning phase is achieved for the model on a training set (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), we want to assess real-time risk prediction performances on a test set.
+
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction --
 
@@
 The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
-The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
-Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), posterior risk classification for subject $i$ and latent class $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+
+posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
 Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
 
-But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in a context where we want to perform real-time risk prediction.
-Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- a natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ would rather be
+But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow
+
+ a natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ would rather be 
 \begin{equation}

### 2022-03-16 manuscript/paper.tex
minor update

@@
 \begin{equation}
-\label{eq:hat_pi_ik-def}
+\label{eq:pi_ik-def}
   \E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}.
@@
 \end{equation}
-It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}).
+It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:pi_ik-def}).
 We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
@@
   Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
-  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
+  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:pi_ik-def}\\
   \tcc{Proximal Quasi-Newton M-step}
@@
 The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
-In practice, we use the snoek2012practical\texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
+In practice, we use the \texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
 SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
@@
 
-posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}).
+posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:pi_ik-def}).
 Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
@@
 where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
-Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:hat_pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
+Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
 \[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
@@
 In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$.
-But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~\citep{bussy2019comparison}), which is not the case of the C-index metric.
+But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~-v\citep{bussy2019comparison}), which is not the case of the C-index metric.
 Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.

### 2022-04-12 slides/JDS2022/short paper/paper.tex
add JDS pres folder

--- 2022-03-16 manuscript/paper.tex
@@
 
-\section{Introduction}
-
-With the increasing expectations to know their customers from account opening throughout the duration of the business relationship, web companies have the luxury of building elaborate systems to help them keep everything on track.
-The amount of recorded data per client is often tremendous and growing through time.
-There is no tool today to take into account simultaneously a huge number of longitudinal signals in a high-dimensional context to perform real-time churn (or satisfaction) risk prediction.
-Similarly, in many clinical studies, it has become increasingly common to record the values of longitudinal features (e.g., biomarkers) until the occurrence of an event of interest for a subject.
-The ``joint modeling'' approaches, namely modeling the longitudinal and survival outcomes through a joint likelihood model rather than separately, has received considerable attention during the past two decades~\citep{tsiatis2004joint}.
-Numerical studies suggest that these approaches are among the most satisfactory to combine information~\citep{yu2004joint}.
-They have the advantage of making more efficient use of the data since information about survival also goes into modeling the longitudinal features.
-In addition, they produce unbiased estimates and do not rely on approximations for incorporating conplex longitudinal trajectories.
-Most developments have either focused on shared random-effect models (SREMs)~\citep{wulfsohn1997joint}, in which characteristics of the longitudinal processes (for instance functions of the random effects) are included as features in the survival model ; or on joint latent class models (JLCMs)~\citep{vermunt2003latent}, in which the population is considered as heterogeneous, with the assumption that there exist homogeneous latent classes sharing the same marker trajectories and the same prognostic.
-
-\paragraph{The high-dimensional longitudinal data context.}
-With the exploding number of daily internet users, or with the development of electronic health records in a medical context, high-dimensional settings are becoming increasingly frequent in various contexts where the number of available features to consider as potential risk factors is tremendous.
-Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical.
-Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}.
-Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
-For instance,~\citet{jaffa2014joint} only considers 3 longitudinal outcomes in the simulation study while mentioning a ``high-dimensional multivariate longitudinal data'' context.
-
-\paragraph{General framework.}
-
-The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model.
-Let us consider the usual survival analysis framework.
-Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
-The event of interest could be for instance the time when a client stops using a company's product or service in a churn prediction setting~ ; or survival time, re-hospitalization, relapse or disease progression in a medical context.
-We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
-\begin{equation*}
-T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
-\end{equation*}
-respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.
-
-Let $X$ denotes the $p$-dimensional vector of time-independent features (e.g., patients characteristics, therapeutic strategy, or omics features recorded at the begining of a medical study), and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
-
-\paragraph{Heterogeneity of the population.}
-
-An assumption of heterogeneity within the subject population is frequently relevant in medical research where several differing profiles of subjects are expected~\citep{bussy2019c}.
-To take account of this, we introduce a latent variable $G \in \{0, \ldots, K-1\}$ modeling the $K \geq 1$ classes of different risk, which is a classical modeling assumption in JLCMs~\citep{lin2002latent, proust2014joint}.
-Let us denote
-\begin{equation}
-  \label{eq:pi}
-  \pi_{\xi_k}(x) = \P[G=k|X=x]
-\end{equation}
-the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
-\begin{equation}
-  \label{eq:pi-softmax} 
-  \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}
-\end{equation}
-where $\xi_k \in \R^p$ denotes a vector of coefficients that quantifies the impact of each time-independent features on the probability that a subject belongs to the $k$-th latent class, with $\xi_0 = \mathbf{0}_p$ for overparameterization purpose, where $\mathbf{0}_p$ stands for the vector of $\R^p$ having all coordinates equal to zero.
-The intercept term is here omitted without loss of generality.
-From now on, all computations are done conditionally on features $x$.
-
-\paragraph{Main contribution.}
-
-In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is inspired by both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a class-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
-To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
-for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
-Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
-Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
-Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.
-
-\paragraph{Organization of the paper.}
-
-A precise description of the model is given in Section~\ref{sec:Method}.
-Section~\ref{sec:inference} focuses on a regularized version of the
-model to exploit dimension reduction and prevent overfitting. Inference is
-presented under this framework, as well as the developed algorithm.
-Section~\ref{sec:Performance evaluation} introduces the C-index metric, as well as a novel evaluation strategy to assess diagnostic prediction performances while mimicking a real-time use of the model in clinical care, and finally the considered competing methods.
-Section~\ref{sec:simulation study} presents the simulation procedure used to evaluate the performance of our method in a high-dimensional context and compares it with state-of-the-art ones.
-In Section~\ref{sec:application}, we apply our method to a publicly available dataset.
-Finally, we discuss the obtained results in Section~\ref{sec:conclusion}.
-
-\paragraph{Notations.}
-
-Throughout the paper, for every $q > 0,$ we denote by $\norm{v}_q$ the usual $\ell_q$-quasi norm of a vector $v \in \R^m,$ namely $\norm{v}_q =(\sum_{k=1}^m|v_k|^q)^{1/q}$.
-We also denote $\norm{v}_0 = |\{k : v_k \neq 0\}|$, where $|A|$ stands for the cardinality of a finite set $A$.
-$\lfloor a \rfloor$ denotes the largest integer less than or equal to $a \in \R$.
-For $u, v \in \R^m$, we denote by $u \odot v$ the Hadamard product $u\odot v =(u_1v_1, \ldots, u_mv_m)^\top.$ 
-For a squared matrix $M$, $\text{vech}(M)$ stacks columns of $M$ one under another in a single vector, starting each column at its diagonal element.
-We write $I_m$ for the identity matrix of $\R^{m \times m}$.
-Finally, we write, for short, $\mathbf{1}_m$  (resp. $\mathbf{0}_m$) for the vector of $\R^m$ having all coordinates equal to one (resp. zero).
-
-\section{Method}
-\label{sec:Method}
-
-In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.
-
-\subsection{Class-specific marker trajectories}
-
-We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
-\begin{equation}
-  \label{eq:link-function}
-  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
- \end{equation}
-where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
-\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
-where $u^l(t) \in \R^{q_l}$ is a row vector of (possibly) time-varying features with corresponding unknown fixed effect parameters $\beta_k^l$, and $v^l(t) \in \R^{r_l}$ is a row vector of (possibly) time-varying features with $r_l \leq q_l$.
-A suitable distributional assumption for the random effects component is a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
-\begin{equation}
-   \label{eq:random-effect-law}
-   b^l \sim \cN(0, D_{ll})
- \end{equation} 
-with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. 
-To account for dependence between the different longitudinal outcome types, we let $\text{Cov}[b^l,b^{l'}] = D_{ll'}$ for $l \ne l'$ and we denote
-\[ D = 
-\begin{bmatrix}
-  D_{11} & \cdots & D_{1L}\\
-  \vdots &  \ddots & \vdots \\
-  D_{1L}^\top & \cdots & D_{LL}
-\end{bmatrix}
-\]
-the global variance-covariance matrix.
-In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.
-
-\subsection{Class-specific risk of event}
-
-To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
-\begin{equation}
-  \label{eq:intensity-model}
-  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
-\end{equation}
-where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
-\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
-the history of the true longitudinal process up to time $t$.
-For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.
-
-Let us finally denote
-\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
-\begin{equation}
-  \label{eq:def-asso-features}
-  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
-\end{equation}
-where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.
-
-\subsection{Generalization of SREMs and JLCMs}
-
-Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way.
-Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
-\begin{table}[htb]
-\centering
-\begin{tabular}{cccc}
-\toprule
-Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
-\midrule
-Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
-Random effects & $b^l$  & \citet{hatfield2011joint} \\
-Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
-Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
- \bottomrule
-\end{tabular}
-\caption[]{Examples of classical shared associations in SREMs.}
-\label{table:shared_associations}
-\end{table}
-
-Our proposition differs from standard SREMs in two respects: $(i)$ representation vector $\Psi(t)$ does not depend on the modeling assumptions in the longitudinal submodel and $(ii)$ it allows to choose multiple high-dimensional representation mappings that characterize time series and concatenate them, since we perform features selection through the regularization strategy described in Section~\ref{sec:penalized-obj}.
-Point $(i)$ is key since for instance the $\beta_k^l$ dependence in $\Psi_a^l$ would lead to complicated updates, while the $b^l$ dependence would lead to untracktable integrals that are common in SREMs and require approximation methods such as Monte Carlo~\citep{hickey2018joinerml}, being computationally intensive and not scalable in a high-dimensional context.
-For point $(ii)$ in practice, we use the \texttt{Python} library \texttt{tsfresh}~\citep{christ2018time} and include many extracted features, such as absolute energy of the time series, statistics on autocorrelation, or Fourier and wavelet basis projections, to name but a few.
-
-\subsection{Graphical model}
-Figure~\ref{fig:graphical-model} below gives the graphical representation for the lights model to get a beter view of the dependence structure for the model.
-\begin{figure}[!htb]
-\centering
-\resizebox{.9\textwidth}{!}{
-\begin{tikzpicture}
-\tikzstyle{main}=[circle, minimum size = 18mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
-\tikzstyle{var}=[rectangle, minimum size = 16mm, ultra thick, draw =black!80, node distance = 13mm,text centered]
-\tikzstyle{connect}=[-latex, thick]
-\tikzstyle{box}=[rectangle, draw=black!100]
-  \node[var, fill = white!100] (beta) [] {\Large$\bm{\beta_k^l}$};
-  \node[main, fill = white!100] (mu) [below=4cm of beta] {$\bm{M_{ik}^l(t)}$};
-  \node[main, fill = black!30] (Y) [below=of mu] {\Large$\bm{Y_i^l(t)}$};
-  \node[main, fill = white!100] (b) [left=of mu] {\Large$\bm{b_i^l}$};
-  \node[var, fill = white!100] (phi) [left=of Y] {\Large$\bm{\phi_l}$};
-  \node[var, fill = white!100] (Sigma) [above=of b] {\Large$\bm{D}$};
-  \node[var, fill = white!100] (xi) [right=4.5cm of beta] {\Large$\bm{\xi_k}$};
-  \node[var, fill = white!100] (gamma) [right=5cm of xi] {\Large$\bm{\gamma_k}$};
-  \node[main, fill = white!100] (G) [below=4cm of xi] {\Large$\bm{G_i}$};
-  \node[main, fill = white!100] (W) [right=8.5cm of Y] {\Large$\bm{\Psi_i(t)}$};
-  \node[main, fill = white!100] (lambda) [below=4cm of gamma] {\Large$\bm{\lambda_i(t)}$};
-  \node[main, fill = black!30] (T) [below=1.3cm of lambda] {\Large$\bm{T_i}$};
-  \node[main, fill = black!30] (X) at (4, -3.3) {\Large$\bm{X_i}$};
-
-  \path [->,draw,ultra thick, blue]
-        (G) edge (mu)
-        (Y) edge (W);
-  \path [->,draw,ultra thick]
-        (beta) edge (mu)
-        (b) edge (mu)
-        (mu) edge (Y)
-        (Sigma) edge (b)
-        (phi) edge (Y)
-        (xi) edge (G)
-        (X) edge (G)
-        (gamma) edge (lambda)
-        (W) edge (lambda)
-        (G) edge (lambda)
-        (lambda) edge (T)
-        ;
-
-  \node[rectangle, inner sep=8mm,draw=black!130, ultra thick, fit= (beta) (xi) (gamma),label={[anchor=west, yshift=-0.4cm, xshift=-8cm]{\Large$\bm{k = 1, \ldots,K}$}}] at (6.8,0.4) {};
-  \node[rectangle, inner sep=4mm,draw=black!100, ultra thick, fit= (beta) (mu) (b) (Y) (Sigma) (phi)] at (-1.6,-4.5) {};
-  \node[rectangle, inner sep=4mm, fit= (beta) (mu) (b) (Y) (Sigma) (phi),ultra thick, label={[anchor=west, yshift=-0.5cm, xshift=-2.8cm]{\Large$\bm{l = 1, \ldots,L}$}}] {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100, label=center:{\color{white}\Large{\textbf{Longitudinal}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (-1.5,2.8) {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-independent}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (5.2,2.8) {};
-
-  \node[rectangle,inner sep=6mm, draw=black!100,  label=center:{\color{white}\Large{\textbf{Time-to-event}}}, minimum width=5cm, minimum height = 1cm, fill = black!50] at (12,2.8) {};
-
-\end{tikzpicture}}
-\caption{Graphical representation of the lights model.
-Square boxes denote observed data while circles denote unobserved (including random) terms.
-One can then observe the two links in blue between the submodels that gives rise to the joint model.}
-\label{fig:graphical-model}
-\end{figure}
-
-\subsection{Likelihood}
-
-Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
-\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
-where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
-\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
-for all $l=1, \ldots, L$. 
-For the $i$-th subject, let us denote 
-\[
-\left\{
-    \begin{array}{ll}
-        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
-        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
-    \end{array}
-\right.
-\]
-with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
-\[ U_i = 
-\begin{bmatrix}
-  U_{i1} & \cdots & 0\\
-  \vdots &  \ddots & \vdots \\
-  0 & \cdots & U_{iL}
-\end{bmatrix} 
-\in \R^{n_i \times q}
-\qquad \text{and} \qquad
-V_i = 
-\begin{bmatrix}
-  V_{i1} & \cdots & 0\\
-  \vdots &  \ddots & \vdots \\
-  0 & \cdots & V_{iL}
-\end{bmatrix}
-\in \R^{n_i \times r}
-\]
-with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
-\[
-\left\{
-    \begin{array}{ll}
-        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
-        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
-    \end{array}
-\right.
-\]
-From now on, all computations are done conditionally on the design matrices $(U_i)_{i=1, \dots, n}$ and $(V_i)_{i=1, \dots, n}$.
-Denoting $(t_1^u, \ldots, t_J^u)$ the $J \in \N_+$ unique failure times, we then extract all representation vectors $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$ for all subjects $i=1, \dots, n$ before inference and from now on, all computations are also done conditionally on them.
-Now, let us make two first hypothesis.
-\begin{assumption}
-\label{assumption1}
-Suppose that the random effects are independent of the latent class membership, and that the latter remain independent conditional on the observed data (namely $T$, $\Delta$ and $Y$).
-\end{assumption}
-Assumption~\ref{assumption1} states that subject-and-longitudinal outcome specific random effects $b^l$ do not depend on the latent class membership, which is not a strong modeling assumption.
-\begin{assumption}
-\label{indep-hyp-2}
-Suppose that for a given subject $i$ and conditional on the random effects and the latent class membership, all $Y_i^l$ are independent for $l=1, \dots, L$.
-\end{assumption}
-Assumption~\ref{indep-hyp-2} is similar to standard modeling hypothesis, see for instance~\citet{tsiatis2004joint}.
-For all $k = 0, \ldots, K-1$, we denote 
-\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
-and 
-\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]
-
-Given~\eqref{eq:link-function} and Assumption~\ref{indep-hyp-2}, each $y_i^l$ is assumed to be from a one-parameter exponential family with respect to a reference measure which is either the Lebesgue measure (e.g., in the Gaussian case) or the counting measure (e.g., in the logistic case).
-The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
-\begin{equation}
-	\label{eq:exp-family-density}
-	f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
-\end{equation}
-with 
-\begin{equation}
-  \label{eq:def-Phi_i} 
-  \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}
-\end{equation} 
-and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
-The density described in~\eqref{eq:exp-family-density} encompasses several distributions, see Table~\ref{table:glm}.
-The functions $c_\phi(\cdot)$ and $d_\phi(\cdot)$ are known as well as the dispersion parameter $\phi_l$, while parameter $\beta_k$ has to be estimated.
-Note that $d_\phi(\cdot)$ is related to the normalizing constant.
-\begin{table}[htb]
-\centering
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{cccccc}
-\toprule
- Model & Support & Use cases & $\phi_l$ & $h_l(\cdot)$ in~\eqref{eq:link-function} & $c_\phi(\cdot)$ \\
- \midrule
- Gaussian & $\R$ & Continuous response data & $\sigma_l^2$ & $z \mapsto z$ & $z \mapsto z^2 / 2 \sigma_l^2$ \\ 
- Categorical & $\{0, 1\}^C$ & Outcome with $C$ modalities & 1 & $z \mapsto \log\Big(\dfrac{z}{1 - z} \Big)$ & $z \mapsto \log(1 + e^z)$ \\
- Poisson & $\N$ & Count of occurrences & 1 &  $z \mapsto \log(z)$ & $z \mapsto e^z$ \\
- \bottomrule
-\end{tabular}}
-\caption[]{Example of standard distributions that fit in the considered setting, given in the univariate case for simplicity of the notations.}
-\label{table:glm}
-\end{table}
-
-Let us denote 
-\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta \]
-the collection of the $\vartheta \in \N_+$ unknown parameters to estimate.
-Note that we include $\lambda_0$ in $\theta$.
-Indeed, we will estimate the baseline using nonparametric maximum likelihood, as a function taking mass at each failure time.
-So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
-To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
-\begin{assumption}
-\label{indep-hyp-3}
-Assume that
-\begin{equation}
-  \label{eq:ind-hyp}
-  f_\theta(t_i, \delta_i, y_i | b_i, G_i = k) = f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | b_i, G_i = k)
-\end{equation}
-for all $i=1, \ldots, n$.
-\end{assumption}
-Assumption~\ref{indep-hyp-3} is a generalization of classical hypothesis used in SREMs and JLCMs~\citep{hickey2016joint}.
-Remember that the densities are also conditionned by all the representation features $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}$.
-Hence Assumption~\ref{indep-hyp-3} says that the random effects vector $b_i$, the latent class membership and the representation features account for the association between the longitudinal and event outcomes, which makes sense given Figure~\ref{fig:graphical-model}.
-Then, one has
-\begin{equation}
-	\label{eq:log-lik}
-	\ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
-\end{equation}
-where 
-\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
-As expected, one has a JLCMs like log-likelihood.
-
-\paragraph{Specification of the design matrices.}
-
-In many practical applications, subjects show highly nonlinear longitudinal trajectories. We consider a flexible representations for $u^l(t)$ using a high-dimensional vector of time monomials, namely 
-\begin{equation}
-  \label{eq:alpha-def}
-  u^l(t) = (1, t, t^2, \ldots, t^\alpha)^\top
-\end{equation}
-with $\alpha \in \N_+$.
-The idea here is to let the practitioner choose a suitable polynomial order $\alpha$ for the representation -- that could also be tuned automatically using a model selection procedure and AIC~\citep{akaike1974new} for instance.
-We then let \[v^l(t) = (1, t)^\top\] so that each trajectory of each subject gets an affine random effect.
-Hence with this choice in practice, one has $q_l=\alpha + 1$ and $r_l=2$ for all $l=1, \ldots, L$.
-
-\section{Inference}
-\label{sec:inference}
-
-In this section, we describe the procedure for estimating the parameters of the lights model.
-Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
-
-\subsection{Penalized objective}
-\label{sec:penalized-obj}
-
-In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
-\begin{equation}
-  \label{eq:pen-log-lik}
-	\ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
-\end{equation}
-where for all $k=0, \ldots, K-1$, we add an elastic net regularization~\citep{zou2005regularization} of the vector $\xi_k$ and a sparse group lasso regularization~\citep{simon2013sparse} of the vector $\gamma_k$, for tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})^\top \in \R_+^2$.
-Here, for fixed $(\eta, \tilde{\eta}) \in [0, 1]^2$, one has
-\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
-for the elastic net penalty of any vector $z$, that is a linear combination of the lasso ($\ell_1$) and ridge (squared $\ell_2$) penalties, and
-\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} \] for the sparse group lasso penalty of any vector $z$, with $\norm{z}_{2,l}=\norm{z^l}_2$, and for which we do not have to account for group sizes since they all have the same one.
-Hence, the resulting optimization problem is written
-\begin{equation}
-  \label{eq:optim-pb}
-   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
- \end{equation}
-
-One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important (time-dependent and time-independent) features relatively to the prediction objective.
-The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership.
-The ridge part allows to handle potential correlation between time-independent features.
-On the other hand, for the sparse group lasso penalty, a group $l$ corresponds to a trajectory, that is one longitudinal outcome.
-So if $\hat \gamma_k^l$ is entirely zeroed (thanks to the group lasso part), it means that the $l$-th longitudinal process is discarded by the model in terms of risk effect for the $k$-th latent class.
-Then, the sparse part of the penalty allows to perform representation features selection for each trajectory : for $\hat \gamma_k^l$ that are not entirely zeroed, its support informs on the representation features involved in the $k$-th latent class risk of event for the $l$-th longitudinal outcome.
-Note that in practice, the intercept terms are not regularized.
-
-\subsection{A Proximal Quasi-Newton EM}
-\label{sec:prox-QNEM}
-
-In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm, being a combination between an EM algorithm~\citep{dempster1977maximum}, an L-BFGS-B algorithm~\citep{zhu1997algorithm} and a proximal gradient descent algorithm~\citep{beck2009fast}.
-EM algorithm has already been used for multivariate data joint modeling (see~\citet{lin2002maximum} for instance), but here we face different and original problems: for each subject $i$, the latent variables are the pairs $(G_i, b_i)$ (not only the random effects); and then, we want to minimize the penalized objective $\ell_n^\text{pen}$ (not ``only'' the negative log-likelihood).
-
-We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
-It can be written
-\begin{align*}
-  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
-  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
-  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
-  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
-\end{align*}
-Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.
-
-\paragraph*{E-step.}
-
-We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
-\begin{equation*}
-  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
-\end{equation*}
-Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
-\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
-for different functions $g$, where we denote 
-\begin{equation}
-  \label{eq:pi_ik-def}
-  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
-\end{equation}
-the posterior probability of the latent class membership using parameters $\theta^{(w)}$. 
-Then, either $g(b_i, G_i) = \tilde g(b_i)$ (either $\tilde g_1: b_i \mapsto b_i$ or $\tilde g_2: b_i \mapsto b_i b_i^\top$), or $g(b_i, G_i) = g_k(G_i)$ with $g_k~:~G_i \mapsto \ind{\{G_i=k\}}$.
-
-In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
-\begin{equation*}
-  \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
-\end{equation*}
-where
-\begin{equation}
-  \label{eq:Lambda_ik-b_i}
-  \Lambda_{ik,\tilde g}^{\theta^{(w)}} =  f_{\theta^{(w)}}(t_i, \delta_i | G_i = k) f_{\theta^{(w)}}(y_i | G_i = k) \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | y_i, G_i = k) \dd b_i,
-\end{equation}
-and in the Gaussian case used in practice one has
-\begin{align*}
-  y_i | G_i = k ; \theta^{(w)} &\sim \cN\big(U_i\beta_k^{(w)}, V_iD^{(w)}V_i^\top + \Sigma_i^{(w)}\big), \\
-  b_i | y_i, G_i=k ; \theta^{(w)} &\sim \cN \big(O_{i,k}^{(w)}, W_i^{(w)}\big),  
-\end{align*}
-with $O_{i,k}^{(w)}=W_i^{(w)}V_i^T{\Sigma_i^{(w)}}^{-1}(y_i - U_i\beta_k^{(w)})$ and $W_i^{(w)} = (V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1})^{-1}$.
-Hence, what is interesting here is that~\eqref{eq:Lambda_ik-b_i} is actually closed-formed since
-\[ \E_{\theta^{(w)}}[ \tilde g_1(b_i) | y_i, G_i = k] = O_{i,k}^{(w)} \quad \text{and} \quad \E_{\theta^{(w)}}[ \tilde g_2(b_i) | y_i, G_i = k] = W_i^{(w)} + O_{i,k}^{(w)}{O_{i,k}^{(w)}}^\top\]
-and then tractable analytically.
-Whereas in classical SREMs, some form of approximation must be used in practice to compute this quantity, like numerical integration techniques (for instance Gaussian quadrature) or Monte Carlo approximation.
-
-Now in the case $g = g_k$, one has
-\begin{equation}
-\label{eq:pi_ik-def}
-  \E_{\theta^{(w)}}[g_k(Gi) | t_i, \delta_i, y_i] = \pi_{ik}^{\theta^{(w)}} = \dfrac{\pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}.
-\end{equation}
-
-\paragraph*{Proximal Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
-We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
-First, the update of $D^{(w)}$ is naturally given in closed-form by
-\begin{equation}
-  \label{eq:update_D}
-  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
-\end{equation}
-Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
-\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
-\begin{equation}
-  \label{eq:minimization-xi_k}
-  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
-\end{equation}
-It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:pi_ik-def}).
-We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
-of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
-In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
-Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
-\begin{equation}
-  \label{eq:sub-problem-xi_k}
-  \begin{split}
-    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
-    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
-  \end{split} 
-\end{equation}
-where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
-\begin{equation}
-  \label{eq:grad-xi_k}
-  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big(\pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
-\end{equation}
-In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}.
-Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.
-
-Concerning the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, let us follow the same strategy by denoting
-\begin{equation*}
-  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[ (y_i \odot \Phi_i^{(w)})^\top \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] - \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big]
-\end{equation*}
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$. The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
-\begin{equation}
-  \label{eq:minimization-beta_k}
-  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k).
-\end{equation}
-The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
-\begin{equation}
-  \label{eq:grad-beta_k}
-  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k} = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[ U_{i}^\top \Sigma_i^{(w)} I_{n_{i}} \big(y_i - U_{i}\beta_k - V_{i} \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i]\big)\Big].
-\end{equation}
-The closed-form update is then given by
-
-\begin{equation}
-  \label{eq:update_beta}
-  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
-\end{equation}
-Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
-\[Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[\delta_i \gamma_k^\top \Psi_i(t_i) - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \exp \big\{\gamma_k^\top \Psi_i(t_j^u)\big\} \ind{\{t_j^u \leq t_i\}} \Big]\]
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$.
-The integration over the survival process has been replaced with a finite sum over the process evaluated at the $J$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
-The update for $\gamma_k^{(w)}$ requires to solve the following convex minimization problem
-\begin{equation}
-  \label{eq:minimization-gamma_k}
-  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\cA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
-\end{equation}
-
-The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-gamma_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}).
-Then, we choose to solve problem~\eqref{eq:minimization-gamma_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}, where we first define
-\[p_{\varrho}(y) = \argmin_u \Big\{\dfrac{1}{2\varrho} \left\lVert u - \big(y - \varrho  \nabla f(y) \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\}.\]
-
-\vspace{.5cm}
-
-\begin{algorithm}[H]
-\DontPrintSemicolon
-\caption{Proximal gradient descent ISTA$_{\text{sg} l_1}$($v$, $f$, $\tilde \eta$, $\zeta$)}
-\label{alg:ISTA}
-\KwInput{A vector $v$, a function $f$ (its gradient $\nabla f$), hyper-parameters $\tilde \eta$, $tol$ and $\zeta$}
-\KwOutput{$\flat^{\breve{w}^{max}}$}
-Initialize $\flat^{(0)}=v$, $\varrho^0 > 0$, some $\epsilon > 1$ \\
-\For{$\breve{w} = 0,\ldots,$ until convergence}    
-{Backtracking line search : find the smallest nonnegative integer $i^{\breve{w}}$ such that
-$f\big(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})})\big) \leq f(\flat^{(\breve{w})}) + \big<\big(p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}\big), \nabla f(\flat^{(\breve{w})})\big> + \dfrac{1}{2\tilde{\varrho}_{\breve{w}}}\norm{p_{\tilde{\varrho}_{\breve{w}}}(\flat^{(\breve{w})}) - \flat^{(\breve{w})}}_2^2$
-with $\tilde{\varrho}_{\breve{w}} = \epsilon^{i^{\breve{w}}}\varrho^{\breve{w}}$\\
-Set $\varrho^{(\breve{w} + 1)} = \tilde{\varrho}_{\breve{w}}$ \\
-$\flat^{(\breve{w}+1)} = p_{\varrho^{(\breve{w} + 1)}}(\flat^{(\breve{w})}) \coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \big(\flat^{(\breve{w})} - \varrho^{(\breve{w} + 1)} \nabla f(\flat^{(\breve{w})}) \big)$ \\
-\If{$\dfrac{\norm{\flat^{(\breve{w}+1)} - \flat^{(\breve{w})}}_2}{\varrho^{(\breve{w} + 1)}} < tol$}
-  {
-    \textbf{Return:} {$\flat^{\breve{w}}$}
-  }
-}
-\end{algorithm}
-
-\vspace{.5cm}
-Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
-\begin{lemma}
-\label{lemma:prox-sgl1}
-The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
-\normalfont
-\begin{equation*}
-  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
-\end{equation*}
-\end{lemma}
-The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable.
-We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our \texttt{lights} package a first \texttt{Python} implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
-A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
-Hence, we update $\gamma_k^{(w)}$ with the following step
-\begin{equation}
-  \label{eq:gamma-update}
-  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
-\end{equation}
-Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
-The gradient of $Q_{n,k}^{(w)}$ is here given by
-\begin{align}
-  \label{eq:grad-Q}
-  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Psi_{i, a}^l(t_i) \Big[ \delta_i - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \exp \big\{\gamma_k^\top \Psi_i(t_j^u) \big\} \ind{\{t_j^u \leq t_i\}} \Big]
-\end{align}
-for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.
-
-For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
-\begin{equation}
-  \label{eq:update_lambda_0}
-  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} \ind{\{t_i \geq t\}} },
-\end{equation}
-for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
-
-Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
-\begin{align}
-  \label{eq:update_phi}
-  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
-  & - 2V_{il} \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
-\end{align}
-Appendix~\ref{sec:prox-QNEM class} gives some details about the way we implement the equations and their computation in practice.
-
-\paragraph*{Initialization.}
-Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced.
-We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNEM}.
-For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\cA}$.
-Then, we initialize $\lambda_0^{(0)}$ like if there is no latent classes ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}.
-Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no latent classes ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.
-
-\paragraph*{The prox-QNEM algorithm.}
-Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting prox-QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.
-
-% \vspace{.5cm}
-
-\begin{algorithm}[!htb]
-\DontPrintSemicolon
-\caption{prox-QNEM algorithm for the lights model inference}
-\label{alg:prox-QNEM}
-\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
-\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
-\tcc{Initialization}
-Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
-\For{$w = 1,\ldots,$ until convergence}    
-{ 
-  \tcc{E-step}
-  Compute $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for $\tilde g_1$ and $\tilde g_2$\\
-  Compute $\pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:pi_ik-def}\\
-  \tcc{Proximal Quasi-Newton M-step}
-  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
-  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
-  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
-  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
-  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
-  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
-}
-\textbf{Return:} {$\hat \theta$}
-\end{algorithm}
-
-Then in practice, the convergence for Algorithm~\ref{alg:prox-QNEM} is declared when the relative objective $\big(\ell_n^\text{pen}(\theta^{(w+1)}) - \ell_n^\text{pen}(\hat\theta^{(w)}) \big) / \ell_n^\text{pen}(\hat\theta^{(w)})$ goes below a chosen threshold.
-
-\subsection{The case $K=2$}
-\label{sec:K=2}
-
-In many practical applications -- including those addressed in Section~\ref{sec:application} -- we are interested in identifying one latent class of the population with a high risk of adverse event compared to the others.
-Then, in the following, we consider $G \in \{ 0, 1\}$ where $G=1$ means high-risk of early adverse event and $G=0$ means low risk.
-To simplify notations, let us set $\xi = \xi_1$, $\pi_\xi(x)$ the conditional probability that a subject belongs to the latent class with high risk of adverse event given its time-independent features $x$, and $\zeta_1 = \zeta_{1,1}$.
-In this context, which is the one we consider in practice in Sections~\ref{sec:simulation study} and~\ref{sec:application}, we suppose that the ``right'' penalty strength for $\gamma_k$ does not depend on $k$, so that we denote $\zeta_2 = \zeta_{2,0} = \zeta_{2,1}$.
-
-\paragraph{Cross-validation procedure.}
-
-The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
-In practice, we use the \texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
-SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
-Moreover, the search interval for hyper-parameter $\zeta_1$ is automatically computed in a data driven way described in Appendix~\ref{sec:auto-grid-cv}, while the search intervals for $\zeta_2$ is determined empirically.
-% TODO: Figure de visualisation de la CV de (\zeta_1, \zeta_2) en 3D
-
-\section{Performance evaluation}
-\label{sec:Performance evaluation}
-
-In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances.
-We then introduce the appropriate metric used to this end: the C-index ; and finally, the considered models for performance comparisons.
-
-\subsection{Evaluation strategy: the real-time prediction paradigm}
-\label{sec:evaluation strategy}
-
-When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
-Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}.
-However, with the prospect of using prognostic prediction in real-time on a daily basis, practitioners will naturally require predictive prognostic tools to evaluate models fit and compare models in terms of discriminative power.
-So we choose to put ourselves in what we call a ``real-time prediction paradigm''.
-
-\paragraph{Real-time prediction paradigm.}
-
-Once the learning phase is achieved for the model on a training set (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), we want to assess real-time risk prediction performances on a test set.
-
-Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction --
-
-\paragraph{Predictive marker.}
-
-The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
-
-posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:pi_ik-def}).
-Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
-
-But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow
-
- a natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ would rather be 
-\begin{equation}
-  \label{eq:marker-def}
-  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
-\end{equation}
-with 
-\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
-
-dire ici qu'on tronque $y$ à $t^{max}$ et le noter $|t^{max}$
-
-where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
-Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
-\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
-
-\subsection{The C-index metric}
-\label{sec:Metrics}
-
-We detail in this section the metric considered to evaluate risk prediction performances. Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}. 
-In this context, the predictive marker for the lights model is
-\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
-Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
-A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
-\begin{equation*}
-  \cC =\P[\hat \cR_i > \hat \cR_j | T^\star_i < T^\star_j],
-\end{equation*}
-with $i \neq j$ two independent subjects (which does not depend on $i, j$ under the i.i.d. sample hypothesis). 
-In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
-\begin{equation*}
-  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
-\end{equation*}
-with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
-A Kaplan-Meier estimator for the censoring distribution leads to a nonparametric and consistent estimator of $\cC_\tau$ \citep{uno2011c}, already implemented in the \texttt{Python} package \texttt{lifelines}.
-Hence in the following, we consider the C-index metric to assess performances.
-
-Note that other metrics have been proposed to compare individual risk predictions.
-In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$.
-But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~-v\citep{bussy2019comparison}), which is not the case of the C-index metric.
-Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.
-
-\subsection{Competing models}
-\label{sec:competing models}
-
-In this section, we briefly introduce the models we consider for performance comparisons in the simulation study as well as in the applications on real datasets in Section~\ref{sec:application}.
-We also give the corresponding marker rule for each model, using our notations.
-
-\paragraph*{Landmark Cox model.}
-
-The first model we consider as a baseline is the well known Cox PH model with time-independent features, also known in the joint modeling context as the ``landmark'' model, in which we include basic time-independent features extracted from longitudinal processes and computed at the prediction time [REF].
-In the Cox PH model introduced in \citet{Cox1972JRSS}, a parameter vector $\beta$ is estimated by minimizing the partial log-likelihood given by
-\begin{equation*}
-  \ell_n^{\text{cox}}(\beta) = n^{-1} \sum_{i=1}^n \delta_i \big( x_i^\top \beta - \log \sum_{i' : t_{i'} \geq t_i} \text{exp}(x_{i'}^\top \beta) \big).
-\end{equation*}
-For each time-dependant feature, we include subject-specific random effects, values of longitudinal processes at baseline and at time $t^{max}$, slope of longitudinal processes at time $t^{max}$ and area under of longitudinal processes from baseline up to $t^{max}$.
-The predictive marker for subject $i$ is the regular risk score~\citep{therneau2000cox}, that is $\hat \cR_i^{\text{cox}} = \exp (x_i^\top\hat\beta^{\text{cox}})$, with
-\[\hat\beta^{\text{cox}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta).\]
-We use the \texttt{Python} package \texttt{tick} for the estimation of the previous quantity.
-Ties are handled via the Breslow approximation of the partial likelihood \citep{breslow1972contribution}.
-
-\paragraph*{Penalized Cox model.}
-
-We then consider a penalized version of the Cox model, including now much more time-independent features automatically extracted from longitudinal processes using the \texttt{Python} package \texttt{tsfresh}~\citep{christ2018time}.
-The predictive marker for the a subject $i$ is $\hat \cR_i^{\text{cox EN}} = \exp (x_i^\top\hat\beta^{\text{cox EN}})$, with 
-\[\hat\beta^{\text{cox EN}} \in \argmin_{\beta \in \R^\vartheta} -\ell_n^{\text{cox}}(\beta) + \xi \big( (1-\eta)\norm{\beta}_1 + \frac{\eta}{2} \norm{\beta}_2^2 \big)\] 
-the estimate obtained using elastic net penalization (still with the \texttt{tick} package), and
-where $\xi$ is chosen by the a 10-fold cross-validation procedure, for a given $\eta \in [0, 1]$.
-
-\paragraph*{The time-dependent Cox model.}
-
-A classical extension of the Cox model supposes that features depend on time~\citep{sueyoshi1992semiparametric} (and then loosing the risk proportionality property), that is with our notations
-\[\lambda_i(t) = \lambda_0(t) \exp\big(x_i^\top \gamma_0 + y_i(t_i^{max})^\top \beta \big). \]
-We use the \texttt{survival} \texttt{R} package~\citep{zhang2018time}, and we consider the natural predictive marker for a subject $i$, that is the corresponding risk score given by
-\[\hat \cR_i^\text{cox t-d} = \exp \big(x_i^\top \hat\gamma_0 + y_i(t)^\top \hat\beta \big).\]
-
-\paragraph*{Multivariate joint latent class model.}
-
-We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
-Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model.
-Given the latent class membership, each submodel is assumed to be independent.
-In this context, the predictive marker for subject $i$ is
-\[\hat \cR_i^\text{lcmm} = \dfrac{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i)\big) f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
-
-\paragraph*{Multivariate shared random effect model.}
-
-In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
-\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\}.\]
-
-\texttt{@Antoine : est-ce qu'il peut prendre des time-indep feat ? et quelles sont les assos ?}
-
-\section{High-dimensional simulation study}
-\label{sec:simulation study}
-
-In this section, we provide details regarding data generation, followed by the results of the extensive Monte Carlo simulation study to examine our method and compare it with state-of-the-art.
-
-\subsection{Simulation design}
-\label{simulation design}
-
-In order to assess the proposed method, we run an extensive Monte Carlo simulation study in the case $K=2$ (see Section~\ref{sec:K=2}). We first choose a coefficient vector 
-\begin{equation}
-  \label{eq:xi-sparse}
-  \xi = (\underbrace{\varsigma_1,\ldots,\varsigma_1}_s,0,\ldots,0) \in \R^p,
-\end{equation} 
-with $\varsigma_1\in\R$ being the value of the active coefficients and $s\in \{1,\dots,p\}$ a sparsity parameter chosen such that a proportion $r_s$ of time-independent features are actually active, that is $s = \lfloor p \times r_s \rfloor$.
-For a desired high-risk subjects proportion $\pi_1 \in [0,1]$, the high-risk subjects index set is given by
-\[\cH = \big\{\lfloor \pi_1 n \rfloor \text{ random samples without replacement} \big\} \subset \{1, \ldots,n\}.\]
-For the generation of the time-independent features matrix, we take 
-\[ [x_{ij}] \in \R^{n \times p} \sim \cN \big(0, \bSigma_1(\rho_1)\big),\] 
-with $\bSigma_1(\rho_1)$ a $(p \times p)$ Toeplitz covariance matrix \citep{mukherjee1988some} with correlation $\rho_1 \in (0, 1)$, that is $\bSigma_1(\rho_1)_{jj'} = \rho_1^{|j - j'|}$. 
-We then add a $gap \in \R^+$ value for subjects $i \in \cH$ and subtract it for subjects $i \notin \cH$, only on active features, that is
-\[x_{ij} \leftarrow x_{ij} + (-1)^{\ind{\{i \notin \cH\}}} gap\ \text{for } j = 1, \dots, s.\]
-Note that this is equivalent to generate the time-independent features according to a gaussian mixture. We finally normalize each time-independent features for practical interpretations.
-Then, we generate $G_i \sim \cB\big(\pi_\xi(x_i)\big)$, where $\pi_\xi(x_i)$ is computed given Equation~\eqref{eq:pi-softmax} and $\cB(\alpha)$ denotes the Bernoulli distribution with parameter $\alpha \in [0,1]$.
-
-Now, concerning the simulation of the longitudinal data, the idea is to sample from multivariate normal distributions. 
-We use linear time-varying features such that
-\[ Y_i^l(t) = \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \big( (1, t)^\top \beta_k^l + (1, t)^\top b_i^l + \epsilon_i^l(t) \big) \]
-where $t \geq 0$, the error term $\epsilon_i^l(t) \sim \cN(0, \sigma_l^2)$, the global variance-covariance matrix for the random effects components~\eqref{eq:random-effect-law} is such that $D = \bSigma_2(\rho_2)$, a $(r \times r)$ Toeplitz covariance matrix with correlation $\rho_2 \in (0, 1)$, and the fixed effect parameters are generated according to
-\[ \beta_k^l \sim \cN\Big( 
-\mu_k, 
-\begin{bmatrix}
-  \rho_3 & 0\\
-  0 & \rho_3
-\end{bmatrix} 
-\Big) \]
-for $k \in \{0, 1\}$ and with correlation $\rho_3 \in (0, 1)$, so that trajectories start on average higher for high-risk subjects ($G_i=1$), and with a higher slope than low-risk subjects ($G_i=0$).
-This choice makes sens regarding that for simplicity, we take joint association parameters being all positive, see~\eqref{eq:gamma-sparse}.
-
-Most joint modeling simulation studies use fixed time points~\citep{hickey2018joinerml}, that is $n_i^l = n_i^{l'}$ and $t_{ij}^l = t_{ij}^{l'}$ for all $(l, l') \in \{1, \ldots, L\}^2$ and $j = 1, \ldots, n_i^l$ for a given subject $i$, but it does not reflect actual behavior in practice, since data measurements of the different longitudinal processes are barely done simultaneously.
-For this reason, the simulation of realistic longitudinal features is not trivial and we use Hawkes processes~\citep{hawkes1974cluster} with exponential kernels to generate measurement times, which is a family of counting process with an autoregressive intensity.
-Hawkes processes model self-exciting behavior, that is in our case, when the measurement of one longitudinal process makes future measurements of the process -- as well as those of others processes, in a multivariate way -- more likely to happen.
-Namely, for a subject $i$, times $\{t_{ij}^l\}_{j \geq 1}$ for processes $l=1, \ldots, L$ are simulated using a multivariate Hawkes process $N_{it} = [N_{it}^1 \cdots N_{it}^L]$ with $t \geq 0$ and $N_{it}^l = \sum_{j \geq 1} \ind{\{t_{ij}^l \leq t\}}$.
-The process $N_{it}$ is a multivariate counting process, whose components $N_{it}^l$ have intensities
-\[ \lambda_i^l(t) = \Upsilon_{l} + \sum_{l'=1}^L \sum_{j \geq 1} A_{ll'} \upsilon \exp\big(-\upsilon(t - t_{ij}^{l'}) \big) \]
-for $l=1, \ldots, L$.
-$\Upsilon_{l} \geq 0$ is called baseline intensity, and corresponds to the exogenous probability of having a measurement for process $l$.
-We sample $\Upsilon_{l} \sim \cU\big([0.1,1]\big)$ ] which produces unbalanced baselines, where $\cU\big([a,b]\big)$ stands for the uniform distribution on a segment $[a,b]$.
-The matrix $A = [A_{ll'}]_{1 \leq l,l' \leq L}$ is the adjacency matrix such that $A_{ll'} \geq 0$ quantifies the impact of past measurement time of process $l'$ on the measurement time of process $l$, and $\upsilon \geq 0$ is a memory parameter.
-We sample $A_{ll'} \sim \cU\big([0.1,0.2]\big)$, but then enforce $A$ to be sparse (with a chosen density of $0.3$) by randomly zeroing non-diagonal coefficients.
-Simulation of the Hawkes processes
-is achieved using the \texttt{tick} library~\citep{bacry2017tick}. An example of simulated longitudinal features for an individual is displayed in Figure~\ref{fig:example-data} below.
-
-Now to generate survival times, we want to induce sparsity in the longitudinal data effects, so we set 
-\[\cS_k=\Big\{ k \big\lfloor \dfrac{L r_s}{K} \big\rfloor + 1, \ldots,  (k + 1) \big\lfloor \dfrac{L r_s}{K} \big\rfloor \Big\}\] 
-and we choose a risk model~\eqref{eq:intensity-model} with
-\begin{equation*}
-  \left\{
-    \begin{array}{ll}
-    \big(\varphi_{k,a}(t, b_i^l)\big)_{a \in \{1,2\}} = \big(\beta_{k,1}^l + \beta_{k,2}^l t + b_{i,1}^l + b_{i,2}^l t\ , {b_i^l}^\top \big)^\top \ind{\{l \in \cS_k\}} + \cN \big(0, \bSigma_3(\rho_3)\big) \ind{\{l \notin \cS_k\}}  \\
-    \big( \varphi_{k,a}(t, b_i^l)\big)_{a=3,\ldots,\cA} = \cN \big(0, \bSigma_4(\rho_3)\big),
-    \end{array}
-  \right.
-\end{equation*}
-
-rmq : plus besoin du sk dans phi vu qu'on l'ajoute directement ds le Y non ?
-
-and 
-\begin{equation}
-  \label{eq:gamma-sparse}
-  \gamma_{k,a}^l = \varsigma_2 \mathbf{1}_{\imath_a} \ind{\big\{l \in \cS_k,\ a\in \{1, 2\} \big\}}
-\end{equation}
-for the choice of the association parameters, with $\cA = 3 + 3(1-r_s)/r_s$ so that $(i)$ there is a proportion $r_s$ of active longitudinal features among the $L$ (with the $\cS_k$), $(ii)$ we induce sparsity within each latent class adding a proportion of $(1-r_s)$ non-active association features ($a \geq 3$) in each latent class (similarly to~\eqref{eq:xi-sparse} for the time-independent features), and $(iii)$ the longitudinal effects are distinct for each latent class $k=0, \ldots, K-1$ (thanks to the $\cS_k$ definition).
-
-In terms of active association features, it corresponds to~\eqref{eq:intensity-model} taking the first two functionals described in Table~\ref{table:shared_associations} as shared associations with $\alpha = 1$ in~\eqref{eq:alpha-def} so that one can write
-\[ \lambda_i(t|G_i = k) = \lambda_0(t) \exp \big\{ \iota_{i,k,1} + \iota_{i,k,2} t \big\},\]
-being a Cox model with a linear time-varying feature that allows the following explicit survival times generation process.
-Then, we do not consider the time-dependent slope as association feature since here with $\alpha = 1$, it would be collinear with the random effects.
-We choose a Gompertz distribution~\citep{gompertz1825xxiv} for the baseline, that is
-\begin{equation}
-  \label{eq:baseline}
-  \lambda_0(t) = \kappa_1 \kappa_2 \exp(\kappa_2t)
-\end{equation}
-with $\kappa_1 > 0$ and $\kappa_2 \in \R$ the scale and shape parameters respectively, which is a common distribution choice in survival analysis~\citep{klein2005survival} with a rich history in describing mortality curves.
-One can now generate survival times explicitly as
-\begin{equation}
-  \label{eq:time-generation}
-  T_i^\star | G_i=k \sim \dfrac{1}{\iota_{i,k,2} + \kappa_2} \log \Big(1 - \dfrac{(\iota_{i,k,2} + \kappa_2) \log U_i}{\kappa_1 \kappa_2 \exp\iota_{i,k,1}} \Big)
-\end{equation}
-where $U_i \sim \cU\big([0,1]\big)$ (see Appendix~\ref{sec:event-time-simu} for the derivation of this expression).
-The distribution of the censoring variable $C_i$ is the geometric distribution $\cG(\alpha_c)$, where $\alpha_c \in (0, 1)$ is empirically tuned to maintain a desired censoring rate $r_c \in [0,1]$. 
-Finally, the prediction time for subject $i$ is draw according to 
-\begin{equation}
-  \label{eq:law-t_i^max}
-  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
-\end{equation}
-with $(\nu_1, \nu_2) \in \R_+^2$ two shape parameters. Figure~\ref{fig:timax-pdf} shows the corresponding probability density function for $t_i^{max}$ with some explanations.
-We then truncate all $Y_i$ measurements being after $t_i^{max}$, since it represents the ``present'' time, so the time up to which one has longitudinal data.
-The choice of all hyper-parameters is driven by the applications on real data presented in Section~\ref{sec:application}, and summarized in Table~\ref{table:parameters choice}.
-Figure~\ref{fig:example-data} gives an example of data generated according to the design we have just described.
-\begin{table}[htb]
-\caption{Hyper-parameter choices for simulation. Let us also precise that we take $n \in [200, 4000]$ and $(p, L) \in [3, 300]^2$.
-\label{table:parameters choice}}
-\centering
-\resizebox{\textwidth}{!}{
-\begin{tabular}{ccccccccccccc}
-\toprule
-$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\nu_1, \nu_2)$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
-\midrule
-0.4 & $(, 10^{-3},)$ & 
-$\begin{pmatrix}
-  1 \\ 0.3
-\end{pmatrix}$ & 
-$\begin{pmatrix}
-  1 \\ 0.5
-\end{pmatrix}$ 
-& 0.5 & 4 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
-\bottomrule
-\end{tabular}}
-\end{table}
-\begin{figure}[!htb]
-\centering
-\subfigure[Density of the law used to simulate $t_i^{max}$ according to~\eqref{eq:law-t_i^max} with $(\nu_1, \nu_2) = (2,5)$, to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.]{
-\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
-\label{fig:timax-pdf}}
-\hspace{.02\textwidth}
-\subfigure[exemple simu]{
-% \includegraphics[width=.38\textwidth]{./figures/time2.pdf}
-\label{fig:example-data}}
-\caption{Illustrations of simulated data.}
-\end{figure}
-
-Following the real-time prediction paradigm presented in Section~\ref{sec:evaluation strategy}, we compare the predictive performances of the lights model with the competing ones introduced in Section~\ref{sec:competing models} in terms of C-index (see Section~\ref{sec:Metrics}).
-We also want to assess the stability of the lights model in terms of feature selection, but we do not compare this aspect with state-of-the-art since other models are not designed to perform feature selection, and comparisons would be unfair.
-To this end, we follow the same simulation procedure explained in the previous lines and for each simulation case, we use the following approach to evaluate the variable selection power of the model.
-Denoting
-\[\tilde{\xi}_j = \dfrac{|\hat{\xi}_j|}{\max_{j=1,\ldots,p}{|\hat{\xi}_j|}} \qquad \text{and} \qquad \tilde{\gamma}_{k,a}^l = \dfrac{|\hat{\gamma}_{k,a}^l|}{\max_{a=1, \ldots, \cA}{|\hat{\gamma}_{k,a}^l|}}\]
-for $k=0,\ldots,K-1$ and $l=1, \ldots, L$, and considering that $\tilde{\xi}_j$ and $\tilde{\gamma}_{k,a}^l$ are the predicted probability that the true $\xi_j$ equals $\varsigma_1$ and the predicted probability that the true $\gamma_{k,a}^l$ equals $\varsigma_2$ respectively, then we are in a binary prediction setting and we use the two resulting AUC scores of this problem to evaluate the selection power of the time-independent and the time-dependent features respectively.
-
-\subsection{Results of simulation}
-\label{sec:simulation results}
-Let us present now the simulation results.
-
-\section{Applications}
-\label{sec:application}
-
-In this section, we apply our lights method on two publicly available datasets and compare its performance with state-of-the-art methods.
-
-\subsection{The PBCseq dataset}
-
-This dataset is a follow-up to the original dataset~\citep{fleming2011counting, murtaugh1994primary}from the Mayo Clinic trial in primary biliary cirrhosis (PBC) of the liver conducted between 1974 and 1984.
-A total of 424 PBC patients, referred to Mayo Clinic during that ten-year interval, met eligibility criteria for the randomized placebo controlled trial of the drug D-penicillamine.
-The first 312 cases in the data set participated in the randomized trial and contain largely complete data.
-The additional 112 cases did not participate in the clinical trial, but consented to have basic measurements recorded and to be followed for survival.
-Six of those cases were lost to follow-up shortly after diagnosis, so the data here are on an additional 106 cases as well as the 312 randomized participants.
-The dataset contains only baseline measurements of the laboratory paramters and contains multiple laboratory results, but only on the first 312 patients.
-
-\subsection{The MIMIC III dataset}
-
-The MIMIC III (Medical Information Mart for Intensive Care III) database is a large, freely-available hospital dataset containing de-identified data from over 40,000 patients.
-This data comes from patients who were admitted in critical care units to Beth Israel Deaconess Medical Center in Boston, Massachusetts from 2001 to 2012~\citep{johnson2016mimic}.
-The dataset was populated with data that had been acquired during routine hospital care, so there was no associated burden on caregivers and no interference with their workflow.
-
-% \url{https://www.nature.com/articles/sdata201635/}\\
-% \url{https://ieeexplore.ieee.org/abstract/document/1166854}\\
-% \url{https://www.sciencedirect.com/science/article/abs/pii/S2213260014702395}\\
-% \url{https://www.sciencedirect.com/science/article/pii/S2352556818302169}\\
-% }
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
-algorithm (prox-QNEM) has been derived, that considers a penalization of the likelihood in order to perform feature
-selection and to prevent overfitting.
-
-\section*{Software}
-All the methodology discussed in this paper is implemented in \texttt{Python}.
-The code is available from \url{https://github.com/Califrais/lights} in the form of annotated programs, together with a notebook tutorial.
-A technical documentation is also provided in Appendix~\ref{sec:package}.
-
-\section*{Acknowledgements}
-\textit{Conflict of Interest}: None declared.
-
-\appendix
-
-\begin{center}
-\LARGE \textbf{Appendices}
-\end{center}
-
-\section{Proof of Lemma~\ref{lemma:prox-sgl1}}
-\label{sec:proof-lemma-prox-sgl1}
-
-The proximal operator~\citep{moreau1962fonctions} for any function $f : \R^p \to \R$ is defined by
-\begin{equation}
-    \label{eq:prox-def}
-    \text{prox}_{f}(u) = \argmin_{x \in \R^p} \big\{\frac{1}{2}\norm{x - u}_2^2 + f(x)\big\}.
-\end{equation}
-To prove Lemma~\ref{lemma:prox-sgl1}, we need to show that
-\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L  \norm{\cdot}_{2,l}} = \text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\lambda_1 \norm{\cdot}_1}\]
-where $\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}$ is the proximal operator for the sparse group lasso regularizer, $\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}$ is the proximal operator for the group lasso regularizer and $\text{prox}_{\lambda_1 \norm{\cdot}_1}$ is the proximal operator for the lasso regularizer.
-$\lambda_1 > 0$ and $\lambda_2 > 0$ are regularization parameters, and $\norm{\cdot}_{2,l}$ denotes the Euclidean norm on features in the $l$-th group.
-
-\begin{proof}
-Fix any $\beta \in \R^p$ and denote 
-\begin{equation*}
-    k_\beta(x) =  \frac{1}{2}\norm{ x - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)}_2^2 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l}
-\end{equation*}
-for all $x \in \R^p$.
-Given that $\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big)_j=\max(0, 1-\lambda_1/|\beta_j|)\beta_j$~\citep{bach2011optimization}, $k_\beta$ is clearly strictly convex.
-Denoting $x^\star$ its unique minimizer,~\eqref{eq:prox-def} entails
-\[\text{prox}_{\lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}\big(\text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta)\big) = \argmin_{x \in \R^p} k_\beta(x) = \{x^\star\}.\]
-Then, it follows that
-\begin{equation}
-    \label{eq:opt_gl}
-    0 \in \partial k_\beta(x^\star) = x^\star - \text{prox}_{\lambda_1 \norm{\cdot}_1}(\beta) + \partial\big( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2,l}\big)
-\end{equation}
-where $\partial k$ denotes the subgradient~\citep{rockafellar1974} of $k$.
-Now denoting
-\[h_\beta(x) = \frac{1}{2}\norm{x - \beta}_2^2 + \lambda_1 \norm{x}_1 + \lambda_2 \sum_{l=1}^L \norm{x}_{2,l} \] 
-for all $x \in \R^p$ such that
-\[\text{prox}_{\lambda_1 \norm{\cdot}_1 + \lambda_2 \sum_{l=1}^L \norm{\cdot}_{2,l}}(\beta) = \argmin_{x\in \R^p} h_\beta(x),\]
-it remains to be proved that $0 \in \partial h_\beta(x^\star)$.
-
-Then, it turns out that 
-\[ \partial h_\beta(x) = x - \beta + \lambda_1 \text{SGN}(x) + \partial( \lambda_2 \sum_{l=1}^L \norm{x}_{2, l})\]
-for all $x \in \R^p$, where SGN($x$) is the subgradient of the $\ell_1$-norm at $x$~\citep{bach2011optimization} defined in a component wise fashion as
-\[\text{SGN}(x) = 
-\left\{
-  \begin{array}{ll}
-    \{1\} &\text{if } x > 0, \\ 
-    $[-1, 1]$ &\text{if } x = 0, \\ 
-    \{-1\} &\text{if } x < 0. 
-  \end{array}
-\right. \]
-Denoting $y = \text{prox}_{\lambda_1\norm{\cdot}_1}(\beta)$, Lemma 1.4) in~\citet{yuan2011efficient} entails $\text{SGN}(y) \subseteq \text{SGN}(x^*)$. Hence,
-\[ y = \argmin_{u \in \R^p} \big\{\frac{1}{2} \norm{u - \beta}_2^2 + \lambda_1 \norm{u}_1\big\} 
-    = \beta - \lambda_1 \text{SGN}(y) \subseteq \beta - \lambda_1 \text{SGN}(x^*), \]
-which gives
-\[0 \in x^\star - y + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ) \subseteq x^\star- \beta + \lambda_1 \text{SGN}(x^\star) + \partial( \lambda_2 \sum_{l=1}^L \norm{x^\star}_{2, l} ).\]
-\end{proof}
-
-\section{Automatic cross-validation grid for $\xi$}
-\label{sec:auto-grid-cv}
-
-Concerning the cross-validation procedure for tuning 
-\[(\zeta_{1,0}, \zeta_{2,0}, \ldots, \zeta_{1,K-1}, \zeta_{2,K-1})^\top \in \R_+^{2K}, \]
-we use a randomized search with the C-index metric (see Section~\ref{sec:evaluation strategy}) where for all $k=0, \ldots, K-1$ and $j \in \{1, 2\}$, each $\zeta_{j,k}$ finds its candidates in the interval
-\[[\zeta_{j,k}^{\text{max}} \times 10^{-4}, \zeta_{j,k}^{\text{max}} ] \subset \R, \]
-with $\zeta_{j,k}^{\text{max}}$ the interval upper bound computed hereafter.
-
-Considering the convex minimization problem~\eqref{eq:minimization-xi_k} at a given step $w$,
-let us denote $\zeta_{1,k}^1 \leq \zeta_{1,k}^2 \leq \dots \leq \zeta_{1,k}^{\text{max}}$ the randomly chosen candidate values for $\zeta_{1,k}$, such that at $\zeta_{1,k}^{\text{max}}$, all coefficients $\hat \xi_{k, j}$ for all $j = 1, \dots, p$ are exactly zero.
-The KKT conditions~\citep{boyd2004convex} claim that
-\begin{center}
-  $\left\{
-      \begin{aligned}
-        &\frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_{k,j}} =  \zeta_{1,k} \big( (\eta-1)\ \text{sgn}(\hat \xi_{k,j}) - \eta \hat \xi_{k,j} \big)  \quad \ \ \ \forall j \in \hat \cA_k \\
-        &\left| \frac{\partial P_{n, k}^{(w)}(\hat \xi_k)}{\partial \xi_{k,j}} \right| < \zeta_{1,k} (1 - \eta) \quad \quad \quad \quad \quad \quad \quad \ \forall j \notin \hat \cA_k
-      \end{aligned}
-    \right.$,
-\end{center}
-where $\hat \cA_k = \big\{j = 1, \dots, p : \hat \xi_{k,j} \neq 0 \big\}$ is the active set of the $\hat \xi_k $ estimator, and for all $x \in~\R\setminus~\{0\},\ \text{sgn}(x) = \mathds{1}_{\{x>0\}} - \mathds{1}_{\{x<0\}} $.
-Then, using~\eqref{eq:grad-xi_k}, one obtains
-\begin{equation*}
-\hat \xi_{k, j} = 0 \ \Rightarrow \ \left| n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_{ij}  \right| < \zeta_{1,k} (1 - \eta)
-\end{equation*}
-for all $j = 1, \dots, p$.
-Hence, we choose the following upper bound for the randomly chosen candidate interval during the cross-validation procedure
-\begin{equation*}
-\zeta_{1,k}^{\text{max}} = \frac{1}{n(1-\eta)} \underset{j = 1, \dots, p}{\text{max}} \sum_{i=1}^n |x_{ij}|.
-\end{equation*}
-Similar strategy to automatically compute $\zeta_{2,k}^{\text{max}}$ is not easy since without additionnal hypothesis, the gradients~\ref{eq:grad-beta_k} and~\ref{eq:grad-Q} are not upper bounded by a quantity that depends only on the data and not on the current iteration parameters estimates.
-We then choose those interval upper bounds empirically.
-
-\section{Event time simulation}
-\label{sec:event-time-simu}
-
-Let us derive here the expression~\eqref{eq:time-generation} used to generate event times.
-Indeed, we can not refer to~\citet{austin2012generating} since the intensity~\eqref{eq:baseline} of the baseline is not correct in this paper.
-Note that other errors had already been pointed in~\citep{austin2013correction}, for instance the fact that there is no closed-form expression with a Weibull baseline.
-Suppose that
-\[\lambda(t) = \lambda_0(t) \exp (a + b t)\]
-with $(a, b) \in \R^2$ and $\lambda_0$ defined in~\eqref{eq:baseline}, then the cumulative hazard function writes
-\[H(t) = \kappa_1 \kappa_2 e^a \int_0^t \exp\{(\kappa_2 + b)s \} \dd s = \dfrac{\kappa_1 \kappa_2 e^a}{\kappa_2 + b} \big( \exp\{(\kappa_2 + b) t\} - 1 \big).\]
-Then, on has
-\[t = \dfrac{1}{\kappa_2 + b} \log \Big( 1 + \dfrac{H(t)(\kappa_2 + b)}{\kappa_1 \kappa_2 e^a} \Big),\]
-and one can generate event times according to
-\[ T \sim \dfrac{1}{\kappa_2 + b} \log \Big( 1 - \dfrac{(\kappa_2 + b)\log U}{\kappa_1 \kappa_2 e^a} \Big) \]
-with $U \sim \cU\big([0,1]\big)$.
-$\hfill\square$
-
-\section{Multivariate linear mixed model}
-\label{sec:MLMM}
-
-Let us derive here the explicit EM algorithm for the multivariate linear mixed model used to initialized the longitudinal parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ in the prox-QNEM algorithm in Section~\ref{sec:prox-QNEM}, acting as if there is no latent classes ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$).
-For the sake of simplicity, let us denote here 
-\[\theta = (\beta^\top, \text{vech}(D), \phi^\top)^\top \in \R^\vartheta\]
-the parameter vector to infer.
-The conditional distribution of $y_i|b_i$ then writes
-\[f(y_i|b_i ; \theta) = \exp \big\{( \Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) \big\},\]
-where $\Sigma_i$ is the diagonal matrix whose diagonal is $\Phi_i$. 
-The complete log-likelihood in this context writes
-\begin{align*}
-\ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}) \\
-&= \sum_{i=1}^n (\Sigma_iy_i)^\top M_{i} - c_\phi(M_{i}) + d_\phi(y_i) -\dfrac12 \big(r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i\big).
-\end{align*}
-
-\paragraph*{E-step.}
-Supposing that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$, we need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters, which is given by
-\[\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].\]
-Here, computing this quantity reduces to computing $\E_{\theta^{(w)}}[b_i | y_i]$ and $\E_{\theta^{(w)}}[b_ib_i^T | y_i]$ for $i = 1,\ldots,n$.
-The marginal distributions of $y_i$ and $b_i$ being both Gaussian, one has from Bayes Theorem
-\[ f(b_i|y_i; \theta^{(w)}) \propto \exp \big\{ -\dfrac12 (b_i - \mu_i^{(w)})^\top {\Omega_i^{(w)}}^{-1} (b_i - \mu_i^{(w)}) \big\}\]
-where
-\[\Omega_i^{(w)} = ({V_i}^\top \Sigma_i^{(w)}V_i + {D^{(w)}}^{-1})^{-1} \quad \text{ and } \quad \mu_i^{(w)} = \Omega_i^{(w)}{V_i}^\top \Sigma_i^{(w)}(y_i - U_i\beta^{(w)}).\]
-Then, on has
-\begin{center}
-  $\left\{
-    \begin{aligned}
-      &\E_{\theta^{(w)}}[b_i | y_i] = \mu_i^{(w)}, \\
-      &\E_{\theta^{(w)}}[b_ib_i^T | y_i] = \Omega_i^{(w)} + \mu_i^{(w)}{\mu_i^{(w)}}^\top.
-    \end{aligned}
-    \right.$
-\end{center}
-
-\paragraph*{M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) .\]
-The parameters updates are then naturally given in closed form by zeroing the gradient. One obtains
-\[\beta^{(w+1)} = \Big(\sum_{i=1}^n U_{i}^\top U_{i}\Big)^{-1}  \sum_{i=1}^n \big[ U_{i}^\top  y_i - U_{i} V_{i} \E_{\theta^{(w)}}[ b_i | y_i]) \big],\]
-\begin{align*}
-\phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1}  \sum_{i=1}^n \Big[(&y_i^l - U_{il}\beta_l^{(w+1)})^\top (y_i^l - U_{il}\beta_l^{(w+1)} - 2V_{il}\E_{\theta^{(w)}}[b_i^l | y_i^l])) \\
-&+\Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | y_i^l]\big) \Big]
-\end{align*}
-and \[D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i {b_i}^\top | y_i].\]
-
-\section{Technical documentation of the \texttt{lights} package}
-\label{sec:package}
-
-\subsection{\texttt{MLMM class}}
-\label{sec:MLMM class}
-
-This class implements an EM algorithm for fitting a multivariate linear mixed model used to initialize the QNEM algorithm.
-Let us introduce the list $\Omega^{(w)} = [\Omega_1^{(w)}, \ldots, \Omega_n^{(w)}]$, the matrices $\mu = [\mu_1, \ldots, \mu_n] \in \R^{r \times n}$, $U^l = [{U_{1l}}^\top \cdots {U_{nl}}^\top]^\top \in \R^{n_l \times q_l}$, $U = [U_1^\top \cdots U_n^\top]^\top \in \R^{\cN \times q}$,
-\[ V^l = \begin{bmatrix}
-    V_{1l} & \cdots & 0\\
-    \vdots &  \ddots & \vdots \\
-    0 & \cdots & V_{nl}
-\end{bmatrix},
-\quad
-V = \begin{bmatrix}
-  V_{1} & \cdots & 0\\
-  \vdots &  \ddots & \vdots \\
-  0 & \cdots & V_{n}
-\end{bmatrix}
-\quad \text{ and } \quad
-{\Omega^l}^{(w)} = \begin{bmatrix}
-    {\Omega_{1}^l}^{(w)} & \cdots & 0\\
-    \vdots &  \ddots & \vdots \\
-    0 & \cdots & {\Omega_{n}^l}^{(w)}
-\end{bmatrix}\]
-that belong respectively in $\R^{n_l \times nr_l}$, $\R^{\cN \times nr}$ and $\R^{nr_l \times nr_l}$, as well as the vectors $\tilde \mu^{(w)} = ({\mu_1^{(w)}}^\top \cdots {\mu_n^\top)^{(w)}}^\top \in \R^{nr}$, $(\tilde \mu^l)^{(w)} = \big({(\mu_1^l)^{(w)}}^\top \cdots {(\mu_n^l)^{(w)}}^\top \big)^\top \in \R^{nr_l}$, $y^l = ({y_1^l}^\top  \cdots {y_n^l}^\top)^\top \in \R^{n_l}$
-with $n_l = \sum_{i=1}^{n} n_i^l$ and $y = (y_1^\top \cdots y_n^\top)^\top \in \R^\cN$ with $\cN = \sum_{i=1}^{n} n_i$.
-The $\beta$ update then rewrites
-\[\beta^{(w+1)} = ({U}^\top U)^{-1}{U}^\top(y-V \tilde \mu^{(w)}).\]
-For the $D$ update, one has
-\[D^{(w+1)} = n^{-1} \big( \text{sum}(\Omega^{(w)}) + \mu^{(w)}{\mu^{(w)}}^\top\big).\]
-And finally for the $\phi$ update, one has
-\begin{align*}
-\phi_l^{(w+1)} = n_l^{-1} \big[&(y^l - U^l{\beta_l^{(w+1)})}^\top \big(y^l - U^l\beta_l^{(w+1)} - 2 V^l (\tilde\mu^l)^{(w)}\big) \\
-&+ \text{Tr}\big\{{V^l}^\top V^l \big({\Omega^l}^{(w)} + (\tilde\mu^l)^{(w)} {(\tilde\mu^l)^{(w)}}^\top\big)\big\} \big].
-\end{align*}
-
-\subsection{\texttt{prox-QNEM class}}
-\label{sec:prox-QNEM class}
-
-Let us introduce the following useful notations
-\begin{equation*}
-  \left\{
-    \begin{array}{ll}
-    \tilde g^1&: b_i \mapsto b_i,\\
-    \tilde g^2&: b_i \mapsto b_ib_i^\top.
-    \end{array}
-  \right. 
-\end{equation*}
+\section{section 1}
 

### 2022-04-13 slides/JDS2022/short paper/paper.tex
Update the short version of lights paper #1

@@
 
-\section{section 1}
+\section{Introduction}
+
+\paragraph{General framework.}
+
+The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model.
+Let us consider the usual survival analysis framework.
+Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
+We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
+\begin{equation*}
+T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}
+\end{equation*}
+respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.
+
+Let $X$ denotes the $p$-dimensional vector of time-independent features and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
+
+\paragraph{Heterogeneity of the population.}
+
+Assume the population that can be devided into a finite number K of latent homogeneous subgroups. Let us denote
+\begin{equation}
+  \label{eq:pi}
+  \pi_{\xi_k}(x) = \P[G=k|X=x]
+\end{equation}
+the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by
+\begin{equation}
+  \label{eq:pi-softmax} 
+  \pi_{\xi_k}(x) = \dfrac{e^{x^\top\xi_k}}{\sum_{k=0}^{K-1}e^{x^\top\xi_k}}
+\end{equation}
+where $\xi_k \in \R^p$ denotes a vector of time-independent parameters.
+
+\section{Method}
+\label{sec:Method}
+
+In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.
+
+\subsection{Class-specific marker trajectories}
+
+We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has
+\begin{equation}
+  \label{eq:link-function}
+  h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)
+ \end{equation}
+where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that
+\[ m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l \]
+where $u^l(t) \in \R^{q_l}$  and $v^l(t) \in \R^{r_l}$ are row vectors of (possibly) time-varying features  with $r_l \leq q_l$.
+The random effects component is assumed to be followed a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is
+\begin{equation}
+   \label{eq:random-effect-law}
+   b^l \sim \cN(0, D_{ll})
+ \end{equation} 
+with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. And we denote the global variance-covariance matrix
+\[ D = 
+\begin{bmatrix}
+  D_{11} & \cdots & D_{1L}\\
+  \vdots &  \ddots & \vdots \\
+  D_{1L}^\top & \cdots & D_{LL}
+\end{bmatrix}
+\]
+
+In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.
+
+\subsection{Class-specific risk of event}
+
+To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
+\begin{equation}
+  \label{eq:intensity-model}
+  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
+\end{equation}
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
+\[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
+the history of the true longitudinal process up to time $t$.
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.
+
+Let us finally denote
+\[\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}, \] so that
+\begin{equation}
+  \label{eq:def-asso-features}
+  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\},
+\end{equation}
+where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.
+
+\subsection{Generalization of SREMs and JLCMs}
+
+Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way.
+Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
+\begin{table}[htb]
+\centering
+\begin{tabular}{cccc}
+\toprule
+Description & $\Psi_a^l(t, \beta_k^l, b^l)$  & Reference \\
+\midrule
+Linear predictor & $m_k^l(t)$ & \citet{chi2006joint} \\ [.15cm]
+Random effects & $b^l$  & \citet{hatfield2011joint} \\
+Time-dependent slope & $\dfrac{\dd}{\dd t} m_k^l(t)$ & \citet{rizopoulos2011bayesian} \\ [.3cm]
+Cumulative effect & $\int_0^t m_k^l(s) \dd s$ & \citet{andrinopoulou2017combined} \\ [.1cm]
+ \bottomrule
+\end{tabular}
+\caption[]{Examples of classical shared associations in SREMs.}
+\label{table:shared_associations}
+\end{table}
+
+\subsection{Likelihood}
+
+Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
+\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
+where 
+\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
+for all $l=1, \ldots, L$. 
+For the $i$-th subject, let us denote 
+\[
+\left\{
+    \begin{array}{ll}
+        y_i &= ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i},\\
+        b_i &= ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r,
+    \end{array}
+\right.
+\]
+with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
+\[ U_i = 
+\begin{bmatrix}
+  U_{i1} & \cdots & 0\\
+  \vdots &  \ddots & \vdots \\
+  0 & \cdots & U_{iL}
+\end{bmatrix} 
+\in \R^{n_i \times q}
+\qquad \text{and} \qquad
+V_i = 
+\begin{bmatrix}
+  V_{i1} & \cdots & 0\\
+  \vdots &  \ddots & \vdots \\
+  0 & \cdots & V_{iL}
+\end{bmatrix}
+\in \R^{n_i \times r}
+\]
+with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes
+\[
+\left\{
+    \begin{array}{ll}
+        U_{il} &= \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l},\\
+        V_{il} &= \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}.
+    \end{array}
+\right.
+\]
+We aslo denote 
+\[ \beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q \] 
+the fixed effect parameter of linear mixed effect model of group $k$ and 
+\[M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}. \]
+the  all linear predictor vector for subject $i$ given the fact that he belongs to group $k$.
+For the collection of the $\vartheta \in \N_+$ unknown parameters to estimate, we denote 
+\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta .\]
+
+The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form
+\begin{equation}
+    \label{eq:exp-family-density}
+    f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\},
+\end{equation}
+with 
+\begin{equation}
+  \label{eq:def-Phi_i} 
+  \Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}
+\end{equation} 
+and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
+
+The likelihood of survival model can be written as
+\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
+Then, the log-likelihood of joint model in type of JLCMs based on these 2 above likelihoods is given by
+\begin{equation}
+    \label{eq:log-lik}
+    \ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
+\end{equation}
+\
+\section{Inference}
+\label{sec:inference}
+
+In this section, we describe the procedure for estimating the parameters of the lights model.
+Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
+
+\subsection{Penalized objective}
+\label{sec:penalized-obj}
+
+In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
+\begin{equation}
+  \label{eq:pen-log-lik}
+    \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
+\end{equation}
+with the elastic net penalty
+\[ \norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2 \]
+and the sparse group lasso penalty
+\[ \norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l} .\]
+Hence, the resulting optimization problem is written
+\begin{equation}
+  \label{eq:optim-pb}
+   \hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).
+ \end{equation}
+
+\subsection{A Proximal Quasi-Newton EM}
+\label{sec:prox-QNEM}
+
+In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm. We first need to compute the negative completed log-likelihood by
+\begin{align*}
+  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) \\ 
+  &= - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) + \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) \\ 
+  & \quad + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) - \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s \\
+  & \quad + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
+\end{align*}
+Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.
+
+\paragraph*{E-step.}
+
+We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
+\begin{equation*}
+  \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
+\end{equation*}
+Then, the previous expression requires to compute expectations of the form
+\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
+for different functions $g$, where we denote 
+\begin{equation}
+  \label{eq:pi_ik-def}
+  \pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i] 
+\end{equation}
+the posterior probability of the latent class membership using parameters $\theta^{(w)}$.
+
+\paragraph*{Proximal Quasi-Newton M-step.}
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+First, the update of $D^{(w)}$ is naturally given in closed-form by
+\begin{equation}
+  \label{eq:update_D}
+  D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
+\end{equation}
+Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
+\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
+based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\begin{equation}
+  \label{eq:minimization-xi_k}
+  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
+\end{equation}
+We then choose to solve using the L-BFGS-B algorithm  which belongs to the class of quasi-Newton optimization routines.
+The update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$ is obtained in closed-form given by
+\begin{equation}
+  \label{eq:update_beta}
+  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
+\end{equation}
+Now, we use proximal gradient (ISTA) for the $\gamma_k^{(w+1)}$ update, based on Lemma 1 that states $\text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}$.
+The closed-form update of $\lambda_0^{(w)}$ is given by
+\begin{equation}
+  \label{eq:update_lambda_0}
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} \ind{\{t_i \geq t\}} },
+\end{equation}
+for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model. 
+Finally, the update of $\phi^{(w)}$ is obtained in closed-form by
+\begin{align}
+  \label{eq:update_phi}
+  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
+  & - 2V_{il} \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
+\end{align}
+
+\section{Performance evaluation}
+
+\paragraph{Real-time prediction paradigm.}
+
+Once the learning phase is achieved for the model on a training set (so that  $\hat\theta$ is obtained, we want to assess real-time risk prediction performances on a test set.
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction.
+
+\paragraph{Predictive marker.}
+
+Posterior risk classification for subject $i$ and latent class $k$ is chosen to be marker rule of the lights 
+\begin{equation}
+  \label{eq:marker-def}
+  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
+\end{equation}
+with 
+\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
+
+\subsection{The C-index metric}
+\label{sec:Metrics}
+
+We use concordance index (C-index) to evaluate risk prediction performances
+\begin{equation*}
+  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
+\end{equation*}
+with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
 

### 2022-04-19 slides/JDS2022/short paper/paper.tex
JDS short version

--- 2022-04-12 slides/JDS2022/short paper/paper.tex
@@
 
-\section{section 1}
+\section{Introduction}
 
+\paragraph{General framework.}
+
+The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model.
+Let us consider the usual survival analysis framework.
+Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
+We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as $T = T^\star \wedge C \quad \text{and} \quad \Delta = \ind{\{T^\star \leq C\}}$
+respectively, where $a \wedge b$ denotes the minimum between two numbers $a$ and $b$, and $\ind{\{\cdot\}}$ the indicator function taking the value $1$ if the condition in $\{\cdot\}$ is satisfied and $0$ otherwise.
+
+Let $X$ denotes the $p$-dimensional vector of time-independent features and let  $Y(t) = \big(Y^1(t), \ldots, Y^L(t) \big)^\top \in \R^L$ denote the value of the $L$-dimensional longitudinal outcome at time point $t \geq 0$, with $L \in \N_+$.
+
+\paragraph{Heterogeneity of the population.}
+
+Assume the population that can be devided into a finite number K of latent homogeneous subgroups.
+Let us denote $\pi_{\xi_k}(x) = \P[G=k|X=x]$ the latent class membership probability given time-independent features $x \in \R^p$, and consider a softmax link function given by $\pi_{\xi_k}(x) = e^{x^\top\xi_k} / \sum_{k=0}^{K-1}e^{x^\top\xi_k}$
+where $\xi_k \in \R^p$ denotes a vector of time-independent parameters.
+
+\section{Method}
+\label{sec:Method}
+
+In this section, we descibe the longitudinal and time-to-event submodels, as well as the required hypothesis in order to write a likelihood and draw inference for the lights model.
+
+\subsection{Class-specific marker trajectories}
+
+We suppose a class-specific marker trajectory and a generalized linear mixed model for each longitudinal marker given latent class $G$, so that for the $l$-th outcome at time $t \geq 0$ one has $h_l\big(\E[Y^l(t)|b^l, G=k]\big) = m_k^l(t)$ where $h_l$ denotes a known one-to-one link function, and $m_k^l$ the linear predictor such that $m_k^l(t) = u^l(t)^\top\beta_k^l + v^l(t)^\top b^l$ where $u^l(t) \in \R^{q_l}$  and $v^l(t) \in \R^{r_l}$ are row vectors of (possibly) time-varying features  with $r_l \leq q_l$.
+The random effects component is assumed to be followed a zero-mean multivariate normal distribution~\citep{hickey2016joint}, that is $b^l \sim \cN(0, D_{ll})$
+with $D_{ll} \in \R^{r_l \times r_l}$ the unstructured variance-covariance matrix. And we denote $D$ the global variance-covariance matrix.
+
+In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.
+
+\subsection{Class-specific risk of event}
+
+To quantify the effect of the longitudinal outcomes on the risk for an event, we use a Cox~\citep{Cox1972JRSS} relative risk model of the form
+\begin{equation*}
+  \label{eq:intensity-model}
+  \lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \Big\{\sum_{l=1}^L \sum_{a=1}^\cA {\gamma_{k,a}^l} \Psi_a^l(t) \Big\},
+\end{equation*}
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote $\cY^l(t) = \{Y^l(u), 0 \leq u < t\}$ and $\cY(t) = \bigcup_{l=1}^L \cY^l(t)$ the history of the true longitudinal process up to time $t$.
+For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.
+
+Let us finally denote $\gamma_k= (\gamma_{k,1}^1, \ldots, \gamma_{k,\cA}^1, \ldots, \gamma_{k,1}^L, \ldots, \gamma_{k,\cA}^L )^\top \in \R^{L\cA}$, so that one has $\lambda(t|\cY(t), G = k) = \lambda_0(t) \exp \big\{\gamma_k^\top \Psi(t) \big\}$, where $\Psi(t) \in \R^{L\cA}$ denotes the representation features vector.
+
+\subsection{Generalization of SREMs and JLCMs}
+
+Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way.
+Our proposition differs from standard SREMs in two respects: $(i)$ representation vector $\Psi(t)$ does not depend on the modeling assumptions in the longitudinal submodel and $(ii)$ it allows to choose multiple high-dimensional representation mappings that characterize time series and concatenate them, since we perform features selection through the regularization strategy described in Section~\ref{sec:penalized-obj}.
+Point $(i)$ is key since for instance the $\beta_k^l$ dependence in $\Psi_a^l$ would lead to complicated updates, while the $b^l$ dependence would lead to untracktable integrals that are common in SREMs and require approximation methods such as Monte Carlo~\citep{hickey2018joinerml}, being computationally intensive and not scalable in a high-dimensional context.
+For point $(ii)$ in practice, we use the \texttt{Python} library \texttt{tsfresh}~\citep{christ2018time} and include many extracted features, such as absolute energy of the time series, statistics on autocorrelation, or Fourier and wavelet basis projections, to name but a few.
+
+\subsection{Likelihood}
+
+Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
+\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
+where $y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l}$ with $y_{ij}^l=Y_i^l(t_{ij}^l)$ for all $l=1, \ldots, L$. 
+For the $i$-th subject, let us denote $y_i = ({y_i^1}^\top \cdots {y_i^L}^\top)^\top \in \R^{n_i}$ and $b_i = ({b_i^1}^\top \cdots {b_i^L}^\top)^\top \in \R^r$, with $n_i = \sum_{l=1}^L n_i^l$ and $r = \sum_{l=1}^L r_l$ the total number of longitudinal measurements (for subject $i$) and the total dimension of the random effects respectively, as well as the following design matrices
+\[ U_i = 
+\begin{bmatrix}
+  U_{i1} & \cdots & 0\\
+  \vdots &  \ddots & \vdots \\
+  0 & \cdots & U_{iL}
+\end{bmatrix} 
+\in \R^{n_i \times q}
+\qquad \text{and} \qquad
+V_i = 
+\begin{bmatrix}
+  V_{i1} & \cdots & 0\\
+  \vdots &  \ddots & \vdots \\
+  0 & \cdots & V_{iL}
+\end{bmatrix}
+\in \R^{n_i \times r}
+\]
+with $q = \sum_{l=1}^L q_l$ and where for all $l=1, \ldots, L$, one writes $U_{il} = \big(u_i^l(t_{i1}^l)^\top \cdots u_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times q_l}$ and $V_{il} = \big(v_i^l(t_{i1}^l)^\top \cdots v_i^l(t_{in_i^l}^l)^\top\big)^\top \in \R^{n_i^l \times r_l}$.
+We also denote $\beta_k = ({\beta_k^1}^\top \cdots {\beta_k^L}^\top)^\top \in \R^q$ the fixed effect parameter of linear mixed effect model of group $k$ and $M_{ik} = U_i\beta_k + V_ib_i \in \R^{n_i}$ the linear predictor vector for subject $i$ given the fact that he belongs to group $k$.
+For the collection of the $\vartheta \in \N_+$ unknown parameters to estimate, we denote 
+\[ \theta = \big(\xi_0^\top \cdots \xi_{K-1}^\top, \beta_0^\top \cdots \beta_{K-1}^\top, \phi^\top, \text{vech}(D), \lambda_0^\top, \gamma_0^\top \cdots \gamma_{K-1}^\top\big)^\top \in \R^\vartheta .\]
+
+The conditional distribution of $y_i|b_i, G_i=k$ is then assumed to be from a distribution with a density of the form $f(y_i|b_i, G_i=k) = \exp \big\{(y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \big\}$,
+with $\Phi_i = (\phi_1^{-1} {\textbf{1}_{n_i^1}}^\top \cdots \phi_L^{-1} {\textbf{1}_{n_i^L}}^\top)^\top \in \R^{n_i}$ and $\phi = (\phi_1, \ldots, \phi_L)^\top \in \R^L$.
+The likelihood of survival model can be written as
+\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
+Then, the log-likelihood of joint model in type of JLCMs based on these 2 above likelihoods is given by
+\begin{equation*}
+    \ell_n(\theta) = \ell_n(\theta ; \cD_n) = n^{-1} \sum_{i=1}^n \log \sum_{k=0}^{K-1} \pi_{\xi_k}(x_i) f_\theta(t_i, \delta_i| G_i = k) f_\theta(y_i | G_i = k),
+\end{equation*}
+\
+\section{Inference}
+\label{sec:inference}
+
+In this section, we describe the procedure for estimating the parameters of the lights model.
+Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
+
+\subsection{Penalized objective}
+\label{sec:penalized-obj}
+
+In order to avoid overfitting and improve the prediction power of our method, we propose to minimize the penalized objective
+\begin{equation*}
+    \ell_n^\text{pen}(\theta) = - \ell_n(\theta) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}
+\end{equation*}
+with the elastic net penalty $\norm{z}_{\text{en}, \eta} = (1-\eta)\norm{z}_1 + \dfrac\eta2 \norm{z}_2^2$ and the sparse group lasso penalty $\norm{z}_{\text{sg} l_1, \tilde{\eta}} = (1-\tilde{\eta})\norm{z}_1 + \tilde{\eta} \sum_{l=1}^L\norm{z}_{2,l}$.
+Hence, the resulting optimization problem is written $\hat \theta \in \argmin_{\theta \in \R^\vartheta} \ell_n^\text{pen}(\theta).$
+
+\subsection{A Proximal Quasi-Newton EM}
+\label{sec:prox-QNEM}
+
+In order to derive an algorithm for this objective, we introduce a so-called prox-QNEM algorithm. We first need to compute the negative completed log-likelihood by
+\begin{align*}
+  \ell_n^\text{comp}(\theta) &= \ell_n^\text{comp}(\theta ; \cD_n, \textbf{\textit{b}}, \textbf{\textit{G}}) = - n^{-1} \sum_{i=1}^n -\dfrac12 (r \log 2\pi + \log|D| + b_i^\top D^{-1}b_i) \\ 
+  &+ \sum_{k=0}^{K-1} \ind{\{G_i=k\}} \Big[ \log \pi_{\xi_k}(x_i) + \delta_i \Big(\log \lambda_0(t_i) + \gamma_k^\top \Psi_i(t_i) \Big) \\ 
+  &- \int_0^{t_i} \lambda_0(s) \exp \big\{ \gamma_k^\top \Psi_i(s) \big\} \dd s  + (y_i \odot \Phi_i)^\top M_{ik} - c_\phi(M_{ik}) + d_\phi(y_i) \Big].
+\end{align*}
+Suppose that we are at step $w + 1$ of the algorithm, with current iterate denoted $\theta^{(w)}$.
+
+\paragraph*{E-step.}
+
+We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by $\cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n]$.
+Then, the previous expression requires to compute expectations of the form
+\[ \E_{\theta^{(w)}}[ g(b_i, G_i) | t_i, \delta_i, y_i] = \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \int_{\R^r} g(b_i, G_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i \]
+for different functions $g$, where we denote $\pi_{ik}^{\theta^{(w)}} = \P_{\theta^{(w)}}[G_i = k | t_i, \delta_i, y_i]$ the posterior probability of the latent class membership using parameters $\theta^{(w)}$.
+
+\paragraph*{Proximal Quasi-Newton M-step.}
+Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+First, the update of $D^{(w)}$ is given in closed-form by $D^{(w+1)} = n^{-1} \sum_{i=1}^n \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i]$.
+Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$.
+We denote $P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i)$ based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$.
+The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
+\begin{equation*}
+  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
+\end{equation*}
+We then choose to solve using the L-BFGS-B algorithm  which belongs to the class of quasi-Newton optimization routines.
+The update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$ is obtained in closed-form.
+Then, we use proximal gradient (ISTA) for the $\gamma_k^{(w+1)}$ update, based on Lemma 1 that states $\text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}$.
+The closed-form update of $\lambda_0^{(w)}$ is given by
+\begin{equation*}
+  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i(t)\big\} \ind{\{t_i \geq t\}} },
+\end{equation*}
+for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model. 
+Finally, the update of $\phi^{(w)}$ is obtained in closed-form.
+
+\section{Performance evaluation}
+
+\paragraph{Real-time prediction paradigm.}
+
+Once the learning phase is achieved for the model on a training set (so that  $\hat\theta$ is obtained, we want to assess real-time risk prediction performances on a test set.
+Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction.
+
+\paragraph{Predictive marker.}
+
+Posterior risk classification for subject $i$ and latent class $k$ is chosen to be marker rule of the lights 
+\begin{equation}
+  \label{eq:marker-def}
+  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
+\end{equation}
+with 
+\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
+
+\subsection{The C-index metric}
+\label{sec:Metrics}
+
+We use concordance index (C-index) to evaluate risk prediction performances
+\begin{equation*}
+  \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
+\end{equation*}
+with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
+
+\small
 \bibliography{biblio}

### 2022-04-19 slides/JDS2022/short paper/paper.tex
Add the figures

--- 2022-04-13 slides/JDS2022/short paper/paper.tex
@@
 \vspace{-.5cm}
-
 \begin{abstract}
@@
 Thus, we propose a powerfull tool with the ability of automatically determining significant prognostic longitudinal features, which is of increasing importance in many areas: for instance personalized medicine, or churn prediction in a customer profile and activity monitoring setting, to name but a few.\\
-
 \noindent
@@
 
-The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model.
-Let us consider the usual survival analysis framework.
-Following~\citet{andersen2012statistical}, let non-negative random variables $T^\star$ and $C$ stand for the times of the event of interest and censoring times respectively.
+With classical setting of survival analysis where we denote $T^\star$ and $C$ are the times of the event of interest and censoring times respectively.
 We then denote $T$ the right-censored time and $\Delta$ the censoring indicator, defined as 
@@
 
-In this section, we describe the procedure for estimating the parameters of the lights model.
-Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
-
 \subsection{Penalized objective}
@@
 \end{equation}
-Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
-\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
-based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
+For the update of $\xi_k^{(w)}$, we choose the L-BFGS-B algorithm to solve the following convex minimization problem
 \begin{equation}
@@
 \end{equation}
-We then choose to solve using the L-BFGS-B algorithm  which belongs to the class of quasi-Newton optimization routines.
+where
+\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) .\]
 The update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$ is obtained in closed-form given by
@@
 Now, we use proximal gradient (ISTA) for the $\gamma_k^{(w+1)}$ update, based on Lemma 1 that states $\text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}$.
-The closed-form update of $\lambda_0^{(w)}$ is given by
+
+\noindent The closed-form update of $\lambda_0^{(w)}$ is given by
 \begin{equation}
@@
 for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.
-Finally, the update of $\phi^{(w)}$ is obtained in closed-form by
-\begin{align}
-  \label{eq:update_phi}
-  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
-  & - 2V_{il} \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il} \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
-\end{align}
 
@@
 \paragraph{Predictive marker.}
-
 Posterior risk classification for subject $i$ and latent class $k$ is chosen to be marker rule of the lights 
@@
   \label{eq:marker-def}
-  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
+  \P_{\hat \theta}[G_i=k | T_i > t^{max}_i, \tilde y_i] = \dfrac{\pi_{\hat \xi_k}(x_i) f_{\hat \theta}(t^{max}_i, \delta_i=0, \tilde y_i | G_i=k)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) f_{\hat \theta}(t^{max}_i, \delta_i=0, \tilde y_i | G_i=k)}
 \end{equation}
-with 
-\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
 
@@
 \label{sec:Metrics}
-
 We use concordance index (C-index) to evaluate risk prediction performances
@@
 
+\subsection{Simulation}
+The below figures are obtained on simulated data with 2 latent groups. They are  learning curve over time of log-likelihood function, estimation of parameters compare to the true ones.
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=.9\textwidth]{./figures/obj_vcg}
+\includegraphics[width=.45\textwidth]{./figures/xi_estimation}
+\includegraphics[width=.4\textwidth]{./figures/baseline_estimation}
+\includegraphics[width=.9\textwidth]{./figures/beta_estimation}
+\includegraphics[width=.9\textwidth]{./figures/gamma_estimation}
+\end{figure}
+
 \bibliography{biblio}

### 2022-05-10 manuscript/paper.tex
add theorem, update real-time prediction paradigme

@@
 \newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
 
@@
 Consider an independent and identically distributed (i.i.d.) cohort of $n$ subjects
-\[ \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\} \]
+\begin{equation}
+  \label{eq:dataset-def}
+  \cD_n = \big\{ (x_1, y_1^1, \ldots, y_1^L, t_1, \delta_1), \ldots, (x_n, y_n^1, \ldots, y_n^L, t_n, \delta_n) \big\}
+\end{equation}
 where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
@@
 So in $\theta$, $\lambda_0$ represents the vector of these baseline hazard quantities, with a dimension equal to $J$.
-To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$, corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
+To write the log-likelihood $\ell_n(\theta)$ (rescaled by $n^{-1}$)  for samples in $\cD_n$ (defined in~\eqref{eq:dataset-def}), corresponding to the joint distribution of the time-to-event and longitudinal outcomes, let us make the following hypothesis.
 \begin{assumption}
@@
 where 
-\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}. \]
-As expected, one has a JLCMs like log-likelihood.
+\[f_\theta(t_i, \delta_i| G_i = k) = \big[\lambda(t_i|\cY_i(t_i), G_i = k)\big]^{\delta_i} S_k(t_i) , \]
+and
+\begin{equation}
+  \label{eq:survival-k-def}
+  S_k(t_i) = \exp \Big\{-\int_0^{t_i} \lambda(s|\cY_i(s), G_i = k) \dd s \Big\}
+\end{equation}
+is the survival probability of subject $i$ given that it belongs to latent class $k$.
+As expected,~\eqref{eq:log-lik} is a JLCMs like log-likelihood.
 
@@
 In this section, we describe the procedure for estimating the parameters of the lights model.
-Let us first present the considered penalized objective, and then focus on the algorithm proposed for inference.
+Let us first present the considered penalized objective, then focus on the algorithm proposed for inference, and on its convergence properties.
 
@@
 We need to compute the expected negative log-likelihood of the complete data conditional on the observed data and the current estimate of the parameters given by 
-\begin{equation*}
+\begin{equation}
+  \label{eq:Q-def}
   \cQ_n(\theta, \theta^{(w)}) = \E_{\theta^{(w)}}[\ell_n^\text{comp}(\theta) | \cD_n].
-\end{equation*}
+\end{equation}
 Given Assumption~\ref{assumption1}, the previous expression requires to compute expectations of the form
@@
 \paragraph*{Proximal Quasi-Newton M-step.}
-Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
+Let us denote
+\begin{equation}
+  \label{eq:M-step-obj}
+  \cQ_n^\text{pen}(\theta, \theta^{(w)}) = \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
+\end{equation}
+Here, we need to compute 
+\[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n^\text{pen}(\theta, \theta^{(w)}).\]
 We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
@@
 Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting prox-QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}. 
-
-% \vspace{.5cm}
-
 \begin{algorithm}[!htb]
@@
 \label{alg:prox-QNEM}
-\KwInput{Training data $\cD_n$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
+\KwInput{Training data $\cD_n$~\eqref{eq:dataset-def}; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
 \KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
@@
 
+\subsection{Convergence to a stationary point}
+
+Let us address here convergence properties of the prox-QNEM algorithm described in the previous section.
+Since in~\eqref{eq:M-step-obj} we only minimize $\cQ_n^\text{pen}(\theta, \theta^{(w)})$ and not directly $\cQ_n(\theta, \theta^{(w)})$ defined in~\eqref{eq:Q-def} as in a classical EM~\citep{wu1983convergence}, we are in a generalized EM (GEM) setting~\citep{dempster1977maximum}.
+For such algorithm, one has the descent property saying that the objective function~\eqref{eq:pen-log-lik} decreases at each iteration, namely
+\[\ell^{\text{pen}}_n(\theta^{(w+1)}) \leq \ell^{\text{pen}}_n(\theta^{(w)}),\]
+which can be observed in Figure ?.
+Now, let us make a classical hypothesis.
+\begin{assumption}
+  \label{hyp:bounded}
+  Suppose that $\ell^{\text{pen}}_n$ is bounded for all $\theta$.
+\end{assumption}
+Assumption~\ref{hyp:bounded} implies that $w \mapsto \ell^{\text{pen}}_n(\theta^{(w)})$ decreases monotically to a finite limit.
+Then, one has the following convexity property with the Gaussian case used in practice.
+\begin{lemma}
+\label{lem:strictly-convex}
+$\cQ_n^{\text{pen}}(\theta, \theta^{(w)})$ is continuous in $\theta$ and $\theta^{(w)}$, and for any fixed $\theta^{(w)}$, it is a convex function in $\theta$, and strictly convex in each coordinate of $\theta$.
+\end{lemma}
+Hence, one can state the following theorem about the convergence of the prox-QNEM algorithm to a stationary point, being here a local minimum.
+\begin{theorem}
+\label{th:cvg}
+Under Assumption~\ref{hyp:bounded}, every cluster point $\bar{\theta}$ of the sequence $\{ \theta^{(w)}\}_{w \geq 1}$ generated by the prox-QNEM algorithm is a stationary point of the objective function~\eqref{eq:pen-log-lik}, that is
+\[\forall \varepsilon > 0, V_\varepsilon(\bar{\theta}) \cap \{ \theta^{(w)}\}_{w \geq 1} \setminus \{\bar{\theta}\} \neq \varnothing \implies \forall \tilde \theta \in \R^\vartheta, \underset{u \rightarrow 0}{\lim} \frac{\ell^{\text{pen}}_n(\bar{\theta} + \tilde \theta u) - \ell^{\text{pen}}_n(\bar{\theta})}{u} \geq 0 ,\]
+with $V_\varepsilon(\bar{\theta})$ the epsilon-neighbourhood of $\bar{\theta}$.
+\end{theorem}
+The proof of Lemma~\ref{lem:strictly-convex} is straightforward, and we follow the proof of Theorem 1 in~\citet{bussy2019c} to prove Theorem~\ref{th:cvg}.
+
 \subsection{The case $K=2$}
@@
 
-The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with predictive marker defined in Section~\ref{sec:evaluation strategy} and the C-index score defined in Section~\ref{sec:Metrics}.
-In practice, we use the \texttt{Python} package \texttt{Hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
+The hyper-parameters $(\zeta_1, \zeta_2)^\top \in \R_+^2$ are then tuned using a Bayesian optimization method~\citep{snoek2012practical}, with the C-index score and the predictive marker both defined in Section~\ref{sec:evaluation strategy}.
+In practice, we use the \texttt{Python} package \texttt{hyperopt}~\citep{bergstra2013hyperopt} and the Tree-structured Parzen Estimator~\citep{bergstra2011algorithms}, which is a sequential model-based optimization (SMBO) approach.
 SMBO methods sequentially construct models to approximate the performance of hyperparameters based on historical measurements, and then subsequently choose new hyperparameters to test based on this model.
@@
 
-In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances.
-We then introduce the appropriate metric used to this end: the C-index ; and finally, the considered models for performance comparisons.
-
-\subsection{Evaluation strategy: the real-time prediction paradigm}
+In this section, we first present the proposed evaluation strategy to assess real-time prognostic prediction performances. We then introduce the considered models for performance comparisons.
+
+\subsection{Evaluation strategy}
 \label{sec:evaluation strategy}
 
-When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:Metrics}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
+When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:evaluation strategy}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
 Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}.
@@
 Once the learning phase is achieved for the model on a training set (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), we want to assess real-time risk prediction performances on a test set.
-
-Denoting $t^{max}_i$ the time for subject $i$ when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction --
+Denoting $t^{max}_i$ the time for subject $i$ in the test set when one wants to perform the risk prediction -- so in practice, the time up to which one has data measurements for $Y_i$, say the ``present'' time for prediction -- one wants then to predict a prognostic marker for subject $i$ using all data available up to $t^{max}_i$, but without using the supervision labels $(t_i, \delta_i)$.
 
@@
 
-The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
-
-posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:pi_ik-def}).
-Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
-
-But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is for risk prediction, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow
-
- a natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ would rather be 
+The question is now to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
+Posterior risk classification for subject $i$ and latent class $k$ can be made through $\pi_{ik}^{\hat\theta}$ (see~\eqref{eq:pi_ik-def}), and such posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.
+
+But in the real-time prediction paradigm, it makes no sense from a practical point of view to use this marker rule as is, since the latter requires to know the survival labels $(t_i, \delta_i)$, being intrinsically unknow in the paradigm.
+A natural and appropriate marker rule of the lights model for subject $i$ being in latent class $k$ in our context would rather be
 \begin{equation}
   \label{eq:marker-def}
-  \hat \cR_{ik} = \dfrac{\pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta)},
-\end{equation}
-with 
-\[ \hat f(t^{max}_i, y_i | G_i=k ; \hat \theta) = \dfrac{1}{2N} \sum_{\breve{b} \in \hat S} f(t^{max}_i, 0, y_i | \breve{b}, G_i = k ; \hat \theta), \]
-
-dire ici qu'on tronque $y$ à $t^{max}$ et le noter $|t^{max}$
-
-where $\hat S$ is obtained with Algorithm~\ref{alg:construct-set-S} using $\hat D$ (that is $S^{(w_{max})}$ for the last iteration $w_{max}$), and the density $f$ in the rigth hand side of the previous equation being defined in~\eqref{eq:ind-hyp}.
-Note that marker~\eqref{eq:marker-def} is in fact obtained from the definition of $\hat \pi_{ik}^{\hat\theta}$ in~\eqref{eq:pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$, and turns out to be an estimate of the following posterior probability
-\[\P_\theta[G_i=k | T^\star_i > t^{max}_i, y_i].\]
-
-\subsection{The C-index metric}
-\label{sec:Metrics}
-
-We detail in this section the metric considered to evaluate risk prediction performances. Let us recall that in practice, one considers the case $K=2$, see Section~\ref{sec:K=2}. 
-In this context, the predictive marker for the lights model is
-\[\hat \cR_i^\text{lights} = \dfrac{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta)}{\pi_{\hat \xi}(x_i) \hat f(t^{max}_i, y_i | G_i=1 ; \hat \theta) + \big(1 - \pi_{\hat \xi}(x_i) \big) \hat f(t^{max}_i, y_i | G_i=0 ; \hat \theta)}.\]
-Then, let us denote by $\hat \cR$ the marker under study in the general case, and assume that it is measured once at $t = t^{max}$.
+  \hat \cR_{ik} = \P_{\hat \theta}[G_i=k | T^\star_i > t^{max}_i, y_i] = \dfrac{\pi_{\hat \xi_k}(x_i) S_k(t^{max}_i) f_{\hat \theta}(y_i | G_i=k)}{\sum_{k=0}^{K-1} \pi_{\hat \xi_k}(x_i) S_k(t^{max}_i) f_{\hat \theta}(y_i | G_i=k)},
+\end{equation}
+where $S_k$ is defined in~\eqref{eq:survival-k-def}.
+Note that marker~\eqref{eq:marker-def} is obtained from $\pi_{ik}^{\hat\theta}$ in~\eqref{eq:pi_ik-def} taking $(t_i, \delta_i) = (t^{max}_i, 0)$.
+Let us recall that in practice, one considers the case $K=2$ for the lights, see Section~\ref{sec:K=2}.
+
+\paragraph{The C-index metric}
+
+We detail in this section the chosen metric to evaluate risk prediction performances. 
+Let us denote by $\hat \cR_i$ the predictive marker for a subject $i$ at $t^{max}_i$ and from a given model (for the lights, $\hat \cR_i = \hat \cR_{ik}$ in~\eqref{eq:marker-def} with $k=1$).
 A common concordance measure that does not depend on time is the C-index~\citep{harrell1996tutorial} defined by
@@
 Note that other metrics have been proposed to compare individual risk predictions.
-In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}$.
-But comparing curves in an automatic way (in a cross-validation process for instance) is never easy, plus the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~-v\citep{bussy2019comparison}), which is not the case of the C-index metric.
-Nevertheless, the idea of chosing $t^{max}$ when it comes to simulate data for models performances comparisons will be discussed in Section~\ref{sec:simulation study}, and it is an interesting notion to consider since it can be sensed that the closer $t^{max}$ gets to $T^\star$, the easier it will be for a model to make good predictions.
+In~\citet{blanche2015quantifying} for instance, authors use dynamic prediction accuracy curves (of AUC or Brier score) with the idea of moving what is called the ``landmark time'', that is the time at which predictions are made and on which the amount of available information depends, corresponding to our $t^{max}_i$.
+But comparing curves in an automatic way (in a cross-validation process for instance) is complicated, and the comparisons proposed in the aforementioned paper require \textit{a priori} choices of time window thresholds (in particular to get back to binary predictions, being always questionable in survival analysis~-v\citep{bussy2019comparison}), which is not the case of the C-index metric.
+Finally, the question of the choice of $t^{max}_i$ in a study to assess a model performances, or to compare models, is discussed in the following paragraph.
+It can be sensed that the closer $t^{max}_i$ gets to $T^\star_i$, the easier it will be for a model to make good predictions.
+
+\paragraph{Procedure.}
+Let us describe here the procedure we follow to evaluate model performances on simulated data in Section~\ref{sec:simulation study} and on real data in Section~\ref{sec:application}.
+\begin{algorithm}[!htb]
+\DontPrintSemicolon
+\caption{Procedure followed to assess model performances in our real-time prediction paradigm.}
+\label{alg:evaluation}
+\KwInput{Dataset $\cD_n$~\eqref{eq:dataset-def} \textbf{only if} \textit{\texttt{simulation = false}} ; a \texttt{model} under study}
+\tcc{We run $\breve{R}=100$ independant experiments}
+$seed_0 = 0$ \\
+\For{$\breve{r} = 1,\ldots,\breve{R}$}
+{
+  $seed_{\breve{r}} = seed_{\breve{r}-1} + 1$ \\
+  \If{\texttt{simulation = true}}
+  {
+    {$\cD_n \leftarrow \texttt{simulate}(seed_{\breve{r}})$ \tcc{Sample a new dataset at each round $\breve{r}$}}
+  }
+  $(\cD_{n_{train}}, \cD_{n_{test}}) \leftarrow \texttt{split\_train\_test}(\cD_n, seed_{\breve{r}})$ \\
+  \tcc{For simplicity, we do not detail the cross-validation procedure here}
+  $\hat \theta \leftarrow$ \texttt{model.fit}$(\cD_{n_{train}})$ \\
+  \For{$i = 1,\ldots,n_{test}$}
+  {
+    $t_i^{max} \sim t_i \times \big(1 - \text{Beta}(2, 5)\big)$ \tcc{See Figure~\ref{fig:timax-pdf}}
+    \tcc{Truncate $y_i$}
+    $\mathcal{X}_i = (x_i, y_i)$ \\
+    \If{\normalfont \texttt{model = lights}}
+    {
+      $\big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J} \leftarrow \texttt{extract\_features}(y_i)$ \tcc{$(t_j^u)_{j=1,\ldots,J}$ from $\cD_{n_{train}}$}
+      $\mathcal{X}_i = \Big(x_i, y_i, \big(\Psi_i(t_j^u)\big)_{j=1,\ldots,J}\Big)$ 
+    }
+    $\hat \cR_{i,\breve{r}} \leftarrow \texttt{model.predict}(\mathcal{X}_i)$
+  }
+}
+\textbf{Return:} {$\big[\hat \cR_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}\ , \ \big[\texttt{time}_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}$}
+\end{algorithm}
+
+\begin{figure}[!htb]
+\centering
+\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
+\caption{Density of the law used to simulate $t_i^{max}$ to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.}
+\label{fig:timax-pdf}
+\end{figure}
+
+tirage sur le test set des timax avec loi Beta
+considérer 2 modèles avec 2 markers pour comparaison
+dire ici qu'on tronque $y$ à $t^{max}$ et le noter $|t^{max}$ 
+dire qu'on follow cette procedure sur donnees simulees et sur donnees reelles (mettre un if simulated-data : then simulate and data has b index)
 
@@
 The distribution of the censoring variable $C_i$ is the geometric distribution $\cG(\alpha_c)$, where $\alpha_c \in (0, 1)$ is empirically tuned to maintain a desired censoring rate $r_c \in [0,1]$. 
-Finally, the prediction time for subject $i$ is draw according to 
-\begin{equation}
-  \label{eq:law-t_i^max}
-  t_i^{max} \sim T_i \times \big(1 - \text{Beta}(\nu_1, \nu_2)\big) 
-\end{equation}
-with $(\nu_1, \nu_2) \in \R_+^2$ two shape parameters. Figure~\ref{fig:timax-pdf} shows the corresponding probability density function for $t_i^{max}$ with some explanations.
-We then truncate all $Y_i$ measurements being after $t_i^{max}$, since it represents the ``present'' time, so the time up to which one has longitudinal data.
 The choice of all hyper-parameters is driven by the applications on real data presented in Section~\ref{sec:application}, and summarized in Table~\ref{table:parameters choice}.
@@
 \resizebox{\textwidth}{!}{
-\begin{tabular}{ccccccccccccc}
+\begin{tabular}{cccccccccccc}
 \toprule
-$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\nu_1, \nu_2)$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
+$\pi_1$ & $(\rho_1, \rho_2, \rho_3)$ & $\mu_0$ & $\mu_1$ & $gap$ & $\sigma_l^2$ & $\upsilon$ & $(\kappa_1, \kappa_2)$ & $(\varsigma_1, \varsigma_2)$ & $r_c$ & $r_s$ & $\eta = \tilde{\eta}$ \\
 \midrule
@@
 \end{pmatrix}$ 
-& 0.5 & 4 & 3 & (2, 5) & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
+& 0.5 & 4 & 3 & $(10^{-3}, 0.1)$ & (1, 0.1) & 0.3 & 0.4 & 0.1 \\
 \bottomrule
@@
 \centering
-\subfigure[Density of the law used to simulate $t_i^{max}$ according to~\eqref{eq:law-t_i^max} with $(\nu_1, \nu_2) = (2,5)$, to mimic the fact that in practice, one has access to a reasonable amount of longitudinal data before making a prediction.]{
-\includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
-\label{fig:timax-pdf}}
+\subfigure[blabla]{
+% \includegraphics[width=.3\textwidth]{./figures/beta_law.pdf}
+\label{fig:yoyo}}
 \hspace{.02\textwidth}
@@
 
-Following the real-time prediction paradigm presented in Section~\ref{sec:evaluation strategy}, we compare the predictive performances of the lights model with the competing ones introduced in Section~\ref{sec:competing models} in terms of C-index (see Section~\ref{sec:Metrics}).
+Following the real-time prediction paradigm presented in Section~\ref{sec:evaluation strategy}, we compare the predictive performances of the lights model with the competing ones introduced in Section~\ref{sec:competing models} in terms of C-index (see Section~\ref{sec:evaluation strategy}).
 We also want to assess the stability of the lights model in terms of feature selection, but we do not compare this aspect with state-of-the-art since other models are not designed to perform feature selection, and comparisons would be unfair.
@@
 \begin{equation*}
-\hat \xi_{k, j} = 0 \ \Rightarrow \ \left| n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_{ij}  \right| < \zeta_{1,k} (1 - \eta)
+\hat \xi_{k, j} = 0 \ \Rightarrow \ \left| n^{-1} \sum_{i=1}^n \pi_{ik}^{\theta^{(w)}} \big( 1 - \pi_{\xi_k}(x_i) \big) x_{ij}  \right| < \zeta_{1,k} (1 - \eta)
 \end{equation*}

### 2022-05-10 manuscript/paper.tex
update ok

@@
 
-tirage sur le test set des timax avec loi Beta
-considérer 2 modèles avec 2 markers pour comparaison
-dire ici qu'on tronque $y$ à $t^{max}$ et le noter $|t^{max}$ 
-dire qu'on follow cette procedure sur donnees simulees et sur donnees reelles (mettre un if simulated-data : then simulate and data has b index)
-
 \subsection{Competing models}

### 2022-05-10 manuscript/paper.tex
update algorithm 3

@@
 where for each subject $i=1, \ldots, n$, process $Y_i^l$ is measured $n_i^l$ times at $t_{i1}^l, \ldots, t_{in_i^l}^l$ (which can differ between subjects and outcomes) with $t_{ij}^l \leq t_{ij+1}^l$ for all $j=1, \ldots, n_i^l-1$ and such that
-\[ y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) \]
+\begin{equation}
+  \label{eq:def-yil}
+  y_i^l=(y_{i1}^l, \ldots, y_{in_i^l}^l)^\top \in \R^{n_i^l} \quad \text{ with } \quad y_{ij}^l=Y_i^l(t_{ij}^l) 
+\end{equation}
 for all $l=1, \ldots, L$. 
@@
 \paragraph{Procedure.}
-Let us describe here the procedure we follow to evaluate model performances on simulated data in Section~\ref{sec:simulation study} and on real data in Section~\ref{sec:application}.
+Let us describe now in Algorithm~\ref{alg:evaluation} the procedure we follow to evaluate model performances on simulated data in Section~\ref{sec:simulation study} and on real data in Section~\ref{sec:application}.
+Here we do not give details on the cross-validation step, but it follows the same idea : for each fold, we sample the $t_i^{max}$ and truncate $y_i$ for $i$ in the test set of the fold.
 \begin{algorithm}[!htb]
 \DontPrintSemicolon
-\caption{Procedure followed to assess model performances in our real-time prediction paradigm.}
+\caption{Procedure followed to assess performances of a given model in our real-time prediction paradigm.}
 \label{alg:evaluation}
 \KwInput{Dataset $\cD_n$~\eqref{eq:dataset-def} \textbf{only if} \textit{\texttt{simulation = false}} ; a \texttt{model} under study}
+\KwOutput{Confidence intervals on C-index metric as well as on running time.}
 \tcc{We run $\breve{R}=100$ independant experiments}
@@
 {
+  $\texttt{start\_time} = \texttt{time}()$ \\
   $seed_{\breve{r}} = seed_{\breve{r}-1} + 1$ \\
@@
   $(\cD_{n_{train}}, \cD_{n_{test}}) \leftarrow \texttt{split\_train\_test}(\cD_n, seed_{\breve{r}})$ \\
-  \tcc{For simplicity, we do not detail the cross-validation procedure here}
-  $\hat \theta \leftarrow$ \texttt{model.fit}$(\cD_{n_{train}})$ \\
+  $\texttt{model.cross\_validate}(\cD_{n_{train}})$ \tcc{Select the best hyper parameters}
+  \texttt{model.fit}$(\cD_{n_{train}})$ \tcc{Learn $\hat \theta$}
   \For{$i = 1,\ldots,n_{test}$}
@@
     $t_i^{max} \sim t_i \times \big(1 - \text{Beta}(2, 5)\big)$ \tcc{See Figure~\ref{fig:timax-pdf}}
-    \tcc{Truncate $y_i$}
+    $y_i \leftarrow \big(Y_i^l(t_{ij}^l) \big)_{\substack{j=1, \ldots, n_i^l-1 \\ l=1, \ldots, L \ \ \ }} \text{ with } t_{ij}^l \leq t_i^{max}$ \tcc{Truncate $y_i$, see~\eqref{eq:def-yil}}
     $\mathcal{X}_i = (x_i, y_i)$ \\
@@
   }
+  $\texttt{score}_{\breve{r}} \leftarrow \texttt{c\_index}\big((\hat \cR_{i,\breve{r}}, t_i, \delta_i)_{i=1,\ldots,n_{test}} \big)$ \\
+  $\texttt{end\_time} = \texttt{time}()$ \\
+  $\texttt{running\_time}_{\breve{r}} = \texttt{end\_time} - \texttt{start\_time}$ \\
 }
-\textbf{Return:} {$\big[\hat \cR_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}\ , \ \big[\texttt{time}_{i,\breve{r}}\big]_{\substack{i=1,\ldots,n_{test} \\ \breve{r}=1,\ldots,\breve{R} \ \ \ }}$}
+\textbf{Return:} {$\big[\texttt{score}_{\breve{r}}\big]_{\breve{r}=1,\ldots,\breve{R}}\ , \ \big[\texttt{running\_time}_{\breve{r}}\big]_{\breve{r}=1,\ldots,\breve{R}}$}
 \end{algorithm}

### 2022-06-25 manuscript/paper.tex
changes after Antoine's review

@@
 Moreover, with an increased focus on personalised medicine, the need to implement multivariate models that account for a large number of longitudinal outcomes is critical.
-Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple univariate joint models separately~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}.
+Despite this, joint models have predominantly focused on univariate data, with attempts to fit multiple joint models separately each one with univariate longitudinal outcome~\citep{wang2012joint}, which is inefficient~\citep{lin2002maximum}.
 Despite many multivariate models being presented in full generality, questions arrising from the high-dimensional context -- e.g., computational power, limits in numerical estimation, or sample size -- are never considered in analyses (to the best of our knowledge), and the number of longitudinal outcomes considered in numerical studies are often very low (see~\citet{hickey2016joint} for a complete review).
@@
 
-The setting of this paper is such that we want to incorporate high-dimensional time-dependent (longitudinal) features measured with error in a survival model.
+The setting of this paper is such that we want to incorporate high-dimensional trajectories, or longitudinal features, measured with error in a survival model.
 Let us consider the usual survival analysis framework.
@@
 \]
-the global variance-covariance matrix.
-In the sequel of the paper, our aim is to associate the true and unobserved value $m_k^l(t)$ of the $l$-th longitudinal outcome at time $t$ with the event outcome $T^\star$.
+the global variance-covariance matrix, which is not class-specific.
 
@@
 \end{equation}
-where $\lambda_0(\cdot)$ is an unspecified baseline hazard function and given that $G = k$, we denote 
+where $\lambda_0(\cdot)$ is an unspecified baseline hazard function that does not depend on $k$, and we denote 
 \[\cY^l(t) = \{Y^l(u), 0 \leq u < t\} \quad \text{and} \quad \cY(t) = \bigcup_{l=1}^L \cY^l(t)\] 
 the history of the true longitudinal process up to time $t$.
-For each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping, and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters.
+Then for each $l$-th longitudinal outcome, we consider $\cA \in \N_+$ known functionals $\Psi_a^l$ extracted from $\cY^l(t)$ through a given representation mapping (also independent from $k$), and $\gamma_{k,a}^l \in \R$ the corresponding joint representation parameters, being the only class-specific object in the risk of event submodel.
 
@@
 Our model is clearly of JLCMs type, and can be viewed as a generalization of SREMs type in a much more flexible way.
-Table~\ref{table:shared_associations} gives common parameterizations for shared associations that are only considered separately ($\cA=1$) in SREMs.
+Table~\ref{table:shared_associations} gives common parameterizations for shared associations in SREMs.
+It is now common to include several of these in the survival sub-model, see for instance~\citet{}, but as far as we know, there is no method to consider simultaneously a huge number of associations and let the model learn which one is predictive for the underlying task it is trained for.
+
+TODO : add REF :
+
+Andrinopoulou, E.-R. and Rizopoulos, D. (2016). Bayesian shrinkage approach for a joint model of
+longitudinal and survival outcomes assuming different association structures. Statistics in Medicine,
+35(26):4813–4823.
+
+Ferrer, L., Rondeau, V., Dignam, J., Pickles, T., Jacqmin-Gadda, H., \& Proust-Lima, C.
+(2016).
+Joint modelling of longitudinal and multi-state processes: application to clinical progressions in prostate cancer.
+Statistics in medicine, 35(22), 3933–3948.
+https://doi.org/10.1002/sim.6972
+
 \begin{table}[htb]
@@
         (G) edge (mu)
+        (G) edge (lambda)
         (Y) edge (W);
@@
         (W) edge (lambda)
-        (G) edge (lambda)
         (lambda) edge (T)
@@
 \end{tikzpicture}}
-\caption{Graphical representation of the lights model.
+\caption{Graphical representation of the lights model, with $\phi_l$ being defined in~\eqref{eq:def-Phi_i} below.
 Square boxes denote observed data while circles denote unobserved (including random) terms.
@@
 As expected,~\eqref{eq:log-lik} is a JLCMs like log-likelihood.
+Note that to write~\eqref{eq:log-lik}, one could simplify Assumption~\ref{indep-hyp-3} by removing the $b_i$ both in the left and right side of Equation~\eqref{eq:ind-hyp} to get a classical JLCMs hypothesis, but the $b_i$ mention is useful later on in the E-step of the inference algorithm, for instance to compute Equation~\eqref{eq:E_g}.
 
@@
 
-One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important (time-dependent and time-independent) features relatively to the prediction objective.
+One advantage of the considered regularization strategy is its ability to perform feature selection and pinpoint the most important features (trajectories and time-independent) relatively to the prediction objective.
 The support of $\hat \xi_k$ informs on the time-independent features involved in the $k$-th latent class membership.
@@
 
-We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1, \ldots, b_n)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
+We first need to compute the negative completed log-likelihood (here scaled by $n^{-1}$), namely the negative joint distribution of $\cD_n$, $\textbf{\textit{b}} = (b_1^\top, \ldots, b_n^\top)$ and $\textbf{\textit{G}} = (G_1, \ldots, G_n)$.
 It can be written
@@
 In the case $g(b_i, G_i) = \tilde g(b_i)$, one has
-\begin{equation*}
+\begin{equation}
+  \label{eq:E_g}
   \E_{\theta^{(w)}}[ \tilde g(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g(b_i) f_{\theta^{(w)}}(b_i | t_i, \delta_i, y_i) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
-\end{equation*}
+\end{equation}
 where
@@
 \end{equation}
-and in the Gaussian case used in practice one has
+and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g}^{\theta^{(w)}}$ taking $\tilde g: b_i \mapsto 1$. 
+In the Gaussian case used in practice one has
 \begin{align*}
@@
 Note that for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the overall algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it.
-To speed up the overall algorithm, we then decrease the L-BFGS-B tolerance with the iterations $w$.
+To speed up the overall algorithm, we then increase the L-BFGS-B tolerance with the iterations $w$.
 
@@
 
-When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in Section~\ref{sec:evaluation strategy}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
+When validating predictions or comparing performances of competing models in a survival context, one can be interested either in the discriminative power of the predictive rule and use concordance measures such as the C-index~\citep{heagerty2005survival} (defined in~\eqref{eq:c-index-def}), or in the predictive accuracy of the rule~\citep{schemper2000predictive}.
 Developments in the field of joint modeling have primarily focused on modeling and estimation, and most studies do not consider goodness-of-fit nor generalization power in a prognostic prediction perspective~\citep{hickey2016joint}.
@@
 In our case,  $T^\star$ is subject to right censoring, so one would typically consider the modified $\cC_\tau$ defined by
-\begin{equation*}
+\begin{equation}
+  \label{eq:c-index-def}
   \cC_\tau =\P[\hat \cR_i > \hat \cR_j | T_i < T_j , T_i < \tau],
-\end{equation*}
+\end{equation}
 with $\tau$ corresponding to the fixed and prespecified follow-up period duration \citep{heagerty2005survival}.
@@
 We consider a multivariate version of JLCMs implemented in the \texttt{R} package \texttt{lcmm}~\citep{2017_lcmm}.
-Contrary to the lights model, there is no shared associations between the longitudinal models and the survival model.
+There is no shared associations between the longitudinal models and the survival model.
 Given the latent class membership, each submodel is assumed to be independent.
@@
 In this context, we consider the \texttt{JMbayes} package~\citep{2017_JMbayes}. The predictive marker associated with this model for subject $i$ is naturally given by the corresponding risk score
-\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\}.\]
-
-\texttt{@Antoine : est-ce qu'il peut prendre des time-indep feat ? et quelles sont les assos ?}
+\[\hat \cR_i^\text{JMbayes} = \exp \Big\{ x_i^\top \hat\gamma_0 + \sum_{l=1}^L \sum_{a=1}^\cA {\widehat{\gamma_{k,a}^l}}^\top \varphi_{k,a}(t_i^{max}, b_i^l) \Big\},\] 
+where the shared associations $\varphi_{k,a}$ are the four given in Table~\ref{table:shared_associations}.
 
@@
 \[\tilde{\xi}_j = \dfrac{|\hat{\xi}_j|}{\max_{j=1,\ldots,p}{|\hat{\xi}_j|}} \qquad \text{and} \qquad \tilde{\gamma}_{k,a}^l = \dfrac{|\hat{\gamma}_{k,a}^l|}{\max_{a=1, \ldots, \cA}{|\hat{\gamma}_{k,a}^l|}}\]
-for $k=0,\ldots,K-1$ and $l=1, \ldots, L$, and considering that $\tilde{\xi}_j$ and $\tilde{\gamma}_{k,a}^l$ are the predicted probability that the true $\xi_j$ equals $\varsigma_1$ and the predicted probability that the true $\gamma_{k,a}^l$ equals $\varsigma_2$ respectively, then we are in a binary prediction setting and we use the two resulting AUC scores of this problem to evaluate the selection power of the time-independent and the time-dependent features respectively.
+for $k=0,\ldots,K-1$ and $l=1, \ldots, L$, and considering that $\tilde{\xi}_j$ and $\tilde{\gamma}_{k,a}^l$ are the predicted probability that the true $\xi_j$ equals $\varsigma_1$ and the predicted probability that the true $\gamma_{k,a}^l$ equals $\varsigma_2$ respectively, then we are in a binary prediction setting and we use the two resulting AUC scores of this problem to evaluate the selection power of the time-independent features and the trajectories respectively.
 

### 2022-06-27 manuscript/paper.tex

@@
 Table~\ref{table:shared_associations} gives common parameterizations for shared associations in SREMs.
-It is now common to include several of these in the survival sub-model, see for instance~\citet{}, but as far as we know, there is no method to consider simultaneously a huge number of associations and let the model learn which one is predictive for the underlying task it is trained for.
-
-TODO : add REF :
-
-Andrinopoulou, E.-R. and Rizopoulos, D. (2016). Bayesian shrinkage approach for a joint model of
-longitudinal and survival outcomes assuming different association structures. Statistics in Medicine,
-35(26):4813–4823.
-
-Ferrer, L., Rondeau, V., Dignam, J., Pickles, T., Jacqmin-Gadda, H., \& Proust-Lima, C.
-(2016).
-Joint modelling of longitudinal and multi-state processes: application to clinical progressions in prostate cancer.
-Statistics in medicine, 35(22), 3933–3948.
-https://doi.org/10.1002/sim.6972
-
+It is now common to include several of these in the survival sub-model, see for instance~\citet{andrinopoulou2016bayesian, ferrer2016joint}, but as far as we know, there is no method to consider simultaneously a huge number of associations and let the model learn which one is predictive for the underlying task it is trained for.
 \begin{table}[htb]
